# Comparison of omics data between NAFLD and CHRIS baseline

For the set of NAFLD participants metabolomics and proteomics measurements is
available for CHRIS baseline and NAFLD. These derive from two different blood
samples (taken at different time points) from the same individuals. Here we
compare the concentrations for these metabolites and proteins between these
samples.

Get the IDs of NAFLD participants and their respective CHRIS AID.

```{r, message = FALSE}
library(chrisData)

gi <- chrisData("general")

nafld <- gi[which(gi$x0_nafld == "Yes"), ]
```

Participation year (in CHRIS baseline) for the `r nrow(nafld)` NAFLD study
participants.

```{r}
barplot(table(nafld$x0_examy))
```

We determine the number of participants with available CHRIS targeted
metabolomics as well as CHRIS ScanningSWATH and Somalogic proteomics data from
the baseline and NAFLD collections.

```{r, message = FALSE}
library(tidyfr)
soma <- data_module(name = "chris_somalogic", version = "1.0.3.2", path = ".")
scan <- data_module(name = "chris_scanningswath_proteins", version = "1.0.0.2",
                    path = ".")
p180 <- data_module(name = "metabolomics_p180", version = "1.0.1.2", path = ".")

soma_aid <- rownames(data(soma))
scan_aid <- rownames(data(scan))
p180_aid <- rownames(data(p180))
```

```{r, results = "asis"}
library(pander)
tmp <- data.frame(baseline = c(sum(nafld$AID %in% p180_aid),
                               sum(nafld$AID %in% scan_aid),
                               sum(nafld$AID %in% soma_aid)),
                  NAFLD = c(sum(nafld$aid_nafld %in% p180_aid),
                            sum(nafld$aid_nafld %in% scan_aid),
                            sum(nafld$aid_nafld %in% soma_aid)))
rownames(tmp) <- c("targeted metabolomics",
                   "ScanningSWATH proteomics",
                   "SomaLogic proteomics")
pandoc.table(tmp, style = "rmarkdown",
             caption = paste0("Number of NAFLD study participants (n = ",
                              nrow(nafld), ") with available omics data"))
```

Note: these numbers don't include any of the drop outs up to version 2 (July
2023).


## Metabolomics data

For NAFLD participants we have metabolite concentrations both for the CHRIS
baseline sample as well as the NAFLD sample. We can thus evaluate changes in
metabolite concentrations for these different time points. We below first
calculate correlations for metabolites (for each metabolite correlating
measurements in baseline to those from NAFLD).

```{r}
p180_data <- data(p180)
p180_labels <- labels(p180)
metabolites <- p180_labels$description
metabolites <- metabolites[-grep("assay flags", metabolites)]
colnames(p180_data) <- p180_labels$description

a <- as.matrix(p180_data[n_data$aid_chris, metabolites])
b <- as.matrix(p180_data[n_data$AID, metabolites])

cors <- vapply(seq_len(ncol(a)), function(i) {
    cor(a[, i], b[, i], use = "pairwise.complete.obs",
        method = "spearman")
}, numeric(1))
```

The table below lists the correlation coefficients (between values from baseline
and NAFLD) for all metabolites. Metabolites are ordered by magnitude of the
correlation coefficient.

```{r}
tmp <- data.frame(p180_labels[match(metabolites, p180_labels$description),
                              c("analyte_name", "analyte_class")],
                  rho = cors)
library(DT)
datatable(tmp[order(tmp$rho, decreasing = TRUE), ])
## pandoc.table(tmp[order(tmp$rho, decreasing = TRUE), ], style = "rmarkdown")
```

Some metabolites yield consistent values while others show much lower
*reproducibility*.

```{r, fig.cap = "Correlation coefficients for metabolites (ordered by magnitude)."}
plot(sort(tmp$rho), pch = 21, col = "#00000080", bg = "#00000040",
     ylab = "rho", xlab = "rank", main = "Metabolites")
grid()
```

We plot next the data for the best and worst correlating metabolite.

```{r, fig.cap = "Correlations between metabolite abundances in CHRIS baseline and NAFLD for the metabolite with the highest and lowest correlation. Correlations calculated for each metabolite on pairs of measurements for the same participant."}
par(mfrow = c(1, 2))

i <- which.max(tmp$rho)
plot(log2(a[, i]), log2(b[, i]),
     main = colnames(a)[i],
     xlab = expression(log[2]~baseline),
     ylab = expression(log[2]~NAFLD), pch = 21,
     col = "#00000080", bg = "#00000040")
grid()
abline(0, 1)

i <- which.min(abs(tmp$rho))
plot(log2(a[, i]), log2(b[, i]),
     main = colnames(a)[i],
     xlab = expression(log[2]~baseline),
     ylab = expression(log[2]~NAFLD), pch = 21,
     col = "#00000080", bg = "#00000040")
grid()
abline(0, 1)
```

For most participants, serotonin concentrations seem to be more or less stable.
Metabolite C14:2-OH showed also previously differences between measurements for
the first 5000 and second 2000 participants - thus, the lack of correlation is
also affected by a technical bias.

Next we calculate correlations for participants (correlate measurements for
each participant between baseline and NAFLD).

```{r}
cors <- vapply(seq_len(nrow(a)), function(i) {
    cor(a[i, ], b[i, ], use = "pairwise.complete.obs",
        method = "spearman")
}, numeric(1))

tmp <- data.frame(AID = rownames(a), rho = cors)
```

```{r, fig.cap = "Correlation coefficients for pairwise correlations of all metabolite measurements from baseline and NAFLD per participant."}
plot(sort(tmp$rho), pch = 21, col = "#00000080", bg = "#00000040",
     ylab = "rho", xlab = "rank", main = "Participants")
grid()
```

Correlation of values between baseline and NAFLD are highly correlated for most
participants, suggesting the metabolic profile (for the set of measured
metabolites) being somewhat stable. As examples we plot below the values for the
best and worst correlating participant.

```{r}
par(mfrow = c(1, 2))

i <- which.max(tmp$rho)
plot(log2(a[i, ]), log2(b[i, ]),
     xlab = expression(log[2]~baseline),
     ylab = expression(log[2]~NAFLD), pch = 21,
     col = "#00000080", bg = "#00000040")
grid()
abline(0, 1)

i <- which.min(tmp$rho)
plot(log2(a[i, ]), log2(b[i, ]),
     xlab = expression(log[2]~baseline),
     ylab = expression(log[2]~NAFLD), pch = 21,
     col = "#00000080", bg = "#00000040")
grid()
abline(0, 1)

```

Values for the first participant are nicely correlated. For the second,
correlation is present, but there seems to be a systematic shift in some values
(eventually e.g. values from FIA measurement).


## Proteomics data

We evaluate also correlations for the available proteomics data.

```{r}
swath_data <- data(scan)
swath_labels <- labels(scan)
proteins <- swath_labels$description
proteins <- proteins[swath_labels$Genes != ""]
colnames(swath_data) <- swath_labels$description

a <- as.matrix(swath_data[n_data$aid_chris, proteins])
b <- as.matrix(swath_data[n_data$AID, proteins])

cors <- vapply(seq_len(ncol(a)), function(i) {
    cor(a[, i], b[, i], use = "pairwise.complete.obs",
        method = "spearman")
}, numeric(1))
```

The table below lists the correlation coefficients (between values from baseline
and NAFLD) for all proteins. Proteins are ordered by magnitude of the
correlation coefficient.

```{r}
tmp <- data.frame(
    Genes = swath_labels[match(proteins, swath_labels$description),
                         c("Genes")],
    rho = cors)
rownames(tmp) <- proteins
## pandoc.table(tmp[order(tmp$rho, decreasing = TRUE), ], style = "rmarkdown")
datatable(tmp[order(tmp$rho, decreasing = TRUE), ])
```

High correlations can be observed for many proteins.

```{r, fig.cap = "Correlation coefficients for proteins (ordered by magnitude)."}
plot(sort(tmp$rho), pch = 21, col = "#00000080", bg = "#00000040",
     ylab = "rho", xlab = "rank", main = "Proteins")
grid()
```

We plot next the data for the best and worst correlating protein.

```{r, fig.cap = "Correlations between protein abundances in CHRIS baseline and NAFLD for the metabolite with the highest and lowest correlation. Correlations calculated for each protein on pairs of measurements for the same participant."}
par(mfrow = c(1, 2))

i <- which.max(tmp$rho)
plot(log2(a[, i]), log2(b[, i]),
     main = tmp[i, "Genes"],
     xlab = expression(log[2]~baseline),
     ylab = expression(log[2]~NAFLD), pch = 21,
     col = "#00000080", bg = "#00000040")
grid()
abline(0, 1)

i <- which.min(abs(tmp$rho))
plot(log2(a[, i]), log2(b[, i]),
     main = tmp[i, "Genes"],
     xlab = expression(log[2]~baseline),
     ylab = expression(log[2]~NAFLD), pch = 21,
     col = "#00000080", bg = "#00000040")
grid()
abline(0, 1)
```

Next we calculate correlations for participants (correlate measurements for
each participant between baseline and NAFLD).

```{r}
cors <- vapply(seq_len(nrow(a)), function(i) {
    cor(a[i, ], b[i, ], use = "pairwise.complete.obs",
        method = "spearman")
}, numeric(1))

tmp <- data.frame(AID = rownames(a), rho = cors)
```

Note that for the in total `r nrow(tmp)` participants we have only for
`r sum(!is.na(tmp$rho))` measurements both in the baseline and NAFLD collection
(mostly because measurements are missing in baseline).

```{r, fig.cap = "Correlation coefficients for pairwise correlations of all protein measurements from baseline and NAFLD per participant."}
plot(sort(tmp$rho), pch = 21, col = "#00000080", bg = "#00000040",
     ylab = "rho", xlab = "rank", main = "Participants")
grid()
```

Correlation of values between baseline and NAFLD are highly correlated. As
examples we plot below the values for the best and worst correlating
participant.

```{r}
par(mfrow = c(1, 2))

i <- which.max(tmp$rho)
plot(log2(a[i, ]), log2(b[i, ]),
     xlab = expression(log[2]~baseline),
     ylab = expression(log[2]~NAFLD), pch = 21,
     col = "#00000080", bg = "#00000040")
grid()
abline(0, 1)

i <- which.min(tmp$rho)
plot(log2(a[i, ]), log2(b[i, ]),
     xlab = expression(log[2]~baseline),
     ylab = expression(log[2]~NAFLD), pch = 21,
     col = "#00000080", bg = "#00000040")
grid()
abline(0, 1)
```


# Session Information

```{r}
sessionInfo()
```
