---
title: "NAFLD_metab_targeted_lm"
author: "Marilyn De Graeve, Nikola Dordevic, Vinicius Verri Hernandes, Clemens Dierks, Essi Hantikainen, Johannes Rainer"
affiliation: "Eurac Research, Bolzano, Italy"
date: "2024-03-11"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
bibliography: references.bib
---

**Modified**: `r file.info("NAFLD_metab_targeted_lm.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r style, message = FALSE, echo = FALSE, warning = FALSE, results = "asis"}
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now, units = "secs")
      # return a character string to show the time
      paste("    Time for this code chunk to run:", round(res,
        2), "seconds")
    }
  }
}))

suppressMessages(library("BiocStyle"))
suppressMessages(library("kableExtra"))
suppressMessages(library("knitr"))
suppressMessages(library("rmarkdown"))
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7, time_it = TRUE)

```

# Introduction

During this workflow, statistical analyis (linear models) are performed on the
targeted metabolomics data from the metabolic dysfunction-associated steatotic
liver disease (MASLD) cohort. MAFLD, previously known as NAFLD is the most
chronic liver disease, affecting more than 30% of global population. The MAFLD
cohort is a subset of approximatley 353 # participants from the Cooperative
Health Research in South Tirol (CHRIS) study, which were requested to return for
a second resampling and additional phenotyping. The NAFLD cohort represent an
unhealty group with various degrees of steatotic (alcoholic) liver disease
classifications.

This document describes the statistical analysis using linear models to identify
associations between metabolites and metabolic dysfunction-associated steatotic
liver disease (MASLD). The next 4 defined types of steatotic liver
disease (SLD) classifications are examined here:

- *none: participants negative for steatotic liver disease.
- *MASLD: (= Metabolic Dysfunction Associated Steatotic Liver Disease)
  participants with SLD, positive for cardiometabolic criteria but low alcohol
  consumption (below 40g/day). This includes also 8 participants without
  information on alcohol consumption. These participants did not provide the
  GA2LEN FFQ (hence missing values for quantified alcohol consumption), but
  according to alcohol-related information in other questionnaire they are not
  heavy drinkers and are thus added to this category.
- *MetALD: (= MASLD and increased alcohol intake) participants with SLD,
  positive for cardiometabolic criteria but daily alcohol consumption > 40g/day.
- *SLD: (= steatotic liver disease) participants with a SLD, but negative for
  cardiometabolic criteria. This includes potentially cryptogenic SLD, alcoholic
  liver disease (ALD), DILI or monogenic diseases.

In contrast to using the different binary variables (SLD, MASLD and MetALD)
seperately, the categories are evaluated together using the single categorical
variable `sld_type`. See the `nafld_data_overview 1.html` report for further
information regarding the SLD sub-classification.


# Setup

## Directories

```{r directories, echo = FALSE}
#' General settings
filename <- " NAFLD_metab_targeted_lm"

#' Path to save images to; remove if exists.
IMAGE_PATH <- paste0("images/", filename, "/")
if (file.exists(IMAGE_PATH))
    unlink(IMAGE_PATH, recursive = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE)

#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)

#' Path to the data
DATA_PATH <- '/home/mdegraeve/Documents/Files/Work_Eurac/Data/NAFLD'  #'/media/mdegraeve/josung/data/NAFLD' #'data' #FINDME!

```

## Packages

```{r packages, message=FALSE}
#' Load libraries
library(BiocParallel)
library(DT)
suppressMessages(library(ggfortify))
library(ggplot2)
library(ggpubr)
suppressMessages(library(magick))
suppressMessages(library(pander))
suppressMessages(library(pheatmap))
suppressMessages(library(plotly))
suppressMessages(library(RColorBrewer))
suppressMessages(library(readxl))
library(reshape2)
options(rgl.useNULL = TRUE)
library(rgl)
suppressMessages(library(tidyfr))
suppressMessages(library(vioplot))
library(writexl)

```

## Computation

```{r parallel_processing_setup, message=FALSE}
#' Set up parallel processing using multiple cores get the number of
#' cpus allocated by queueing system or fall back to 7
NR_CORES <- as.integer(Sys.getenv("SLURM_JOB_CPUS_PER_NODE", 8)) - 1L
if (.Platform$OS.type == "unix") {
    register(MulticoreParam(NR_CORES))
} else {
    register(bpstart(SnowParam(NR_CORES)))
}

```


# Data import

The data from metabolites was converted into the TDF format and can hence be
loaded and imported using the `tidyfr` R package. Now we read the full data
which includes general information for each measurement.


```{r load_dataset, message = FALSE}
#' load TDF using tidyfr
#list_data_modules(path = DATA_PATH)
p180 <- data_module(name = "metabolomics_p180", version = "1.0.1.2", path = DATA_PATH)
met_data <- data(p180)
met_ann <- labels(p180)
stopifnot(all(colnames(met_data) == rownames(met_ann)))
colnames(met_data) <- met_ann$description
rownames(met_ann) <- met_ann$description

#' define the columns with metabolite concentrations
metabolites <- rownames(met_ann)[met_ann$type != "categorical"]

#' problem in calculating lm below for
#'   'PC aa C30:2' #NOK, only 8 col!!
#'   'Histamine' NOK, only 8 col!!
#' because no values for each typeSLD, due to lot NA values
#' option 1: rm metabolites, not done
#metabolites <- metabolites[metabolites != 'PC aa C30:2']
#metabolites <- metabolites[metabolites != 'Histamine']
#' option2: retain only none and MASLD, see done later

#' select metabolites dataframe
met_data <- met_data[, metabolites]
#head(met_data)    #row samples, cols are metab name, matrix are absol conc
print('nr of  metabolites included:')
ncol(met_data)
print('included metabolites are:')
colnames(met_data)
#dim(met_data)

```

The loaded data contains the samples as rows, and the metabolite names as
colums. The measurements of the `met_data` object are the absolute
concentrations of metabolite per sample.

To ensure a normal distribution of the metabololits, we perform log2
transformation.

```{r log2-transformation, message = FALSE}
#' log2 scaling of absolute abundances metabolites
met_data <- log2(met_data)

```

Before proceeding with the data analysis we remove the POOL samples. In addition
we restrict to samples from the NAFLD study.

```{r subset to NAFLD set, message = FALSE, warning = FALSE}
met_data <- met_data[grep("^008", rownames(met_data)), ]

```

All available traid (phenotype) data, previously saved in the file, was loaded
and merged with the log2 transformed metabolomics data of the NAFLD cohort.

```{r add phenotype dataa}
#' load xlsx file instead of below
handle <- 'masld_phenotypes.xlsx'
masld_phenotype_df <- read_excel(file.path(DATA_PATH, handle))
#head(masld_phenotype_df, 5)
#colnames(masld_phenotype_df)
#dim(masld_phenotype_df)

#' make sure sld_type is a factor with group 'none' as the baseline
masld_phenotype_df$sld_type <- as.factor(masld_phenotype_df$sld_type)
masld_phenotype_df$sld_type <- factor(masld_phenotype_df$sld_type, levels = c('none', 'MASLD', 'MetALD', 'SLD'))
print('the sld types are:')
levels(masld_phenotype_df$sld_type)
print('nr of samples per sld types:')
table(masld_phenotype_df$sld_type)

#' merge with metabolomics data
met_data$AID <- rownames(met_data)
#dim(met_data)
met_data <- merge(met_data, masld_phenotype_df, by = 'AID')
#dim(met_data)

#' write to file for manuel inspection
#handle <- 'data/met_data.txt'
#write.table(met_data, file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)

```

Because for 2 metabolites, 'PC aa C30:2' and 'Histamine', only NAs are present
for the less populated `sld_type` groups (MetALD end SLD); only none and MASLD
types are considered further.

```{r keep only none and MASLD types}
#' because problem 2 metabolites only NAs in some sld types,
#' keep only none and MASLD
met_data <- met_data[met_data$sld_type == 'none' | met_data$sld_type == 'MASLD',]

met_data$sld_type <- as.factor(met_data$sld_type)
met_data$sld_type <- factor(met_data$sld_type, levels = c('none', 'MASLD'))

print('the sld types are:')
levels(met_data$sld_type)
print('nr of samples per sld types:')
table(met_data$sld_type)

```


Create additional variable for age per 10y, because coefficients and effect
sizes correspond to one year difference, and the anual difference is
neglectible. Dividing age with 10, difference in coefficents and effect sizes
would correspond to 10 years difference and it would be easier to define
meaningful cut-off.

```{r define-age10-variable, message = FALSE}
#' define age per 10y group
met_data$Age_10 <- met_data$Age / 10

```

# Linear model to evaluate SLD types

To identify metabolites related with the 2 (none, MASLD) of the 4 SLD types
(none, MASLD, MetALD, SLD), seperate linear regression models are fitted for all
metabolites separately (as a response) with sex, age and the (categorical)
sld_type as a covariates.

Model coefficients and p-values are provided. Raw p-values are adjusted for
multiple hypothesis testing with the method from Benjamini & Hochberg.

```{r linear model, message=FALSE}
#' Define a function to fit the linear model:
#' - fit the linear model
#' - extract the coefficients and p-values skipping the intercept
#' - return the result as a one-row matrix, columns being coefficients
#'   and p-values
funForLinReg <- function(analyte, data) {
    y <- data[, analyte]
    data <- data[, c("Sex", "Age_10", "sld_type")]
    res <- coef(summary(lm(y ~ ., data = data)))[-1, ]
    pvals <- matrix(c(res[, 1L], res[, 4L]), nrow = 1)
    rownames(pvals) <- analyte
    colnames(pvals) <- c(paste0("coef_", rownames(res)),
                         paste0("p-value_", rownames(res)))
    return(pvals)
}

#' tests
#met_data[1:5, c("Sex", "Age", "sld_type")]
#met_data[1:5, metabolites[1]]
#data <- met_data
#analyte <- metabolites[1]
#funForLinReg(analyte, met_data)
#analyte <- metabolites[14]
#funForLinReg(analyte, met_data)

#' Fit linear models and extract coefficients and p-values on each metabolite
lm_output <- do.call(rbind,
                     lapply(metabolites, function(x) funForLinReg(x, met_data)))

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)

```

Next we calculate the effect sizes.

To obtain comparable coefficients, prior to linear modelling all the metabolites
are standardized to zero mean and unit variance (note that this does not
influence the linear models since they are affine equivariant).
In autoscaled data, difference of 1 corresponds to standard deviation of 1.
This parameter will be called "effect size" and will be use to evaluate group
differences.

```{r effectsize, message=FALSE}
#' Perform autoscaling for effect size (es) calculation
met_data_scaled <- met_data
met_data_scaled[, metabolites] <- met_data[, metabolites] |>
    as.matrix() |>
    scale(scale = TRUE)

#' fit linear models and extract the coefficients
es <- lapply(metabolites, function(x) funForLinReg(x, met_data_scaled)) |>
    do.call(what = rbind)
es <- es[, grep("coef", colnames(es))]
colnames(es) <- sub("coef", "effectSize", colnames(es))
lm_output <- cbind(lm_output, es)
#head(lm_output)

```

To call a metabolites *significant* we require that it has, in addition to an
adjusted p-value smaller than 0.05, also a difference in abundances (expressed
in percentage) which is at least once as high as the coefficient of variation
(CV) for that metabolite. The coefficient of variation was calculated on the QC
CHRIS Pool samples (cv_qc_chris column in the p180 TDF labels), hence it
represents the technical variability that was observed for each metabolite.

Below we add thus evaluate for each metabolite and comparison whether it fulfills
that criteria.

```{r define significance criteria}
#' define the different coefficients as comparisons
comps <- colnames(lm_output)[grep("coef", colnames(lm_output))]

#' define which metabolites are considered significant using p-adj only
sign_p <- lm_output[, grep("p-adj", colnames(lm_output))] < 0.05
#colSums(sign_p)

#' Convert the (absolute) difference in abundance into a difference in
#' percentage based on a (log2) coefficient.
diff_percentage <- function(x) {
    (2^abs(x) - 1) * 100
}
diff_perc <- apply(lm_output[, comps],
                   MARGIN = 2, diff_percentage)
colnames(diff_perc) <- sub("coef", "diff_perc", colnames(diff_perc))

#' define which metabolites are considered significant using p-adj and cv qc chris
sign_cv <- lapply(comps, function(z) {
    z <- sub("coef_", "", z)
    sign_p[, paste0("p-adj_", z)] < 0.05 &
        diff_perc[, paste0("diff_perc_", z)] > 50 * #findme!!! 2*100 of *100? EVAL!!
        met_ann[metabolites, "cv_qc_chris"]
}) |>
    do.call(what = cbind)
colnames(sign_cv) <- paste0("significant_", comps)

#' Use statistical significance for non-categorical traits instead of bool
sign_cv[, "significant_coef_Age_10"] <- sign_p[, "p-adj_Age_10"]

lm_output <- cbind(lm_output, sign_cv)

```

The number of metabolites called significant or with an adjusted p-value smaller
than 0.05 for each of the extracted coefficients are shown below. For a complete
comparison, the number of metabolites complying to either the p-adjusted value
or either the CV-value, as single threshold criteria, are calculated and
summarized here as well.

```{r table-sig-mets, results = "asis"}
#' nr of covariates with p-adj < 0.05
sig_p <- apply(lm_output[, grep("adj", colnames(lm_output))], 2,
                 function(z) sum(z < 0.05))
names(sig_p) <- sub("p-adj_", "", names(sig_p))

#' only cv significance (wo p-adj) to be complete
cv <- lapply(comps, function(z) {
    z <- sub("coef_", "", z)
    diff_perc[, paste0("diff_perc_", z)] > 50 * #findme!!! 2*100 of *100? EVAL!!
        met_ann[metabolites, "cv_qc_chris"]
}) |>
    do.call(what = cbind)
colnames(cv) <- paste0("cv_", comps)

#' merge to p-adj to cv significance summary
sig_tab <- data.frame(
    `p_adj < 0.05` = sig_p,
    `cv thresh` = colSums(cv),
    significant = colSums(sign_cv[, paste0("significant_coef_", names(sig_p))]),
    check.names = FALSE)

pandoc.table(data.frame(sig_tab, check.names = FALSE),
             style = "rmarkdown",
             caption = paste0("Number of significant metabolites per tested covariate. ",
                     " *p_adj < 0.05*: number of statistically significant metabolites.",
                     " *cv*: number of metabolites with percentage abundance difference ",
                     " above the CV-based threshold.",
                     " *significant*: number of metabolites that fulfill ",
                     "the CV-based criteria in addition to the p-adj. threshold"))

```

Export results for inspection:

```{r export-results}
results <- data.frame(
    analyte_class = met_ann[rownames(lm_output), c("analyte_class")],
    lm_output,
    #es,
    diff_perc,
    met_ann[metabolites, "cv_qc_chris"], #check manual cv threshold
    rownames(lm_output))
    #sign_cv)

#colnames(results)
#colnames(lm_output)

#' write to file for manual inspection,
#' inclusing inspection of the CV-value of each metabolite
handle <- 'data/lm_output_results.txt'
write.table(results, file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)

```

we visualize the associations in a heatmap. The effect size are added for the
associations, for the significant metabolites (min 1 significance to
covariate). also for the not significant associations, the effect size is shown.

```{r heatmap, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for metabolites with at least one significant association."}
#' Extract effect sizes
es_results <- results[, grep("effectSize_", colnames(results))] |>
    as.matrix()

#' Restrict to those significant for at least one of the groups (sex, age, SLD type)
sig_results <- results[, grep("significant_", colnames(results))] |>
    as.matrix()
es_results <- es_results[rowSums(sig_results) > 0, ]
print('nr of metabolites with >= 1 significant association:')
nrow(es_results)

#' hierarchical clustering
clust1 <- hclust(dist((es_results)), method = "ward.D")
clust2 <- hclust(dist(t(es_results)), method = "ward.D")

#' clean roundup nrs for plot
es_results[es_results == 0] <- NA
es_results <- round(es_results, 2)

#' info regarding sld_type
#print('nr of metabolites with a not-zero effect size in the sld_type group:')
#sum(es_results[, 'effectSize_sld_typeMASLD'] != 0)

#' in long format
es_results_long <- melt(as.matrix(es_results))

#' ggplot
plt <- es_results_long |>
    ggplot(aes(Var2, Var1, fill = value, label = value)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Effect Size") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 2.5) +
    theme_classic() +
    scale_x_discrete(
        expand = c(0, 0), limits = colnames(es_results)[clust2$order]) +
    scale_y_discrete(
        expand = c(0, 0), limits = rownames(es_results)[clust1$order]) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 10),
          axis.text.y = element_text(hjust = 1, size = 8),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.position = "none")

ggsave(paste0(IMAGE_PATH, "metabolite_SLDtypes_effect_sizes.png"),
       plot = plt, dpi = 600, width = 12,
       height = 12, units = "cm", scale = 2.5)
plt

```

In second plot, only the significant effect size labels were added to the
heat map. also, the dendograms after hierarchical clustering are shown.

```{r heatmap2, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for significant associations the value of the effect size is shown."}
#' keep just the effect sizes of the significant associations
es_results_sig <- es_results
es_results_sig[as.matrix(sig_results)[rownames(es_results), ] == 0] <- ""

#' info regarding sld_type (check=ok)
print('nr of significant metabolites in the sld_type group:')
sum(es_results_sig[, 'effectSize_sld_typeMASLD'] != "")

#' plot
pheatmap(es_results, display_numbers = es_results_sig,
         number_color = "#444444", fontsize_number = 7)

```

We investigate the associations for the SLD types 'none vs. MASLD' further, by
visualizing the metabolites in a volcano plot.  Red colored points are those with an adjusted
p-value smaller than 0.05.


```{r volcano, FIG-volcano-sld_type, fig.path = IMAGE_PATH, fig.width = 8, fig.height = 8, fig.cap = "Volcano plot for the liver stiffness. Red colored points are those with an adjusted p-value smaller than 0.05."}
plot_volcano <- function(x, comparison = character(), ...) {
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    plot(X, -log10(Y), pch = 21,
         col = "#000000ce", bg = "#00000060",
         xlab = "coefficient", ylab = expression(-log[10](p[adjusted])),
         ...)
}
significant_points <- function(x, comparison = character(),
                               diff_perc = 2, col = "#E41A1CCE",
                               bg = "#E41A1C40", ...) {
    sign_cv <- x[, paste0("significant_coef_", comparison)]
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    if (any(sign_cv))
        points(X[sign_cv], -log10(Y[sign_cv]), col = col, bg = bg, pch = 21)
}
plot_volcano(results, "sld_typeMASLD", main = "sld_typeMASLD")
abline(h = -log10(0.05), lty = 2, col = "grey")
significant_points(results, comparison = "sld_typeMASLD")
                                        #!!!findme warning due to double colnames, if 0 signif...

```

The (static) table with the significant metabolites is shown below:
#FINDME with signif ipv p-adj only better here???!!! also >0.05 so NOK???

```{r table-sig-sex, results = "asis", echo = FALSE}
significant_table <- function(x, comparison = character(),
                              columns = c("analyte_class")) {
 sigs <- which(x[, paste0("significant_coef_", comparison)] == 1) #get rownr of sign metab
 columns <- c(columns, paste0(c("coef_", "p.adj_", "effectSize_"), comparison))
 tmp <- x[sigs, columns]
 tmp <- tmp[order(tmp[, paste0("p.adj_", comparison)]), ]
 colnames(tmp) <- sub(paste0("_", comparison), "", colnames(tmp))
 return(tmp)
}
sig_tab <- significant_table(results, comparison = "sld_typeMASLD")

pandoc.table(sig_tab,
             style = "rmarkdown", split.table = Inf,
             caption = "Significant metabolites for sld_typeMASLD")  #NOK bug FINDME!!! wrong if see excel and 2nd table???

```

Show in table with the results for all metabolites from the test ordered with the
tests p-values and effect size:

```{r table-top-20, results = "asis"}
##
oorder <- results[order(results[, "p.adj_sld_typeMASLD"]), ]

pandoc.table(oorder[1:5, c(1, 4, 7, 10, 13, 17)], style = "rmarkdown",
             split.table = Inf, caption = "Top 5 metabolites for type MASLD")

#datatable(Order[,c(1, 4, 7, 10, 13, 17)]) %>%
#    formatRound(columns=c(2:5), digits = 6)

```

There are `r nrow(sig_tab)` significantly associated metabolites for SLD type none vs. MASLD.


Significant metabolites adjusted for age and sex (top 3) are visualized in
boxplot in next figure. Points are colored according to sample sex.

```{r, BP_ADMA, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
box_plot <- function(data, aes, analyte) {
    ggplot(data, aes) +
        geom_jitter(position = position_jitter(0.24),
                    aes(colour = 'black' , alpha = 0.5)) + #HCU
        geom_violin(trim = FALSE, aes(alpha = 0.5)) +
        stat_summary(fun = median, geom = "point", shape = 23, size = 2) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) + labs(title = analyte) +
        theme_bw ()
}
PLOTLY <- FALSE
ggplotly_or_not <- function(x, interactive = TRUE) {
    if (interactive) ggplotly(x)
    else print(x)
}

#' boxplot for metabolite
s_1 <- box_plot(met_data, aes(x = sld_type, y = ADMA),
                  analyte = "ADMA")
ggplotly_or_not(s_1, PLOTLY)

```

2nd example

```{r, BP_Glu, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
#' boxplot for metabolite
s_1 <- box_plot(met_data, aes(x = sld_type, y = Glu),
                  analyte = "Glu")
ggplotly_or_not(s_1, PLOTLY)

```


# Acknowledgments

Consulted sources used as basis for code:

- https://github.com/EuracBiomedicalResearch/chris_plasma_proteome/blob/main/sex_age_bmi_associated_proteins.Rmd
- https://github.com/EuracBiomedicalResearch/metabo_chris_targeted_MetaboEnv/blob/main/medication_effect_atc4.Rmd


# Session information

R packages used for the analysis:

```{r}
sessionInfo()
```

# References