---
title: "Overview and characterization of the NAFLD substudy"
author: "Johannes Rainer, Essi Hantikainen"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
bibliography: references.bib
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 10, fig.height = 7)
```

# Introduction

In this document we evaluate and analyze the clinical/phenotypic data available
for participants of the NAFLD sub-study of CHRIS. In addition, we
stratify/diagnose individuals following recent clinical guidelines.


# Data overview

Demographics of the NAFLD cohort and evaluation of available trait data.

```{r}
library(chrisData)
library(nafldData)
library(vioplot)

n_anns <- nafldTraitAnnotations()

#' personal data
n_perso <- nafldData("personal")
n_data <- n_perso[, c("AID", "aid_chris", "na_ager", "na_sex", "na_fwup")]
#' na_ager: rounded age
#' na_fwup: follow-up duration CHRIS->NAFLD

tmp <- nafldData("recruitment")
rownames(tmp) <- tmp$AID
n_data$case <- tmp[n_data$AID, "na_case"]

```

The data set is nicely balanced by sex:

```{r}
table(n_data$na_sex)
```

The age distribution of participants (split by sex) is shown below.

```{r, fig.cap = "Age distribution."}
vioplot(split(n_data$na_ager, n_data$na_sex), ylab = "age")
grid()
```

The distribution of time between participants' CHRIS baseline and NAFLD
participation is shown below.

```{r, fig.cap = "Distribution of the time span between CHRIS baseline and NAFLD."}
plot(density(n_data$na_fwup))
grid()
```

The time span between the two studies is thus between 2 to 5 years. We next
evaluate BMI and fat composition for the study participants, split by
sex.

```{r}
#' BMI, fat etc.
n_anthro <- nafldData("anthro")
rownames(n_anthro) <- n_anthro$AID

n_data$BMI <- n_anthro[n_data$AID, "naan03a"]
n_data$body_fat <- n_anthro[n_data$AID, "naan04"]
n_data$visc_fat <- n_anthro[n_data$AID, "naan05"]

```

```{r, fig.cap = "Distribution of BMI and body fat per sex."}
par(mfrow = c(1, 3))
#' split by sex
vioplot(split(n_data$BMI, n_data$na_sex), main = "BMI", ylab = "BMI")
grid()
vioplot(split(n_data$body_fat, n_data$na_sex), main = "Body fat",
        ylab = "Body fat")
grid()
vioplot(split(n_data$visc_fat, n_data$na_sex), main = "Visceral fat",
        ylab = "Visceral fat")
grid()

```

While the distribution of the BMI is balanced between female and male
participants, females have on average a much higher body fat percentage, while
males have on average a higher proportion of visceral fat.

Note that there seems to be no dependency between age and BMI in the present
data set (data not shown).

We next evaluate the clinical examinations performed on all participants.

```{r}
#' Clinical examinations
tmp <- nafldData("clinical")
rownames(tmp) <- tmp$AID
```

We will use and evaluate the following parameters defined by a clinical
examination:

- *naab11*: liver stiffness (higher values are worse). Values for participants
  that have a *naab12* (liver stiffness IQR) > 30 should be either removed or
  taken with care.
- *naab12*: liver stiffness IQR: the IQR of repeated liver stiffness
  measurements. Values below 30 are considered OK.
- *naab13*: liver fat. Can be binarized at an absolute value of 288.
- *naab14*: subjective, visual evaluation of the liver fat. Categorical from 1
  to 3. Values 2 and 3 can be considered steatotic.
- *nascfib4*: fibrosis 4 index. Linear score calculated from standard clinical
  parameters.
- *nascmets*: metabolic syndrome, binary category.

```{r}
n_data$liver_stiffness <- tmp[n_data$AID, "naab11"]
n_data$liver_stiffness_iqr <- tmp[n_data$AID, "naab12"]
n_data$liver_fat <- tmp[n_data$AID, "naab13"]
n_data$visual_fat <- tmp[n_data$AID, "naab14"]
n_data$fib4 <- tmp[n_data$AID, "nascfib4"]
n_data$metsym <- tmp[n_data$AID, "nascmets"]
```

From the `r nrow(n_data)` participants,
`r sum(n_data$liver_stiffness_iqr >= 30, na.rm = TRUE)`
have a liver stiffness IQR `>= 30`.

At first we evaluate the distribution of the liver stiffness score in the data
set.

```{r, fig.cap = "Distribution of liver stiffness measurements. All values considered. Problematic values are highlighted in red in the lower panel."}
stiff_ok <- which(n_data$liver_stiffness_iqr < 30)
par(mfrow = c(1, 3))
vioplot(split(log2(n_data$liver_stiffness), n_data$na_sex),
        main = "Liver stiffness", ylab = expression(log[2](Liver~stiffness)))
grid()
col <- rep("#ff0000", nrow(n_data))
col[stiff_ok] <- "#000000"
x <- n_data$na_ager
y <- log2(n_data$liver_stiffness)
plot(x, y, pch = 21, bg = paste0(col, "40"),
     col = paste0(col, "80"), main = "Liver stiffness",
     xlab = "age", ylab = expression(log[2](Liver~stiffness)))
grid()
abline(lm(y ~ x), lty = 1, col = "#00000080")
x <- n_data$BMI
plot(x, y, pch = 21, bg = paste0(col, "40"),
     col = paste0(col, "80"), main = "Liver stiffness",
     xlab = "BMI", ylab = expression(log[2](Liver~stiffness)))
grid()
abline(lm(y ~ x), lty = 1, col = "#00000080")
```

While there is a clear average difference for liver stiffness between male and
female participants, there seems to be no general age dependency of the liver
stiffness. Only for a subset of participants there seems to be a (log linear)
increase with age. In contrast, a (log-linear) relationship between BMI and
liver stiffness can be observed.

We next evaluate the distribution of liver fat values in the study cohort.

```{r, fig.cap = "Distribution of liver fat values in the study cohort."}
par(mfrow = c(1, 3))
vioplot(split(n_data$liver_fat, n_data$na_sex),
        main = "Liver fat", ylab = "Liver fat")
grid()
x <- n_data$na_ager
y <- n_data$liver_fat
plot(x, y, pch = 21, main = "Liver fat", xlab = "age", ylab = "Liver fat",
     bg = "#00000040", col = "#00000080")
grid()
abline(lm(y ~ x), lty = 1, col = "#00000080")
x <- n_data$BMI
plot(x, y, pch = 21, main = "Liver fat", xlab = "BMI", ylab = "Liver fat",
     bg = "#00000040", col = "#00000080")
grid()
abline(lm(y ~ x), lty = 1, col = "#00000080")
```

Similar to the liver stiffness, we see differences in average liver fat values
between male and female study participants, while there seems no age
dependency. A strong relationship between BMI and liver fat can be observed.

For completeness we also plot the visual fat categorization.

```{r, fig.cap = "Overview of visual fat categories."}
par(mfrow = c(1, 3))
split(n_data$visual_fat, n_data$na_sex) |>
    lapply(table) |>
    do.call(what = cbind) |>
    barplot()
vioplot(split(n_data$na_ager, n_data$visual_fat), main = "Visual fat",
        ylab = "age")
grid()
vioplot(split(n_data$BMI, n_data$visual_fat), main = "Visual fat",
        ylab = "BMI")
grid()
```

A clear relationship can be seen for the visual fat evaluation with the cases as
well as BMI.

Next we evaluate the FIB4 score.

```{r, fig.cap = "Distribution of FIB4 scores."}
par(mfrow = c(1, 3))
vioplot(split(log2(n_data$fib4), n_data$na_sex),
        main = "FIB4", ylab = expression(log[2](FIB4)))
grid()
x <- n_data$na_ager
y <- log2(n_data$fib4)
plot(x, y, pch = 21, main = "FIB4", xlab = "age",
     ylab = expression(log[2](FIB4)),
     bg = "#00000040", col = "#00000080")
grid()
abline(lm(y ~ x), lty = 1, col = "#00000080")
x <- n_data$BMI
plot(x, y, pch = 21, main = "FIB4", xlab = "BMI",
     ylab = expression(log[2](FIB4)),
     bg = "#00000040", col = "#00000080")
grid()
abline(lm(y ~ x), lty = 1, col = "#00000080")
```

There seems to be only a small difference in average FIB4 values between male
and female participants. A strong (positive) relationship is however visible
between FIB4 and age. A slight negative relationship seems to be present between
FIB4 and BMI.

```{r, fig.cap = "Overview of metabolic syndrome categories."}
par(mfrow = c(1, 3))
split(n_data$metsym, n_data$na_sex) |>
    lapply(table) |>
    do.call(what = cbind) |>
    barplot()
vioplot(split(n_data$na_ager, n_data$metsym), main = "Metabolic syndrome",
        ylab = "age")
grid()
vioplot(split(n_data$BMI, n_data$metsym), main = "Metabolic syndrome",
        ylab = "BMI")
grid()
```

While no cases of metabolic syndrome seem to be balanced between male and female
participants and seem to be independent of the participants' ages, a strong
difference can be seen for cases and controls as well as for BMI.


# Classifying NAFLD/MASLD and diabetes

In this section we define different clinical phenotypes for study participants
based on the data collected in the study.


## Type 2 Diabetes

Type 2 Diabetes (T2B) is diagnosed for study participants using the information
and data from the study according to [@american_diabetes_association_2_2018]
[@resl_guidelines_2023]. The criteria are:

- HbA1C >= 6.5%
- OR Fasting glucose >= 126 mg/dl
- OR Anti-diabetic medication

We are loading in addition CHRIS baseline data to enable comparison of these
values in the baseline and the NAFLD study collection.

```{r, message = FALSE}
library(chrisData)
chris_med <- chrisData("drugs_summary")
rownames(chris_med) <- chris_med$AID
chris_med <- chris_med[n_data$aid_chris, ]

chris_gi <- chrisData("general")
rownames(chris_gi) <- chris_gi$AID
chris_gi <- chris_gi[n_data$aid_chris, ]

chris_int <- chrisData("interview")
rownames(chris_int) <- chris_int$AID
chris_int <- chris_int[n_data$aid_chris, ]

chris_lab <- chrisData("labdata")
rownames(chris_lab) <- chris_lab$AID
chris_lab <- chris_lab[n_data$aid_chris, ]

n_data$chris_dm <- chris_int$x0dm00 # self-reported diabetes diagnosis
n_data$chris_HbA1c <- chris_lab$x0lp06a # Glycated Hemoglobin
n_data$chris_glucose <- chris_lab$x0lp05 # fasting glucose
n_data$chris_dia_med <- chris_med$x0dd21 # diabetes medication
n_data$chris_sr_fasting <- chris_lab$x0bc12 # self-reported fasting
```

We next extract the same data from the NAFLD collection. We also update the
self-reported diabetes diagnosis considering all participants that reported
diabetes at baseline OR at NAFLD follow-up. Finally, we extract the information
on self-reported fasting and no drinking.

```{r}
med <- nafldData("Drug")
rownames(med) <- med$AID
med <- med[n_data$AID, ]

int <- nafldData("interview")
rownames(int) <- int$AID
int <- int[n_data$AID, ]

lab <- nafldData("Labdata")
rownames(lab) <- lab$AID
lab <- lab[n_data$AID, ]

#' self-reported diabetes considering also baseline
n_data$self_dm <- n_data$chris_dm
n_data$self_dm[which(int$nadm00 == "Yes")] <- "Yes"
n_data$HbA1c <- lab$nalp06a
n_data$glucose <- lab$nalp05
n_data$dia_med <- med$nadd21

#' self-reported fasting.
n_data$sr_fasting <- lab$nabc12
n_data$sr_no_drinking <- lab$nabc13
```

Glucose levels can be affected by the (missing) fasting status. Thus we next
evaluate glucose levels in participants with or without self-reported fasting
(and no drinking).

```{r, fig.cap = "Glucose concentrations for self-reported fasting (or no drinking) status.", fig.width = 12, fig.height = 6}
library(beeswarm)

par(mfrow = c(1, 2))
beeswarm(split(n_data$glucose, n_data$sr_fasting), pch = 21,
         bg = "#00000020", main = "Self-reported fasting",
         ylab = "Glucose")
grid()
abline(h = 126)
beeswarm(split(n_data$glucose, n_data$sr_no_drinking), pch = 21,
         bg = "#00000020", main = "Self-reported no drinking",
         ylab = "Glucose")
grid()
abline(h = 126)
```

Drinking did not affect the glucose levels, while for some participant that did
not fast the glucose levels are higher than the cut-off of 126 used in diabetes
diagnosis. Thus, we below only consider glucose levels for participants that did
not eat (self-reported) prior to the blood drawing.

And we are next diagnosing the NAFLD study participants based on the clinical
and medication information. We ignore the self-reported diagnosis.

```{r}
#' Diagnosing in NAFLD
t2d <- rep(FALSE, nrow(n_data))

#' Anti-diabetic medication: nadd21
t2d[which(n_data$dia_med == "Yes")] <- TRUE

#' Glycated hemoglobin %: nalp06a
t2d[which(n_data$HbA1c >= 6.5)] <- TRUE

#' fasting glucose: nalp05 + fasting
t2d[which(n_data$glucose >= 126 & n_data$sr_fasting == "Yes")] <- TRUE

#' finally define those without clear information.
no_lab <- is.na(n_data$HbA1c)
t2d[is.na(n_data$dia_med) & no_lab] <- NA

n_data$t2d <- t2d
```

The same criteria are applied to CHRIS baseline samples for the NAFLD study
participants.

```{r}
#' Diagnosing in CHRIS
t2d <- rep(FALSE, nrow(n_data))

#' Anti-diabetic medication: nadd21
t2d[which(n_data$chris_dia_med == "Yes")] <- TRUE

#' Glycated hemoglobin %: nalp06a
t2d[which(n_data$chris_HbA1c >= 6.5)] <- TRUE

#' fasting glucose: nalp05
t2d[which(n_data$chris_glucose >= 126 &
          n_data$chris_sr_fasting == "Yes")] <- TRUE

n_data$chris_t2d <- t2d
```

Note that for two participants no glycated hemoglobin measurement was
available: one of them was (in CHRIS baseline) diabetic and one was selected as
control (i.e., not diabetic). According to the available medication information
in the NAFLD substudy these two participants were still diagnosed the same
(i.e. one as diabetic and one not).

```{r, results = "asis"}
tmp <- data.frame(
    count = as.integer(table(n_data$t2d)),
    females = vapply(split(n_data$na_sex, n_data$t2d),
                     function(x) sum(x == "Female"), integer(1)),
    age = vapply(split(n_data$na_ager, n_data$t2d), mean, numeric(1)),
    BMI = vapply(split(n_data$BMI, n_data$t2d), mean, numeric(1), na.rm = TRUE)
)
library(pander)
pandoc.table(
    tmp, style = "rmarkdown",
    caption = paste0("Overview of NAFLD participants with and without type 2",
                     " diabetes (diagnosed based on laboratory parameters ",
                     "and medication)."))

```

Below we compare the total number of diagnosed T2D in baseline and NAFLD as well
as the self-reported diabetes status.

```{r, results = "asis"}
data.frame(status = c("No", "Yes"),
           NAFLD_T2D = as.integer(table(n_data$t2d)),
           CHRIS_T2D = as.integer(table(n_data$chris_t2d)),
           NAFLD_SR_DM = as.integer(table(n_data$self_dm)),
           CHRIS_SR_DM = as.integer(table(n_data$chris_dm))) |>
    pandoc.table(
        style = "rmarkdown",
        caption = paste0("Type 2 diabetes diagnosis for NAFLD study ",
                         "participants based on data from the NAFLD study ",
                         "(NAFLD) or CHRIS baseline samples (CHRIS). ",
                         "Columns SR indicate self-reported diagnosis."))
```

The number of positive cases are higher in CHRIS baseline than in the NAFLD
study because of the different instruments/protocols that were used to quantify
the glycated hemoglobin (HbA1c; data not shown here). Also interestingly,
self-reported diabetes is much higher than diagnosed (type 2) diabetes in NAFLD.

HbA1c measurements were conducted on blood samples for all NAFLD participants on
a Tosoh Automated Glycohemoglobin Analyzer HLC-723 G8 (TOSOH CORPORATION
SHIBA-KOEN FIRST BLDG) while for the measurement on blood samples from the NAFLD
samples the ARKRAY ADAMS A1c HA-8180V instrument was used (instrument change was
performed 2015-05-15). The HbA1c (%) measurements are systematically (slightly)
higher with the old instrument.

For the present analysis we use however the T2D diagnosis performed on the
samples collected during the NAFLD study - for comparison with baseline data we
will however need to normalize the values.

We next evaluate distribution of age, BMI, sex etc for participants with and
without T2D.

```{r, fig.cap = "Age distribution for female and male study participants with and without T2D."}
f <- paste0(n_data$na_sex, n_data$t2d)
vioplot(split(n_data$na_ager, f), main = "Age/T2D", ylab = "age [years]")
grid()
```

Age distribution is similar for female or male study participants with or
without T2D. We next evaluate the distribution of BMI values for T2D and control
(again stratified by sex).

```{r, fig.cap = "Distribution of BMI values for female and male study participants with and without T2D."}
vioplot(split(n_data$BMI, f), main = "BMI/T2D", ylab = "BMI")
grid()
```

Participants with T2D have a consistently higher BMI, both in female and male
participants. We next evaluate the liver stiffness measurements (removing
measurements that don't fulfill the quality requirements).

```{r, fig.cap = "Distribution of liver stiffness values for female and male study participants with and without T2D."}
lstiff <- n_data$liver_stiffness
lstiff[n_data$liver_stiffness_iqr > 30] <- NA
vioplot(split(log2(lstiff), f), main = "liver stiffness/T2D",
        ylab = expression(log[2]~liver~stiffness))
grid()
```

Participants with T2D have slightly higher liver stiffness measurements than
those without. This might however be a consequence of their higher BMI.

```{r, fig.cap = "Distribution of visceral fat values for female and male study participants with and without T2D."}
vioplot(split(n_data$visc_fat, f), main = "Visceral fat/T2D",
        ylab = expression(visceral~fat))
grid()
```

Again, participants with T2D have on average higher visceral fat.

```{r, fig.cap = "Distribution of FIB4 values for female and male study participants with and without T2D."}
vioplot(split(n_data$fib4, f), main = "FIB4/T2D",
        ylab = expression(FIB4))
grid()
```

FIB4 values are higher for participants without T2D.


## Categorization of steatotic liver disease

Clinical diagnosis of metabolic dysfunction associated steatotic liver disease
(MASLD [@rinella_multisociety_2024], formerly NAFLD
[@european_association_for_the_study_of_the_liver_easl_easl-easd-easo_2016])
is based on a range of parameters:

1) Presence of steatotic liver disease (**SLD**): sonographic evaluation,
   steatose score of 2 - 3.

2) Presence of cardiometabolic risk factors/criteria (at least one out of 5):
   a) BMI >= 25 OR waist circumference > 94 (M) 80 (F) cm.
   b) Fasting glucose >= 100 mg/dl OR HbA1c >= 5.7% OR type 2 diabetes OR
      treatment for type 2 diabetes.
   c) Blood pressure >= 130/85 mmHg OR specific antihypertensive drug treatment.
   d) Triglycerides >= 150 mg/dl OR lipid lowering treatment.
   e) HDL <= 40 (M) <= 50 (F) mg/dl OR lipid lowering treatment.

3) Rule out alcoholic liver disease or MetALD. Alcohol consumption < 40g/day
   (one *standard drink* is 10g alcohol).

This diagnosis includes also metabolic syndrome (MetS) which is diagnosed based
on the presence of 3 risk factors:

- (must have): central obesity (WC >= 94 (M) >= 80 (F) cm.
- (two out of four):
  - TG >= 150 mg/dl OR specific treatment for this lipid abnormality (**d**).
  - HDL < 40 (M) 50 (F) mg/dl OR specific treatment for this lipid abnormality
    (**e**).
  - systolic BP >= 130 or diastolic >= 85 mmHG OR treatment for hypertension
    (**c**).
  - fasting glucose >= 100 mg/dl OR diagnosed T2D (**b**).

Below we perform the diagnosis based on the above criteria.

Presence of steatotic liver disease.

```{r}
#' steatose score: naab14, visual scoring of liver fat.
n_data$SLD <- n_data$visual_fat > 1
summary(n_data$SLD)
```

**a)** BMI >= 25 or waist circumference > 94 (M) 80 (F) cm.

```{r}
#' Waist circumference
n_data$waist_circumference <- n_anthro[n_data$AID, "naan12"]
#' Classify
n_data$cardiometabolic_a <-
    n_data$BMI >= 25 |
    (n_data$waist_circumference >= 94 & n_data$na_sex == "Male") |
    (n_data$waist_circumference >= 80 & n_data$na_sex == "Female")
```

A quite large proportion of participants fulfill this criteria - also maybe not
unexpected based on the selection criteria for study participants.

**b)** Fasting glucose >= 100 mg/dl OR HbA1c >= 5.7% OR type 2 diabetes OR
treatment for type 2 diabetes (already included in the diagnosis of T2D
performed in the previous section).

```{r}
#' Classify
n_data$cardiometabolic_b <-
    n_data$glucose >= 100 |
    n_data$HbA1c > 5.7 |
    n_data$t2d
```

**c)** Blood pressure >= 130/85 mmHg OR specific antihypertensive drug
treatment.

```{r}
#' Add blood pressure measurements.
n_data$systolic_bp <- n_anthro[n_data$AID, "nabp01"] # systolic BP [mm/Hg]
n_data$diastolic_bp <- n_anthro[n_data$AID, "nabp02"] # diastolic BP [mm/Hg]
n_data$systolic_bp[n_data$systolic_bp < 0] <- NA_real_
n_data$diastolic_bp[n_data$diastolic_bp < 0] <- NA_real_
#' Add blood pressure lowering therapy
n_data$pb_med <- med$nadd20 # blood pressure lowering agents
#' Classify
n_data$cardiometabolic_c <-
    n_data$systolic_bp >= 130 |
    n_data$diastolic_bp >= 85 |
    n_data$pb_med == "Yes"
```

**d)** Triglycerides >= 150 mg/dl OR lipid lowering treatment.

```{r}
#' triglycerides [mg/dl]
n_data$triglycerides <- lab[n_data$AID, "nalp19"]
n_data$lm_med <- med$nadd22 # lipid modifying agents
n_data$ms_med <- med$nadd23 # metabolic syndrome medication
#' Classify
n_data$cardiometabolic_d <-
    n_data$triglycerides >= 150 |
    n_data$lm_med == "Yes"
```

**e)** HDL <= 40 (M) <= 50 (F) mg/dl OR lipid lowering treatment.

```{r}
#' HDL [mg/dl]: nalp17
n_data$HDL <- lab[n_data$AID, "nalp17"]
#' Classify
n_data$cardiometabolic_e <-
    (n_data$HDL <= 40 & n_data$na_sex == "Male") |
    (n_data$HDL <= 50 & n_data$na_sex == "Female") |
    n_data$lm_med == "Yes"
```

Summary of the numbers of numbers of cases for each sub-criteria is provided
below.

```{r, results = "asis"}
cardiom <- n_data[, c("cardiometabolic_a", "cardiometabolic_b",
                      "cardiometabolic_c", "cardiometabolic_d",
                      "cardiometabolic_e")]
lapply(cardiom, summary) |>
    do.call(what = rbind) |>
    pandoc.table(
        style = "rmarkdown",
        caption = paste0("Number of cases for each cardiometabolic sub-criteria"))

```

Most participants scored positive in subcategories **c** (hypertension) and
**a** (BMI and waist circumference). We visually inspect the overlap between the
sub-criteria also with an upset plot.

```{r, fig.cap = "Overview and overlap of participants scoring positive in the individual cardiometabolic sub-criteria."}
library(UpSetR)
cardiom <- as.data.frame(lapply(cardiom, as.integer))
upset(cardiom, order.by = "degree")
```

A high number of participants score positive in all sub-criteria.

**3)** We next evaluate alcohol consumption, based on the (self reported) GA2LEN
FFQ and the derived daily alcohol consumption (variable *nanu016*).

```{r}
ffq <- nafldData("FFQ")
rownames(ffq) <- ffq$AID
ffq <- ffq[n_data$AID, ]
n_data$daily_total_alcohol <- ffq$nanu016
```

The distribution of daily total alcohol consumption for female and male
participants is shown below.

```{r, fig.cap = "Distribution of total daily alcohol for male and female participants."}
vioplot(split(n_data$daily_total_alcohol, n_data$na_sex),
        main = "Daily total alcohol", ylab = "daily total alcohol [g]")
grid()
abline(h = 40)
```

In total `r sum(n_data$daily_total_alcohol > 40, na.rm = TRUE)` participants
have a higher alcohol consumption (all male).
For `r sum(is.na(n_data$daily_total_alcohol))` no alcohol consumption is
available.

At last we combine the scores from the sub-criteria to diagnose MASLD based on
the criteria 1), 2) and 3) from the definition.

1) Presence of steatotic liver disease: sonographic evaluation, steatose score
   of 2 - 3.

2) Presence of cardiometabolic risk factors/criteria (at least one out of 5):
   a) BMI >= 25 OR waist circumference > 94 (M) 80 (F) cm.
   b) Fasting glucose >= 100 mg/dl OR HbA1c >= 5.7% OR type 2 diabetes OR
      treatment for type 2 diabetes.
   c) Blood pressure >= 130/85 mmHg OR specific antihypertensive drug treatment.
   d) Triglycerides >= 150 mg/dl OR lipid lowering treatment.
   e) HDL <= 40 (M) <= 50 (F) mg/dl OR lipid lowering treatment.

3) Alcohol consumption < 40g/day. Discriminate between MASLD and MetALD. We
consider participants without alcohol information as MASLD cases.

```{r}
#' MASLD:
#' - steatotic liver disease
#' - one out of 5 of a), b), c), d) and e)
n_data$MASLD <-
    n_data$SLD &
    rowSums(n_data[, c("cardiometabolic_a", "cardiometabolic_b",
                       "cardiometabolic_c", "cardiometabolic_d",
                       "cardiometabolic_e")], na.rm =TRUE) > 0
n_data$MetALD <- n_data$MASLD & n_data$daily_total_alcohol >= 40
#' - alcohol consumption < 40g/day; participants with missing values are,
#'   according to other questionnaires, no heavy drinkers and are thus also
#'   considered MASLD
n_data$MASLD[which(n_data$daily_total_alcohol >= 40)] <- FALSE
summary(n_data$MASLD)
```

The MASLD positive/negative cases are thus almost balanced in the present
study. Also, maybe interesting, there were in total `r sum(n_data$SLD)`
individuals scored positive for steatotic liver disease, but 2 of them did not
fulfill the additional criteria for MASLD.

Also, some participants score positive for MetALD:

```{r}
summary(n_data$MetALD)
```

In contrast to the different binary variables (SLD, MetALD and MASLD) we create
a single categorical variable *sld_type* with the categories:

- **none**: participants negative for steatotic liver disease.
- **SLD**: participants with a steatotic liver disease, but negative for
  cardiometabolic criteria. This includes potentially cryptogenic SLD, alcoholic
  liver disease (ALD), DILI or monogenic diseases.
- **MASLD**: participants with SLD, positive for cardiometabolic criteria but
  low alcohol consumption (below 40g/day). This includes also 8 participants
  **without information on alcohol consumption**. These participants did not
  provide the GA2LEN FFQ (hence missing values for quantified alcohol
  consumption), but according to alcohol-related information in other
  questionnaire they are not heavy drinkers and are thus added to this category.
- **MetALD**: participants with SLD, positive for cardiometabolic criteria but
  daily alcohol consumption > 40g/day.

Below we define this variable.

```{r}
#' Define the SDL categorical variable
n_data$sld_type <- "none"
n_data$sld_type[n_data$SLD] <- "SLD"
n_data$sld_type[n_data$MASLD] <- "MASLD"
n_data$sld_type[which(n_data$MetALD)] <- "MetALD"
n_data$sld_type <- factor(
    n_data$sld_type, levels = c("none", "MASLD", "MetALD", "SLD"))
summary(n_data$sld_type)
```

Note that all MetALD cases are exclusively male participants.

We next evaluate distribution of age, BMI and other variables for participants
in the various SLD categories.

```{r, fig.cap = "Age distribution for female and male study participants with MASLD, SLD, MetALD or non-steatotic liver."}
f <- n_data[n_data$na_sex == "Female", ]
f$sld_type <- droplevels(f$sld_type)
m <- n_data[n_data$na_sex == "Male", ]
par(mfrow = c(1, 2))
vioplot(split(f$na_ager, f$sld_type), main = "Age, females",
        ylab = "age [years]")
grid()
vioplot(split(m$na_ager, m$sld_type), main = "Age, males",
        ylab = "age [years]")
grid()

```

Male participants with MASLD seem to be on average younger than apparently
healthy controls. We next evaluate the distribution of BMI values for the SLD
categories.

```{r, fig.cap = "Distribution of BMI values for female and male study participants with MASLD, SLD, MetALD or non-steatotic liver."}
par(mfrow = c(1, 2))
vioplot(split(f$BMI, f$sld_type), main = "BMI, females", ylab = "BMI")
grid()
vioplot(split(m$BMI, m$sld_type), main = "BMI, males", ylab = "BMI")
grid()
```

Participants with MASLD have a consistently higher BMI, both in female and male
participants. We next evaluate the liver stiffness measurements (removing
measurements that don't fulfill the quality requirements).

```{r, fig.cap = "Distribution of liver stiffness values for female and male study participants with MASLD, SLD, MetALD or non-steatotic liver."}
lstiff <- f$liver_stiffness
lstiff[f$liver_stiffness_iqr > 30] <- NA
par(mfrow = c(1, 2))
vioplot(split(log2(lstiff), f$sld_type), main = "liver stiffness, females",
        ylab = expression(log[2]~liver~stiffness))
grid()
lstiff <- m$liver_stiffness
lstiff[m$liver_stiffness_iqr > 30] <- NA
vioplot(split(log2(lstiff), m$sld_type), main = "liver stiffness, males",
        ylab = expression(log[2]~liver~stiffness))
grid()
```

Participants with (any) steatotic liver disease have higher liver stiffness
measurements than those without.

```{r, fig.cap = "Distribution of visceral fat values for female and male study participants with MASLD, SLD, MetALD or non-steatotic liver."}
par(mfrow = c(1, 2))
vioplot(split(f$visc_fat, f$sld_type), main = "Visceral fat, females",
        ylab = expression(visceral~fat))
grid()
vioplot(split(m$visc_fat, m$sld_type), main = "Visceral fat, males",
        ylab = expression(visceral~fat))
grid()
```

Again, participants with MASLD have on average higher visceral fat.

```{r, fig.cap = "Distribution of FIB4 values for female and male study participants with MASLD, SLD, MetALD or non-steatotic liver."}
par(mfrow = c(1, 2))
vioplot(split(f$fib4, f$sld_type), main = "FIB4, females",
        ylab = expression(FIB4))
grid()
vioplot(split(m$fib4, m$sld_type), main = "FIB4, males",
        ylab = expression(FIB4))
grid()
```

FIB4 values are higher for participants without MASLD.

At last we evaluate also the T2D cases with and without any steatotic liver
disease.

```{r}
table(n_data[, c("t2d", "sld_type")])
```

About half of the participants with MASLD are also positive for T2D. In the
present data set, most of the T2D positive cases do also have MASLD.


## Classification of Metabolic Syndrome

Metabolic syndrome (MetS) is diagnosed based on the presence of 3 risk factors
(and partially overlaps with the MASLD definition above):

- (must have): central obesity (WC >= 94 (M) >= 80 (F) cm.
- (two out of four):
  - TG >= 150 mg/dl OR specific treatment for this lipid abnormality (**d**).
  - HDL < 40 (M) 50 (F) mg/dl OR specific treatment for this lipid abnormality
    (**e**).
  - systolic BP >= 130 or diastolic >= 85 mmHG OR treatment for hypertension
    (**c**).
  - fasting glucose >= 100 mg/dl OR diagnosed T2D (**b**).

```{r}
n_data$MetS <-
    ((n_data$waist_circumference >= 94 & n_data$na_sex == "Male") |
     (n_data$waist_circumference >= 80 & n_data$na_sex == "Female")) &
    rowSums(n_data[, c("cardiometabolic_b", "cardiometabolic_c",
                       "cardiometabolic_d", "cardiometabolic_e")],
            na.rm =TRUE) > 1
```

The overlap between participants with metabolic syndrome and MASLD are shown
below. Also, almost 3 times more participants with MASLD score also positive for
MetS.

```{r}
table(n_data[, c("MetS", "sld_type")])
```

For the biggest part, individuals in the present data set with MetS are also
positive for MASLD.

```{r}
table(n_data[, c("t2d", "MetS")])
```

Again, we next evaluate distribution of age, BMI, sex etc for participants with
and without MetS.

```{r, fig.cap = "Age distribution for female and male study participants with and without MetS."}
f <- paste0(n_data$na_sex, n_data$MetS)
vioplot(split(n_data$na_ager, f), main = "Age/MetS", ylab = "age [years]")
grid()
```

For females the average age of MetS positive cases is higher while for male
participant the opposite effect can be observed. We next evaluate the
distribution of BMI values for MetS and control (again stratified by sex).

```{r, fig.cap = "Distribution of BMI values for female and male study participants with and without MetS."}
vioplot(split(n_data$BMI, f), main = "BMI/MetS", ylab = "BMI")
grid()
```

Participants with MetS have a consistently higher BMI, both in female and male
participants. We next evaluate the liver stiffness measurements (removing
measurements that don't fulfill the quality requirements).

```{r, fig.cap = "Distribution of liver stiffness values for female and male study participants with and without MetS."}
lstiff <- n_data$liver_stiffness
lstiff[n_data$liver_stiffness_iqr > 30] <- NA
vioplot(split(log2(lstiff), f), main = "liver stiffness/MetS",
        ylab = expression(log[2]~liver~stiffness))
grid()
```

Participants with MetS have slightly higher liver stiffness measurements than
those without. This might however be a consequence of their higher BMI.

```{r, fig.cap = "Distribution of visceral fat values for female and male study participants with and without MetS."}
tmp <- split(n_data$visc_fat, f)
tmp <- lapply(tmp, function(z) z[!is.na(z)])
tmp <- tmp[lengths(tmp) > 0]
vioplot(tmp, main = "Visceral fat/MetS",
        ylab = expression(visceral~fat))
grid()
```

Again, participants with MetS have on average higher visceral fat.

```{r, fig.cap = "Distribution of FIB4 values for female and male study participants with and without MetS."}
vioplot(split(n_data$fib4, f), main = "FIB4/MetS",
        ylab = expression(FIB4))
grid()
```

FIB4 values are higher for participants without MetS.

```{r}
#' At last we export the phenodata defined in this analysis.
library(writexl)
tmp <- n_data[, c("AID", "aid_chris", "na_ager", "na_sex",
                  "case", "BMI", "body_fat", "visc_fat", "liver_stiffness",
                  "liver_stiffness_iqr", "liver_fat", "visual_fat", "fib4",
                  "self_dm", "HbA1c", "glucose", "dia_med", "sr_fasting",
                  "sr_no_drinking", "t2d", "SLD", "waist_circumference",
                  "systolic_bp", "diastolic_bp", "pb_med", "triglycerides",
                  "lm_med", "ms_med", "HDL", "daily_total_alcohol", "MASLD",
                  "MetALD", "sld_type", "MetS")]
#' renaming some columns.
colnames(tmp) <- c("AID", "AID_chris", "Age", "Sex",
                  "case", "BMI", "body_fat", "visc_fat", "liver_stiffness",
                  "liver_stiffnes_iqr", "liver_fat", "visual_fat", "FIB4",
                  "self_DM", "HbA1c", "glucose", "med_DM", "fasting",
                  "no_drinking", "T2D", "SLD", "waist_circumference",
                  "systolic_BP", "diastolic_BP", "med_BP", "triglycerides",
                  "med_lip", "med_MS", "HDL", "daily_total_alcohol", "MASLD",
                  "MetALD", "sld_type", "MetS")
write_xlsx(tmp, path = "data/masld_phenotypes.xlsx")

```



# NAFLD omics data overview


Get the IDs of NAFLD participants and their respective CHRIS AID.

```{r, message = FALSE}
library(chrisData)

gi <- chrisData("general")

nafld <- gi[which(gi$x0_nafld == "Yes"), ]
```

Participation year (in CHRIS baseline) for the `r nrow(nafld)` NAFLD study
participants.

```{r}
barplot(table(nafld$x0_examy))
```

We determine the number of participants with available CHRIS targeted
metabolomics as well as CHRIS ScanningSWATH and Somalogic proteomics data from
the baseline and NAFLD collections.

```{r, message = FALSE}
library(tidyfr)
soma <- data_module(name = "chris_somalogic", version = "1.0.3.2", path = ".")
scan <- data_module(name = "chris_scanningswath_proteins", version = "1.0.0.2",
                    path = ".")
p180 <- data_module(name = "metabolomics_p180", version = "1.0.1.2", path = ".")

soma_aid <- rownames(data(soma))
scan_aid <- rownames(data(scan))
p180_aid <- rownames(data(p180))
```

```{r, results = "asis"}
library(pander)
tmp <- data.frame(baseline = c(sum(nafld$AID %in% p180_aid),
                               sum(nafld$AID %in% scan_aid),
                               sum(nafld$AID %in% soma_aid)),
                  NAFLD = c(sum(nafld$aid_nafld %in% p180_aid),
                            sum(nafld$aid_nafld %in% scan_aid),
                            sum(nafld$aid_nafld %in% soma_aid)))
rownames(tmp) <- c("targeted metabolomics",
                   "ScanningSWATH proteomics",
                   "SomaLogic proteomics")
pandoc.table(tmp, style = "rmarkdown",
             caption = paste0("Number of NAFLD study participants (n = ",
                              nrow(nafld), ") with available omics data"))
```

Note: these numbers don't include any of the drop outs up to version 2 (July
2023).


```{r}
do <- read.table(system.file("txt", "drop-outs.txt", package = "chrisData"),
                 sep = "\t", header = TRUE, colClasses = "character")
any(do$aid %in% nafld$AID)
any(do$aid %in% nafld$aid_nafld)
```


# Questions/comments

- [X] DONE: check instruments between baseline and NAFLD study - was there a
      change that could explain differences in HbA1c?
- [X] TODO: discuss with Christoph, what is his opinion on that? Use the data
      from NAFLD. In a clinical settings, two measurements are required for
      diagnosis.
- [X] TODO: how to normalize the values. No need as long as we don't compare
      against baseline data.
- [X] TODO: should we consider a categorical variable with CTRL, SLD, MASLD,
      MetALD instead just a binary for MASLD?
- [X] TODO: how to categorize potential MASLD cases without alcohol information?
      (in total 8). According to other alcohol related information from other
      questionnaires they are no heavy drinkers and thus added to the MASLD
      category.


# Session Information

```{r}
sessionInfo()
```

# References