---
title: "Steatosis - check MASH"
author: "Marilyn De Graeve, Johannes Rainer, Christoph Grander"
affiliation: "Eurac Research, Bolzano, Italy"
date: "2025-02-27"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
---

**Modified**: `r file.info("NAFLD_MASH.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r style, message = FALSE, echo = FALSE, warning = FALSE, results = "asis"}
library("BiocStyle")
library("kableExtra")
library("knitr")
library("rmarkdown")
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now, units = "secs")
      # return a character string to show the time
      paste("    Time for this code chunk to run:", round(res,
        2), "seconds")
    }
  }
}))
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7, time_it = TRUE)

```

# Introduction

During this workflow, exploratory analyis (Spearman correlations) are performed
on the relatively quantified Scanning SWATH proteomics data (148 proteins) from
the metabolic dysfunction-associated steatotic liver disease (MASLD)
cohort. MASLD, previously known as NAFLD is the most chronic liver disease,
affecting more than 30% of global population. The MAFLD cohort is a subset of
approximatley 353 # participants from the Cooperative Health Research in South
Tyrol (CHRIS) study, which were requested to return for a second resampling and
additional phenotyping. The NAFLD cohort represent an unhealty group with
various degrees of steatotic (non-alcoholic) liver disease classifications. See
[nafld_data_overview.Rmd](nafld_data_overview.Rmd) (respectively the rendered
html report) for an overview of the data set and the definition of the
phenotype.

This documents describes the association analysis between circulating
metabolites, lipids and proteins with liver steatosis.

See the `nafld_data_overview 1.html` report for further
information regarding the SLD sub-classification.

Regarding steatosis, i.e. Abnormal retention of fat (lipids) within a cell or
organ. In this scope: liver steatosis.
The variables containing this parameter is:

Presence of steatotic liver disease (SLD): sonographic evaluation, steatose
score of 2 – 3.

- *naab13*: liver fat. Can be binarized at an absolute value of 288. "liver_fat"
- *naab14*: subjective, visual evaluation of the liver fat. Categorical from 1
  to 3. Values 2 and 3 can be considered steatotic. "visual_fat"

This parameter will be used as the target for a biomarker search in *MASLD*
participants.


# Setup

## Directories

```{r directories, echo = FALSE}
#' General settings
filename <- "NAFLD_MASH"

#' Path to save images to; remove if exists.
IMAGE_PATH <- paste0("images/", filename, "/")
dir.create(IMAGE_PATH, recursive = TRUE)

#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)

#' Path to the data
DATA_PATH <- '/home/mdegraeve/Documents/Files/Work_Eurac/Data/NAFLD' #FINDME!
#DATA_PATH <- "."


DATASET <- 'CHRIS-NAFLD, none and MASLD'

```

## Packages

```{r packages, message=FALSE}
#' Load libraries
library(chrisUtils)
library(chrisData)
library(DT)
library(ggfortify)
library(ggplot2)
library(pander)
library(pheatmap)
library(plotly)
library(RColorBrewer)
library(readxl)
library(reshape2)
library(tidyfr)
library(vioplot)
library(writexl)

```


# Data import

## Circulating proteins (tprot) dataset (Scanning SWATH)

The data from proteins was converted into the TDF format and can hence be
loaded and imported using the `tidyfr` R package. Now we read the full data
which includes general information for each measurement.

```{r load_prot_dataset, message = FALSE}
#' load TDF using tidyfr
#list_data_modules(path = DATA_PATH)
md1 <- data_module(name = "chris_scanningswath_proteins", version = "1.0.0.3",
                   path = DATA_PATH)
prot_data <- data(md1)
prot_ann <- labels(md1)
stopifnot(all(colnames(prot_data) == rownames(prot_ann)))
colnames(prot_data) <- prot_ann$description
rownames(prot_ann) <- prot_ann$description

#' define the columns with protein concentrations
proteins <- rownames(prot_ann)[prot_ann$type != "categorical"]
proteins <- rownames(prot_ann)[prot_ann$Genes != ""]

#' select proteins dataframe
prot_data <- prot_data[, proteins]

```

The loaded data contains the samples as rows, and the protein names as
colums. The measurements of the `prot_data` object are the relative
concentrations of protein per sample. No data transformation is performed as the
abundances are already reported in log2 scale, a normal distribution of the
proteomics is assumed.

Before proceeding with the data analysis we remove the POOL samples. In addition
we restrict to samples from the NAFLD study. The number of samples and proteins
are:

```{r subset to NAFLD set, message = FALSE, warning = FALSE}
prot_data <- prot_data[grep("^008", rownames(prot_data)), ]
dim(prot_data)
```


## Targeted metabolomics (tmetab) dataset (Biocrates p180 kit)

The data from metabolites was converted into the TDF format and can hence be
loaded and imported using the `tidyfr` R package. Now we read the full data
which includes general information for each measurement.

```{r load_tmetb_dataset, message = FALSE}
#' load TDF using tidyfr
#list_data_modules(path = DATA_PATH)
p180 <- data_module(name = "metabolomics_p180", version = "1.0.1.3",
                    path = DATA_PATH)
met_data <- data(p180)
met_ann <- labels(p180)
stopifnot(all(colnames(met_data) == rownames(met_ann)))
colnames(met_data) <- met_ann$description
rownames(met_ann) <- met_ann$description

#' define the columns with metabolite concentrations
metabolites <- rownames(met_ann)[met_ann$type != "categorical"]

#' select metabolites dataframe
met_data <- met_data[, metabolites]
#' Remove the individuals with more than 80% missing values
rem <- apply(met_data, MARGIN = 1, function(z) {
    sum(is.na(z)) > length(z) * 0.8
})
met_data <- met_data[!rem, ]


#' Impute missing values using an uniform distribution, after gap filling
na_unidis <- function(z) {
    na <- is.na(z)
    if (any(na)) {
        min = min(z, na.rm = TRUE)
        z[na] <- runif(sum(na), min = min/2, max = min)
    }
    z
}

#' Row-wise impute missing values and add the data as a new assay
tmp <- apply(met_data, MARGIN = 1, na_unidis)
met_data <- t(tmp)
```

The loaded data contains the samples as rows, and the metabolite names as
colums. The measurements of the `met_data` object are the absolute
concentrations of metabolite per sample, in natural scale.

To ensure a normal distribution of the metabololits, we perform log2
transformation.

```{r m log2-transformation, message = FALSE}
#' log2 scaling of absolute abundances metabolites
met_data <- log2(met_data)
```

Before proceeding with the data analysis we remove the POOL samples. In addition
we restrict to samples from the NAFLD study. The number of samples and
metabolites are listed below.

```{r m subset to NAFLD set, message = FALSE, warning = FALSE}
met_data <- met_data[grep("^008", rownames(met_data)), ]
dim(met_data)
```


## Diet GA2LEN data

Dietary information, i.e. GA2LEN food frequency questionaires, from the MASLD
study were loaded and included in the analysis.

DO NOT RUN next 2 chuncks, we use the newly created dietary indices from Essi
(10/07/2025). 

```{r diet-data, eval = FALSE}
#' 1. load the metadata regarding the GA2LEN FFQs
library(nafldData)
#nafldTraitAnnotations()|> head()

galan <- nafldData("FFQ")
rownames(galan) <- galan$AID

#variables, subset:
#naff001 - naff229 are galan FFQ, and the rest = not!

galan <- galan[, 1:230] #incl AID
ffq_items <- colnames(galan)[2:230]

#' change to numerical scores instead of FFQ answers
#' solve using category encoder, in R see:
#' https://www.r-bloggers.com/2020/02/a-guide-to-encoding-categorical-features-using-r/

#' select categor variables
categor_variables <- NULL
for (x in ffq_items){
  if (class(galan[[x]]) == 'factor' | class(galan[[x]]) == 'character'){  #factor and character
    categor_variables <- c(categor_variables, x)
  }
}

#' using ordinal encoding on categorical vars
encode_ordinal <- function(x) {
  if (is.factor(x)) {
    order <- levels(x)   #retain defined factor level order, wo NA but next step ignored so fine
    x <- as.numeric(factor(x, levels = order, exclude = NA)) #as numeric, keep NA in dataset
  } else if (is.character(x)) {
    print(x) #*
  }
  x
}

#* text info: NOT
    #naff246 - naff249b
    #naff260 - naff263a

#' testing
summary(galan$naff001)

#' apply encoding on categor vars
galan[, categor_variables] <- lapply(galan[, categor_variables], encode_ordinal)
```

Corresponding fake CV-values (no 'pool' samples) for diet.

```{r kora-cv, eval = FALSE}
#' 1. load the metadata regarding the GA2LEN FFQs
library(nafldData)
#nafldTraitAnnotations()|> head()

nafldAnn <- nafldTraitAnnotations()
kora_ann <- nafldAnn[ffq_items, ]
rownames(kora_ann) <- ffq_items
#head(kora_ann)

#' TEMP, just accept all features now as having good cv
cv_values <- 0
#' save in the met_ann as column
kora_ann$cv_qc_chris <- cv_values
```

The newly created dietary indices from Essi are saved under 
NAFLD_dietary_indices_data_Marilyn. And note in the
lm: "you should also correct your models for total energy intake (just have it
as covariate in the model, nanu006a)". TODO!!!

```{r NAFLD-indices}
#' load the NAFLD_dietary_indices_data
handle <- file.path(DATA_PATH, 'GALAN', 'NAFLD_dietary_indices_data_Marilyn.txt')
galan <- read.table(handle, header = TRUE, sep = "\t")  |>
    as.data.frame()

#'galan is with , instead of decimal point so change
comma_to_point <- function(x) {
  x <- scan(text = x, dec = ",", sep = ".")
  x
}
galan <- apply(galan, MARGIN = 2, comma_to_point) |>
    as.data.frame()
rownames(galan) <- paste0('00', as.character(galan$AID))
#dim(galan)
#colnames(galan)
#galan$AID[1:5]
#head(galan)

#replace the aids columns with 00 start and as a string
galan$AID <- rownames(galan)

```

Also add the 4 macros and add to the `ffq_items`.

```{r NAFLD_food_macros}
library(nafldData)
#need these one
#nanu003 (protein), nanu004 (total fat), nanu005 (total carbs), nanu019b (saturated fatty acids), nanu023b (monounsaturated fatty acids), nanu025b (polyunsaturated fatty acids), nanu028b (trans fatty acids)

ffq <- nafldData("FFQ")
rownames(ffq) <- ffq$AID

galan$protein <- ffq[rownames(galan), 'nanu003']
galan$total_fat <- ffq[rownames(galan), 'nanu004']
galan$total_carbs <- ffq[rownames(galan), 'nanu005']
galan$saturated_fatty_acids <- ffq[rownames(galan), 'nanu019b']
galan$monounsaturated_fatty_acids <- ffq[rownames(galan), 'nanu023b']
galan$polyunsaturated_fatty_acids <- ffq[rownames(galan), 'nanu025b']
galan$trans_fatty_acids <- ffq[rownames(galan), 'nanu028b']

#' variables
ffq_items <- colnames(galan)[2:ncol(galan)] #excl AID col

```

## Metadata (excel)

All available trait (phenotype) data was defined in the
[nafld_data_overview.Rmd](nafld_data_overview.Rmd), is loaded below and merged
with the log2 transformed proteomics data of the NAFLD cohort.

```{r add phenotype dataa}
#' load xlsx file instead of below
handle <- 'masld_phenotypes.xlsx'
masld_phenotype_df <- read_excel(file.path(DATA_PATH, handle)) |>
    as.data.frame()
rownames(masld_phenotype_df) <- masld_phenotype_df$AID
#head(masld_phenotype_df, 5)
metadata <- colnames(masld_phenotype_df)[3:ncol(masld_phenotype_df)] #info wo AIDs
#dim(masld_phenotype_df)

#' make sure sld_type is a factor with group 'none' as the baseline
masld_phenotype_df$sld_type <- factor(masld_phenotype_df$sld_type,
                                      levels = c('none', 'MASLD',
                                                 'MetALD', 'SLD'))
print('the sld types are:')
levels(masld_phenotype_df$sld_type)
print('nr of samples per sld types:')
table(masld_phenotype_df$sld_type)
```


# Merge tmetab, tprot, kora and metadata

Merging all data tables. Note that restrict the data to participants for which
we have proteomics **and** metabolomics data. This also ensures that any
potential study drop-outs are no longer considered in the data (as their
proteomics or metabolomics data was removed).

There were less AIDs present in the dietary data, so we restrict with the data
to participants for which we have proteomics **and** metabolomics **and** 
dietary data. ??? TODO findme!!!

```{r merge-dfs}
#' keep only AIDs of overlapping
AIDs1 <- intersect(rownames(prot_data), rownames(met_data))
AIDs2 <- intersect(rownames(masld_phenotype_df), rownames(galan))
AIDs <- intersect(AIDs1, AIDs2)

#' Subset all datasets to the same IDs
prot_data <- prot_data[AIDs, ]
met_data <- met_data[AIDs, ]
masld_phenotype_df <- masld_phenotype_df[AIDs, ]
galan <- galan[AIDs, ]

#' merge with proteomics data
prot_data$AID <- rownames(prot_data)
prot_data <- merge(prot_data, masld_phenotype_df, by = 'AID')

#' merge all sets
met_data <- as.data.frame(met_data)
met_data$AID <- rownames(met_data)
met_prot_data <- merge(prot_data, met_data, by = 'AID')

#' merge with metabolomics data only
met_data <- merge(met_data, masld_phenotype_df, by = 'AID')

#' merge all sets, kora too
met_prot_data <- as.data.frame(met_prot_data)
rownames(met_prot_data) <- met_prot_data$AID
gal_met_prot_data <- merge(met_prot_data, galan, by = 'AID') #kora has only 001 AIDs
rownames(gal_met_prot_data) <- gal_met_prot_data$AID
#gal_met_prot_data[1:5, 1:5]

#' write to file for manuel inspection
handle <- 'data/gal_met_prot_data.txt'
write.table(gal_met_prot_data, file = handle, sep = "\t",
            row.names = FALSE, col.names = TRUE)
```

All data sets are subset to `r length(AIDs)` study participants.


# Clean data and restrict to relevant participants

The data is further processed and cleaned to remove unreliable
measurements. Subsequently the data is restricted to relevant participants. We
thus next:

- remove unreliable liver fat measurements
- exclude participants with MetALD and SLD to ensure a more homogeneous data set
- remove participants with missing values in relevant variables
- define variables and phenotype(s) for the analysis

Add the information from the [nafld_data_overview.Rmd](nafld_data_overview.Rmd)
report regarding the steatosis, using the parameter naab11
(higher values are worse). Values for participants that have a naab12
(liver stiffness IQR) > 30 should be either removed or taken with care. Below we
remove these unreliable measurements.

```{r load-liver-param2}
#' remove samples with  naab12 (liver stiffness IQR) > 30
gal_met_prot_data$liver_stiffness[
                      which(gal_met_prot_data$liver_stiffnes_iqr > 30)] <- NA
```

We next remove the MetALD and SLD participants.

```{r keep only none and MASLD types}
#' because problem 2 proteins only NAs in some sld types,
#' keep only none and MASLD
gal_met_prot_data <- gal_met_prot_data[gal_met_prot_data$sld_type %in%
                                       c('none', 'MASLD'), ] #
gal_met_prot_data$sld_type <- droplevels(gal_met_prot_data$sld_type)
gal_met_prot_data$sld_type <- factor(gal_met_prot_data$sld_type)
```

We next remove participants with missing values for relevant variables:
- liver fat measurement

```{r}
#' Define variables for which we don't allow missing values
no_na_var <- c("liver_fat")
has_na <- lapply(gal_met_prot_data[, no_na_var, drop = FALSE], is.na) |>
    do.call(what = cbind)
#' Remove rows (participants) with a missing value in any of the above
#' variables
gal_met_prot_data <- gal_met_prot_data[rowSums(has_na) == 0, , drop = FALSE]
```

Remove fibrosis outlier values: **CURRENTLY NOT DONE**.
Previous outlier removal removed a participant with a FIB4 score of 7.8. This
participant was originally removed. The second outlier removal excluded
a participant with a liver_stiffness value of 34. Both participants are still
present, as fibrosis is not the phenotype of interest anymore.

We next define the categorical variables from numeric measurements:

Add a categoric liver_stiffness category too, using both the definition of
Christoph (7, 15 cutoffs) and in literature (7, 12.5 kPa cutoffs).

```{r liver-stiffness-category}
#' naab11 according to Christoph
gal_met_prot_data$liver_stiffness_cat1 <- 2
gal_met_prot_data$liver_stiffness_cat1[
                      is.na(gal_met_prot_data$liver_stiffness)] <- NA
gal_met_prot_data$liver_stiffness_cat1[
                      which(gal_met_prot_data$liver_stiffness < 7)] <- 1
gal_met_prot_data$liver_stiffness_cat1[
                      which(gal_met_prot_data$liver_stiffness > 15)] <- 3
gal_met_prot_data$liver_stiffness_cat1 <-
    factor(gal_met_prot_data$liver_stiffness_cat1)

#' naab11 according to literature
gal_met_prot_data$liver_stiffness_cat2 <- 2
gal_met_prot_data$liver_stiffness_cat2[
                      is.na(gal_met_prot_data$liver_stiffness)] <- NA
gal_met_prot_data$liver_stiffness_cat2[
                      which(gal_met_prot_data$liver_stiffness < 7)] <- 1     #mild or absent fibrosis
gal_met_prot_data$liver_stiffness_cat2[
                      which(gal_met_prot_data$liver_stiffness > 12.5)] <- 3  #cirrhosis
gal_met_prot_data$liver_stiffness_cat2 <-
    factor(gal_met_prot_data$liver_stiffness_cat2)
#' appearantly differnet cutoffs, depending what background...
#' https://www.mskcc.org/cancer-care/patient-education/understanding-your-fibroscan-results
```

Add a categoric liver_fat (i.e. steatosis) category, using the following
cutoffs:

- "normal": A score below 238 dB/m means the amount of fatty change in your
  liver is not higher than normal.
- grade steatosis S1: 238 to 260 dB/m
- grade steatosis S2: 260 to 290 dB/m
- grade steatosis S3: 290 to 400 dB/m

Depending on the distribution, look only at the extreme normal - steatosis.

```{r liver-fat-category}
#' naab13 make 4 groups
gal_met_prot_data$liver_fat_cat1 <- NA
gal_met_prot_data$liver_fat_cat1[which(gal_met_prot_data$liver_fat < 238)] <- 1 #normal
gal_met_prot_data$liver_fat_cat1[which(gal_met_prot_data$liver_fat >= 238)] <- 2
gal_met_prot_data$liver_fat_cat1[which(gal_met_prot_data$liver_fat > 260)] <- 3
gal_met_prot_data$liver_fat_cat1[which(gal_met_prot_data$liver_fat > 290)] <- 4
gal_met_prot_data$liver_fat_cat1 <- factor(gal_met_prot_data$liver_fat_cat1)

#' naab13 make 2 groups
gal_met_prot_data$liver_fat_cat2 <- NA
gal_met_prot_data$liver_fat_cat2[which(gal_met_prot_data$liver_fat < 238)] <- 1 #normal
gal_met_prot_data$liver_fat_cat2[which(gal_met_prot_data$liver_fat > 290)] <- 2  #steatosis, high S3

#' appearantly differnet cutoffs, depending what background...
#' https://www.mskcc.org/cancer-care/patient-education/understanding-your-fibroscan-results
```

At last we ensure categorical variables are encoded as a factor. Also, we create
additional variable for age per 10y, because coefficients and effect sizes
correspond to one year difference, and the anual difference is
neglectible. Dividing age with 10, difference in coefficents and effect sizes
would correspond to 10 years difference and it would be easier to define
meaningful cut-off. Finally, we add a visc_subcut_fat variable on the dataset,
comprising the ratio of visceral to subcutaneous fat thickness.


```{r define-age10-variable, message = FALSE}
#' Convert categorical variables to factors
col_cat <- c("case", "self_DM", "med_DM", "fasting", "no_drinking",
             "med_BP", "med_lip", "med_MS", "T2D", "Sex", "MASLD")
for (i in col_cat)
    gal_met_prot_data[, i] <- as.factor(gal_met_prot_data[, i])

#' define age per 10y group
gal_met_prot_data$Age_10 <- gal_met_prot_data$Age / 10
gal_met_prot_data$visc_subcut_fat <- gal_met_prot_data$visc_fat /
    gal_met_prot_data$body_fat
```


## Summary of the final data set

```{r}
summary(gal_met_prot_data[, c("Age", "Sex", "T2D", "MASLD", "FIB4", "BMI",
                              "liver_stiffness", "liver_fat", "liver_fat_cat1",
                              "visc_subcut_fat")])
```

The variables that are not allowed to conatin NAs are: sex, age and liver_fat.
Participants with NA in BMI, and those 2 with extreme liver stiffness/FIB4
scores are not removed.


# overview TODOs just here quick

TODO

- [x] if mash: xxx, but not relevant to further research
- [x] TABLE 2: demographics !!!
- [x] barplot splitted cat nr of signif prot/metab
- [x] if diet: xxx, but not relevant to further research
- [x] heatmap 3 redo, add in code jo
- [x] fibrosis: % male, age ... sentence add
- [x] fib4 score corr SI also mention in results


# Getting MASH classification

metabolic dysfunction-associated steatohepatitis (MASH).
MASH is stage between steatosis and fibrosis (in the masld phenotype). control
<-> steatosis -> MASH -> cirrhosis (severe fibrosis) -> liver cancer.

fyi: NASH = MASH

feedback christoph: use FAST TM score

FAST = 3 param:
liver stiffness from fibroscan, CAP (fiver fat), AST

but not in the list...:

Liver	Liver stiffness
Liver	Interquartile range (IQR) of liver stiffness
Liver	Controlled Attenuation Parameter
Liver	Steatose Score
Fat thickness	Subcutaneous fat thickness
Fat thickness	Visceral fat thickness
Clinical examination comment	NA
Scores	Fibrosis-4 index
Scores	Fatty liver index
Scores	Fatty liver index
Scores	Hepatic steatosis index
Scores	NAFLD liver fat score
Scores	Metabolic syndrome
Scores	NAFLD Fibrose Score
Scores	NAFLD Fibrose Score
Scores	Visceral Adipositas Index


```{r}
#' load AST
library(nafldData)
#nafldTraitAnnotations()|> head()

nafldTraitAnnotations()[c("nalp10", "nalp10n", "nalp10u"), ]
df <- nafldTraitAnnotations() |> as.data.frame()
dim(df)
df[1:5,]


labdata <- nafldData("Labdata")
rownames(labdata) <- labdata$AID
labdata$nalp10[1:5]  #ok, correct one
colnames(labdata)
dim(labdata)

#' check unit AST
labdata[1,'nalp10u']

```

add to the main df, as AST variable

```{r}

#' keep only the retained AIDs
labdata <- labdata[rownames(labdata) %in% rownames(gal_met_prot_data),]
dim(labdata)

#' add AST variable for the FAST score
dim(gal_met_prot_data)
gal_met_prot_data$AST <- labdata$nalp10

summary(gal_met_prot_data$AST)

```

Calculate the FAST score

FibroScan-AST (FAST) score for the non-invasive identification of patients with non-alcoholic steatohepatitis with significant activity and fibrosis.

FAST = 3 param:
liver stiffness from fibroscan (LSM), fiver fat (CAP), AST

formula: 

FAST=
e-1·65+1·07×ln⁡(LSM)+2·66*10-8×CAP3-63·3×AST-1
1+e-1·65+1·07×ln⁡(LSM)+2·66*10-8×CAP3-63·3×AST-1

in R: exp(2) represents e^2
in R: log() is natural logarithm

from article: https://www.thelancet.com/journals/langas/article/PIIS2468-1253%2819%2930383-8/fulltext 

art from Christoph:
https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0266859


```{r, eval = FALSE}
#' testing:
#' 3 needed param for FAST score
gal_met_prot_data$liver_stiffness[1]   #kPa
gal_met_prot_data$liver_fat[1]         #dB/m
gal_met_prot_data$AST[1]               #U/L

#' formula FAST
FAST <- (  exp( -1.65 + 1.07 * log(gal_met_prot_data$liver_stiffness[1]) + 2.66 * 10^-8 * gal_met_prot_data$liver_fat[1]^3 - 63.3 * gal_met_prot_data$AST[1]^-1 )) / ( 1 + exp(-1.65 + 1.07 * log(gal_met_prot_data$liver_stiffness[1]) + 2.66 * 10^-8 * gal_met_prot_data$liver_fat[1]^3 - 63.3 * gal_met_prot_data$AST[1]^-1 ))

FAST <- round(FAST, 2)
FAST

#ok, checked via https://www.mashscan.com/#calculators
#form from article: https://www.thelancet.com/journals/langas/article/PIIS2468-1253%2819%2930383-8/fulltext 
```

checked was correct using the FAST Calculator

The FAST score combines results from the FibroScan® with an easily accessible blood biomarker (AST) to help identify patients with active fibrotic-MASH.

Interpretation:
• A score <0.35 indicates that your liver is likely healthy, with little or no significant fibrosis or inflammation.
• If your score falls between 0.35 and 0.67, it means your results are inconclusive, and your doctor may recommend additional tests to get a clearer picture of your liver health.
• A score >0.67 suggests there may be significant fibrosis in your liver, which could indicate advanced liver disease.

```{r}
#' calcualte and define FAST score column
gal_met_prot_data$FAST <- (  exp( -1.65 + 1.07 * log(gal_met_prot_data$liver_stiffness) + 2.66 * 10^-8 * gal_met_prot_data$liver_fat^3 - 63.3 * gal_met_prot_data$AST^-1 )) / ( 1 + exp(-1.65 + 1.07 * log(gal_met_prot_data$liver_stiffness) + 2.66 * 10^-8 * gal_met_prot_data$liver_fat^3 - 63.3 * gal_met_prot_data$AST^-1 ))

gal_met_prot_data$FAST

#' give classifications
gal_met_prot_data$MASH <- 'inconclusive'                          # gray area
gal_met_prot_data$MASH[
                      is.na(gal_met_prot_data$FAST)] <- NA
gal_met_prot_data$MASH[
                      which(gal_met_prot_data$FAST < 0.35)] <- 'Not_at_risk_MASH'     #no MASH
gal_met_prot_data$MASH[
                      which(gal_met_prot_data$FAST > 0.67)] <- 'At_risk_MASH'  #MASH
gal_met_prot_data$MASH <-
    factor(gal_met_prot_data$MASH)

summary(gal_met_prot_data$MASH)

```

from art chrisoph:
the rule-out cutoff of 0.35 and a rule-in cutoff of 0.67, diagnostic performance measures of the FAST™ score for the identification of at-risk NASH 

- < 0.35 NO MASH
- >= 0.67 == MASH
- middle, ignore

FAST™ score for identifying patients with at-risk NASH 




# Table 2: distribution data set

```{r}
#' 1. load the labdata
library(nafldData)
#nafldTraitAnnotations()|> head()

#nafldTraitAnnotations()[c("nalp10", "nalp10n", "nalp10u", "nalp11", "nalp11q", "nalp11n", "nalp11u"), ]
df <- nafldTraitAnnotations() |> as.data.frame()
dim(df)
df[1:5,]
colnames(df)[1:5]
#paste0(df$variable, "_", df$name)

write.table(df, "data/traits_nafld.csv", col.names = TRUE)


labdata <- nafldData("Labdata")

rownames(labdata) <- labdata$AID
#rownames(labdata) #AIDs

lab_items <- colnames(labdata)
#lab_items #nalp... names

#lab_ann <- nafldAnn[lab_items, ]
#rownames(lab_ann) <- lab_items
#rownames(lab_ann) #nalp... names

```

the variables are AST: nalp10, nalp10n and nalp10u and for ALT nalp11, nalp11n, nalp11q and nalp11u - the funny thing is that nalp11 is marked as "text" - and only the nalp11q is numeric. no idea why actually.


        variable  source version    domain            name   note_detail      
nalp10    nalp10 Labdata       1 AST (GOT) AST (GOT) (U/L)         value
nalp10n  nalp10n Labdata       1 AST (GOT) AST (GOT) (U/L)          note
nalp10u  nalp10u Labdata       1 AST (GOT) AST (GOT) (U/L)          unit
nalp11    nalp11 Labdata       1 ALT (GPT) ALT (GPT) (U/L)         value
nalp11q  nalp11q Labdata       1 ALT (GPT) ALT (GPT) (U/L) Numeric value
nalp11n  nalp11n Labdata       1 ALT (GPT) ALT (GPT) (U/L)          note
nalp11u  nalp11u Labdata       1 ALT (GPT) ALT (GPT) (U/L)          unit
           type format unit
nalp10  numeric     NA   NA
nalp10n    text     NA   NA
nalp10u    text     NA   NA
nalp11     text     NA   NA
nalp11q numeric     NA   NA
nalp11n    text     NA   NA
nalp11u    text     NA   NA



naan02	naan02	Anthro	1	Anthropometry	Weight (kg)

nadd20	nadd20	Drug Summary	1	Drug usage	Usage of blood pressure lowering therapy (y/n)
nadd22	nadd22	Drug Summary	1	Drug usage	Usage of lipid modifying agents (y/n)
nalp18	nalp18	Labdata	1	LDL	LDL (mg/dl)
nalp18n	nalp18n	Labdata	1	LDL	LDL (mg/dl)
nalp18u	nalp18u	Labdata	1	LDL	LDL (mg/dl)



## Get all chris-nafld data there is and write to excel for inspection

get all data

```{r all-nafld-data}
#' 1. load the metadata
library(nafldData)
#nafldTraitAnnotations()|> head()

sources <- c("Anthro", "Clinical Exam", "Drug Summary", "Labdata", "FFQ", "Interview", "Personal Data", "Recruitment", "Self-admin")

#' Subset all to the same IDs
AIDs <- rownames(gal_met_prot_data)
all_metadata <- gal_met_prot_data

for(source in sources){
  metadata <- nafldData(source)
  #c("Anthro", "Clinical Exam", "Drug Summary", "Labdata", "FFQ", "Interview", "Personal Data", "Recruitment", "Self-admin") #only takes 1...
  rownames(metadata) <- metadata$AID

  metadata <- metadata[AIDs, ]
  
  all_metadata <- cbind(all_metadata, metadata)

}

dim(all_metadata)

#' change colnames to better understandable
# define wanted names
df$new_var_names <- paste0(df$variable, "_", df$name)

# Copy current column names
col_names <- colnames(all_metadata)

# Match column names in big_df to small_df$old_names
match_idx <- match(col_names, df$variable)

# Replace matched names with new names
col_names[!is.na(match_idx)] <- df$new_var_names[match_idx[!is.na(match_idx)]]

# Assign the updated column names
colnames(all_metadata) <- col_names

#colnames(all_metadata)


#nothing filtered, converted: contains notes, units, values, text...

#' write to file for manuel inspection
handle <- 'data/all_metadata.txt'
write.table(all_metadata, file = handle, sep = "\t",
            row.names = FALSE, col.names = TRUE)

```

## make table with all information

```{r pivot-table1}

#' column splitted sex per liver_fat_cat1
all_metadata$sex_liver_fat_cat1 <- paste0(gal_met_prot_data$Sex, gal_met_prot_data$liver_fat_cat1)

#' pivot table with 8 columns = 2 sex * 4 liver_tat categ
#' want a average as value
#' want an SD as value
#' if categorical (eg sex, sld), provide percentage per group -- first do numbers
#' 


df <- all_metadata

# Sample input: df (data frame), with one column called 'group_col' indicating the 8 categories
group_col <- "sex_liver_fat_cat1"  # Replace with your actual column name

# Ensure the group column is a factor
df[[group_col]] <- as.factor(df[[group_col]])

# Identify variable types
num_vars <- sapply(df, is.numeric)
fact_vars <- sapply(df, is.factor)
char_vars <- sapply(df, is.character)

# Exclude the grouping column from processing
num_vars[group_col] <- FALSE
fact_vars[group_col] <- FALSE

# ---- 1. Numeric summary: mean ± SD ----
numeric_summary <- sapply(names(df)[num_vars], function(var) {
  tapply(df[[var]], df[[group_col]], function(x) {
    m <- mean(x, na.rm = TRUE)
    s <- sd(x, na.rm = TRUE)
    sprintf("%.2f ± %.2f", m, s)
  })
})

numeric_summary <- t(numeric_summary)
colnames(numeric_summary) <- levels(df[[group_col]])


# ---- 2. Categorical summary: counts ----
categorical_summary <- lapply(names(df)[fact_vars], function(var) {
  tab <- table(df[[var]], df[[group_col]])
  as.data.frame.matrix(tab)
})
names(categorical_summary) <- names(df)[fact_vars]


# ---- 2. Categorical summary: counts and percentages ----
categorical_summary_with_percent <- lapply(names(df)[fact_vars], function(var) {
  # Create the count table
  tab <- table(df[[var]], df[[group_col]])
  df_counts <- as.data.frame.matrix(tab)

  # Compute row-wise percentages
  df_perc <- round(prop.table(as.matrix(tab), margin = 1) * 100, 1)
  df_perc <- as.data.frame.matrix(df_perc)

  # Rename percentage columns
  colnames(df_perc) <- paste0(colnames(df_perc), "_pct")

  # Combine counts and percentages
  df_combined <- cbind(df_counts, df_perc)

  # Add rownames as a column (factor levels)
  df_combined <- cbind(Level = rownames(df_combined), df_combined)
  rownames(df_combined) <- NULL

  df_combined
})

# Name the list entries
names(categorical_summary_with_percent) <- names(df)[fact_vars]



# ---- 3. Output ----
cat("Summary of numeric variables (mean ± SD):\n")
#print(numeric_summary)

pandoc.table(
    numeric_summary, style = "rmarkdown", split.table = Inf,
    caption = paste0("Table 1 - nummeric variables"))

#' write to file for manuel inspection
handle <- 'data/Table 1 - numeric variables.csv'
write.table(numeric_summary, file = handle, sep = "\t",
            row.names = TRUE, col.names = TRUE)



#cat("\nSummary of categorical variables (counts per category):\n")
#for (name in names(categorical_summary_with_percent)) {
#  cat("\nVariable:", name, "\n")
#  print(categorical_summary_with_percent[[name]])
#}

# Combine all into a single wide-format data frame
cat_summary_wide <- do.call(rbind, lapply(names(categorical_summary_with_percent), function(var_name) {
  df <- categorical_summary_with_percent[[var_name]]
  df$variable <- var_name
  df
}))

# Reorder columns: variable, level, group counts..., group %...
# Detect count and percentage columns
group_cols <- grep("^[^_]+$", names(cat_summary_wide), value = TRUE)        # e.g. A, B, C...
percent_cols <- grep("^[^_]+_pct$", names(cat_summary_wide), value = TRUE)  # e.g. A_pct, B_pct...

# Arrange columns
cat_summary_wide <- cat_summary_wide[, c("variable", "Level", group_cols, percent_cols)]

# View result
head(cat_summary_wide)

#' write to file for manuel inspection
handle <- 'data/Table 1 - categorical variables.csv'
write.table(cat_summary_wide, file = handle, sep = "\t",
            row.names = TRUE, col.names = TRUE)

```


# barplot

todo update if numbers changes...

|   | G1 | G2 | G3 |
|---|----|----|----|
| Significant proteins | 1 | 6 | 23 |
| Significant metabolites | 0 | 51 | 50 |

After using macronutrients and PDI index, instead of the scewed food groups.
nr of participants changed to 341#, so update.

|   | G1 | G2 | G3 |
|---|----|----|----|
| Significant proteins | 2 | 5 | 24 |
| Significant metabolites | 0 | 58 | 48 |

This is also Figure 2A.

```{r barplot-significant-nr-lm, fig.path = IMAGE_PATH, res = 150, fig.cap = "Number of significant circulating proteins, metabolites, and lipids between the control and grade 2 steatosis, and the control and grade 3 steatosis comparisons, respectively"}

sig_tab <- as.matrix(x = rbind(c(2, 5, 24), c(0, 58, 48)))
rownames(sig_tab) <- c("Significant proteins", "Significant metabolites")
colnames(sig_tab) <- c("Grade 1 steatosis", "Grade 2 steatosis", "Grade 3 steatosis")

sig_tab

# Create side-by-side barplot
barplot(sig_tab,
        beside = TRUE,           # Side-by-side (not stacked)
        col = c("black", "grey"),  # Colors for each group (row)
        legend.text = rownames(sig_tab),  # Add legend for groups
        args.legend = list(x = "topleft"),
        ylab = "Number of significant molecules",
        cex.names = 0.9)


#barplot with numbers on top
values <- c(2, 0, 5, 58, 24, 48) # added line
barobject <- barplot(sig_tab,
        beside = TRUE,           # Side-by-side (not stacked)
        col = c("black", "grey"),  # Colors for each group (row)
        legend.text = rownames(sig_tab),  # Add legend for groups
        args.legend = list(x = "topleft"),
        ylab = "Number of significant molecules",
        cex.names = 0.9)
text(x = barobject, y = (values+1), label = values, cex = 0.9, col = "black", xpd = TRUE) # added line



```


# fibrosis clinical info

Not for in the paper, but for the sentence regarding if age/gender with 
fibrosis vs control 

```{r violin-fibr-age-vs-sex-target, fig.path = IMAGE_PATH, fig.height=8, fig.width=7, dpi = 300}
vioplot(split(gal_met_prot_data$Age,
              paste0(gal_met_prot_data$Sex, gal_met_prot_data$liver_stiffness_cat2)),
        ylab = "Age")
grid()
```


violin plot to show age

```{r violin-fibr-bmi-vs-sex-target, fig.path = IMAGE_PATH, fig.height=8, fig.width=7, dpi = 300}
vioplot(split(gal_met_prot_data$BMI,
              paste0(gal_met_prot_data$Sex, gal_met_prot_data$liver_stiffness_cat2)),
        ylab = "BMI")
grid()
```





```{r}


#' column of interest:
#all_metadata$liver_stiffness_cat2 

#' pivot table with 2 liver_stiffness_cat2 categ
#' want a average as value
#' want an SD as value
#' if categorical (eg sex, sld), provide percentage per group


df <- all_metadata

# Sample input: df (data frame), with one column called 'group_col' indicating the 8 categories
group_col <- "liver_stiffness_cat2"  # Replace with your actual column name

# Ensure the group column is a factor
df[[group_col]] <- as.factor(df[[group_col]])

# Identify variable types
num_vars <- sapply(df, is.numeric)
fact_vars <- sapply(df, is.factor)
char_vars <- sapply(df, is.character)

# Exclude the grouping column from processing
num_vars[group_col] <- FALSE
fact_vars[group_col] <- FALSE

# ---- 1. Numeric summary: mean ± SD ----
numeric_summary <- sapply(names(df)[num_vars], function(var) {
  tapply(df[[var]], df[[group_col]], function(x) {
    m <- mean(x, na.rm = TRUE)
    s <- sd(x, na.rm = TRUE)
    sprintf("%.2f ± %.2f", m, s)
  })
})

numeric_summary <- t(numeric_summary)
colnames(numeric_summary) <- levels(df[[group_col]])


# ---- 2. Categorical summary: counts ----
categorical_summary <- lapply(names(df)[fact_vars], function(var) {
  tab <- table(df[[var]], df[[group_col]])
  as.data.frame.matrix(tab)
})
names(categorical_summary) <- names(df)[fact_vars]


# ---- 2. Categorical summary: counts and percentages ----
categorical_summary_with_percent <- lapply(names(df)[fact_vars], function(var) {
  # Create the count table
  tab <- table(df[[var]], df[[group_col]])
  df_counts <- as.data.frame.matrix(tab)

  # Compute row-wise percentages
  df_perc <- round(prop.table(as.matrix(tab), margin = 1) * 100, 1)
  df_perc <- as.data.frame.matrix(df_perc)

  # Rename percentage columns
  colnames(df_perc) <- paste0(colnames(df_perc), "_pct")

  # Combine counts and percentages
  df_combined <- cbind(df_counts, df_perc)

  # Add rownames as a column (factor levels)
  df_combined <- cbind(Level = rownames(df_combined), df_combined)
  rownames(df_combined) <- NULL

  df_combined
})

# Name the list entries
names(categorical_summary_with_percent) <- names(df)[fact_vars]



# ---- 3. Output ----
cat("Summary of numeric variables (mean ± SD):\n")
#print(numeric_summary)

pandoc.table(
    numeric_summary, style = "rmarkdown", split.table = Inf,
    caption = paste0("Table 1 - nummeric variables"))

#' write to file for manuel inspection
handle <- 'data/fibrosis - numeric variables.csv'
write.table(numeric_summary, file = handle, sep = "\t",
            row.names = TRUE, col.names = TRUE)



#cat("\nSummary of categorical variables (counts per category):\n")
#for (name in names(categorical_summary_with_percent)) {
#  cat("\nVariable:", name, "\n")
#  print(categorical_summary_with_percent[[name]])
#}

# Combine all into a single wide-format data frame
cat_summary_wide <- do.call(rbind, lapply(names(categorical_summary_with_percent), function(var_name) {
  df <- categorical_summary_with_percent[[var_name]]
  df$variable <- var_name
  df
}))

# Reorder columns: variable, level, group counts..., group %...
# Detect count and percentage columns
group_cols <- grep("^[^_]+$", names(cat_summary_wide), value = TRUE)        # e.g. A, B, C...
percent_cols <- grep("^[^_]+_pct$", names(cat_summary_wide), value = TRUE)  # e.g. A_pct, B_pct...

# Arrange columns
cat_summary_wide <- cat_summary_wide[, c("variable", "Level", group_cols, percent_cols)]

# View result
head(cat_summary_wide)

#' write to file for manuel inspection
handle <- 'data/fibrosis - categorical variables.csv'
write.table(cat_summary_wide, file = handle, sep = "\t",
            row.names = TRUE, col.names = TRUE)



```