---
title: "NAFLD preprocessing pos"
author: "Philippine Louail"
affiliation: "Eurac Research, Bolzano, Italy"
date: "2023-10-05"
output: html_document
---

```{r style, message = FALSE, echo = FALSE, warning = FALSE, results = "asis"}
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now, units = "secs")
      # return a character string to show the time
      paste("Time for this code chunk to run:", round(res,
        2), "seconds")
    }
  }
}))

library("BiocStyle")
library("knitr")
library("rmarkdown")
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7, time_it = TRUE)
```

# Packages

```{r packages}
library(MsExperiment)
library(RSQLite)
library(xcms)
library(Spectra)
library(RColorBrewer)
library(pander)
library(readxl)
library(MetaboCoreUtils)
library(pheatmap)
library(MsBackendSql)
library(SummarizedExperiment)
```

# Data import

Create a `MsExperiment` representing the data
```{r create MsExperiment}
#' Load the spectra data from the database
sps <- Spectra("data/NAFLD_untargeted.sqlite", drv = SQLite(),
               source = MsBackendOfflineSql())

nafld <- MsExperiment(spectra = sps)

#' Subset for pos
nafld <- nafld[sampleData(nafld)$polarity == "POS"]
```


```{r parallel processing}
#' Set up parallel processing using 10 cores
if (.Platform$OS.type == "unix") {
    register(bpstart(MulticoreParam(10)))
} else {
    register(bpstart(SnowParam(10)))
}
```

# Data organisation

The experimental data is now a `MsExperiment` object:

```{r}
nafld
```

```{r phenodata, echo=FALSE, results='asis'}
sampleData(nafld)[, c(5, 10, 11, 13)] |>
    as.data.frame() |>
    head() |>
    pandoc.table(style = "rmarkdown",
                 caption = "Some samples from the data set.")
```

```{r define-colors, include=FALSE}
#' Define colors for the groups.
#' Sample type
col_phenotype <- brewer.pal(8, "Accent")[c(1, 3, 5, 4, 8)]
names(col_phenotype) <- c("Study", "Pool_IS_Dil", "Pool", "Pool_Dil", "Blank")

col_sample <- col_phenotype[sampleData(nafld)$sample_type]

#' Batches
col_batch_id <- brewer.pal(7, "Paired")[c(4, 3, 1, 5, 2, 6, 7)]
names(col_batch_id) <- c("BATCH0125", "BATCH0130", "BATCH0143",
                        "BATCH0159", "BATCH0165", "BATCH0169",
                        "BATCH0186")

col_batches <- col_batch_id[sampleData(nafld)$batch_id]
```

We therefore have an data set of `r length(nafld)` samples for a total
of `r length(spectra(nafld))` spectra.

The retention time range for the entire data set is: `r range(rtime(spectra(nafld)))`

Get some data overview to be sure everything is fine:

```{r some healthy stuff}
#' check number of spectra per file
fromFile(nafld) |>
    table()
```

No sample with below average number of spectra

# Visualisation

## BPC

```{r filter rtime}
#' Filter for ret time
nafld <- filterRt(nafld, c(10, 250))
```

```{r plot again, include=FALSE}
#' Plot again
chromatogram(nafld, aggregationFun = "max") |>
    plot(col = col_sample, main = "BPC after filtering",  lwd = 0.8)
    grid()
    legend("topright", col = col_phenotype, legend = names(col_phenotype),
           lty = 1)
```

## Internal standards

Here just generate EICs for all standard, see the EIC_IS folder for the images.
Code will be displayed for the first one as an example but not for the others.

### All

```{r EIC extract for internal standard, include=FALSE}
#' get the list
intern_standard <- read.delim("internal_standards.txt")
intern_standard <- intern_standard[!is.na(intern_standard$POS), ]
rownames(intern_standard) <- intern_standard$abbreviation

#'generate calcualte formula
intern_standard$mz <- mapply(intern_standard$formula,
                             intern_standard$POS,
                             FUN = mass2mz)

#' Fit for each standards
intern_standard$mzmin <- intern_standard$mz - 0.05
intern_standard$mzmax <- intern_standard$mz + 0.05
intern_standard$rtmin <- intern_standard$RT - 15
intern_standard$rtmax <- intern_standard$RT + 15

#' Extract the EICs
eics <- chromatogram(nafld,
                     mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]),
                     rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]))

dr <- "EIC_IS/full/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics[i], main =intern_standard$name[i],
         col = col_sample)
    grid()
    legend("topright", col = col_phenotype,
           legend = names(col_phenotype), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}

#' Add info on the EICs:
#' - ...
```

### Batch all sample

All samples colored by different col_batches

```{r EIC-IS, include=FALSE}
dr <- "EIC_IS/batch/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics[i], main =intern_standard$name[i], col = col_batches)
    grid()
    legend("topright", col = col_batch_id,
           legend = names(col_batch_id), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```

### Only study samples

Only the study samples colored by batches

```{r EIC-study, include=FALSE}
#' select for study samples
study_sample <-grep("Study", sampleData(nafld)$sample_type)
study_sample <- nafld[study_sample]

batches_study <- col_batch_id[sampleData(study_sample)$batch_id]

dr <- "EIC_IS/sample/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)

eics <- chromatogram(study_sample,
                     mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]),
                     rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]))

for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics[i], main =intern_standard$name[i], col = batches_study)
    grid()
    legend("topright",
           col = col_batch_id,
           legend = names(col_batch_id),
           lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```

### Only pool

The pool samples colored by batches.

```{r EIC-pool, include=FALSE}
pool_sample <- nafld[sampleData(nafld)$sample_type == "Pool"]
batches_pool <- col_batch_id[sampleData(pool_sample)$batch_id]

eics <- chromatogram(pool_sample,
                     mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]),
                     rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]))

dr <- "EIC_IS/Pool_batch/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)

for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics[i], main =intern_standard$name[i], col = batches_pool)
    grid()
    legend("topright",
           col = col_batch_id,
           legend = names(col_batch_id),
           lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}

#' get the list
standard <- read.delim("standards_dilution_nafld.txt")
standard <- standard[!is.na(standard$POS), ]

#'generate calcualte formula
standard$mass <- mapply(standard$formula,
                             FUN = calculateMass)

standard$mz <- mapply(standard$mass,
                             standard$POS,
                             FUN = mass2mz)

#' Fit for each standards
standard$mzmin <- standard$mz - 0.05
standard$mzmax <- standard$mz + 0.05
standard$rtmin <- standard$RT - 15
standard$rtmax <- standard$RT + 15

pool_sample <- nafld[sampleData(nafld)$sample_type == "Pool"]
batches_pool <- col_batch_id[sampleData(pool_sample)$batch_id]

eics_s <- chromatogram(pool_sample,
                     mz = as.matrix(standard[, c("mzmin", "mzmax")]),
                     rt = as.matrix(standard[, c("rtmin", "rtmax")]))

dr <- "EIC_standard/Pool_batch/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)

for (i in seq_len(nrow(standard))) {
    png(paste0(dr, "EIC_", standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics_s[i], main =standard$name[i], col = batches_pool)
    grid()
    legend("topright",
           col = col_batch_id,
           legend = names(col_batch_id),
           lty = 1)
    abline(v = standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```

## Chromatographic info

We therefore have an data set of `r length(nafld)` samples for a total of
`r length(spectra(nafld))` spectra.

The retention time range for the entire data set is: `r range(rtime(spectra(nafld)))`

```{r heatmap, echo=TRUE, fig.height=8, fig.width=7}
#' Heatmap from total ion chromatogram
tic <- chromatogram(nafld, aggregationFun = "sum") |>
  bin(binSize = 2)

# Data frame with column annotations.
ticmap <- do.call(cbind, lapply(tic, intensity)) |>
    cor()

#' yes the code is weird but pheatmap has a weird  problems
#' with rownames/colnames handling.
col_hm <- data.frame(sample_type = sampleData(nafld)[, "sample_type"])
rownames(col_hm) <- colnames(ticmap)
row_hm <- data.frame(batch = sampleData(nafld)[, "batch_id"])
rownames(row_hm) <- rownames(ticmap)
ann_color <- list(
    sample_type = col_phenotype,
    batch = col_batch_id
)

rownames(ticmap) <- rownames(row_hm)
colnames(ticmap) <- rownames(col_hm)

pheatmap(ticmap, annotation_row = row_hm, annotation_col = col_hm,
         annotation_colors = ann_color, annotation_names_row = FALSE,
         annotation_names_col = FALSE, show_rownames = FALSE,
         show_colnames = FALSE, annotation_legend = TRUE)
```

## Spectra
```{r combine spectra, echo=TRUE}
#' Combine spectra
processingChunkSize(nafld@spectra) <- 100000
#' Combine spectra
bps <- spectra(nafld) |>
    bin(binSize = 0.01, zero.rm = TRUE) |>
    combineSpectra(f = fromFile(nafld), p = fromFile(nafld), intensityFun = max,
                   ppm = 5)
```

```{r compare, echo=TRUE}
sim_matrix <- compareSpectra(bps) # check which faster
```

```{r heatmap_sim_matrix, fig.height=8, fig.width=7, include=FALSE}
pheatmap(sim_matrix, annotation_row = row_hm, annotation_col = col_hm,
         annotation_colors = ann_color, annotation_names_row = FALSE,
         annotation_names_col = FALSE, show_rownames = FALSE,
         show_colnames = FALSE, annotation_legend = TRUE)
processingChunkSize(nafld@spectra) <- Inf
```

```{r}
save(nafld, file = "after_visu")
```

# Preprocessing
## Chromatographic peak detection

finding cwp parameter

- peakwidth: observing previous EIC, expecting between 2 to 20s
- ppm:

```{r ppm parameter }
#' choose a nice compound here
cystine_mz <- calculateMass("C6H12N2O4S2") |>
  mass2mz("[M+H]+")

cystine_mz <- cystine_mz[1, 1]

#' plot to see rt range for one sample
cst <- chromatogram(nafld[2],
                    mz = cystine_mz + c(-0.01, 0.01),
                    rt = c(200, 250))
plot(cst)

#' Restrict the data to signal from Cystine
cst <- nafld[2L] |>
  spectra() |>
  filterRt(rt = c(210, 220)) |>
  filterMzRange(mz = cystine_mz + c(-0.01, 0.01))

lengths(cst)

#' Calculate the difference in m/z values between scans
mz_diff <- cst |>
    mz() |>
    unlist() |>
    diff() |>
    abs()

#' Express it in ppm
range(mz_diff * 1e6 / mean(unlist(mz(cst))))
```

Chose larger ppm of 30
But can also see very strong right skew on the EIC. So choose a larger
peak width to accommodate.

```{r echo=TRUE}
#' extract test ion
#' for cystine
eic_cystine <- chromatogram(nafld,
                            rt = c(200, 235),
                            mz = cystine_mz + c(-0.01, 0.01))
#' histidine

his_mz <- calculateMass("C6H9N3O2") |>
  mass2mz("[M+H]+")

his_mz <- his_mz[1, 1]

eic_his <- chromatogram(nafld,
                        rt = c(170, 210),
                        mz = his_mz + c(-0.01, 0.01))
```


```{r echo=TRUE, warning=FALSE}
param <- CentWaveParam(peakwidth = c(2, 20), ppm = 50, snthresh = 7,
                       integrate = 2)
# i tested other param these where the best in my opinion
```

```{r echo=TRUE}
#' test for some compounds:
#' cystine
cystine_test <- findChromPeaks(eic_cystine, param = param, chunkSize = 10L)
head(chromPeaks(cystine_test))

#' histidine
his_test <- findChromPeaks(eic_his, param = param, chunkSize = 10L)
head(chromPeaks(his_test))
```


```{r echo=TRUE}
#' Plot test chromatogram
par(mfrow = c(1, 2))
plot(cystine_test,
     main = "Cystine",
     col = paste0(col_sample, 80),
     peakCol = col_sample[chromPeaks(cystine_test)[, "column"]],
     peakBg = paste0(col_sample, 40)[chromPeaks(cystine_test)[, "column"]])
grid()

plot(his_test,
     main = "Histidine",
     col = paste0(col_sample, 80),
     peakCol = col_sample[chromPeaks(his_test)[, "column"]],
     peakBg = paste0(col_sample, 40)[chromPeaks(his_test)[, "column"]])
grid()
legend("topright", col = col_phenotype, cex = 0.50,
       horiz = TRUE,inset = c(-0.18, -0.1), xpd = TRUE,
       text.width = 7, bty = "n",
       legend = names(col_phenotype), lty = 1)
```

In my opinion looks pretty good, should apply to all data:

```{r findchrompkeaks all data, echo=TRUE}
nafld <- findChromPeaks(nafld, param = param, chunkSize = 10L)

save(nafld, file = "after_peakdetect")
```


```{r echo=TRUE}
#' test for ions of interest
eic_cystine <- chromatogram(nafld,
                            rt = c(202, 217),
                            mz = cystine_mz + c(-0.01, 0.01),
                            aggregationFun = "max")

eic_his <- chromatogram(nafld,
                        rt = c(172, 203),
                        mz = his_mz + c(-0.01, 0.01),
                        aggregationFun = "max")

par(mfrow = c(1, 2))
plot(eic_cystine, main = "Cystine", col = paste0(col_sample, 80),
     peakCol = col_sample[chromPeaks(eic_cystine)[, "sample"]],
     peakBg = paste0(col_sample[chromPeaks(eic_cystine)[, "sample"]], 40))
grid()
plot(eic_his, main = "Histidine", col = paste0(col_sample, 80),
     peakCol = col_sample[chromPeaks(eic_his)[, "sample"]],
     peakBg = paste0(col_sample[chromPeaks(eic_his)[, "sample"]], 40))
grid()
legend("topright", col = col_phenotype, cex = 0.50,
       horiz = TRUE,inset = c(-0.18, -0.1), xpd = TRUE,
       text.width = 7, bty = "n",
       legend = names(col_phenotype), lty = 1)
```

```{r echo=TRUE}
#' All EICs in pool
pool_sample <- nafld[sampleData(nafld)$sample_type == "Pool"]
batches_pool <- col_batch_id[sampleData(pool_sample)$batch_id]

eics <- chromatogram(pool_sample,
                     mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]),
                     rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]))

dr <- "EIC_IS/Pool_batch/afterchrompeak/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)

for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics[i], main =intern_standard$name[i], col = paste0(batches_pool,80),
         peakCol = col_batches[chromPeaks(eics[i])[, "sample"]],
         peakBg = paste0(col_batches[chromPeaks(eics[i])[, "sample"]], 40))
    grid()
    legend("topright", col = col_batch_id,
           legend = names(col_batch_id), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}

eics_s <- chromatogram(pool_sample,
                     mz = as.matrix(standard[, c("mzmin", "mzmax")]),
                     rt = as.matrix(standard[, c("rtmin", "rtmax")]))

dr <- "EIC_standard/Pool_batch/afterchrompeak/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)

for (i in seq_len(nrow(standard))) {
    png(paste0(dr, "EIC_", standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics_s[i], main =standard$name[i], col = batches_pool)
    grid()
    legend("topright",
           col = col_batch_id,
           legend = names(col_batch_id),
           lty = 1)
    abline(v = standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```

Look at the nb of peak per files
```{r echo=TRUE}
#' Count peaks per file
chromPeaks(nafld)[, "sample"] |>
    table()

#' Maybe could be nice to see per sample type here
chromPeaks(nafld) |>
    head()
```

```{r}
save(nafld, file= "after_peakdetect")
```

### Refine chromatographic peaks

```{r eval=FALSE, include=FALSE}
load("after_peakdetect")
```


```{r merging}
#' set up the parameter
param <- MergeNeighboringPeaksParam(expandRt = 5,
                                    expandMz = 0.01,
                                    minProp = 0.75)
```

```{r refine dataset,  message=TRUE}
#' apply to all dataset
nafld <- refineChromPeaks(nafld, param = param, chunkSize = 10)
save(nafld, file = "nafld_after_refine")

chromPeakData(nafld)$merged |>
    table() |>
    head()
```

```{r}
#' All EICs in pool
pool_sample <- nafld[sampleData(nafld)$sample_type == "Pool"]
batches_pool <- col_batch_id[sampleData(pool_sample)$batch_id]

eics <- chromatogram(pool_sample,
                     mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]),
                     rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]))

dr <- "EIC_IS/Pool_batch/afterrefine/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)

for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics[i], main =intern_standard$name[i], col = paste0(batches_pool, 80),
         peakCol = col_batches[chromPeaks(eics[i])[, "sample"]],
         peakBg = paste0(col_batches[chromPeaks(eics[i])[, "sample"]], 40))
    grid()
    legend("topright", col = col_batch_id,
           legend = names(col_batch_id), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}

eics_s <- chromatogram(pool_sample,
                     mz = as.matrix(standard[, c("mzmin", "mzmax")]),
                     rt = as.matrix(standard[, c("rtmin", "rtmax")]))

dr <- "EIC_standard/Pool_batch/afterrefine/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)

for (i in seq_len(nrow(standard))) {
    png(paste0(dr, "EIC_", standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics_s[i], main =standard$name[i], col = batches_pool)
    grid()
    legend("topright",
           col = col_batch_id,
           legend = names(col_batch_id),
           lty = 1)
    abline(v = standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```


```{r include=FALSE}
save(nafld, file = "nafld_after_refine")
```

## Retention time alignment

Multiple strategies were tested in a separate Rmd file

```{r plot-before-alignment,  echo=FALSE}
load("nafld_after_refine")

pool_index <- sampleData(nafld)$sample_type == "Pool"
pool_sample <- nafld[pool_index]

bpc_raw <- chromatogram(pool_sample, aggregationFun = "max",
                        chromPeaks = "none") |>
    plot(main = "BPC of Pool sample",
         col = paste0(batches_pool, 80),
         lwd = 0.8)
    grid()
    legend("topright", col = col_batch_id, cex = 0.50,
           horiz = TRUE,inset = c(-0.18, -0.1),
           xpd = TRUE, text.width = 7,
           bty = "n",
           legend = names(col_batch_id),
           lty = 1)

#' standard
par(mfrow = c(1, 2))
plot(eic_cystine[1, pool_index],
     main = "Cystine",
     col = paste0(batches_pool, 80))
grid()
abline(v = 211 , col = "red", lty = 3)

plot(eic_his[1, pool_index],
     main = "Histidine",
     col = paste0(batches_pool, 80))
grid()
abline(v = 188 , col = "red", lty = 3)
legend("topright", col = col_batch_id,
       cex = 0.50,
       horiz = TRUE,
       inset = c(-0.18, -0.1),
       xpd = TRUE,
       text.width = 7, bty = "n",
       legend = names(col_batch_id), lty = 1)

#' keep a raw nafld object to compare before/after alignment
nafld_raw <- nafld
```

```{r}
#'Reorganising samples order for alignment
#' saving old order
sampleData(nafld)$original_index <- seq_along(nafld)

#' determining new order
sd <- as.data.frame(sampleData(nafld))
tmp <- lapply(split(sd, sd$batch_id), function(batch){
    qcs <- which(batch$sample_type == "Pool")
    indices <- batch$original_index
    first <- min(qcs)
    last <- max(qcs)
    c(indices[first], indices[-c(first, last)], indices[last])
})

index_qc <- unlist(tmp, use.names = FALSE)

#' applying it
nafld <- nafld[index_qc]
```

One run based on a pre-defined peak matrix of IS standard with rt for each Pool samples

```{r}
#' creating matrix for rt alignment
standard <- read.delim("Mix_alignment.txt")

#' loop results
ID_table <- matrix(
    ncol = length(nafld),
    nrow = nrow(standard),
    dimnames = list(c(row.names(standard)), c(seq_len(length(nafld))))
)

cpks <- as.data.frame(chromPeaks(nafld))
cpks$peak_id <- rownames(cpks)

#' Keep only matches with highest intensity
rem_dups <- function(x) {
    if (nrow(matches(x)) > 1)
        filterMatches(x, index = which.max(x$target_maxo))
    else x
}

#' get ID for peaks matching with IS for each samples ( minus Blanks)
for (i in which(sampleData(nafld)$sample_type != "Blank")) {
    tmp <- cpks[cpks$sample == i, ]
    match_intern_standard <- matchValues(
        query = standard,
        target = tmp,
        mzColname = c("mz", "mz"),
        rtColname = c("RT", "rt" ),
        param = MzRtParam(ppm = 50, toleranceRt = 20))
        match_intern_standard <- MetaboAnnotation::endoapply(match_intern_standard,
                                                             FUN = rem_dups)
        ID_table[, i] <- match_intern_standard$target_peak_id
}

#' Function to create rt dataframe;
#' avoiding subset with NA turns out to be much more efficient
rtdf <- function(nafld, ID_table) {
    index <- as.vector(ID_table)
    nna <- !is.na(index)
    x <- rep(NA, length(index))
    x[nna] <- chromPeaks(nafld)[index[nna], "rt"]
    dim(x) <- dim(ID_table)
    rownames(x) <- rownames(ID_table)
    colnames(x) <- colnames(ID_table)
    x
}

#' run for nafld
RT_raw <- rtdf(nafld, ID_table)

#' Subset for Pool samples
index <- which(sampleData(nafld)$sample_type == "Pool")
final_table <- RT_raw[, index]

#' run with that first
#' Define parameters of choice
param <- PeakGroupsParam(minFraction = 0.9,
                         span = 0.6,
                         peakGroupsMatrix = final_table,
                         subset = index,
                         subsetAdjust = "average")
nafld <- adjustRtime(nafld, param = param, chunkSize = 10L)

#' See result
plotAdjustedRtime(nafld, col = paste0(col_sample, 80))
grid()
legend("topright", col = col_phenotype,
       legend = names(col_phenotype), lty = 1)

plotAdjustedRtime(nafld, col = paste0(col_batches, 80))
grid()
legend("topright", col = col_batch_id,
       legend = names(col_batch_id), lty = 1)

#' Replace the Rtime by the adjusted ones
nafld <- applyAdjustedRtime(nafld)
```

```{r}
#' Plot EICs for the mix standards.
eics_s_raw <- eics_s
eics_s <- chromatogram(nafld[pool_index],
                       mz = as.matrix(standard[, c("mzmin", "mzmax")]),
                       rt = as.matrix(standard[, c("rtmin", "rtmax")]))

dr <- "EIC_standard/Pool_batch/after_first_alignment/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)

for (i in seq_len(nrow(standard))) {
    png(paste0(dr, "EIC_", standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    eic <- eics_s[i]
    plot(eic, main = standard$name[i], col = paste0(batches_pool, 80)LLLLL)
    grid()
    legend("topright",
           col = col_batch_id,
           legend = names(col_batch_id),
           lty = 1)
    abline(v = standard$RT[i], col = "red", lty = 3)
    dev.off()
}


```

Second run this timebased on correspondence step on Pool samples.

```{r adjustRtime, message=FALSE, warning=FALSE}
## second run
# Grouping the peaks according to
param <- PeakDensityParam(
    sampleGroups = sampleData(nafld)$sample_type == "Pool",
    minFraction = 0.9,
    binSize = 0.01,
    bw = 2.5)

nafld <- groupChromPeaks(nafld, param = param)

#' Define parameters of choice
subset <- which(sampleData(nafld)$sample_type == "Pool")

param <- PeakGroupsParam(minFraction = 0.9,
                         extraPeaks = 50,
                         span = 0.3,
                         subsetAdjust = "average",
                         subset = subset)

#' Input in the function
nafld <- adjustRtime(nafld, param = param, chunkSize = 10L)

#' See result
plotAdjustedRtime(nafld, col = col_sample)
grid()
legend("topright", col = col_phenotype,
       legend = names(col_phenotype), lty = 1)

plotAdjustedRtime(nafld, col = col_batches)
grid()
legend("topright", col = col_batch_id,
       legend = names(col_batch_id), lty = 1)
```

```{r}
#' get RT table after alignment
RT_aligned <- rtdf(nafld, ID_table)

#' Get RT table only for study samples
index_s <- sampleData(nafld)$sample_type == "Study"
RT_raw <- RT_raw[, index_s]
RT_aligned <- RT_aligned[, index_s]

Sdsdf <- data.frame(
    Raw = rowSds(RT_raw, na.rm = TRUE),
    Aligned = rowSds(RT_aligned, na.rm = TRUE)
)

pandoc.table(Sdsdf, style = "rmarkdown")
```

```{r}
boxplot(Sdsdf, outline = FALSE, ylab = "RT standard deviation")
```

Looks pretty cool

```{r}
 #' remove reordering
nafld <- nafld[order(sampleData(nafld)$original_index), keepFeatures = TRUE, keepAdjustedRtime = TRUE]
```


```{r bpc-before-and-after, echo=FALSE}
pool_index <- sampleData(nafld)$sample_type == "Pool"
pool_sample <- nafld[pool_index, keepAdjustedRtime = TRUE]
batches_pool <- col_batch_id[sampleData(pool_sample)$batch_id]

#' Plot the BPC before and after alignment
par(mfrow = c(2,1), mar = c(2, 1, 1, 0.5))
pool_sample_raw <- dropAdjustedRtime(pool_sample)
chromatogram(pool_sample_raw,
             aggregationFun = "max",
             chromPeaks = "none") |>
    plot(main = "BPC of pool before alignment",
         col = batches_pool, lwd = 0.8)
    grid()

chromatogram(pool_sample,
             aggregationFun = "max",
             chromPeaks = "none") |>
    plot(main = "BPC of pool after alignment",
         col = batches_pool, lwd = 0.8)
    grid()
    legend("topright", col = col_batch_id, cex = 0.50,
           horiz = TRUE,inset = c(-0.18, -0.1), xpd = TRUE,
           text.width = 7, bty = "n",
           legend = names(col_batch_id), lty = 1)

#' replace value to not have problems when indexing

nafld <- applyAdjustedRtime(nafld)
pool_sample <- nafld[pool_index]
```

```{r}
#' standard after alignment
par(mfrow = c(1, 2), mar = c(4, 4.5, 2, 0.5))
plot(eic_cystine[1, pool_index], peakType = "none",
     main = "Cystine in pool before alignment",
     col = paste0(batches_pool,80))
grid()
abline(v = 211 , col = "red", lty = 3)

eic_cystine <- chromatogram(nafld, rt = c(201, 220),
                            mz = cystine_mz + c(-0.01, 0.01))
plot(eic_cystine[1, pool_index], peakType = "none",
     main = "Cystine in pool after alignment",
     col = paste0(batches_pool,80))
     grid()
     abline(v = 211 , col = "red", lty = 3)
legend("topright", col = col_batch_id, cex = 0.50,
       horiz = TRUE,inset = c(-0.18, -0.1), xpd = TRUE,
       text.width = 7, bty = "n",
       legend = names(col_batch_id), lty = 1)


par(mfrow = c(1, 2), mar = c(4, 4.5, 2, 0.5))
plot(eic_his[1, pool_index], peakType = "none",
     main = "Histidine in pool before alignment",
     col = paste0(batches_pool,80))
     grid()
     abline(v = 188 , col = "red", lty = 3)

eic_his <- chromatogram(nafld, rt = c(170, 200),
                        mz = his_mz + c(-0.01, 0.01))
plot(eic_his[1, pool_index], peakType = "none",
     main = "Histidine in pool after alignment",
     col = paste0(batches_pool,80))
     grid()
     abline(v = 188 , col = "red", lty = 3)
     legend("topright", col = col_batch_id, cex = 0.50,
            horiz = TRUE,inset = c(-0.18, -0.1), xpd = TRUE,
            text.width = 7, bty = "n",
            legend = names(col_batch_id), lty = 1)
```

```{r EIC-pool-after-alignment, include=FALSE}
eics <- chromatogram(pool_sample,
                     mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]),
                     rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                     chromPeaks = "none")

dr <- "EIC_IS/Pool_batch/afteralignment/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)

for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics[i], main =intern_standard$name[i],
         col = batches_pool,
         peakType = "none")
    grid()
    legend("topright", col = col_batch_id,
           legend = names(col_batch_id), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}

#' Extract the EICs
eics <- chromatogram(nafld,
                     mz = as.matrix(intern_standard[, c("mzmin", "mzmax")]),
                     rt = as.matrix(intern_standard[, c("rtmin", "rtmax")]),
                     chromPeaks = "none")

dr <- "EIC_IS/full/afteralignment/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(intern_standard))) {
    png(paste0(dr, "EIC_", intern_standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics[i],
         main =intern_standard$name[i],
         col = col_sample,
         peakType = "none")
    grid()
    legend("topright",
           col = col_phenotype,
           legend = names(col_phenotype), lty = 1)
    abline(v = intern_standard$RT[i], col = "red", lty = 3)
    dev.off()
}

#' eic standard
eics_s <- chromatogram(pool_sample,
                     mz = as.matrix(standard[, c("mzmin", "mzmax")]),
                     rt = as.matrix(standard[, c("rtmin", "rtmax")]))

dr <- "EIC_standard/Pool_batch/afteralignment/"
dir.create(dr, recursive = TRUE, showWarnings = FALSE)

for (i in seq_len(nrow(standard))) {
    png(paste0(dr, "EIC_", standard$abbreviation[i], ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eics_s[i], main =standard$name[i], col = batches_pool)
    grid()
    legend("topright",
           col = col_batch_id,
           legend = names(col_batch_id),
           lty = 1)
    abline(v = standard$RT[i], col = "red", lty = 3)
    dev.off()
}
```


```{r include=FALSE}
save(nafld, file = "nafld_after_Alignment")
```

## Correspondence

prob need to check for binwidth
```{r first plots test}
#'Updating parameters
param <- PeakDensityParam(sampleGroups = sampleData(nafld)$sample_type[pool_index],
                        minFraction = 0.75, binSize = 0.015, bw = 2)


plotChromPeakDensity(
    eic_cystine[1, pool_index],
    param = param,
    col = paste0(batches_pool, 80),
    peakCol = col_batches[chromPeaks(eic_cystine[1, pool_index])[, "sample"]],
    peakBg = paste0(col_batches[chromPeaks(eic_cystine[1, pool_index])[, "sample"]], 20),
    peakPch = 16
    )
legend("bottomright", col = col_batch_id, cex = 0.50,
       horiz = TRUE, inset = c(0, -0.5), xpd = TRUE,
       text.width = 4, bty = "n",
       legend = names(col_batch_id), lty = 1)

plotChromPeakDensity(
    eic_his[1, pool_index], param = param,
    col = paste0(batches_pool, 80),
    peakCol = col_batches[chromPeaks(eic_his[1, pool_index])[, "sample"]],
    peakBg = paste0(col_batches[chromPeaks(eic_his[1, pool_index])[, "sample"]], 20),
    peakPch = 16
    )
legend("bottomright", col = col_batch_id, cex = 0.50,
       horiz = TRUE,inset = c(0, -0.5), xpd = TRUE,
       text.width = 4, bty = "n",
       legend = names(col_batch_id), lty = 1)
```

```{r}
#' Now apply to whole data
param <- PeakDensityParam(sampleGroups = sampleData(nafld)$sample_type,
                          minFraction = 0.75, binSize = 0.015, bw = 1.8)
nafld <- groupChromPeaks(nafld, param = param)

#' Extract pool_sample for better visualization of analysis
pool_index <- sampleData(nafld)$sample_type == "Pool"
pool_sample <- nafld[pool_index, keepAdjustedRtime = TRUE, keepFeatures = TRUE]
```


```{r}
#' Extract chromatogram with signal for isomers 1-Methylhistidine and
#' 3-Methylhistidine
met_mz <- calculateMass("C7H11N3O2")
met_mz <- mass2mz(met_mz, adduct = "[M+H]+")[1,]
chr_test <- chromatogram(nafld, mz = met_mz + c(-0.01, 0.01), rt = c(160, 200))

plotChromPeakDensity(chr_test[1, pool_index], col = paste0(batches_pool, 80),
    peakCol = col_batches[chromPeaks(chr_test[1, pool_index])[, "sample"]],
    peakBg = paste0(col_batches[chromPeaks(chr_test[1, pool_index])[, "sample"]], 20),
    peakPch = 16)
legend("bottomright", col = col_batch_id, cex = 0.50,
       horiz = TRUE,inset = c(0, -0.5), xpd = TRUE,
       text.width = 4, bty = "n",
       legend = names(col_batch_id), lty = 1)
```

```{r include=FALSE}
save(nafld, file = "nafld_after_Cor")
```


## Gap filling

Process to try to rescue some (pre-)selected peaks that are only detected in a subset of samples.
```{r peaks found in a few QC not all, echo=FALSE}
#' Number of missing values
sum(is.na(featureValues(nafld)))
```

We can see quite a large amount of NA values in our data set.

```{r}
nafld <- fillChromPeaks(nafld, param = ChromPeakAreaParam(), chunkSize = 10)

#' How many missing values after
sum(is.na(featureValues(nafld)))
```

With `fillChromPeaks` we could thus rescue signal for all but `r sum(is.na(featureValues(nafld)))` features.

Another way to confirm gap-filling rescued signals. use QC sample signals as basis

```{r Detected vs filled signal}
#' Get only detected signal
vals_detect <- featureValues(nafld, filled = FALSE)[, pool_index]

#' Get detected and filled-in signal
vals_filled <- featureValues(nafld)[, pool_index]

#' Replace detected signal with NA
vals_filled[!is.na(vals_detect)] <- NA

#' Identify features with at least one filled peak
has_filled <- is.na(rowSums(vals_detect))

#' Calculate row averages
avg_detect <- rowMeans(vals_detect, na.rm = TRUE)
avg_filled <- rowMeans(vals_filled, na.rm = TRUE)

#' Restrict to features with at least one filled peak
avg_detect <- avg_detect[has_filled]
avg_filled <- avg_filled[has_filled]

#' plot the values against each other (in log2 scale)
plot(log2(avg_detect), log2(avg_filled),
     xlim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     ylim = range(log2(c(avg_detect, avg_filled)), na.rm = TRUE),
     pch = 21, bg = "#00000080")
grid()
abline(0, 1)
```


Then calculate statistics on these values. below we fit a linear regression line to the data and summarize its results
```{r}
#' fit a linear regression line to the data
l <- lm(log2(avg_filled) ~ log2(avg_detect))
summary(l)
```

```{r include=FALSE}
save(nafld, file = "nafld_after_GF")
```

## Removing features detected in blank
```{r}
blank_index <- sampleData(nafld)$sample_type == "Blank"
blank_sample <- nafld[blank_index, keepAdjustedRtime = TRUE, keepFeatures = TRUE]

#' index of features that are detected in blank samples
idx_fts_blk <- which(featureDefinitions(nafld)$Blank != 0)

#' get something like the average intensity for each feature in blank samples

#' also group the intensity for

#
```

# Next steps

separate blanks from others
```{r}
blank_index <- sampleData(nafld)$sample_type == "Blank"
blank_sample <- nafld[blank_index, keepAdjustedRtime = TRUE, keepFeatures = TRUE]
nafld <- nafld[!blank_index, keepAdjustedRtime = TRUE, keepFeatures = TRUE]

save(blank_sample, file = "blank_nafld.RData")
save(nafld, file = "samples_nafld.RData")
```


Summarized experiment
```{r}
#' Extract results as a SummarizedExperiment
res <- quantify(nafld, method = "sum", filled = FALSE)
assays(res)$raw_filled <- featureValues(nafld, method = "sum",
filled = TRUE )

res

save(res, file = "SumExp_NAFLD.RData")

# also for blanks
res_blank <- quantify(blank_sample, method = "sum", filled = FALSE)
assays(res_blank)$raw_filled <- featureValues(blank_sample, method = "sum",
filled = TRUE )
save(res_blank, file = "SumExp_NAFLD_blank.RData")
```
