---
title: "Exploratory data analysis of the NAFLD metabolomics data set"
author: "Nikola Dordevic, Essi Hantikainen, Johannes Rainer"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
bibliography: references.bib
---

**Modified**: `r file.info("NAFLD_metabolomics_exploratory_analysis.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r biocstyle, echo = FALSE, results = "asis", message = FALSE}
#' rmarkdown format settings
library(BiocStyle)
library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8, fig.height = 7)
BiocStyle::markdown()
```



```{r settings, echo = FALSE}
#' General settings
filename <- "NAFLD_metabolomics_exploratory_analysis"
#' Path to save images to
IMAGE_PATH <- paste0("images/", filename, "/")
if (file.exists(IMAGE_PATH))
    unlink(IMAGE_PATH, recursive = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE)
#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)
```


# Introduction

In this document we perform an unsupervised *exploratory* data analysis of the
metabolomics data set of the NAFLD study to evaluate and identify any relations 
between metabolites and/or participants.

# Data import

Below we load all required packages for the analysis and retrieve the data.

```{r libraries}
library(CompMetaboTools)
library(tidyfr)
library(chrisUtils)
library(chrisData)
library(nafldData)
library(RColorBrewer)
library(pander)
library(pheatmap)
library(DT)
library(ggfortify)
library(ggplot2)
library(reshape2)
library(plotly)
library(rrcov)
library(WGCNA)
library(flashClust)
options(rgl.useNULL = TRUE)
library(rgl)
library(writexl)
```

We next load the data the TDFF data module. 

```{r, message = FALSE}
##
p180 <- data_module(name = "metabolomics_p180", version = "1.0.1.1", path = ".")
met_data <- data(p180)
met_ann <- labels(p180)
stopifnot(all(colnames(met_data) == rownames(met_ann)))
colnames(met_data) <- met_ann$description
rownames(met_ann) <- met_ann$description
## define the columns with peptide concentrations
metabolites <- rownames(met_ann)[met_ann$type != "categorical"] 
met_data <- met_data[, metabolites]
```

Perform log2 transformation!

```{r log2-transformation, message = FALSE}
##
met_data <- log2(met_data)
```

Before proceeding with the data analysis we remove the POOL samples. In addition
we restrict to samples from the NAFLD study.

```{r remove-POOLs-NAFLDs, message = FALSE, warning = FALSE}
met_data <- met_data[grep("^008", rownames(met_data)), ]
```

# General data overview

In this dataset `r ncol(met_data)` metabolites measured from `r nrow(met_data)`
samples collected from the NAFLD study are evaluated. Note that all metabolites 
are log2 transformed.

Distributions of all 175 metabolites are visualized in histograms.

```{r Distributions}
PlotHist <- function(AnlName) {
    Anl <- met_data[, AnlName]
    plot(density(Anl, na.rm = T), col = 2, main = AnlName,
                                               xlab = "Conc.")
    }
anlts <- colnames(met_data)
splits <- rep(seq_len(ceiling(length(anlts) / 12)), each = 12)
splits <- splits[1:length(anlts)]
anlts <- split(anlts, splits)
for (i in seq_along(anlts)) {
    anlts_i <- anlts[[i]]
    par(mfrow = c(3, 4))
    for (anlt in anlts_i)
        PlotHist(anlt)
}
```


Demographics of the NAFLD cohort and evaluation of available trait data.

```{r}
n_anns <- nafldTraitAnnotations()

#' personal data
n_perso <- nafldData("personal")
rownames(n_perso) <- n_perso$AID
#met_data <- n_perso[rownames(met_data), c("AID", "aid_chris", "na_ager", 
#                                          "na_sex", "na_fwup")]
met_data$AID <- n_perso[rownames(met_data), "AID"]
met_data$AID_CHRIS <- n_perso[rownames(met_data), "aid_chris"]
met_data$Age <- n_perso[rownames(met_data), "na_ager"]
met_data$Sex <- n_perso[rownames(met_data), "na_sex"]
met_data$fwup <- n_perso[rownames(met_data), "na_fwup"]


#' na_ager: rounded age
#' na_fwup: follow-up duration CHRIS->NAFLD

tmp <- nafldData("recruitment")
rownames(tmp) <- tmp$AID
met_data$case <- tmp[rownames(met_data), "na_case"]
```

The time span between the two studies is thus between 2 to 5 years. We next
evaluate BMI and fat composition for the study participants, first split by
gender and then by case/control definition.

```{r}
#' BMI, fat etc.
n_anthro <- nafldData("anthro")
rownames(n_anthro) <- n_anthro$AID

met_data$BMI <- n_anthro[rownames(met_data), "naan03"]
met_data$body_fat <- n_anthro[rownames(met_data), "naan04"]
met_data$visc_fat <- n_anthro[rownames(met_data), "naan05"]
```

We next evaluate the clinical examinations performed on all participants.

```{r}
#' Clinical examinations
tmp <- nafldData("clinical")
rownames(tmp) <- tmp$AID
```

We will use and evaluate the following parameters defined by a clinical
examination:

- *naab11*: liver stiffness (higher values are worse). Values for participants
  that have a *naab12* (liver stiffness IQR) > 30 should be either removed or
  taken with care.
- *naab12*: liver stiffness IQR: the IQR of repeated liver stiffness
  measurements. Values below 30 are considered OK.
- *naab13*: liver fat. Can be binarized at an absolute value of 288.
- *naab14*: subjective, visual evaluation of the liver fat. Categorical from 1
  to 3. Values 2 and 3 can be considered steatotic.
- *nascfib4*: fibrosis 4 index. Linear score calculated from standard clinical
  parameters.
- *nascmets*: metabolic syndrome, binary category.

```{r}
met_data$liver_stiffness <- tmp[rownames(met_data), "naab11"]
met_data$liver_stiffness_iqr <- tmp[rownames(met_data), "naab12"]
met_data$liver_fat <- tmp[rownames(met_data), "naab13"]
met_data$visual_fat <- tmp[rownames(met_data), "naab14"]
met_data$fib4 <- tmp[rownames(met_data), "nascfib4"]
met_data$metsym <- tmp[rownames(met_data), "nascmets"]
```

In next table, general overview for all metabolites are presented (mean, 
standard deviation, coefficients of variation of QC samples):

```{r}
Table_1 <- data.frame(
    #Metabolite = metabolites,
    Class = met_ann[metabolites,"analyte_class"],
    Mean = colMeans(met_data[, metabolites], na.rm = TRUE),
    SD = sapply(met_data[, metabolites], sd, na.rm = TRUE))
    #CV_QC = rowData(chris_plasma_proteins)$cv_qc)
datatable(Table_1) |>
formatRound(columns = seq(2, ncol(Table_1)), digits = 3)
```

# Missing value imputation


Percentage of missing values per metabolite:

```{r NAs}

NAsPerc <- sort( 100 * colSums(is.na(met_data[, metabolites])) / 
                 nrow(met_data), decreasing = TRUE)
round(NAsPerc, 2) [1:10]
```

For PCA analyses we impute missing values (missing because of technical 
problems) replacing them with random values from the same distribution than the
actual measurements are.

```{r pca-imputation}
## To perform PCA, full data matrix without NAs is needed. Remaining missing
## values with a flag of -9 are replaced with random numbers from the normal
## distribution, considering all non-missing metabolites (considering their mean
## and standard deviation).

met_data_imp <- met_data

##  define metabolites to impute (only numeric part of data):
MetToImpute_3 <- names(which(colSums(
    is.na(met_data[, metabolites])) != 0))
## make function to impute NAs with random values from normal distribution:
ImputeOneByOne_NDist <- function(VarWithNAs, DataWithNAs) {
    Mean <- mean(DataWithNAs[,VarWithNAs], na.rm = TRUE)
    SDv <- sd(DataWithNAs[,VarWithNAs], na.rm = TRUE)
    MissVal_2 <- which(is.na(DataWithNAs[, VarWithNAs]))
    set.seed(7) ##to have repeatible rundom numbers ## to avoid negative values
    DataWithNAs[MissVal_2, VarWithNAs] <- rnorm(length(MissVal_2),
                                                mean = Mean, sd = SDv)
    DataWithNAs[, VarWithNAs]
}
## impute all at once:
met_data_imp[, MetToImpute_3] <- sapply(MetToImpute_3, function(x)
    ImputeOneByOne_NDist(x, met_data_imp))
```


# Outlier detection

Outlier detection was performed using robust PCA as described by Hubert et al.
2005 and implemented in the rrcov package for R
[@todorovObjectOrientedFrameworkRobust2009].
It is important to be aware of possible outliers, because they can affect the
results of some possible future analyses (e.g. some parametric multivariate
methods like PLS is highly sensitive to them).
In addition, they can provide us insight into some possible technical issues
during measurements or they can detect samples with some interestin biological
properties (with high or low abundances of some metabolites).
In next figure, diagnostic plot of the dataset based on two robust principal
components is presented:

```{r, PCA_Hubert, fig.path = IMAGE_PATH, fig.width=12, fig.height=10}
## outlier identification:
par(mfrow=c(1,1))
pca_Hubert <- PcaHubert(met_data_imp[, metabolites])
plot(pca_Hubert)
```

Looking at the plot there are no clear outliers.


# General unsupervised grouping of samples (PCA)

In next figures, scores and loadings from the PCA are presented considering all
metabolites.
Prior the PCA, data is normalized to 0 mean and standard deviation of 1
(z-score or autoscaling).
Plot shows the scores and loadings of all subjects in the space of the first 3
PCs, with almost 39% of total variances explained.

In next PCA, samples are colored acordin sample sex:

```{r}
##
scaledData <- scale(met_data_imp[, metabolites], scale = TRUE)
scPCA <- svd(scaledData)
scPCA.scores <- scPCA$u %*% diag(scPCA$d)    ## scores
scPCA.loadings <- scPCA$v                    ## loadings
scPCA.variances <- round(100*((scPCA$d^2)/sum(scPCA$d^2)),1)[1:10]
rownames(scPCA.loadings) <- metabolites

plot3d(scPCA.scores[,1], scPCA.scores[,2], scPCA.scores[,3], xlab=paste("PC 1(",
       scPCA.variances[1], "%)", sep = ""), ylab = paste("PC 2(",
       scPCA.variances[2], "%)", sep = ""), zlab = paste("PC 3(",
       scPCA.variances[3], "%)", sep = ""), pch = 19,
       col = as.integer(met_data_imp$Sex))
#text3d(39*scPCA.loadings[,1:3], texts=rownames(scPCA.loadings), col="darkgrey")
#coords <- NULL
#for (i in 1:nrow(scPCA.loadings)) {
#  coords <- rbind(coords, rbind(c(0,0,0),39*scPCA.loadings[i,1:3]))}
#lines3d(coords, col="darkgrey", lwd=1)
rglwidget()
```

There is a discrimination between males and females!


In next PCA, study samples are colored by case/control definition:

```{r}

plot3d(scPCA.scores[,1], scPCA.scores[,2], scPCA.scores[,3], xlab=paste("PC 1(",
       scPCA.variances[1], "%)", sep = ""), ylab = paste("PC 2(",
       scPCA.variances[2], "%)", sep = ""), zlab = paste("PC 3(",
       scPCA.variances[3], "%)", sep = ""), pch = 19,
       col = as.integer(met_data_imp$case))
rglwidget()
```

Also here, there is moderate shift beetween 2 groups!


# General clustering of analytes

In this section we aim to identify metabolites with highly correlated (or anti
correlated) concentrations to potentially get some more detailed insides into
the dataset. For that, Pearson correlation coefficients between all metabolites
are computed and visualized.  To detect groups with correlated metabolites which
might belong from the same class, correlation network analysis is
performed. 


## Correlation analyses

Pearson's correlation coefficients between all metabolites are computed, grouped
and visualized:

```{r, corr-plot-features, fig.path = IMAGE_PATH, fig.width = 12, fig.height = 10}
cormat <- cor(met_data[, metabolites], method = "pearson", 
              use = "pairwise.complete.obs")
reorder_cormat <- function(cormat){
    ## Use correlation between variables as distance
    dd <- as.dist((1-cormat)/2)
    hc <- hclust(dd)
    cormat <- cormat[hc$order, hc$order]
}
## Get upper triangle of the correlation matrix
get_upper_tri <- function(cormat){
    cormat[lower.tri(cormat)]<- NA
    return(cormat)
}
# Reorder the correlation matrix
cormat <- reorder_cormat(cormat)
upper_tri <- get_upper_tri(cormat)
# Melt the correlation matrix
melted_cormat <- melt(upper_tri, na.rm = TRUE)
# Create a ggheatmap
ggpHMap <- ggplot(melted_cormat, aes(Var2, Var1, fill = value)) +
geom_tile(color = "white") + scale_fill_gradient2(low = "blue", high = "red",
mid = "white", midpoint = 0, limit = c(-1,1), space = "Lab",
name="Pearson\nCorrelation") + theme_minimal() +
theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 10, hjust = 1),
axis.text.y = element_text(vjust = 1, size = 10, hjust = 1),
axis.title.x = element_blank(), axis.title.y = element_blank(),
panel.grid.major = element_blank(), panel.border = element_blank(),
panel.background = element_blank(), axis.ticks = element_blank(),
legend.justification = c(1, 0), legend.position = c(0.6, 0.7),
legend.direction = "horizontal") + coord_fixed()
ggplotly(ggpHMap)
```

There is a large number of metabolites and figure is not very clear. Please zoom 
in the part of figure which you want to analyse further.
All correlation coefficients are summariezed in table and ordered according
the Pearson correlation coefficients:

```{r, message = FALSE}
CorTable <- melted_cormat[order(melted_cormat[,3],decreasing=TRUE),]
colnames(CorTable) <- c("Prot. 1", "Prot. 2", "Corr")
## exclude the results between the same metabolites
CorTable_2 <- CorTable[-c(which(CorTable[, 1] == CorTable[,2])),]
datatable(CorTable_2, caption = "Pearson correlations between all proteins") %>%
formatRound(columns = "Corr", digits = 4)
```

In addition we cluster the metabolites based on their pairwise correlation
coefficient.

```{r corr-heatmap-features, fig.path = IMAGE_PATH, fig.width = 16, fig.height = 14}
cormat2 <- cormat
rownames(cormat2) <- met_ann[rownames(cormat),"analyte_class"]
pheatmap(cormat2)
```


## Correlation network analysis

To identify groups of highly corraleted metabolites, correlation network 
analysis is performed. Weighted gene coexpression network analysis (WGCNA) 
implemented in the WGCNA R package is used for determining groups (modules) of 
higlhy correlated metabolites
[@langfelderWGCNAPackageWeighted2008][@langfelderFastFunctionsRobust2012].
Modules are identified using unsupervised clustering Dynamic Branch Cut method
with a minimal number of 3 metabolites per module.
Note that modules are randomly assigned colors, while unassigned metabolites are
grouped in the grey module.
Results are visualized with hierarchical clustering dendrogram and correlations
between all metabolites grouped together are computed.

In next figure, there is a hierarchical cluster tree of the 175 metabolites,
color-coded according to the identified modules (groups):

```{r WGCNA_figure, fig.path = IMAGE_PATH, results='hide', message=FALSE, warning=FALSE, fig.width=32, fig.height=20, out.width='1600px'}
## WGCNA
# Choose a set of soft thresholding powers
powers <- c(1:20) # in practice this should include powers up to 20.
# choose power based on SFT criterion
sft <- pickSoftThreshold(met_data[, metabolites], powerVector = powers,
                         networkType = "unsigned")
## https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/faq.html
## point 6: low connectivity; soft-thresholding power = 6
adjacencyA1 <- adjacency(met_data[, metabolites], power = sft$powerEstimate, 
                         type = "unsigned")
diag(adjacencyA1) <- 0
dissTOMA1 <- 1 - TOMsimilarity(adjacencyA1, TOMType = "unsigned")
colnames(dissTOMA1) <- colnames(adjacencyA1)
biomTreeA1 <- flashClust(as.dist(dissTOMA1), method = "average")
mColorh <- NULL
tree <- cutreeHybrid(dendro = biomTreeA1, pamStage = FALSE, minClusterSize = 3,
                     cutHeight = 0.99, deepSplit = 3, distM = dissTOMA1)
mColorh <- cbind(mColorh,labels2colors(tree$labels))
plotDendroAndColors(biomTreeA1, mColorh, paste("dpSplt =", 3),
                    main = "Hierarchical cluster",dendroLabels = NULL)
```

If parameters are in the same and long branch, means that they are correlated.
Because of large number of metabolites with a long names figure is not clear, 
thus tables are provided.
Numbers of the metabolites associatet to each module are presented here:

```{r}
# To take out module information
GroupVar <- cbind(metabolites, mColorh)
orderedGV <- GroupVar[order(GroupVar[, 2]), ]
## number of variables per module:
sort(table(orderedGV[, 2]))
```

We next create tables for all groups of correlated metabolites. Shown are the 
analytical class and the mean correlation of the metabolite against all 
metabolites within the cluster (including itself).

```{r, results = "asis"}
clusters <- split(GroupVar[, 1], GroupVar[, 2])
clusters <- clusters[order(lengths(clusters))]

clust_table <- function(x) {
    cormat_x <- cormat[x, x]
    data.frame(
        Class  = met_ann[x,"analyte_class"],
        mean_cor = rowMeans(cormat_x))
}

for (cluster in clusters) {
    pandoc.table(clust_table(cluster), style = "rmarkdown", split.table = "Inf")
}
```

Most of the clusters with highly correlated metabolites make biologically
sense. 

In next table, identified correlated metabolites per module with correlation
coefficients are presented:

```{r}
##  Take module membership info and check correlations per module!
corPerModul <- function (Data, colour) {
colourModule <- round(cor(Data[,orderedGV[which(orderedGV[,2]==colour),1]],
                          use = "pairwise.complete.obs"), 4)
colourModule[lower.tri(colourModule)] <- NA
diag(colourModule) <- NA
colourModule = as.data.frame(as.table(colourModule))
colourModule = na.omit(colourModule)
colourModule[,"Module"] <- colour
return(colourModule)
}
cluster_colors <- unique(mColorh)
cluster_colors <- cluster_colors[cluster_colors != "grey"]
CorrPerModule <- do.call(rbind.data.frame,
                         lapply(cluster_colors, function(z)
                             corPerModul(met_data[, metabolites], z)))
colnames(CorrPerModule)[3] <- "Corr"
datatable(CorrPerModule, caption = "Correlations per module")
```

Seems that there are moderate to strong correlations between metabolites which
belongs to the same module (group).


# Session information

R packages used for the analysis:

```{r}
sessionInfo()
```

# References
