---
title: "Steatosis associated metabolites, lipids and proteins"
author: "Marilyn De Graeve, Essi Hantikainen, Johannes Rainer, Christoph Grander"
affiliation: "Eurac Research, Bolzano, Italy"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
editor_options:
  markdown:
    wrap: 78
---

**Modified**: `r file.info("NAFLD_steatosis_associations.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r style, message = FALSE, echo = FALSE, warning = FALSE, results = "asis"}
library("BiocStyle")
library("kableExtra")
library("knitr")
library("rmarkdown")
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7)
set.seed(1)
```

# Introduction

During this workflow, exploratory analyis (Spearman correlations) are performed
on the relatively quantified Scanning SWATH proteomics data (148 proteins) from
the metabolic dysfunction-associated steatotic liver disease (MASLD)
cohort. MASLD, previously known as NAFLD is the most chronic liver disease,
affecting more than 30% of global population. The MAFLD cohort is a subset of
approximately 353  participants from the Cooperative Health Research in South
Tyrol (CHRIS) study, which were requested to return for a second resampling and
additional phenotyping. The NAFLD cohort represent an unhealty group with
various degrees of steatotic (non-alcoholic) liver disease classifications. See
[nafld_data_overview.Rmd](nafld_data_overview.Rmd) (respectively the rendered
html report) for an overview of the data set and the definition of the
phenotype.

This documents describes the association analysis between circulating
metabolites, lipids and proteins with liver steatosis.

See the `nafld_data_overview 1.html` report for further
information regarding the SLD sub-classification.

Regarding steatosis, i.e., abnormal retention of fat (lipids) within a cell or
organ. In this scope: liver steatosis.

The variables containing this parameter are:

Presence of steatotic liver disease (SLD): sonographic evaluation, steatose
score of 2 – 3.

- *naab13*: liver fat. Can be binarized at an absolute value of 288. "liver_fat"
- *naab14*: subjective, visual evaluation of the liver fat. Categorical from 1
  to 3. Values 2 and 3 can be considered steatotic. "visual_fat"

This parameter will be used as the target for a biomarker search in *MASLD*
participants.


# Setup

## Directories

```{r directories, echo = FALSE}
#' General settings
filename <- "NAFLD_steatosis_associations"

#' Path to save images to; remove if exists.
IMAGE_PATH <- paste0("images/", filename, "/")
if (dir.exists(IMAGE_PATH)) unlink(IMAGE_PATH, recursive = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE)

#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)

#' Path to store xls results files
XLSX_PATH <- file.path("data", "xlsx", filename)
if (dir.exists(XLSX_PATH)) unlink(XLSX_PATH, recursive = TRUE)
dir.create(XLSX_PATH, recursive = TRUE)

#' Path to the data
#DATA_PATH <- '/home/mdegraeve/Documents/Files/Work_Eurac/Data/NAFLD' #FINDME!
DATA_PATH <- "."

DATASET <- "CHRIS-NAFLD, none and MASLD"
```

## Packages

```{r packages, message=FALSE}
#' Load libraries
library(chrisUtils)
library(DT)
library(ggfortify)
library(ggplot2)
library(pander)
library(pheatmap)
library(plotly)
library(RColorBrewer)
library(readxl)
library(reshape2)
library(tidyfr)
library(vioplot)
library(writexl)
```


# Data preparation and setup

## Circulating proteins (tprot) dataset (Scanning SWATH)

The data from proteins was converted into the TDF format and can hence be
loaded and imported using the `tidyfr` R package. Now we read the full data
which includes general information for each measurement.

```{r load_prot_dataset, message = FALSE}
#' load TDF using tidyfr
md1 <- data_module(name = "chris_scanningswath_proteins", version = "1.0.0.3",
                   path = DATA_PATH)
prot_data <- data(md1)
prot_ann <- labels(md1)
stopifnot(all(colnames(prot_data) == rownames(prot_ann)))
colnames(prot_data) <- prot_ann$description
rownames(prot_ann) <- prot_ann$description

#' define the columns with protein concentrations
proteins <- rownames(prot_ann)[prot_ann$type != "categorical"]
proteins <- rownames(prot_ann)[prot_ann$Genes != ""]

#' select proteins dataframe
prot_data <- prot_data[, proteins]
```

The loaded data contains the samples as rows, and the protein names as
colums. The measurements of the `prot_data` object are the relative
concentrations of protein per sample. No data transformation is performed as the
abundances are already reported in log2 scale, a normal distribution of the
proteomics is assumed.

Before proceeding with the data analysis we remove the POOL samples. In addition
we restrict to samples from the NAFLD study. The number of samples and proteins
are:

```{r subset to NAFLD set, message = FALSE, warning = FALSE}
prot_data <- prot_data[grep("^008", rownames(prot_data)), ]
dim(prot_data)
```

## Targeted metabolomics (tmetab) dataset (Biocrates p180 kit)

The data from metabolites was converted into the TDF format and can hence be
loaded and imported using the `tidyfr` R package. Now we read the full data
which includes general information for each measurement.

```{r load_tmetb_dataset, message = FALSE}
#' load TDF using tidyfr
p180 <- data_module(name = "metabolomics_p180", version = "1.0.1.3",
                    path = DATA_PATH)
met_data <- data(p180)
met_ann <- labels(p180)
stopifnot(all(colnames(met_data) == rownames(met_ann)))
colnames(met_data) <- met_ann$description
rownames(met_ann) <- met_ann$description

#' define the columns with metabolite concentrations
metabolites <- rownames(met_ann)[met_ann$type != "categorical"]

#' select metabolites dataframe
met_data <- met_data[, metabolites]
#' Remove the individuals with more than 80% missing values
rem <- apply(met_data, MARGIN = 1, function(z) {
    sum(is.na(z)) > length(z) * 0.8
})
met_data <- met_data[!rem, ]
```

The loaded data contains the samples as rows, and the metabolite names as
colums. The measurements of the `met_data` object are the absolute
concentrations of metabolite per sample, in natural scale.

To ensure a normal distribution of the metabololits, we perform log2
transformation.

```{r m log2-transformation, message = FALSE}
#' log2 scaling of absolute abundances metabolites
met_data <- log2(met_data)
```

Before proceeding with the data analysis we remove the POOL samples. In addition
we restrict to samples from the NAFLD study. The number of samples and
metabolites are listed below.

```{r m subset to NAFLD set, message = FALSE, warning = FALSE}
met_data <- met_data[grep("^008", rownames(met_data)), ]
dim(met_data)
```

## GA2LEN diet data

Dietary information, i.e., GA2LEN food frequency questionaires, from the MASLD
study were loaded and included in the analysis.

The newly created dietary indices from Essi are saved under
NAFLD_dietary_indices_data_Marilyn. And note for linear modeling: "you should
also correct your models for total energy intake (just have it as covariate in
the model, nanu006a)". This was added in the section of dietary lms.

```{r NAFLD-indices}
#' load the NAFLD_dietary_indices_data
fl <- file.path(DATA_PATH, "GALAN", "NAFLD_dietary_indices_data_Marilyn.txt")
galan <- read.table(fl, header = TRUE, sep = "\t", dec = ",")  |>
    as.data.frame()
rownames(galan) <- paste0("00", as.character(galan$AID))

#' replace the aids columns with 00 start and as a string
galan$AID <- rownames(galan)

#' variables
ffq_items <- colnames(galan)[2:ncol(galan)] #excl AID col
```

Add the 4 macro nutrient categories from GA2LEN and add to the `ffq_items`.

```{r NAFLD_food_macros}
library(nafldData)
#' - nanu003 (protein)
#' - nanu004 (total fat)
#' - nanu005 (total carbs),
#' - nanu019b (saturated fatty acids)
#' - nanu023b (monounsaturated fatty acids)
#' - nanu025b (polyunsaturated fatty acids)
#' - nanu028b (trans fatty acids)

ffq <- nafldData("FFQ")
rownames(ffq) <- ffq$AID

galan$protein <- ffq[rownames(galan), "nanu003"]
galan$total_fat <- ffq[rownames(galan), "nanu004"]
galan$total_carbs <- ffq[rownames(galan), "nanu005"]
galan$saturated_fatty_acids <- ffq[rownames(galan), "nanu019b"]
galan$monounsaturated_fatty_acids <- ffq[rownames(galan), "nanu023b"]
galan$polyunsaturated_fatty_acids <- ffq[rownames(galan), "nanu025b"]
galan$trans_fatty_acids <- ffq[rownames(galan), "nanu028b"]

#' variables
ffq_items <- colnames(galan)[2:ncol(galan)] #excl AID col
```

## Metadata (excel)

All available trait (phenotype) data was defined in the
[nafld_data_overview.Rmd](nafld_data_overview.Rmd), is loaded below and merged
with the log2 transformed proteomics data of the NAFLD cohort.

```{r add phenotype dataa}
#' load xlsx file instead of below
handle <- file.path("data", "masld_phenotypes.xlsx")
masld_phenotype_df <- read_excel(file.path(DATA_PATH, handle)) |>
    as.data.frame()
rownames(masld_phenotype_df) <- masld_phenotype_df$AID
metadata <- colnames(masld_phenotype_df)[3:ncol(masld_phenotype_df)]

#' make sure sld_type is a factor with group 'none' as the baseline
masld_phenotype_df$sld_type <- factor(masld_phenotype_df$sld_type,
                                      levels = c('none', 'MASLD',
                                                 'MetALD', 'SLD'))
print('nr of samples per sld types:')
table(masld_phenotype_df$sld_type)
```

## Merge all data sets

Merging all data tables. Note that we restrict the data to participants for
which we have proteomics **and** metabolomics data. This also ensures that any
potential study drop-outs are no longer considered in the data (as their
proteomics or metabolomics data was removed).

There were less AIDs present in the dietary data, so we restrict with the data
to participants for which we have proteomics **and** metabolomics **and**
dietary data.

```{r merge-dfs}
AIDs1 <- intersect(rownames(prot_data), rownames(met_data))
AIDs2 <- intersect(rownames(masld_phenotype_df), rownames(galan))
AIDs <- intersect(AIDs1, AIDs2)

#' Subset all datasets to the same IDs
prot_data <- prot_data[AIDs, ]
met_data <- met_data[AIDs, ]
masld_phenotype_df <- masld_phenotype_df[AIDs, ]
galan <- galan[AIDs, ]

#' merge with proteomics data
prot_data$AID <- rownames(prot_data)
prot_data <- merge(prot_data, masld_phenotype_df, by = "AID")

#' merge all sets
met_data <- as.data.frame(met_data)
met_data$AID <- rownames(met_data)
met_prot_data <- merge(prot_data, met_data, by = "AID")

#' merge with metabolomics data only
met_data <- merge(met_data, masld_phenotype_df, by = "AID")

#' merge all sets
met_prot_data <- as.data.frame(met_prot_data)
rownames(met_prot_data) <- met_prot_data$AID
gal_met_prot_data <- merge(met_prot_data, galan, by = "AID")
rownames(gal_met_prot_data) <- gal_met_prot_data$AID
```

All data sets are subset to `r length(AIDs)` study participants.

## Clean data and restrict to relevant participants

The data is further processed and cleaned to remove unreliable
measurements. Subsequently the data is restricted to relevant participants. We
thus next:

- remove unreliable liver fat measurements
- exclude participants with MetALD and SLD to ensure a more homogeneous data set
- remove participants with missing values in relevant variables
- define variables and phenotype(s) for the analysis

Add the information from the [nafld_data_overview.Rmd](nafld_data_overview.Rmd)
report regarding the steatosis, using the parameter naab11
(higher values are worse). Values for participants that have a naab12
(liver stiffness IQR) > 30 should be either removed or taken with care. Below we
remove these unreliable measurements.

```{r load-liver-param2}
#' remove samples with  naab12 (liver stiffness IQR) > 30
gal_met_prot_data$liver_stiffness[
                      which(gal_met_prot_data$liver_stiffnes_iqr > 30)] <- NA
```

We next remove the MetALD and SLD participants.

```{r keep only none and MASLD types}
#' because problem 2 proteins only NAs in some sld types,
#' keep only none and MASLD
gal_met_prot_data <- gal_met_prot_data[gal_met_prot_data$sld_type %in%
                                       c('none', 'MASLD'), ] #
gal_met_prot_data$sld_type <- droplevels(gal_met_prot_data$sld_type)
gal_met_prot_data$sld_type <- factor(gal_met_prot_data$sld_type)
```

We next remove participants with missing values for relevant variables:
- liver fat measurement

```{r}
#' Define variables for which we don't allow missing values
no_na_var <- c("liver_fat")
has_na <- lapply(gal_met_prot_data[, no_na_var, drop = FALSE], is.na) |>
    do.call(what = cbind)
#' Remove rows (participants) with a missing value in any of the above
#' variables
gal_met_prot_data <- gal_met_prot_data[rowSums(has_na) == 0, , drop = FALSE]
```

Remove fibrosis outlier values: **CURRENTLY NOT DONE**.
Previous outlier removal removed a participant with a FIB4 score of 7.8. This
participant was originally removed. The second outlier removal excluded
a participant with a liver_stiffness value of 34. Both participants are still
present, as fibrosis is not the phenotype of interest anymore.

We next define the categorical variables from numeric measurements:

Add a categoric liver_stiffness category too, using both the definition of
Christoph (7, 15 cutoffs) and in literature (7, 12.5 kPa cutoffs).

```{r liver-stiffness-category}
#' naab11 according to Christoph
gal_met_prot_data$liver_stiffness_cat1 <- 2
gal_met_prot_data$liver_stiffness_cat1[
                      is.na(gal_met_prot_data$liver_stiffness)] <- NA
gal_met_prot_data$liver_stiffness_cat1[
                      which(gal_met_prot_data$liver_stiffness < 7)] <- 1
gal_met_prot_data$liver_stiffness_cat1[
                      which(gal_met_prot_data$liver_stiffness > 15)] <- 3
gal_met_prot_data$liver_stiffness_cat1 <-
    factor(gal_met_prot_data$liver_stiffness_cat1)

#' naab11 according to literature
gal_met_prot_data$liver_stiffness_cat2 <- 2
gal_met_prot_data$liver_stiffness_cat2[
                      is.na(gal_met_prot_data$liver_stiffness)] <- NA
gal_met_prot_data$liver_stiffness_cat2[
                      which(gal_met_prot_data$liver_stiffness < 7)] <- 1     #mild or absent fibrosis
gal_met_prot_data$liver_stiffness_cat2[
                      which(gal_met_prot_data$liver_stiffness > 12.5)] <- 3  #cirrhosis
gal_met_prot_data$liver_stiffness_cat2 <-
    factor(gal_met_prot_data$liver_stiffness_cat2)
#' appearantly differnet cutoffs, depending what background...
#' https://www.mskcc.org/cancer-care/patient-education/understanding-your-fibroscan-results
```

Add a categoric liver_fat (i.e. steatosis) category, using the following
cutoffs:

- "normal": A score below 238 dB/m means the amount of fatty change in your
  liver is not higher than normal.
- grade steatosis S1: 238 to 260 dB/m
- grade steatosis S2: 260 to 290 dB/m
- grade steatosis S3: 290 to 400 dB/m

Depending on the distribution, look only at the extreme normal - steatosis.

```{r liver-fat-category}
#' naab13 make 4 groups
gal_met_prot_data$liver_fat_cat <- NA
gal_met_prot_data$liver_fat_cat[which(gal_met_prot_data$liver_fat < 238)] <- 1 #normal
gal_met_prot_data$liver_fat_cat[which(gal_met_prot_data$liver_fat >= 238)] <- 2
gal_met_prot_data$liver_fat_cat[which(gal_met_prot_data$liver_fat > 260)] <- 3
gal_met_prot_data$liver_fat_cat[which(gal_met_prot_data$liver_fat > 290)] <- 4
gal_met_prot_data$liver_fat_cat <- factor(gal_met_prot_data$liver_fat_cat)
```

At last we ensure categorical variables are encoded as a factor. Also, we create
additional variable for age per 10y, because coefficients and effect sizes
correspond to one year difference, and the annual difference is
neglectible. Dividing age with 10, difference in coefficents and effect sizes
would correspond to 10 years difference and it would be easier to define
meaningful cut-off. Finally, we add a visc_subcut_fat variable on the dataset,
comprising the ratio of visceral to subcutaneous fat thickness.

```{r define-age10-variable, message = FALSE}
#' Convert categorical variables to factors
col_cat <- c("case", "self_DM", "med_DM", "fasting", "no_drinking",
             "med_BP", "med_lip", "med_MS", "T2D", "Sex", "MASLD")
for (i in col_cat)
    gal_met_prot_data[, i] <- as.factor(gal_met_prot_data[, i])

#' define age per 10y group
gal_met_prot_data$Age_10 <- gal_met_prot_data$Age / 10
gal_met_prot_data$visc_subcut_fat <- gal_met_prot_data$visc_fat /
    gal_met_prot_data$body_fat
```

Add additional variables.

```{r}
lab <- nafldData("Labdata")
rownames(lab) <- lab$AID
lab <- lab[gal_met_prot_data$AID, ]

#' LDL
gal_met_prot_data$LDL <- lab$nalp18
```

## Summary of the final data set

```{r}
summary(gal_met_prot_data[, c("Age", "Sex", "T2D", "MASLD", "FIB4", "BMI",
                              "liver_stiffness", "liver_fat", "liver_fat_cat",
                              "visc_subcut_fat")])
```

The variables that are not allowed to conatin NAs are: sex, age and liver_fat.
Participants with NA in BMI, and those 2 with extreme liver stiffness/FIB4
scores are not removed.

# Demographics

We create next an overview table (Table 1 demographics) for the data set.

```{r, results = "asis"}
tmp <- split(gal_met_prot_data, gal_met_prot_data$liver_fat_cat)

tab <- lapply(tmp, function(z) {
    data.frame(n = nrow(z),
               Age = mean(z$Age),
               Age_sd = sd(z$Age),
               Women = sum(z$Sex == "Female") / nrow(z) * 100,
               Men = sum(z$Sex == "Male") / nrow(z) * 100,
               BMI = mean(z$BMI, na.rm = TRUE),
               BMI_sd = sd(z$BMI, na.rm = TRUE),
               visc_fat = mean(z$visc_fat, na.rm = TRUE),
               visc_fat_sd = sd(z$visc_fat, na.rm = TRUE),
               body_fat = mean(z$body_fat, na.rm = TRUE),
               body_fat_sd = sd(z$body_fat, na.rm = TRUE),
               waist_circ = mean(z$waist_circumference, na.rm = TRUE),
               waist_circ_sd = sd(z$waist_circumference, na.rm = TRUE),
               Glucose = mean(z$glucose, na.rm = TRUE),
               Glucose_sd = sd(z$glucose, na.rm = TRUE),
               HbA1c = mean(z$HbA1c, na.rm = TRUE),
               HbA1c_sd = sd(z$HbA1c, na.rm = TRUE),
               T2D = sum(z$T2D == "TRUE") / nrow(z) * 100,
               T2D_med = sum(z$med_DM == "Yes" , na.rm = TRUE) / nrow(z) * 100,
               Triglycerides = mean(z$triglycerides, na.rm = TRUE),
               Triglycerides_sd = sd(z$triglycerides, na.rm = TRUE),
               HDL = mean(z$HDL, na.rm = TRUE),
               HDL_sd = sd(z$HDL, na.rm = TRUE),
               LDL = mean(z$LDL, na.rm = TRUE),
               LDL_sd = sd(z$LDL, na.rm = TRUE),
               lip_med = sum(z$med_lip == "Yes", na.rm = TRUE) / nrow(z) * 100,
               MASLD = sum(z$MASLD == "TRUE", na.rm = TRUE) / nrow(z) * 100,
               daily_alcohol = mean(z$daily_total_alcohol, na.rm = TRUE),
               daily_alcohol_sd = sd(z$daily_total_alcohol, na.rm = TRUE),
               liver_fat = mean(z$liver_fat, na.rm = TRUE),
               liver_fat_sd = sd(z$liver_fat, na.rm = TRUE),
               liver_stiffness = mean(z$liver_stiffness, na.rm = TRUE),
               liver_stiffness_sd = sd(z$liver_stiffness, na.rm = TRUE),
               FIB4 = mean(z$FIB4, na.rm = TRUE),
               FIB4_sd = sd(z$FIB4, na.rm = TRUE)
               )
}) |>
    do.call(what = rbind) |>
    t()

pandoc.table(tab, style = "rmarkdown", split.tables = Inf, digits = 3)
```

# Exploring steatosis

We first evaluate relationships between steatosis and other
parameters, first for anthropometric variables age, sex and BMI.

```{r, fig.width = 8, fig.height = 8}
col_liver_fat_cat <- brewer.pal(7, "Set1")[c(7, 1, 5, 6)]

supp_figure_1 <- function() {
    par(mfrow = c(2, 3))
    vioplot(split(gal_met_prot_data$Age,
                  gal_met_prot_data$liver_fat_cat),
            col = paste0(col_liver_fat_cat, "ce"))
    mtext(side = 1, text = "steatosis grade", line = 3, cex = par("cex.axis"))
    mtext(side = 2, text = "Age", line = 2.5, cex = par("cex.axis"))
    mtext(side = 3, text = "A", line = 1.5, cex = 2, adj = -0.3)
    grid(nx = NA, ny = NULL)
    l <- split(gal_met_prot_data$liver_fat_cat, gal_met_prot_data$Sex)
    m <- do.call(cbind, lapply(l, function(z) as.numeric(table(z)) * 100 / length(z)))
    barplot(m, ylab = "%", col = paste0(col_liver_fat_cat, "ce"))
    mtext(side = 3, text = "B", line = 1.5, cex = 2, adj = -0.3)
    vioplot(split(gal_met_prot_data$BMI,
                  gal_met_prot_data$liver_fat_cat),
            col = paste0(col_liver_fat_cat, "ce"))
    mtext(side = 1, text = "steatosis grade", line = 3, cex = par("cex.axis"))
    mtext(side = 2, text = "BMI", line = 2.5, cex = par("cex.axis"))
    mtext(side = 3, text = "C", line = 1.5, cex = 2, adj = -0.3)
    grid(nx = NA, ny = NULL)

    vioplot(split(gal_met_prot_data$body_fat,
                  gal_met_prot_data$liver_fat_cat),
            col = paste0(col_liver_fat_cat, "ce"))
    grid(nx = NA, ny = NULL)
    mtext(side = 1, text = "steatosis grade", line = 3, cex = par("cex.axis"))
    mtext(side = 2, text = "% body fat", line = 2.5, cex = par("cex.axis"))
    mtext(side = 3, text = "D", line = 1.5, cex = 2, adj = -0.3)

    vioplot(split(gal_met_prot_data$visc_fat,
                  gal_met_prot_data$liver_fat_cat),
            col = paste0(col_liver_fat_cat, "ce"))
    grid(nx = NA, ny = NULL)
    mtext(side = 1, text = "steatosis grade", line = 3, cex = par("cex.axis"))
    mtext(side = 2, text = "% visceral fat", line = 2.5, cex = par("cex.axis"))
    mtext(side = 3, text = "E", line = 1.5, cex = 2, adj = -0.3)

    vioplot(split(gal_met_prot_data$waist_circumference,
                  gal_met_prot_data$liver_fat_cat),
            col = paste0(col_liver_fat_cat, "ce"))
    grid(nx = NA, ny = NULL)
    mtext(side = 1, text = "steatosis grade", line = 3, cex = par("cex.axis"))
    mtext(side = 2, text = "waist circumference", line = 2.5,
          cex = par("cex.axis"))
    mtext(side = 3, text = "F", line = 1.5, cex = 2, adj = -0.3)
}

## l <- lm(Age ~ liver_fat_cat, data = gal_met_prot_data)
## summary(l)

## l <- lm(BMI ~ liver_fat_cat, data = gal_met_prot_data)
## summary(l)
## l <- lm(BMI ~ liver_fat_cat + Sex + Age_10, data = gal_met_prot_data)
## summary(l)


supp_figure_1()

png(file.path(IMAGE_PATH, "Supp_Figure_1.png"), width = 10, height = 8,
    units = "cm", res = 600, pointsize = 6)
supp_figure_1()
dev.off()

```

- No clear relationship between steatosis grade and age.
- Clear and strongly significant relationship between BMI and steatosis grade.
- Higher percentage of highest grade steatosis in men compared to women.

Conclusion: will not adjust for BMI, but will consider Sex and age.

Other fat-related measurements: body fat, visceral fat and waist circumference.

```{r}
l <- lm(body_fat ~ Age_10 + Sex + liver_fat_cat, data = gal_met_prot_data)
summary(l)

l <- lm(visc_fat ~ Age_10 + Sex + liver_fat_cat, data = gal_met_prot_data)
summary(l)

l <- lm(waist_circumference ~ Age_10 + Sex + liver_fat_cat,
        data = gal_met_prot_data)
summary(l)
```

- Strong association between body fat and steatosis grade, in particular if
  adjusted for age and sex.
- Same for visceral fat and waist circumference. Even a bit more obvious.


## Compare steatosis with other liver health related traits

FIB4, liver stiffness, T2D and MASLD.

```{r, fig.width = 10, fig.height = 4}

figure_1 <- function() {
    par(mfrow = c(1, 4), mar = c(4.8, 4.1, 4.1, 1))
    vioplot(split(gal_met_prot_data$FIB4,
                  gal_met_prot_data$liver_fat_cat),
            col = paste0(col_liver_fat_cat, "ce"), xaxt = "n")
    grid(nx = NA, ny = NULL)
    axis(side = 1, at = 1:4, labels = c(0, 1, 2, 3))
    mtext(side = 1, text = "steatosis grade", line = 3, cex = par("cex.axis"))
    mtext(side = 2, text = "FIB4", line = 2.5, cex = par("cex.axis"))
    mtext(side = 3, text = "A", line = 1.5, cex = 2, adj = -0.4)

    vioplot(split(gal_met_prot_data$liver_stiffness,
                  gal_met_prot_data$liver_fat_cat),
            col = paste0(col_liver_fat_cat, "ce"), xaxt = "n")
    grid(nx = NA, ny = NULL)
    axis(side = 1, at = 1:4, labels = c(0, 1, 2, 3))
    mtext(side = 1, text = "steatosis grade", line = 3, cex = par("cex.axis"))
    mtext(side = 2, text = "liver stiffness", line = 2.5, cex = par("cex.axis"))
    mtext(side = 3, text = "B", line = 1.5, cex = 2, adj = -0.4)

    l <- split(gal_met_prot_data$liver_fat_cat, gal_met_prot_data$T2D)
    m <- do.call(cbind, lapply(l, function(z) as.numeric(table(z)) * 100 / length(z)))
    b <- barplot(m, ylab = "", col = paste0(col_liver_fat_cat, "ce"),
                 main = "T2D", xaxt = "n")
    mtext(side = 2, line = 2.5, "%", cex = par("cex.axis"))
    mtext(side = 1, line = 2, at = b, c("No", "Yes"), cex = par("cex.axis"))
    mtext(side = 3, text = "C", line = 1.5, cex = 2, adj = -0.4)

    l <- split(gal_met_prot_data$liver_fat_cat, gal_met_prot_data$MASLD)
    m <- do.call(cbind, lapply(l, function(z) as.numeric(table(z)) * 100 / length(z)))
    b <- barplot(m, ylab = "", col = paste0(col_liver_fat_cat, "ce"),
                 main = "MASLD", xaxt = "n")
    mtext(side = 2, line = 2.5, "%", cex = par("cex.axis"))
    mtext(side = 1, line = 2, at = b, c("No", "Yes"), cex = par("cex.axis"))
    mtext(side = 3, text = "D", line = 1.5, cex = 2, adj = -0.4)
}

figure_1()

png(file.path(IMAGE_PATH, "Figure_1.png"), width = 10, height = 5,
    units = "cm", res = 600, pointsize = 6)
figure_1()
dev.off()

```

Results from the linear models:

```{r}
l <- lm(liver_stiffness ~ liver_fat_cat, data = gal_met_prot_data)
summary(l)

l <- lm(liver_stiffness ~ liver_fat_cat + Age_10 + Sex,
        data = gal_met_prot_data)
summary(l)
```

- No association between steatosis grade and FIB4; note that is a strong
  relationship between FIB4 and age.
- Relationship between steatosis grade and liver stiffness: grade 3 and 4 both
  have significantly higher stiffness than control group 1.
- Higher percentage of T2D cases in highest steatosis grade.
- Highest proportion of MASLD cases in highest steatosis grade, but there are
  also MASLD cases in the control group.

For completeness: check medication.

```{r, fig.width = 4, fig.height = 4}
par(mfrow = c(1, 2))
l <- split(gal_met_prot_data$liver_fat_cat, gal_met_prot_data$med_DM)
m <- do.call(cbind, lapply(l, function(z) as.numeric(table(z)) * 100 / length(z)))

barplot(m, ylab = "%", col = paste0(col_liver_fat_cat, "ce"), main = "med DM")

l <- split(gal_met_prot_data$liver_fat_cat, gal_met_prot_data$med_lip)
m <- do.call(cbind, lapply(l, function(z) as.numeric(table(z)) * 100 / length(z)))

barplot(m, ylab = "%", col = paste0(col_liver_fat_cat, "ce"), main = "med lip")
```

- More diabetes medication in the highest steatosis grade, not unexpected,
  because it contains the highest proportion of T2D cases.
- Lipid modifying therapy is about balanced between the categories.


## PCA of omics data

```{r, fig.width = 8, fig.height = 8}
cols <- col_liver_fat_cat[as.integer(gal_met_prot_data$liver_fat_cat)]

figure_2 <- function(cols = "#000000") {
    tmp <- gal_met_prot_data[, proteins]

    pc_all <- prcomp(scale(tmp, center = TRUE, scale = TRUE))
    library(CompMetaboTools)
    par(mfrow = c(2, 2), mar = c(4.5, 4.5, 3, 1.5), cex.lab = 1.3)
    plot_pca(pc_all, pch = 21, col = "#000000ce", bg = paste0(cols, "ce"))
    mtext(side = 3, text = "A", line = 1, cex = 2, adj = -0.2)
    plot_pca(pc_all, pch = 21, col = "#000000ce", bg = paste0(cols, "ce"),
             pc_x = 3, pc_y = 4)
    mtext(side = 3, text = "B", line = 1, cex = 2, adj = -0.2)

    tmp <- gal_met_prot_data[, metabolites] |>
        as.matrix() |>
        t() |>
        na.omit() |>
        t()

    pc_all <- prcomp(scale(tmp, center = TRUE, scale = TRUE))
    plot_pca(pc_all, pch = 21, col = "#000000ce", bg = paste0(cols, "ce"))
    mtext(side = 3, text = "C", line = 1, cex = 2, adj = -0.2)
    plot_pca(pc_all, pch = 21, col = "#000000ce", bg = paste0(cols, "ce"),
             pc_x = 3, pc_y = 4)
    mtext(side = 3, text = "D", line = 1, cex = 2, adj = -0.2)
}

figure_2(cols)

png(file.path(IMAGE_PATH, "Figure_2.png"), width = 10, height = 10,
    units = "cm", res = 600, pointsize = 6)
figure_2(cols)
dev.off()

```

```{r}
col_sex <- brewer.pal(3, "Set1")[1:2]
cols <- col_sex[as.integer(gal_met_prot_data$Sex)]

figure_2(cols)

```

# Association analyses

We next perform the association analyses between food items, metabolites and
proteins with the liver fat categories.

## Association liver fat (steatosis) with diet

We perform linear regression analyses to evaluate whether there are differences
in dietary habits between the liver fat categories. We first test associations
for the macronutrients and later test for association with the PDI for adherence
for a healthy plant based diet. For the diet associations we need to adjust in
addition for the estimated daily nutrient intake, which we add as a covariate to
the models.

```{r}
library(nafldData)
ffq <- nafldData("FFQ")
rownames(ffq) <- ffq$AID

gal_met_prot_data$estim_daily_nutr_intake <-
    ffq[rownames(gal_met_prot_data), "nanu006a"]
```

We next define the macronutrients for which we test for associations. These were
generated from the GA2LEN FFQ data. The same model is fitted to the PDI score,
but this is evaluated separately from the macronutrients results.

```{r}
#' selection of diet items of interest
ffq_items <- c("protein",
               "total_fat",
               "total_carbs",
               "saturated_fatty_acids",
               "monounsaturated_fatty_acids",
               "polyunsaturated_fatty_acids",
               "trans_fatty_acids",
               "pdi_score")

#' print final selection of diet items
print(ffq_items)
```

linear models setup:

```{r define-lm-covariates-and-target-diet, results = "hold"}
COVARIATES <- c("Sex", "Age_10", "estim_daily_nutr_intake")
TARGET <- "liver_fat_cat"
DATASET <- "NAFLD"

print("Covariates of the linear regression model:")
print(COVARIATES)

print("Target of the linear regression model:")
print(TARGET)

print(paste0("final variables of the ", DATASET, " dataset used:"))

print("nr of variables:")
print(length(ffq_items))

print("nr of individuals:")
print(nrow(gal_met_prot_data))
```

Model coefficients and p-values are provided. Raw p-values are adjusted for
multiple hypothesis testing with the method from Benjamini & Hochberg.

```{r linear model-diet with CIs, message=FALSE}
#' Define a function to fit the linear model:
#' - fit the linear model
#' - extract the coefficients and p-values skipping the intercept
#' - return the result as a one-row matrix, columns being coefficients
#'   and p-values
funForLinReg <- function(analyte, data) {
    y <- data[, analyte]
    data <- data[, c(COVARIATES, TARGET)]
    l <- lm(y ~ ., data = data)
    res <- coef(summary(l))[-1, ]
    pvals <- matrix(c(res[, 1L], res[, 4L]), nrow = 1)
    rownames(pvals) <- analyte
    colnames(pvals) <- c(paste0("coef_", rownames(res)),
                         paste0("p-value_", rownames(res))
                         )
    pvals
}

#' Define a function to fit the linear model:
#' - fit the linear model
#' - extract the coefficients and p-values skipping the intercept
#' - extract the 95% confidence interval for coefficients in lm
#' - return the result as a one-row matrix, columns being coefficients
#'   and p-values
funForLinReg_CIs <- function(analyte, data) {
    y <- data[, analyte]
    data <- data[, c(COVARIATES, TARGET)]
    l <- lm(y ~ ., data = data)
    res <- coef(summary(l))[-1, ]
    CIs <- confint(l)[-1, ]
    pvals <- matrix(c(res[, 1L], res[, 4L], CIs[, 1L], CIs[, 2L]), nrow = 1)
    rownames(pvals) <- analyte
    colnames(pvals) <- c(paste0("coef_", rownames(res)),
                         paste0("p-value_", rownames(res)),
                         paste0("CI_lower_", rownames(res)),
                         paste0("CI_upper_", rownames(res))
                         )
    pvals
}

#' Fit linear models and extract coefficients and p-values on each protein
lm_output <- lapply(ffq_items, funForLinReg_CIs, data = gal_met_prot_data) |>
    do.call(what = rbind)

#' Separate the PDI results.
lm_pdi <- lm_output["pdi_score", , drop = FALSE]
lm_output <- lm_output[rownames(lm_output) != "pdi_score", ]

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)
```

Next we calculate the effect sizes.

To obtain comparable coefficients, prior to linear modelling all the ffq_items
are standardized to zero mean and unit variance (note that this does not
influence the linear models since they are affine equivariant). In such
autoscaled data, difference of 1 corresponds to a standard deviation of 1. This
parameter will be called *effect size* and will be use to evaluate group
differences.

```{r effectsize-diet, message=FALSE}
#' Perform autoscaling for effect size (es) calculation
gal_met_prot_data_scaled <- gal_met_prot_data
gal_met_prot_data_scaled[, ffq_items] <- gal_met_prot_data[, ffq_items] |>
    as.matrix() |>
    scale(scale = TRUE)

#' fit linear models and extract the coefficients
es <- lapply(ffq_items, FUN = funForLinReg, data = gal_met_prot_data_scaled) |>
    do.call(what = rbind)
es <- es[, grep("coef", colnames(es))]
colnames(es) <- sub("coef", "effectSize", colnames(es))

lm_pdi <- data.frame(lm_pdi, es["pdi_score", , drop = FALSE])

lm_output <- cbind(lm_output, es[rownames(es) != "pdi_score", ])
```

For diet, we can not have cv-based cut-off, therefore, we use "just" the
statistical significance criteria (just the adjusted p-value).

```{r define significance criteria-diet}
#' define the different coefficients as comparisons
comps <- colnames(lm_output)[grep("coef", colnames(lm_output))]

#' define which ffq_items are considered statistically significant
sign_p <- lm_output[, grep("p-adj", colnames(lm_output))] < 0.05
colnames(sign_p) <- sub("p-adj_", "significant_", colnames(sign_p))

#' Combine with result from linear model converting to a data.frame to
#' avoid logicals being converted to numeric
lm_output <- data.frame(lm_output, sign_p)
```

The number of ffq_items called significant or with an adjusted p-value smaller
than 0.05 are shown below.

```{r table-sig-diet, results = "asis"}
#' merge to p-adj to cv significance summary
sig_tab <- data.frame(
    significant = colSums(lm_output[, sub("coef_", "significant_", comps)]),
    check.names = FALSE)

pandoc.table(
    sig_tab, style = "rmarkdown",
    caption = paste0("Number of significant ffq_items per tested covariate. ",
                     " *significant*: number of ffq_items with an adjusted ",
                     "p-value smaller than 0.05"))
```

For ease of evaluation, add a column describing if the CI is not in zero.

```{r CI_eval}
#' nice output for paper, CIs
lm_output$CI_range_fat_cat4 <- paste0(
    "[",
    format(round(lm_output$CI_lower_liver_fat_cat4, 4), 4),
    ", ",
    format(round(lm_output$CI_upper_liver_fat_cat4, 4), 4),
    "]"
)

#' we want, for each variable, to see if the CI is not zero (for singif
#' difference variables).
lm_output$CI_No_0_liver_fat_cat4 <-((lm_output$CI_lower_liver_fat_cat4 > 0 &
                                     lm_output$CI_upper_liver_fat_cat4 > 0) |
                                    (lm_output$CI_lower_liver_fat_cat4 < 0 &
                                     lm_output$CI_upper_liver_fat_cat4 < 0))
#' returns true if CI NOT in 0, wanted to find 'sign' different vars!

#' nice explain:
#' https://www.econometrics-with-r.org/5.2-cifrc.html
#' if the interval [CI_lower, CI_upper] does not contain the value zero,
#' this leads to the rejection of the null hypothesis β1,0=0.
#' A Confidence Interval for βi
#' Imagine you could draw all possible random samples of given size. The
#' interval that contains the true value βi in 95% of all samples is given
#' by the expression CIβi0.95=[^βi−1.96×SE(^βi),^βi+1.96×SE(^βi)].
#' Equivalently, this interval can be seen as the set of null hypotheses for
#' which a 5% two-sided hypothesis test does not reject.

```

Export the results to an xls file.

```{r}
results_diet <- data.frame(lm_output)
write_xlsx(results_diet,
           path = file.path(XLSX_PATH, "diet_steatosis_associations.xlsx"))
```

The results for the association of FFQ-estimated macronutrient intake are shown
below.

```{r, results = "asis"}
tmp <- results_diet[, grepl("coef_liver|effectSize_liver|p.adj_liver",
                            colnames(results_diet))]
tmp <- tmp[, c(1, 4, 7, 2, 5, 8, 3, 6, 9)]
pandoc.table(tmp, style = "rmarkdown", split.table = Inf,
             digits = c(2, 2, 2, 2, 2, 2, 2, 2, 2))
```

And the result for the PDI.

```{r, results = "asis"}
tmp <- lm_pdi[, grepl("coef_liver|effectSize_liver|p.value_liver",
                            colnames(lm_pdi)), drop = FALSE]
tmp <- tmp[, c(1, 4, 7, 2, 5, 8, 3, 6, 9), drop = FALSE]
pandoc.table(tmp, style = "rmarkdown", split.table = Inf,
             digits = c(2, 2, 2, 2, 2, 2, 2, 2, 2))
```

```{r diet-lm-heatmap, fig.width = 3.5, fig.height = 4, fig.cap = "Associations between diet and steatosis grades. Shown are the effect size for the associations."}
#' Extract effect sizes
es <- results_diet[, grep("effectSize_liver", colnames(results_diet))] |>
    rbind(lm_pdi[, grep("effectSize_liver", colnames(lm_pdi)),
                  drop = FALSE]) |>
    as.matrix()

#' hierarchical clustering
clust1 <- hclust(dist((es)), method = "ward.D")
clust2 <- hclust(dist(t(es)), method = "ward.D")

#' clean roundup nrs for plot
es <- round(es, 2)
colnames(es) <- c("grade 1", "grade 2", "grade 3")
rownames(es) <- c("protein", "total fat", "total carb", "SFA",
                  "MUFA", "PUFA", "TFA", "PDI")

#' in long format
es_long <- melt(es)
#' add labels - only the significant ones
es_long$label <- es_long$value
es_long$label[1:(nrow(es_long) - 1)] <- ""

#' ggplot
plt <- es_long |>
    ggplot(aes(Var2, Var1, fill = value, label = label)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Effect Size") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 2.5) +
    theme_classic() +
    scale_y_discrete(
        expand = c(0, 0), limits = rownames(es)) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 10),
          axis.text.y = element_text(hjust = 1, size = 8),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.position = "right")

ggsave(file.path(IMAGE_PATH, "diet_steatosis_associations_es_heatmap.png"),
       plot = plt, dpi = 600, width = 3,
       height = 2.3, units = "cm", scale = 3)
plt
```


### Summary

- Only PDI score is significant for grade 4 steatosis.
- No other macro nutrient variable is significant.

## Association between circulating proteins and liver fat (steatosis)

For the association analysis between protein abundances and steatosis grades we
include the participant's age and sex as covariates. Other potential influence
factors such as BMI, T2D (including T2D treatment) are almost colinear with
steatosis grades, including them into the model, we would thus reduce the effect
of steatosis. On the downside, we can not exclude that observed differences are
related to any of these not considered potential confounders.

Model setup:

```{r define-lm-covariates-and-target, results = "hold"}
COVARIATES <- c("Sex", "Age_10")
TARGET <- "liver_fat_cat"
DATASET <- "NAFLD"

print("covariates of the linear regression model:")
print(COVARIATES)

print("target of the linear regression model:")
print(TARGET)

print(paste0("final variables of the ", DATASET, " dataset used:"))

print("nr of variables (proteins):")
print(length(proteins))

print("nr of individuals:")
print(nrow(gal_met_prot_data))

```

To identify proteins associated with liver steatosis, seperate linear regression
models are fitted for all proteins separately using their (log2) abundance as
response variable with sex, age and the (categorical) steatosis as a covariates.

Model coefficients and p-values are provided. Raw p-values are adjusted for
multiple hypothesis testing with the method from Benjamini & Hochberg.

```{r linear-model-prot}
#' Fit linear models and extract coefficients and p-values on each protein
lm_output <- lapply(proteins, FUN = funForLinReg, data = gal_met_prot_data) |>
    do.call(what = rbind)

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)
```

Next we calculate the effect sizes.

To obtain comparable coefficients, prior to linear modelling all the proteins
are standardized to zero mean and unit variance (note that this does not
influence the linear models since they are affine equivariant). In such
autoscaled data, difference of 1 corresponds to a standard deviation of 1. This
parameter will be called "effect size" and will be use to evaluate group
differences.

```{r lm-effectsize, message=FALSE}
#' Perform autoscaling for effect size (es) calculation
gal_met_prot_data_scaled[, proteins] <- gal_met_prot_data[, proteins] |>
    as.matrix() |>
    scale(scale = TRUE)

#' fit linear models and extract the coefficients
es <- lapply(proteins, FUN = funForLinReg, data = gal_met_prot_data_scaled) |>
    do.call(what = rbind)
es <- es[, grep("coef", colnames(es))]
colnames(es) <- sub("coef", "effectSize", colnames(es))
lm_output <- cbind(lm_output, es)
```

To call a proteins *significant* we consider, in addition to the *statistical*
significance levels also the magnitude of its coefficient, relative to the
coefficient of variation (CV) for that protein. The CV, calculated on
study-internal QC samples (pools of all samples) are a measure of the technical
noise specific for that protein in the present data set. Thus, with this method
we prioritize proteins with a larger difference between the observed
(biological) difference and the (separately estimated) technical variance.

Below we add thus evaluate for each protein and comparison whether it
fulfills that criteria.

```{r lm-define significance criteria}
#' define the different coefficients as comparisons
comps <- colnames(lm_output)[grep("coef", colnames(lm_output))]

#' define which proteins are considered statistically significant
sign_p <- lm_output[, grep("p-adj", colnames(lm_output))] < 0.05
colnames(sign_p) <- sub("p-adj_", "significant_", colnames(sign_p))

#' Convert the (absolute) difference in abundance into a difference in
#' percentage based on a (log2) coefficient.
diff_percentage <- function(x) {
    (2^abs(x) - 1) * 100
}
diff_perc <- apply(lm_output[, comps], MARGIN = 2, diff_percentage)
colnames(diff_perc) <- sub("coef", "diff_perc", colnames(diff_perc))

#' Define for which proteins the difference in concentration is larger
#' than a certain proportion of the CV
cv_cut <- lapply(comps, function(z) {
    z <- sub("coef_", "", z)
    diff_perc[, paste0("diff_perc_", z)] > 50 *  ##after eval, set to 50
        prot_ann[proteins, "cv_qc_chris"]
}) |>
    do.call(what = cbind)
colnames(cv_cut) <- sub("coef_", "cv_cut_", comps)

#' Use statistical significance for non-categorical traits instead of bool
sign_cv <- sign_p & cv_cut
sign_cv[, "significant_Age_10"] <- sign_p[, "significant_Age_10"]

#' Combine with result from linear model converting to a data.frame to
#' avoid logicals being converted to numeric
lm_output <- data.frame(lm_output, sign_cv)
```

The number of proteins called significant or with an adjusted p-value smaller
than 0.05 for each of the extracted coefficients are shown below. For a complete
comparison, the number of proteins complying to either the adjusted p value
or either the CV-value, as single threshold criteria, are calculated and
summarized here as well.

```{r lm-table-sig, results = "asis"}
#' merge to p-adj to cv significance summary
sig_tab <- data.frame(
    `p_adj < 0.05` = colSums(sign_p),
    `cv thresh` = colSums(cv_cut),
    significant = colSums(lm_output[, sub("coef_", "significant_", comps)]),
    check.names = FALSE)

pandoc.table(
    sig_tab, style = "rmarkdown",
    caption = paste0("Number of significant proteins per tested covariate. ",
                     " *p_adj < 0.05*: number of statistically significant",
                     " proteins.",
                     " *cv*: number of proteins with percentage abundance",
                     " difference above the CV-based threshold.",
                     " *significant*: number of proteins that fulfill ",
                     "the CV-based criteria in addition to the p-adj. ",
                     "threshold"))
```

Export results for inspection, and for loading e.g. PDI influence as covariate:

```{r lm-export-results}
results_prot <- data.frame(
    prot_ann[rownames(lm_output),
             c("description", "long_description", "PeptideNN",
               "Protein.Names", "Genes", "cv_qc_chris")],
    lm_output,
    diff_perc
)
colnames(results_prot) <- sub("^description$", "Uniprot ID",
                              colnames(results_prot))
colnames(results_prot) <- sub("^long_description$", "Description",
                              colnames(results_prot))
colnames(results_prot) <- sub("^Genes$", "Gene",
                              colnames(results_prot))
write_xlsx(results_prot,
           path = file.path(XLSX_PATH, "protein_steatosis_associations.xlsx"))
```

The significantly associated proteins are shown below.

```{r, results = "asis"}
s <- rowSums(results_prot[, c("significant_liver_fat_cat2",
                              "significant_liver_fat_cat3",
                              "significant_liver_fat_cat4")]) > 0
tmp <- results_prot[s, grepl("Gene|coef_liver|effectSize_liver|p.adj_liver",
                             colnames(results_prot))]
tmp <- tmp[order(tmp$p.adj_liver_fat_cat4), ]
tmp <- tmp[, c(1, 2, 5, 8, 3, 6, 9, 4, 7, 10)]
pandoc.table(tmp, style = "rmarkdown", split.table = Inf,
             digits = c(1, 2, 2, 2, 2, 2, 2, 2, 2, 2))
```

We visualize the associations in a heatmap. The effect size are added for the
associations, for the significant proteins (min 1 significance to
covariate).

```{r prot-lm-heatmap, fig.width = 3.5, fig.height = 4, fig.cap = "Associations between protein abundances and steatosis grades. Shown are the effect size for the associations, for proteins with at least one significant association."}
#' Extract effect sizes
es <- results_prot[, grep("effectSize_liver", colnames(results_prot))] |>
    as.matrix()
#' Get the significance information
s <- results_prot[, grep("significant_liver", colnames(results_prot))] |>
    as.matrix()
keep <- rowSums(s) > 0
es <- es[keep, ]
s <- s[keep, ]

#' hierarchical clustering
clust1 <- hclust(dist((es)), method = "ward.D")
clust2 <- hclust(dist(t(es)), method = "ward.D")

#' clean roundup nrs for plot
es <- round(es, 2)
colnames(es) <- c("grade 1", "grade 2", "grade 3")

#' in long format
es_long <- melt(es)
#' add labels - only the significant ones
es_long$label <- es_long$value
es_long$label[!melt(s)$value] <- ""

#' ggplot
plt <- es_long |>
    ggplot(aes(Var2, Var1, fill = value, label = label)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Effect Size") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 2.5) +
    theme_classic() +
    ## scale_x_discrete(
    ##     expand = c(0, 0), limits = colnames(es)[clust2$order]) +
    scale_y_discrete(
        expand = c(0, 0), limits = rownames(es)[clust1$order]) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 10),
          axis.text.y = element_text(hjust = 1, size = 8),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.position = "right")

ggsave(file.path(IMAGE_PATH, "protein_steatosis_associations_es_heatmap.png"),
       plot = plt, dpi = 600, width = 3.5,
       height = 6, units = "cm", scale = 3)
plt
```

We next create volcano plots showing the results for the individual steatosis
categories.


```{r prot-FIG-volcano-steatosis, fig.path = IMAGE_PATH, fig.width = 15, fig.height = 5, fig.cap = "Volcano plots for the protein associations with the various steatosis categories (compared to category 1)."}
plot_volcano <- function(x, comparison = character(), ...) {
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    plot(X, -log10(Y), pch = 21,
         col = "#00000080", bg = "#00000040",
         xlab = "coefficient", ylab = expression(-log[10](p[adjusted])),
         ...)
}
significant_points <- function(x, comparison = character(),
                               diff_perc = 2, col = "#E41A1CCE",
                               bg = "#E41A1C40", ...) {
    s <- x[, paste0("significant_", comparison)]
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    if (any(s))
        points(X[s], -log10(Y[s]), col = col, bg = bg, pch = 21)
}


supp_figure_2 <- function() {
    par(mfrow = c(1, 3))
    xl <- range(results_prot[, c("coef_liver_fat_cat2", "coef_liver_fat_cat3",
                                 "coef_liver_fat_cat4")])
    yl <- c(0, -log10(min(results_prot[, c("p.adj_liver_fat_cat2",
                                           "p.adj_liver_fat_cat3",
                                           "p.adj_liver_fat_cat4")])))
    sigs <- results_prot$significant_liver_fat_cat2 |
        results_prot$significant_liver_fat_cat3 |
        results_prot$significant_liver_fat_cat4
    plot_volcano(results_prot, "liver_fat_cat2", main = "Grade 1",
                 xlim = xl, ylim = yl)
    points(results_prot$coef_liver_fat_cat2[sigs],
           -log10(results_prot$p.adj_liver_fat_cat2[sigs]),
           pch= 21, col = "#E41A1CCC", bg = "#E41A1C40")
    grid()
    plot_volcano(results_prot, "liver_fat_cat3", main = "Grade 2",
             xlim = xl, ylim = yl)
    points(results_prot$coef_liver_fat_cat3[sigs],
           -log10(results_prot$p.adj_liver_fat_cat3[sigs]),
           pch= 21, col = "#E41A1CCC", bg = "#E41A1C40")
    grid()
    plot_volcano(results_prot, "liver_fat_cat4", main = "Grade 3",
                 xlim = xl, ylim = yl)
    points(results_prot$coef_liver_fat_cat4[sigs],
           -log10(results_prot$p.adj_liver_fat_cat4[sigs]),
           pch= 21, col = "#E41A1CCC", bg = "#E41A1C40")
    grid()
}

supp_figure_2()

png(file.path(IMAGE_PATH, "Supp_Figure_2.png"), width = 12, height = 4.5,
    units = "cm", res = 600, pointsize = 6)
supp_figure_2()
dev.off()

```

Difference of abundance increases with the steatosis grade. Also, the
significant proteins for steatosis grade 4 show a trend of difference already in
steatosis grade 2 and 3 participants.

Significant proteins associated with steatosis (top 5 strongest associated
proteins) are visualized below. Points are colored according to sample sex.

```{r lm-prep for plotting}
#' prep for plotting
PLOTLY <- FALSE
ggplotly_or_not <- function(x, interactive = TRUE) {
    if (interactive) ggplotly(x)
    else print(x)
}
myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))

box_plot <- function(data, aes, analyte) {
    ggplot(data, aes) +
        geom_jitter(position = position_jitter(0.24),
                    aes(colour = Sex, alpha = 0.5 , size = (BMI/10))) +
        geom_violin(trim = FALSE, aes(alpha = 0.5)) +
        stat_summary(fun = median, geom = "point", shape = 23, size = 2) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) + labs(title = analyte) +
        theme_bw ()
}

scatter_plot <- function(data, aes, analyte) {
    ggplot(data, aes) +
        geom_jitter(position = position_jitter(0.24),
                    aes(colour = Sex, alpha = 0.5 , size = 0.1)) +
        stat_smooth(method = lm, se = TRUE) + #with standard error
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) + labs(title = analyte) +
        theme_bw ()
}

#' for boxplot func to work automatically
tmp_colnr <- which(colnames(gal_met_prot_data) == TARGET)
gal_met_prot_data$tmp_target <- gal_met_prot_data[, tmp_colnr]   #value of eg BMI
```

```{r prot-lm-top1, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[1]
gal_met_prot_data$abundance <- gal_met_prot_data[, top_analyte]  #intensities of metabolite
s_1 <- box_plot(gal_met_prot_data, aes(x = tmp_target, y = abundance),
                    analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r prot-lm-top2, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[2]
gal_met_prot_data$abundance <- gal_met_prot_data[, top_analyte]
s_1 <- box_plot(gal_met_prot_data, aes(x = tmp_target, y = abundance),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r prot-lm-top3, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[3]
gal_met_prot_data$abundance <- gal_met_prot_data[, top_analyte]
s_1 <- box_plot(gal_met_prot_data, aes(x = tmp_target, y = abundance),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r prot-lm-top4, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[4]
gal_met_prot_data$abundance <- gal_met_prot_data[, top_analyte]
s_1 <- box_plot(gal_met_prot_data, aes(x = tmp_target, y = abundance),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r prot-lm-top5, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[5]
gal_met_prot_data$abundance <- gal_met_prot_data[, top_analyte]
s_1 <- box_plot(gal_met_prot_data, aes(x = tmp_target, y = abundance),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

### Summary

- A considerable number of proteins is associated with steatosis.
- The extend of difference, compared to steatosis grade 1 increases with the
  steatosis category. Consequently, the largest number of proteins with
  significant differences were found between steatosis grade 4 and 1.
- We restrict thus interpretation and further analyses on these.

### Sensitivity analyses

We first evaluate the number of T2D cases in each steatosis category:

```{r}
t2d <- split(gal_met_prot_data$T2D, gal_met_prot_data$liver_fat_cat)
lapply(t2d, summary)
```

The ratio between T2D and non T2D seems to change with the steatosis category -
this might have an impact on the analysis, in particular for the higher
categories.

We apply the same linear regression analyses as above, but with the *T2D*
variable as additional covariate.

linear models (classification) setup:

```{r define-lm-covariates-and-target-t2d, results = "hold"}
COVARIATES <- c("Sex", "Age_10", "T2D")
TARGET <- "liver_fat_cat"
DATASET <- "NAFLD"

print("covariates of the linear regression model:")
print(COVARIATES)

print("target of the linear regression model:")
print(TARGET)

print(paste0("final variables of the ", DATASET, " dataset used:"))

print("nr of variables inspected:")
print(length(proteins))

print("nr of individuals inspected:")
print(nrow(gal_met_prot_data))
```

```{r linear-model-t2d}
#' Fit linear models and extract coefficients and p-values on each protein
lm_output <- lapply(proteins, FUN = funForLinReg, data = gal_met_prot_data) |>
    do.call(what = rbind)

es <- lapply(proteins, FUN = funForLinReg, data = gal_met_prot_data_scaled) |>
    do.call(what = rbind)
es <- es[, grep("coef", colnames(es))]
colnames(es) <- sub("coef", "effectSize", colnames(es))
lm_output <- cbind(lm_output, es)
lm_output <- as.data.frame(lm_output)

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)
lm_output <- as.data.frame(lm_output)
```

The impact of including *T2D* in the linear models on the steatosis-related
protein abundance differences are shown below.

```{r sensitivity-proteomics, fig.path = IMAGE_PATH, fig.width = 12, fig.height = 4}
supp_figure_5 <- function() {
    par(mfrow = c(1, 3))
    lim <- range(lm_output$effectSize_liver_fat_cat2_liver_fat_cat2,
                 results_prot$effectSize_liver_fat_cat2)
    plot(results_prot$effectSize_liver_fat_cat2,
         lm_output$effectSize_liver_fat_cat2,
         xlim = lim, ylim = lim, xlab = expression(ES),
         ylab = expression(ES[+T2D]), main = "Grade 1",
         pch = 21, col = "#00000060", bg = "#00000020")
    grid()
    l <- lm(lm_output$effectSize_liver_fat_cat2 ~
                results_prot$effectSize_liver_fat_cat2)
    abline(l)
    lim <- range(lm_output$effectSize_liver_fat_cat3,
                 results_prot$effectSize_liver_fat_cat3)
    plot(results_prot$effectSize_liver_fat_cat3,
         lm_output$effectSize_liver_fat_cat3,
         xlim = lim, ylim = lim, xlab = expression(ES),
         ylab = expression(ES[+T2D]), main = "Grade 2",
         pch = 21, col = "#00000060", bg = "#00000020")
    grid()
    l <- lm(lm_output$effectSize_liver_fat_cat3 ~
                results_prot$effectSize_liver_fat_cat3)
    abline(l)
    lim <- range(lm_output$effectSize_liver_fat_cat4,
                 results_prot$effectSize_liver_fat_cat4)
    plot(results_prot$effectSize_liver_fat_cat4,
         lm_output$effectSize_liver_fat_cat4,
         xlim = lim, ylim = lim, xlab = expression(ES),
         ylab = expression(ES[+T2D]), main = "Grade 3",
         pch = 21, col = "#00000060", bg = "#00000020")
    grid()
    l <- lm(lm_output$effectSize_liver_fat_cat4 ~
                results_prot$effectSize_liver_fat_cat4)
    abline(l)
}


supp_figure_5()

png(file.path(IMAGE_PATH, "Supp_Figure_5.png"), width = 12, height = 4.5,
    units = "cm", res = 600, pointsize = 6)
supp_figure_5()
dev.off()

```

- No strong deviation from the identify line, hence, results are comparable
  with - or without adjustment for T2D.


## Association between serum metabolites and liver fat (steatosis)

Perform multiple linear regression analyses for each metabolite to identify
those with significant differences in abundances between steatosis categories.

```{r define-lm-covariates-and-target-met, results = "hold"}
COVARIATES <- c("Sex", "Age_10")
TARGET <- "liver_fat_cat"
DATASET <- "NAFLD"

print("covariates of the linear regression model:")
print(COVARIATES)
print("target of the linear regression model:")
print(TARGET)

print(paste0("final variables of the ", DATASET, " dataset used:"))

print("nr of variables (metabolites):")
print(length(metabolites))

print("nr of individuals:")
print(nrow(gal_met_prot_data))
```

Linear regression models are fitted for all metabolites separately (as a
response) with sex, age and the (categorical) steatosis category as a
covariates. The analysis is conducted in the same way (with the same settings)
as the analysis for protein abundances above and hence the code is not displayed
again.

Model coefficients and p-values are provided. Raw p-values are adjusted for
multiple hypothesis testing with the method from Benjamini & Hochberg.

```{r metabolites-lm, message = FALSE, echo = FALSE}
#' 1) Fit linear models and extract coefficients and p-values on each protein
lm_output <- lapply(metabolites, FUN = funForLinReg,
                    data = gal_met_prot_data) |>
    do.call(what = rbind)

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)

#' 2) Calculate effect sizes:
#' Perform autoscaling for effect size (es) calculation
gal_met_prot_data_scaled[, metabolites] <- gal_met_prot_data[, metabolites] |>
    as.matrix() |>
    scale(scale = TRUE)

#' fit linear models and extract the coefficients
es <- lapply(metabolites, FUN = funForLinReg,
             data = gal_met_prot_data_scaled) |>
    do.call(what = rbind)
es <- es[, grep("coef", colnames(es))]
colnames(es) <- sub("coef", "effectSize", colnames(es))
lm_output <- cbind(lm_output, es)

#' 3) define significance
#' define the different coefficients as comparisons
comps <- colnames(lm_output)[grep("coef", colnames(lm_output))]

#' define which metabolites are considered statistically significant
sign_p <- lm_output[, grep("p-adj", colnames(lm_output))] < 0.05
colnames(sign_p) <- sub("p-adj_", "significant_", colnames(sign_p))
#colSums(sign_p)

diff_perc <- apply(lm_output[, comps], MARGIN = 2, diff_percentage)
colnames(diff_perc) <- sub("coef", "diff_perc", colnames(diff_perc))

#' Define for which metabolites the difference in concentration is larger
#' than a certain proportion of the CV
cv_cut <- lapply(comps, function(z) {
    z <- sub("coef_", "", z)
    diff_perc[, paste0("diff_perc_", z)] > 50 *  ##after eval, set to 50
        met_ann[metabolites, "cv_qc_chris"]
}) |>
    do.call(what = cbind)
colnames(cv_cut) <- sub("coef_", "cv_cut_", comps)

#' Use statistical significance for non-categorical traits instead of bool
sign_cv <- sign_p & cv_cut
sign_cv[, "significant_Age_10"] <- sign_p[, "significant_Age_10"]

#' Combine with result from linear model converting to a data.frame to
#' avoid logicals being converted to numeric
lm_output <- data.frame(lm_output, sign_cv)
```

The number of metabolites called significant or with an adjusted p-value smaller
than 0.05 for each of the extracted coefficients are shown below. For a complete
comparison, the number of metabolites complying to either the adjusted p value
or either the CV-value, as single threshold criteria, are calculated and
summarized here as well.

```{r metabolites-table-sig, results = "asis"}
#' merge to p-adj to cv significance summary
sig_tab <- data.frame(
    `p_adj < 0.05` = colSums(sign_p),
    `cv thresh` = colSums(cv_cut),
    significant = colSums(lm_output[, sub("coef_", "significant_", comps)]),
    check.names = FALSE)

pandoc.table(
    sig_tab, style = "rmarkdown",
    caption = paste0("Number of significant metabolites per tested covariate. ",
                     " *p_adj < 0.05*: number of statistically significant",
                     " metabolites.",
                     " *cv*: number of metabolites with percentage abundance",
                     " difference above the CV-based threshold.",
                     " *significant*: number of metabolites that fulfill ",
                     "the CV-based criteria in addition to the p-adj. threshold"))
```

Export results for inspection, and for loading e.g. PDI influence as covariate:

```{r metabolites-export-results}
results_met <- data.frame(
    met_ann[rownames(lm_output), c("analyte_name", "analyte_class",
                                   "biochemical_name", "cv_qc_chris")],
    lm_output,
    diff_perc
)
write_xlsx(results_met,
           path = file.path(XLSX_PATH, "metabolite_steatosis_associations.xlsx"))
```

The significantly associated metabolites and lipids are shown below.

```{r, results = "asis"}
s <- rowSums(results_met[, c("significant_liver_fat_cat2",
                              "significant_liver_fat_cat3",
                              "significant_liver_fat_cat4")]) > 0
tmp <- results_met[s, grepl("coef_liver|effectSize_liver|p.adj_liver",
                             colnames(results_met))]
tmp <- tmp[order(tmp$p.adj_liver_fat_cat4), ]
tmp <- tmp[, c(1, 4, 7, 2, 5, 8, 3, 6, 9)]
pandoc.table(tmp, style = "rmarkdown", split.table = Inf,
             digits = c(2, 2, 2, 2, 2, 2, 2, 2, 2))
```

We visualize the associations in a heatmap. The effect size are added for the
associations, for the significant proteins (min 1 significance to
covariate).

```{r met-lm-heatmap, fig.width = 3.5, fig.height = 4, fig.cap = "Associations between metabolite and lipid abundances and steatosis grades. Shown are the effect size for the associations, for metabolites or lipids with at least one significant association."}
#' Extract effect sizes
es <- results_met[, grep("effectSize_liver", colnames(results_met))] |>
    as.matrix()
#' Get the significance information
s <- results_met[, grep("significant_liver", colnames(results_met))] |>
    as.matrix()
keep <- rowSums(s) > 0
es <- es[keep, ]
s <- s[keep, ]

#' hierarchical clustering
clust1 <- hclust(dist((es)), method = "ward.D")
clust2 <- hclust(dist(t(es)), method = "ward.D")

#' clean roundup nrs for plot
es <- round(es, 2)
colnames(es) <- c("grade 1", "grade 2", "grade 3")

#' in long format
es_long <- melt(es)
#' add labels - only the significant ones
es_long$label <- es_long$value
es_long$label[!melt(s)$value] <- ""

#' ggplot
plt <- es_long |>
    ggplot(aes(Var2, Var1, fill = value, label = label)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Effect Size") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 2.5) +
    theme_classic() +
    ## scale_x_discrete(
    ##     expand = c(0, 0), limits = colnames(es)[clust2$order]) +
    scale_y_discrete(
        expand = c(0, 0), limits = rownames(es)[clust1$order]) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 10),
          axis.text.y = element_text(hjust = 1, size = 8),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.position = "right")

ggsave(file.path(IMAGE_PATH, "metabolite_steatosis_associations_es_heatmap.png"),
       plot = plt, dpi = 600, width = 3.5,
       height = 9, units = "cm", scale = 3)
plt
```

We next create volcano plots showing the results for the individual steatosis
categories.

```{r met-FIG-volcano-steatosis, fig.path = IMAGE_PATH, fig.width = 15, fig.height = 5, fig.cap = "Volcano plots for the metabolite and lipid associations with the various steatosis categories (compared to category 1)."}
supp_figure_6 <- function() {
    par(mfrow = c(1, 3))
    xl <- range(results_met[, c("coef_liver_fat_cat2", "coef_liver_fat_cat3",
                                "coef_liver_fat_cat4")])
    yl <- c(0, -log10(min(results_met[, c("p.adj_liver_fat_cat2",
                                          "p.adj_liver_fat_cat3",
                                          "p.adj_liver_fat_cat4")])))
    sigs <- results_met$significant_liver_fat_cat2 |
        results_met$significant_liver_fat_cat3 |
        results_met$significant_liver_fat_cat4
    plot_volcano(results_met, "liver_fat_cat2", main = "Grade 1",
                 xlim = xl, ylim = yl)
    points(results_met$coef_liver_fat_cat2[sigs],
           -log10(results_met$p.adj_liver_fat_cat2[sigs]),
           pch= 21, col = "#E41A1CCC", bg = "#E41A1C40")
    grid()
    plot_volcano(results_met, "liver_fat_cat3", main = "Grade 2",
             xlim = xl, ylim = yl)
    points(results_met$coef_liver_fat_cat3[sigs],
           -log10(results_met$p.adj_liver_fat_cat3[sigs]),
           pch= 21, col = "#E41A1CCC", bg = "#E41A1C40")
    grid()
    plot_volcano(results_met, "liver_fat_cat4", main = "Grade 3",
                 xlim = xl, ylim = yl)
    points(results_met$coef_liver_fat_cat4[sigs],
           -log10(results_met$p.adj_liver_fat_cat4[sigs]),
           pch= 21, col = "#E41A1CCC", bg = "#E41A1C40")
    grid()
}

supp_figure_6()

png(file.path(IMAGE_PATH, "Supp_Figure_6.png"), width = 12, height = 4.5,
    units = "cm", res = 600, pointsize = 6)
supp_figure_6()
dev.off()

```


Significant metabolites associated with steatosis (top 5 strongest associated
metabolites) are visualized below. Points are colored according to sample sex.

```{r met-lm-top1, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[1]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]  #intensities of metabolite
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r met-lm-top2, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[2]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r met-lm-top3, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[3]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r met-lm-top4, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[4]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r met-lm-top5, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[5]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

### Summary

- Similar to the proteomics analysis we see the most significant differences
  between steatosis grade 4 and 1.
- However, we see already differences between grade 3 and 1.

### Sensitivity analyses

We perform the same sensitivity analysis to evaluate influence of T2D on the
observed results as conducted for the proteomics data (code is therefore not
shown again).

```{r met-define-lm-covariates-and-target-t2d, echo = FALSE}
COVARIATES <- c("Sex", "Age_10", "T2D")
TARGET <- "liver_fat_cat"
DATASET <- "NAFLD"
```

```{r met-sensitivity-linear-model-t2d, echo = FALSE}
#' Fit linear models and extract coefficients and p-values on each protein
lm_output <- lapply(metabolites, FUN = funForLinReg,
                    data = gal_met_prot_data) |>
    do.call(what = rbind)

es <- lapply(metabolites, FUN = funForLinReg,
             data = gal_met_prot_data_scaled) |>
    do.call(what = rbind)
es <- es[, grep("coef", colnames(es))]
colnames(es) <- sub("coef", "effectSize", colnames(es))
lm_output <- cbind(lm_output, es)
lm_output <- as.data.frame(lm_output)

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)
lm_output <- as.data.frame(lm_output)
```

The impact of including *T2D* in the linear models on the steatosis-related
metabolite abundance differences are shown below.

```{r sensitivity-metabolomics, fig.path = IMAGE_PATH, fig.width = 12, fig.height = 4}
supp_figure_7 <- function() {
    par(mfrow = c(1, 3))
    lim <- range(lm_output$effectSize_liver_fat_cat2_liver_fat_cat2,
                 results_met$effectSize_liver_fat_cat2)
    plot(results_met$effectSize_liver_fat_cat2,
         lm_output$effectSize_liver_fat_cat2,
         xlim = lim, ylim = lim, xlab = expression(ES),
         ylab = expression(ES[+T2D]), main = "Grade 1",
         pch = 21, col = "#00000060", bg = "#00000020")
    grid()
    l <- lm(lm_output$effectSize_liver_fat_cat2 ~
                results_met$effectSize_liver_fat_cat2)
    abline(l)
    lim <- range(lm_output$effectSize_liver_fat_cat3,
                 results_met$effectSize_liver_fat_cat3)
    plot(results_met$effectSize_liver_fat_cat3,
         lm_output$effectSize_liver_fat_cat3,
         xlim = lim, ylim = lim, xlab = expression(ES),
         ylab = expression(ES[+T2D]), main = "Grade 2",
         pch = 21, col = "#00000060", bg = "#00000020")
    grid()
    l <- lm(lm_output$effectSize_liver_fat_cat3 ~
                results_met$effectSize_liver_fat_cat3)
    abline(l)
    lim <- range(lm_output$effectSize_liver_fat_cat4,
                 results_met$effectSize_liver_fat_cat4)
    plot(results_met$effectSize_liver_fat_cat4,
         lm_output$effectSize_liver_fat_cat4,
         xlim = lim, ylim = lim, xlab = expression(ES),
         ylab = expression(ES[+T2D]), main = "Grade 3",
         pch = 21, col = "#00000060", bg = "#00000020")
    grid()
    l <- lm(lm_output$effectSize_liver_fat_cat4 ~
                results_met$effectSize_liver_fat_cat4)
    abline(l)
}


supp_figure_7()

png(file.path(IMAGE_PATH, "Supp_Figure_7.png"), width = 12, height = 4.5,
    units = "cm", res = 600, pointsize = 6)
supp_figure_7()
dev.off()

```

- No strong deviation from the identify line, hence, results are comparable
  with - or without adjustment for T2D.
- Only H1 is (expectedly) deviating (blood sugars).


## Sensitivity analysis for dependence on PDI

We apply the same linear regression analyses as above, but with the *PDI*
variable as additional covariate.

linear models (classification) setup:

```{r define-lm-covariates-and-target-pdi}
## gal_met_prot_data$PDI <- log2(gal_met_prot_data$pdi_score)
## gal_met_prot_data_scaled$PDI <- log2(gal_met_prot_data$pdi_score)

COVARIATES <- c("Sex", "Age_10", "pdi_score")
TARGET <- "liver_fat_cat"
DATASET <- "NAFLD"
```

```{r linear-model-pdi}
#' Fit linear models and extract coefficients and p-values on each protein
lm_output <- lapply(proteins, FUN = funForLinReg, data = gal_met_prot_data) |>
    do.call(what = rbind)

es <- lapply(proteins, FUN = funForLinReg, data = gal_met_prot_data_scaled) |>
    do.call(what = rbind)
es <- es[, grep("coef", colnames(es))]
colnames(es) <- sub("coef", "effectSize", colnames(es))
lm_output <- cbind(lm_output, es)
lm_output <- as.data.frame(lm_output)

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)
lm_output <- as.data.frame(lm_output)
```

The impact of including *PDI* in the linear models on the steatosis-related
protein abundance differences are shown below.

```{r sensitivity-proteomics-pdi, fig.path = IMAGE_PATH, fig.width = 12, fig.height = 4}
supp_figure_10 <- function() {
    par(mfrow = c(1, 3))
    lim <- range(lm_output$effectSize_liver_fat_cat2_liver_fat_cat2,
                 results_prot$effectSize_liver_fat_cat2)
    plot(results_prot$effectSize_liver_fat_cat2,
         lm_output$effectSize_liver_fat_cat2,
         xlim = lim, ylim = lim, xlab = expression(ES),
         ylab = expression(ES[+PDI]), main = "Grade 1",
         pch = 21, col = "#00000060", bg = "#00000020")
    grid()
    l <- lm(lm_output$effectSize_liver_fat_cat2 ~
                results_prot$effectSize_liver_fat_cat2)
    abline(l)
    lim <- range(lm_output$effectSize_liver_fat_cat3,
                 results_prot$effectSize_liver_fat_cat3)
    plot(results_prot$effectSize_liver_fat_cat3,
         lm_output$effectSize_liver_fat_cat3,
         xlim = lim, ylim = lim, xlab = expression(ES),
         ylab = expression(ES[+PDI]), main = "Grade 2",
         pch = 21, col = "#00000060", bg = "#00000020")
    grid()
    l <- lm(lm_output$effectSize_liver_fat_cat3 ~
                results_prot$effectSize_liver_fat_cat3)
    abline(l)
    lim <- range(lm_output$effectSize_liver_fat_cat4,
                 results_prot$effectSize_liver_fat_cat4)
    plot(results_prot$effectSize_liver_fat_cat4,
         lm_output$effectSize_liver_fat_cat4,
         xlim = lim, ylim = lim, xlab = expression(ES),
         ylab = expression(ES[+PDI]), main = "Grade 3",
         pch = 21, col = "#00000060", bg = "#00000020")
    grid()
    l <- lm(lm_output$effectSize_liver_fat_cat4 ~
                results_prot$effectSize_liver_fat_cat4)
    abline(l)
}

supp_figure_10()

png(file.path(IMAGE_PATH, "Supp_Figure_10.png"), width = 12, height = 4.5,
    units = "cm", res = 600, pointsize = 6)
supp_figure_10()
dev.off()

```

We perform the same sensitivity analysis to evaluate influence of PDI on the
observed results as conducted for the proteomics data (code is therefore not
shown again).

```{r met-define-lm-covariates-and-target-pdi, echo = FALSE}
COVARIATES <- c("Sex", "Age_10", "pdi_score")
TARGET <- "liver_fat_cat"
DATASET <- "NAFLD"
```

```{r met-sensitivity-linear-model-pdi, echo = FALSE}
#' Fit linear models and extract coefficients and p-values on each protein
lm_output <- lapply(metabolites, FUN = funForLinReg,
                    data = gal_met_prot_data) |>
    do.call(what = rbind)

es <- lapply(metabolites, FUN = funForLinReg,
             data = gal_met_prot_data_scaled) |>
    do.call(what = rbind)
es <- es[, grep("coef", colnames(es))]
colnames(es) <- sub("coef", "effectSize", colnames(es))
lm_output <- cbind(lm_output, es)
lm_output <- as.data.frame(lm_output)

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)
lm_output <- as.data.frame(lm_output)
```

The impact of including *PDI* in the linear models on the steatosis-related
metabolite abundance differences are shown below.

```{r sensitivity-metabolomics-pdi, fig.path = IMAGE_PATH, fig.width = 12, fig.height = 4}
supp_figure_11 <- function() {
    par(mfrow = c(1, 3))
    lim <- range(lm_output$effectSize_liver_fat_cat2_liver_fat_cat2,
                 results_met$effectSize_liver_fat_cat2)
    plot(results_met$effectSize_liver_fat_cat2,
         lm_output$effectSize_liver_fat_cat2,
         xlim = lim, ylim = lim, xlab = expression(ES),
         ylab = expression(ES[+PDI]), main = "Grade 1",
         pch = 21, col = "#00000060", bg = "#00000020")
    grid()
    l <- lm(lm_output$effectSize_liver_fat_cat2 ~
                results_met$effectSize_liver_fat_cat2)
    abline(l)
    lim <- range(lm_output$effectSize_liver_fat_cat3,
                 results_met$effectSize_liver_fat_cat3)
    plot(results_met$effectSize_liver_fat_cat3,
         lm_output$effectSize_liver_fat_cat3,
         xlim = lim, ylim = lim, xlab = expression(ES),
         ylab = expression(ES[+PDI]), main = "Grade 2",
         pch = 21, col = "#00000060", bg = "#00000020")
    grid()
    l <- lm(lm_output$effectSize_liver_fat_cat3 ~
                results_met$effectSize_liver_fat_cat3)
    abline(l)
    lim <- range(lm_output$effectSize_liver_fat_cat4,
                 results_met$effectSize_liver_fat_cat4)
    plot(results_met$effectSize_liver_fat_cat4,
         lm_output$effectSize_liver_fat_cat4,
         xlim = lim, ylim = lim, xlab = expression(ES),
         ylab = expression(ES[+PDI]), main = "Grade 3",
         pch = 21, col = "#00000060", bg = "#00000020")
    grid()
    l <- lm(lm_output$effectSize_liver_fat_cat4 ~
                results_met$effectSize_liver_fat_cat4)
    abline(l)
}


supp_figure_11()

png(file.path(IMAGE_PATH, "Supp_Figure_11.png"), width = 12, height = 4.5,
    units = "cm", res = 600, pointsize = 6)
supp_figure_11()
dev.off()

```


# Other visualization of `r TARGET` phenotype

## Pathway analysis: proteins and metab, other

plan:

- [x] find related pathways and perform pathway enrichment with Ramp-DB.
      https://rampdb.nih.gov/pathway-enrichment. see rampdb-test.ods
- [x] manually curate and define the pathway clusters, also split again proteins
      and metabolites (since no overlap).
- [x] floorplan of coagulation and AA metabolism, make/paste in ppt.
- [x] add barplots of measured metab/prot with background (prot/metab) and
     control, grade S3 steatosis 2 bars, with '*' if significant (lm's).
     For now, boxplots and only of the significant ones, can change in the
     future.
- [x] optional: add umetab results from mummichog for validation of the
     targeted p180 metabolites results. Or just as a sentence (incl in M&M)
     that this was confirmed and SI table of found pathways.


Box plots for on the plot pathway:

Proteins:

1) Response to elevated platelet cytosolic Ca2+

```{r}
#' Response to elevated platelet cytosolic Ca2+
gal_met_prot_data$liver_fat_cat2 <- gal_met_prot_data$liver_fat_cat
gal_met_prot_data$liver_fat_cat2[gal_met_prot_data$liver_fat_cat2 %in% c(2, 3)] <- NA
gal_met_prot_data$liver_fat_cat2 <- droplevels(gal_met_prot_data$liver_fat_cat2)

plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P01011, main = 'SERPINA3')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Q14624, main = 'ITIH4')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`Q14624;Q14624-2`, main = 'ITIH4')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`Q06033;Q06033-2`, main = 'ITIH3')

```


2) Network map of SARS-CoV-2 signaling pathway

```{r}
#' Network map of SARS-CoV-2 signaling pathway
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Q14624, main = 'ITIH4')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`Q14624;Q14624-2`, main = 'ITIH4')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`Q06033;Q06033-2`, main = 'ITIH3')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P02750, main = 'LRG1')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P05156, main = 'CFI')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P08603, main = 'CFH')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P05090, main = 'APOD')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$O95445, main = 'APOM')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`P06396;P06396-2`, main = 'GSN')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P04217, main = 'A1BG')

```

3) Post-translational protein phosphorylation

```{r}#' Post-translational protein phosphorylation
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P01008, main = 'SERPINC1')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P05546, main = 'SERPIND1')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P01024, main = 'C3')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P00450, main = 'CP')

```

4) Complement and coagulation cascades

```{r}
#' Complement and coagulation cascades
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P01008, main = 'SERPINC1', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P05546, main = 'SERPIND1', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P01024, main = 'C3', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P02748, main = 'C9', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P04004, main = 'VTN', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P05156, main = 'CFI', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P08603, main = 'CFH', col = c("#0C6291", "#A63446"))

```

Metabolites and lipids:

5) Lipids biosynthesis and degradation

As all except one has same direction, and only 1 dot on the pathways, we only
add one PC and one lysoPC for on the map. We take the top-ranking ones. idem for
the top SM.

```{r}
#' Lipids synthesis/degradation
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`PC ae C34:3`, main = 'PC ae C34:3')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`lysoPC a C24:0`, main = 'lysoPC a C24:0')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`lysoPC a C20:4`, main = 'lysoPC a C20:4')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`SM C16:0`, main = 'SM C16:0')

```

6) AAs biosynthesis and degradation

```{r}
#' AAs synthesis/degradation
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Glu, main = 'Glu', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Tyr, main = 'Tyr', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Val, main = 'Val', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Ile, main = 'Ile', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Ala, main = 'Ala', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Leu, main = 'Leu', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Gly, main = 'Gly', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Phe, main = 'Phe', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Trp, main = 'Trp', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Gln, main = 'Gln', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`alpha-AAA`, main = 'alpha-AAA', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$SDMA, main = 'SDMA', col = c("#0C6291", "#A63446"))

```

7) carnitines beta oxidation

```{r}
#' carnitines beta-oxidat
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$C0, main = 'C0')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$C3, main = 'C3')

```

8) BCAA and carnitines

```{r}
#' BCAA and carnitines
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$C5, main = 'C5')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$C4, main = 'C4')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$C3, main = 'C3')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`C3:1`, main = 'C3:1')
#plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`C4:1`, main = 'C4:1')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Leu, main = 'Leu')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Val, main = 'Val')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Ile, main = 'Ile')

```

9) SM metabolism

```{r}
#' SM metabolism
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`SM (OH) C22:2`, main = 'SM (OH) C22:2')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`SM C24:1`, main = 'SM C24:1')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`SM C16:0`, main = 'SM C16:0')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`SM C20:2`, main = 'SM C20:2')

```


# Acknowledgments

Consulted sources used as basis for code:

- ~/nafld_proteo_metabolomics/NAFLD_metab_targeted_lm.Rmd
- ~?metabo_chris_untargeted/CHRIS_lm_pos.Rmd
- kora: CHRIS_KORA_tmetab_lm_binary.Rmd


# Session information

R packages used for the analysis:

```{r}
sessionInfo()
```
