---
title: "NAFLD_steatosis_explore, splitted per T2D type"
author: "Marilyn De Graeve, Johannes Rainer, Christoph Grander"
affiliation: "Eurac Research, Bolzano, Italy"
date: "2025-02-27"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
---

**Modified**: `r file.info("NAFLD_steatosis_T2Dsplit.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r style, message = FALSE, echo = FALSE, warning = FALSE, results = "asis"}
library("BiocStyle")
library("kableExtra")
library("knitr")
library("rmarkdown")
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now, units = "secs")
      # return a character string to show the time
      paste("    Time for this code chunk to run:", round(res,
        2), "seconds")
    }
  }
}))
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7, time_it = TRUE)

```

# Introduction

During this workflow, exploratory analyis (Spearman correlations) are performed on the
relatively quantified Scanning SWATH proteomics data (148 proteins) from the
metabolic dysfunction-associated steatotic liver disease (MASLD) cohort. MASLD,
previously known as NAFLD is the most chronic liver disease, affecting more than
30% of global population. The MAFLD cohort is a subset of approximatley 353 #
participants from the Cooperative Health Research in South Tirol (CHRIS) study,
which were requested to return for a second resampling and additional
phenotyping. The NAFLD cohort represent an unhealty group with various degrees
of steatotic (alcoholic) liver disease classifications. See
[nafld_data_overview.Rmd](nafld_data_overview.Rmd) (respectively the rendered
html report) for an overview of the data set and the definition of the
phenotype.

This document describes the exploratory analysis using Spearman correlations to
identify associations between proteins and metabolic dysfunction-associated
steatotic liver disease (MASLD). The definition of control and MASLD types are:

- *none*: participants negative for steatotic liver disease.
- *MASLD*: (= Metabolic Dysfunction Associated Steatotic Liver Disease)
  participants with SLD, positive for cardiometabolic criteria but low alcohol
  consumption (below 40g/day). This includes also 8 participants without
  information on alcohol consumption. These participants did not provide the
  GA2LEN FFQ (hence missing values for quantified alcohol consumption), but
  according to alcohol-related information in other questionnaire they are not
  heavy drinkers and are thus added to this category.

The MetALD and SLD types are not included in this analysis. FYI:

- *MetALD*: (= MASLD and increased alcohol intake) participants with SLD,
  positive for cardiometabolic criteria but daily alcohol consumption > 40g/day.
- *SLD*: (= steatotic liver disease) participants with a SLD, but negative for
  cardiometabolic criteria. This includes potentially cryptogenic SLD, alcoholic
  liver disease (ALD), DILI or monogenic diseases.

See the `nafld_data_overview 1.html` report for further
information regarding the SLD sub-classification.

Regarding steatosis, i.e. Abnormal retention of fat (lipids) within a cell or 
organ. In this scope: liver steatosis. 
The variables containing this parameter is:

Presence of steatotic liver disease (SLD): sonographic evaluation, steatose 
score of 2 – 3.

- *naab13*: liver fat. Can be binarized at an absolute value of 288. "liver_fat"
- *naab14*: subjective, visual evaluation of the liver fat. Categorical from 1 
  to 3. Values 2 and 3 can be considered steatotic. "visual_fat"

This parameter will be used as the target for a biomarker search in *MASLD* 
participants.


# Setup

## Directories

```{r directories, echo = FALSE}
#' General settings
filename <- "NAFLD_steatosis_T2Dsplit"

#' Path to save images to; remove if exists.
IMAGE_PATH <- paste0("images/", filename, "/")
dir.create(IMAGE_PATH, recursive = TRUE)

#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)

#' Path to the data
DATA_PATH <- '/home/mdegraeve/Documents/Files/Work_Eurac/Data/NAFLD' #FINDME!
#DATA_PATH <- "."


DATASET <- 'CHRSI-NAFLD, none and MASLD'

```

## Packages

```{r packages, message=FALSE}
#' Load libraries
library(chrisUtils)
library(chrisData)
library(DT)
library(ggfortify)
library(ggplot2)
library(pander)
library(pheatmap)
library(plotly)
library(RColorBrewer)
library(readxl)
library(reshape2)
library(tidyfr)
library(vioplot)
library(writexl)

```


# Data import

## tprot dataset (SWATH)

The data from proteins was converted into the TDF format and can hence be
loaded and imported using the `tidyfr` R package. Now we read the full data
which includes general information for each measurement.

```{r load_prot_dataset, message = FALSE}
#' load TDF using tidyfr
#list_data_modules(path = DATA_PATH)
md1 <- data_module(name = "chris_scanningswath_proteins", version = "1.0.0.2",
                    path = DATA_PATH)
prot_data <- data(md1)
prot_ann <- labels(md1)
stopifnot(all(colnames(prot_data) == rownames(prot_ann)))
colnames(prot_data) <- prot_ann$description
rownames(prot_ann) <- prot_ann$description

#' define the columns with protein concentrations
proteins <- rownames(prot_ann)[prot_ann$type != "categorical"]
proteins <- rownames(prot_ann)[prot_ann$Genes != ""]

#' problem in calculating lm below for
#'   'PC aa C30:2' #NOK, only 8 col!!
#'   'Histamine' NOK, only 8 col!!
#' because no values for each typeSLD, due to lot NA values
#' option 1: rm proteins, not done
#proteins <- proteins[proteins != 'PC aa C30:2']
#proteins <- proteins[proteins != 'Histamine']
#' option2: retain only none and MASLD, see done later

#' select proteins dataframe
prot_data <- prot_data[, proteins]
#head(prot_data)    #row samples, cols are protein name, matrix are absol conc
print('nr of  proteins included:')
ncol(prot_data)
print('included proteins are:')
#colnames(prot_data)
#dim(prot_data)
#head(prot_data)

```

The loaded data contains the samples as rows, and the protein names as
colums. The measurements of the `prot_data` object are the relative
concentrations of protein per sample. No data transformation is performed, a
normal distribution of the proteomics is assumed.

Before proceeding with the data analysis we remove the POOL samples. In addition
we restrict to samples from the NAFLD study.

```{r subset to NAFLD set, message = FALSE, warning = FALSE}
prot_data <- prot_data[grep("^008", rownames(prot_data)), ]

```


## tmetab dataset (Biocratis)

The data from metabolites was converted into the TDF format and can hence be
loaded and imported using the `tidyfr` R package. Now we read the full data
which includes general information for each measurement.

```{r load_tmetb_dataset, message = FALSE}
#' load TDF using tidyfr
#list_data_modules(path = DATA_PATH)
p180 <- data_module(name = "metabolomics_p180", version = "1.0.1.2",
                    path = DATA_PATH)
met_data <- data(p180)
met_ann <- labels(p180)
stopifnot(all(colnames(met_data) == rownames(met_ann)))
colnames(met_data) <- met_ann$description
rownames(met_ann) <- met_ann$description

#' define the columns with metabolite concentrations
metabolites <- rownames(met_ann)[met_ann$type != "categorical"]

#' problem in calculating lm below for
#'   'PC aa C30:2' #NOK, only 8 col!!
#'   'Histamine' NOK, only 8 col!!
#' because no values for each typeSLD, due to lot NA values
#' option 1: rm metabolites, not done
#metabolites <- metabolites[metabolites != 'PC aa C30:2']
#metabolites <- metabolites[metabolites != 'Histamine']
#' option2: retain only none and MASLD, see done later

#' select metabolites dataframe
met_data <- met_data[, metabolites]

#' Impute missing values using an uniform distribution, after gap filling
na_unidis <- function(z) {
    na <- is.na(z)
    if (any(na)) {
        min = min(z, na.rm = TRUE)
        z[na] <- runif(sum(na), min = min/2, max = min)
    }
    z
}

#' Row-wise impute missing values and add the data as a new assay
tmp <- apply(met_data, MARGIN = 1, na_unidis)
met_data <- t(tmp)

#' checks
#head(met_data)    #row samples, cols are metab name, matrix are absol conc
print('nr of  metabolites included:')
ncol(met_data)
print('included metabolites are:')
#colnames(met_data)
#dim(met_data) #ok!

```

The loaded data contains the samples as rows, and the metabolite names as
colums. The measurements of the `met_data` object are the absolute
concentrations of metabolite per sample, in natural scale.

To ensure a normal distribution of the metabololits, we perform log2
transformation.

```{r m log2-transformation, message = FALSE}
#' log2 scaling of absolute abundances metabolites
met_data <- log2(met_data)

```

Before proceeding with the data analysis we remove the POOL samples. In addition
we restrict to samples from the NAFLD study.

```{r m subset to NAFLD set, message = FALSE, warning = FALSE}
met_data <- met_data[grep("^008", rownames(met_data)), ]

```


## Diet GALAN data

Dietary information, i.e. GALAN food frequency questionaires, from the MASLD
study were loaded and included in the analysis.

```{r diet-data}
#' 1. load the metadata regarding the KORA FFQs
library(nafldData)
#nafldTraitAnnotations()|> head()

galan <- nafldData("FFQ")
rownames(galan) <- galan$AID

#variables, subset:
#naff001 - naff229 are galan FFQ, and the rest = not!

galan <- galan[, 1:230] #incl AID
koras <- colnames(galan)[2:230]

#' change to numerical scores instead of FFQ answers
#' solve using category encoder, in R see:
#' https://www.r-bloggers.com/2020/02/a-guide-to-encoding-categorical-features-using-r/

#' select categor variables
categor_variables <- NULL
for (x in koras){
  if (class(galan[[x]]) == 'factor' | class(galan[[x]]) == 'character'){  #factor and character
    categor_variables <- c(categor_variables, x)
  }
}

#' using ordinal encoding on categorical vars
encode_ordinal <- function(x) {
  if (is.factor(x)) {
    order <- levels(x)   #retain defined factor level order, wo NA but next step ignored so fine
    x <- as.numeric(factor(x, levels = order, exclude = NA)) #as numeric, keep NA in dataset
  } else if (is.character(x)) {
    print(x) #*
  }
  x
}

#* text info: NOT
    #naff246 - naff249b
    #naff260 - naff263a

#' testing
summary(galan$naff001)

#' apply encoding on categor vars
galan[, categor_variables] <- lapply(galan[, categor_variables], encode_ordinal)

```


Corresponding fake CV-values (no 'pool' samples) for diet.

```{r kora-cv}
chrisAnn <- chrisTraitAnnotations()
#dim(chrisAnn)
#head(chrisAnn)
kora_ann <- chrisAnn[koras, ]
rownames(kora_ann) <- koras
#head(kora_ann)

#' TEMP, just accept all features now as having good cv
cv_values <- 0
#' save in the met_ann as column
kora_ann$cv_qc_chris <- cv_values

#kora_ann[koras, "cv_qc_chris"]

```


## Metadata (excel)

All available trait (phenotype) data was defined in the
[nafld_data_overview.Rmd](nafld_data_overview.Rmd), is loaded below and merged
with the log2 transformed proteomics data of the NAFLD cohort.

```{r add phenotype dataa}
#' load xlsx file instead of below
handle <- 'masld_phenotypes.xlsx'
masld_phenotype_df <- read_excel(file.path(DATA_PATH, handle)) |>
    as.data.frame()
#head(masld_phenotype_df, 5)
metadata <- colnames(masld_phenotype_df)[3:ncol(masld_phenotype_df)] #info wo AIDs
#dim(masld_phenotype_df)

#' make sure sld_type is a factor with group 'none' as the baseline
masld_phenotype_df$sld_type <- factor(masld_phenotype_df$sld_type,
                                      levels = c('none', 'MASLD',
                                                 'MetALD', 'SLD'))
print('the sld types are:')
levels(masld_phenotype_df$sld_type)
print('nr of samples per sld types:')
table(masld_phenotype_df$sld_type)

```


# Merge tmetab, tprot, kora and metadata

```{r merge-dfs}
#' merge with proteomics data
prot_data$AID <- rownames(prot_data)
#dim(prot_data)
prot_data <- merge(prot_data, masld_phenotype_df, by = 'AID')
#dim(prot_data)

#' write to file for manuel inspection
#handle <- file.path(RDATA_PATH, 'prot_data.txt')
#write.table(prot_data, file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)


#' merge all sets
met_data <- as.data.frame(met_data)
met_data$AID <- rownames(met_data)
met_prot_data <- merge(prot_data, met_data, by = 'AID')
dim(met_prot_data) #lost 1s by merge prot, ok

#' merge with metabolomics data only
met_data <- merge(met_data, masld_phenotype_df, by = 'AID')
#dim(met_data)
#head(met_data)

#' merge all sets, kora too
met_prot_data <- as.data.frame(met_prot_data)
rownames(met_prot_data) <- met_prot_data$AID
gal_met_prot_data <- merge(met_prot_data, galan, by = 'AID') #kora has only 001 AIDs
rownames(gal_met_prot_data) <- gal_met_prot_data$AID
dim(gal_met_prot_data)
#gal_met_prot_data[1:5, 1:5]

#' write to file for manuel inspection
handle <- 'data/gal_met_prot_data.txt'
write.table(gal_met_prot_data, file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)

```

All of 351 with prot+metab were kept after merging with GA2LEN from nasld.


# Subset only relevant participants

Only the *MASLD* type are considered further.

```{r keep only none and MASLD types}
#' because problem 2 proteins only NAs in some sld types,
#' keep only none and MASLD
gal_met_prot_data <- gal_met_prot_data[gal_met_prot_data$sld_type %in% c('none', 'MASLD'), ] # 
gal_met_prot_data$sld_type <- droplevels(gal_met_prot_data$sld_type)
gal_met_prot_data$sld_type <- factor(gal_met_prot_data$sld_type)
gal_met_prot_data$Sex <- factor(gal_met_prot_data$Sex)

#' Summary of available data:
summary(gal_met_prot_data[, c("sld_type", "Age", "Sex", "BMI")])

```

Create additional variable for age per 10y, because coefficients and effect
sizes correspond to one year difference, and the anual difference is
neglectible. Dividing age with 10, difference in coefficents and effect sizes
would correspond to 10 years difference and it would be easier to define
meaningful cut-off.

```{r define-age10-variable, message = FALSE}
#' define age per 10y group
gal_met_prot_data$Age_10 <- gal_met_prot_data$Age / 10

```

```{r load-liver-param}
#' from same excel loaded above, already added to gal_met_prot_data
#colnames(gal_met_prot_data)
gal_met_prot_data$T2D <- factor(gal_met_prot_data$T2D)
summary(gal_met_prot_data[, c("liver_fat", "T2D")])

```

Add a categoric liver_stiffness category too, using both the definition of 
Christoph (7, 15 cutoffs) and in literature (7, 12.5 kPa cutoffs).

```{r liver-stiffness-category}
#' naab11 according to Christoph
gal_met_prot_data$liver_stiffness_cat1 <- 2
gal_met_prot_data$liver_stiffness_cat1[which(gal_met_prot_data$liver_stiffness < 7)] <- 1
gal_met_prot_data$liver_stiffness_cat1[which(gal_met_prot_data$liver_stiffness > 15)] <- 3

#' naab11 according to literature
gal_met_prot_data$liver_stiffness_cat2 <- 2
gal_met_prot_data$liver_stiffness_cat2[which(gal_met_prot_data$liver_stiffness < 7)] <- 1     #mild or absent fibrosis
gal_met_prot_data$liver_stiffness_cat2[which(gal_met_prot_data$liver_stiffness > 12.5)] <- 3  #cirrhosis

#' appearantly differnet cutoffs, depending what background...
#' https://www.mskcc.org/cancer-care/patient-education/understanding-your-fibroscan-results

```

FIB4 is numeric, continuous ranging between 0.3 - 7.9. A higher score is more 
liver fibrosis. After initial run, the 1 participant with the upper value of 
FIB4 = 7.9 was removed, as the median = 1.2 with the second highest 
participant with a FIB4 = 3.0.

```{r subset-fib4-outlier}
#' remove 1 participant with max fib4 value

median(na.omit(gal_met_prot_data$FIB4))
mean(na.omit(gal_met_prot_data$FIB4))
sd(na.omit(gal_met_prot_data$FIB4))
summary(gal_met_prot_data$FIB4)

dim(gal_met_prot_data)
gal_met_prot_data <- gal_met_prot_data[gal_met_prot_data$FIB4 < 7.8876, ]
dim(gal_met_prot_data)

```

Add the information from the [nafld_data_overview.Rmd](nafld_data_overview.Rmd) 
report regarding the steatosis, using the parameter naab11: liver stiffness 
(higher values are worse). Values for participants that have a naab12 
(liver stiffness IQR) > 30 should be either removed or taken with care. 
Further called "liver_stiffness".

```{r load-liver-param2}
#' from same excel loaded above, already added to gal_met_prot_data
#colnames(gal_met_prot_data)

#' remove samples with  naab12 (liver stiffness IQR) > 30
dim(gal_met_prot_data)
gal_met_prot_data <- gal_met_prot_data[gal_met_prot_data$liver_stiffnes_iqr <= 30.0 & !is.na(gal_met_prot_data$liver_stiffnes_iqr), ] #no one above 30, have some NAs though that also need to be removed to be sure ok
dim(gal_met_prot_data)

gal_met_prot_data$T2D <- factor(gal_met_prot_data$T2D)
summary(gal_met_prot_data[, c("liver_stiffness", 'liver_stiffness_cat2', "T2D")])
dim(gal_met_prot_data)

```

liver_stiffness is numeric, continuous ranging between 1.5 - 34.4. A higher 
score is more liver stiffness, aka more steatosis. 
After initial run, the 1 participant with the upper value of 
liver_stiffness = 34.4 was removed, as the median = 4.8 with the second highest 
participant with a liver_stiffness = 18.4.

```{r subset-liver_stiffness-outlier}
#' remove 1 participant with max liver_stiffness value

median(na.omit(gal_met_prot_data$liver_stiffness))
mean(na.omit(gal_met_prot_data$liver_stiffness))
sd(na.omit(gal_met_prot_data$liver_stiffness))
summary(gal_met_prot_data$liver_stiffness)

dim(gal_met_prot_data)
gal_met_prot_data <- gal_met_prot_data[gal_met_prot_data$liver_stiffness < 34.4, ]
dim(gal_met_prot_data)

```

Add a categoric liver_fat category, using the following cutoffs:

- "normal": A score below 238 dB/m means the amount of fatty change in your 
  liver is not higher than normal.
- grade steatosis S1: 238 to 260 dB/m
- grade steatosis S2: 260 to 290 dB/m
- grade steatosis S3: 290 to 400 dB/m

depending on the distribution, look only at the extreme normal - steatosis.

```{r liver-fat-category}
#' naab13 make 4 groups
gal_met_prot_data$liver_fat_cat1 <- NA
gal_met_prot_data$liver_fat_cat1[which(gal_met_prot_data$liver_fat < 238)] <- 1 #normal
gal_met_prot_data$liver_fat_cat1[which(gal_met_prot_data$liver_fat > 238)] <- 2
gal_met_prot_data$liver_fat_cat1[which(gal_met_prot_data$liver_fat > 260)] <- 3
gal_met_prot_data$liver_fat_cat1[which(gal_met_prot_data$liver_fat > 290)] <- 4

#' naab13 make 2 groups
gal_met_prot_data$liver_fat_cat2 <- NA
gal_met_prot_data$liver_fat_cat2[which(gal_met_prot_data$liver_fat < 238)] <- 1 #normal
gal_met_prot_data$liver_fat_cat2[which(gal_met_prot_data$liver_fat > 290)] <- 2  #steatosis, high S3

#' appearantly differnet cutoffs, depending what background...
#' https://www.mskcc.org/cancer-care/patient-education/understanding-your-fibroscan-results

summary(gal_met_prot_data[, "liver_fat"])
table(gal_met_prot_data[, "liver_fat_cat1"])
table(gal_met_prot_data[, "liver_fat_cat2"])
dim(gal_met_prot_data)

```

We add a visc_subcut_fat variable on the dataset, comprising the ratio of 
visceral to subcutaneous fat thickness.

```{r}
gal_met_prot_data$visc_subcut_fat <- gal_met_prot_data$visc_fat / gal_met_prot_data$body_fat
summary(gal_met_prot_data$visc_subcut_fat)
sd(gal_met_prot_data$visc_subcut_fat, na.rm = TRUE)

```


## Keep only T2D 

Run the rmarkdown twice, one with T2D == TRUE, rename the lm_output tables to 
"W" (lm_output_results_proteins, lm_output_results_metabolites), rerun with T2D
== FALSE. rename the lm_output tables to "WO"
(lm_output_results_proteins, lm_output_results_metabolites). The senititvity
analysis will also be done under "NAFLD_T2D_sensititvity_analysis". FINDME!!!

```{r subset-t2d}
#' keep only none and MASLD
dim(gal_met_prot_data)
gal_met_prot_data <- gal_met_prot_data[gal_met_prot_data$T2D == TRUE, ]
dim(gal_met_prot_data)
table(gal_met_prot_data$T2D)
table(gal_met_prot_data$liver_fat_cat2)

```


# Logistic model with naab13_cat2 - proteins

Done for proteomics. FYI: here glm func, but actually same as lm is using 
factors.

```{r}
gal_met_prot_data$liver_fat_cat1 <- as.factor(gal_met_prot_data$liver_fat_cat1)
summary(gal_met_prot_data$liver_fat_cat1)

gal_met_prot_data <- gal_met_prot_data[!is.na(gal_met_prot_data$liver_fat_cat2), ]
gal_met_prot_data$liver_fat_cat2 <- as.factor(gal_met_prot_data$liver_fat_cat2)
summary(gal_met_prot_data$liver_fat_cat2)

```

Logistic models (classification) setup:

```{r define-glm-covariates-and-target}
print('covariates of the linear regression model:')
COVARIATES <- c("Sex", "Age_10")
print(COVARIATES)

print('target of the linear regression model:')
TARGET <- 'liver_fat_cat2'
print("liver_fat_cat2")

DATASET <- 'NAFLD'
print(paste0('final variables of the ', DATASET, ' dataset used:'))
#' variables are the output y in lm(y ~ .):
#print(proteins)

print('nr of variables inspected:')
print(length(proteins))

print('nr of individuals inspected:')
print(dim(gal_met_prot_data))

```

To identify proteins and metabolites related with the 2 (none, MASLD) of the 4 
SLD types (none, MASLD, MetALD, SLD), seperate linear regression models are 
fitted for all proteins and metabolites separately (as a response) with sex, 
age and the (categorical) FIB4 as a covariates.

Model coefficients and p-values are provided. Raw p-values are adjusted for
multiple hypothesis testing with the method from Benjamini & Hochberg.

```{r logistic-model}
funForLogReg <- function(analyte, data) {
    y <- data[, analyte]
    data <- data[, c("Sex", "Age_10", TARGET)]
    res <- coef(summary(glm(y ~ ., data = data)))[-1, ]
    pvals <- matrix(c(res[, 1L], res[, 4L]), nrow = 1)
    rownames(pvals) <- analyte
    colnames(pvals) <- c(paste0("coef_", rownames(res)),
                         paste0("p-value_", rownames(res)))
    pvals
}

#' tests
#gal_met_prot_data[1:5, c("Sex", "Age", "FIB4")]
#gal_met_prot_data[1:5, proteins[1]]
#data <- gal_met_prot_data
#analyte <- proteins[1]
#funForLinReg(analyte, gal_met_prot_data)
#analyte <- proteins[14]
#funForLinReg(analyte, gal_met_prot_data)

#' Fit linear models and extract coefficients and p-values on each protein
lm_output <- lapply(proteins, FUN = funForLogReg, data = gal_met_prot_data) |>
    do.call(what = rbind)

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)

```

Next we calculate the effect sizes.

To obtain comparable coefficients, prior to linear modelling all the 
proteins
are standardized to zero mean and unit variance (note that this does not
influence the linear models since they are affine equivariant). In such
autoscaled data, difference of 1 corresponds to a standard deviation of 1. This
parameter will be called "effect size" and will be use to evaluate group
differences.

```{r glm-effectsize, message=FALSE}
#' Perform autoscaling for effect size (es) calculation
gal_met_prot_data_scaled <- gal_met_prot_data
gal_met_prot_data_scaled[, proteins] <- gal_met_prot_data[, proteins] |>
    as.matrix() |>
    scale(scale = TRUE)

#' fit linear models and extract the coefficients
es <- lapply(proteins, FUN = funForLogReg, data = gal_met_prot_data_scaled) |>
    do.call(what = rbind)
es <- es[, grep("coef", colnames(es))]
colnames(es) <- sub("coef", "effectSize", colnames(es))
lm_output <- cbind(lm_output, es)
#head(lm_output)

```

To call a proteins *significant* we consider, in addition to the
*statistical* significance levels also the magnitude of its coefficient,
relative to the coefficient of variation (CV) for that protein. The CV,
calculated on study-internal QC samples (pools of all samples) are a measure of
the technical noise specific for that protein in the present data set. Thus,
with this method we prioritize proteins with a larger difference between the
observed (biological) difference and the (separately estimated) technical variance.

Below we add thus evaluate for each protein and comparison whether it
fulfills that criteria.

```{r glm-define significance criteria}
#' define the different coefficients as comparisons
comps <- colnames(lm_output)[grep("coef", colnames(lm_output))]

#' define which proteins are considered statistically significant
sign_p <- lm_output[, grep("p-adj", colnames(lm_output))] < 0.05
colnames(sign_p) <- sub("p-adj_", "significant_", colnames(sign_p))
#colSums(sign_p)

#' Convert the (absolute) difference in abundance into a difference in
#' percentage based on a (log2) coefficient.
diff_percentage <- function(x) {
    (2^abs(x) - 1) * 100
}
diff_perc <- apply(lm_output[, comps], MARGIN = 2, diff_percentage)
colnames(diff_perc) <- sub("coef", "diff_perc", colnames(diff_perc))

#' Define for which proteins the difference in concentration is larger
#' than a certain proportion of the CV
cv_cut <- lapply(comps, function(z) {
    z <- sub("coef_", "", z)
    diff_perc[, paste0("diff_perc_", z)] > 50 *  ##after eval, set to 50
        prot_ann[proteins, "cv_qc_chris"]
}) |>
    do.call(what = cbind)
colnames(cv_cut) <- sub("coef_", "cv_cut_", comps)

#' Use statistical significance for non-categorical traits instead of bool
sign_cv <- sign_p & cv_cut
sign_cv[, "significant_Age_10"] <- sign_p[, "significant_Age_10"]

#' Combine with result from linear model converting to a data.frame to
#' avoid logicals being converted to numeric
lm_output <- data.frame(lm_output, sign_cv)

```

The number of proteins called significant or with an adjusted p-value smaller
than 0.05 for each of the extracted coefficients are shown below. For a complete
comparison, the number of proteins complying to either the adjusted p value
or either the CV-value, as single threshold criteria, are calculated and
summarized here as well.

```{r glm-table-sig, results = "asis"}
#' merge to p-adj to cv significance summary
sig_tab <- data.frame(
    `p_adj < 0.05` = colSums(sign_p),
    `cv thresh` = colSums(cv_cut),
    significant = colSums(lm_output[, sub("coef_", "significant_", comps)]),
    check.names = FALSE)

pandoc.table(
    sig_tab, style = "rmarkdown",
    caption = paste0("Number of significant proteins per tested covariate. ",
                     " *p_adj < 0.05*: number of statistically significant",
                     " proteins.",
                     " *cv*: number of proteins with percentage abundance",
                     " difference above the CV-based threshold.",
                     " *significant*: number of proteins that fulfill ",
                     "the CV-based criteria in addition to the p-adj. threshold"))

```

Export results for inspection:

```{r glm-export-results}
results <- data.frame(
    prot_ann[rownames(lm_output), c("description", "long_description", "PeptideNN",
                         "PeptideNames", "Protein.Ids", "Protein.Names",
                         "Genes", "Modified.Sequence", "Stripped.Sequence",
                         "Precursor.Id", "cv_qc_chris")],
    lm_output,
    diff_perc
)

#colnames(results)
#colnames(lm_output)

#' write to file for manual inspection,
#' inclusing inspection of the CV-value of each protein
handle <- file.path(RDATA_PATH, 'lm_output_results_proteins.txt')
write.table(results, file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)

lm_output_prot <- results

```

We visualize the associations in a heatmap. The effect size are added for the
associations, for the significant proteins (min 1 significance to
covariate). also for the not significant associations, the effect size is shown.

```{r glm-heatmap, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for proteins with at least one significant association."}
#' Extract effect sizes
es_results <- results[, grep("effectSize_", colnames(results))] |>
    as.matrix()

#' Restrict to those significant for at least one of the groups (sex, age, SLD type)
sig_results <- results[, grep("significant_", colnames(results))] |>
    as.matrix()
es_results <- es_results[rowSums(sig_results) > 0, ]
print('nr of proteins with >= 1 significant association:')
nrow(es_results)

#' hierarchical clustering
clust1 <- hclust(dist((es_results)), method = "ward.D")
clust2 <- hclust(dist(t(es_results)), method = "ward.D")

#' clean roundup nrs for plot
es_results[es_results == 0] <- NA
es_results <- round(es_results, 2)

#' info regarding FIB4
#print('nr of proteins with a not-zero effect size in the FIB4 group:')
#sum(es_results[, 'effectSize_FIB4'] != 0)

#' in long format
es_results_long <- melt(as.matrix(es_results))

#' ggplot
plt <- es_results_long |>
    ggplot(aes(Var2, Var1, fill = value, label = value)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Effect Size") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 2.5) +
    theme_classic() +
    scale_x_discrete(
        expand = c(0, 0), limits = colnames(es_results)[clust2$order]) +
    scale_y_discrete(
        expand = c(0, 0), limits = rownames(es_results)[clust1$order]) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 10),
          axis.text.y = element_text(hjust = 1, size = 8),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.position = "none")

ggsave(paste0(IMAGE_PATH, "protein_SLDtypes_effect_sizes.png"),
       plot = plt, dpi = 600, width = 12,
       height = 12, units = "cm", scale = 2.5)
plt

```

In second plot, only the significant effect size labels were added to the
heat map. also, the dendograms after hierarchical clustering are shown.

```{r glm-heatmap2, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for significant associations the value of the effect size is shown."}
#' keep just the effect sizes of the significant associations
es_results_sig <- es_results
es_results_sig[as.matrix(sig_results)[rownames(es_results), ] == 0] <- ""

#' info regarding FIB4 (check=ok)
print('nr of significant proteins in the FIB4 group:')
sum(es_results_sig[, 'effectSize_liver_fat_cat22'] != "")

#' plot
colnames(es_results) <- sub("effectSize_", "", colnames(es_results))
pheatmap(es_results, display_numbers = es_results_sig,
         number_color = "#444444", fontsize_number = 7)

```

A considerable number of proteins are significant both for MASLD and for sex,
in many cases even with the same sign.

We investigate the associations for MASLD, i.e. the difference in average
concentrations between participants of SLD types *MASLD* and *none*. We show the
results for that coefficient using a volcano plot, coloring significant
associations in red.

```{r glm-volcano, FIG-volcano-FIB4, fig.path = IMAGE_PATH, fig.width = 8, fig.height = 8, fig.cap = "Volcano plot for the liver stiffness. Red colored points are those with an adjusted p-value smaller than 0.05."}
plot_volcano <- function(x, comparison = character(), ...) {
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    plot(X, -log10(Y), pch = 21,
         col = "#000000ce", bg = "#00000060",
         xlab = "coefficient", ylab = expression(-log[10](p[adjusted])),
         ...)
}
significant_points <- function(x, comparison = character(),
                               diff_perc = 2, col = "#E41A1CCE",
                               bg = "#E41A1C40", ...) {
    s <- x[, paste0("significant_", comparison)]
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    if (any(s))
        points(X[s], -log10(Y[s]), col = col, bg = bg, pch = 21)
}
plot_volcano(results, "liver_fat_cat22", main = "liver_fat_cat22")
grid()
abline(h = -log10(0.05), lty = 1, col = "grey")
significant_points(results, comparison = "liver_fat_cat22")

```

The (static) table with the significant proteins is shown below:

```{r glm-table-sig-sex, results = "asis", echo = FALSE}
significant_table <- function(x, comparison = character(),
                              columns = c("Genes", "description", "cv_qc_chris")) {
    sigs <- x[, paste0("significant_", comparison)]
    columns <- c(columns, paste0(c("coef_", "p.adj_", "effectSize_"),
                                 comparison))
    tmp <- x[sigs, columns]
    tmp <- tmp[order(tmp[, paste0("p.adj_", comparison)]), ]
    colnames(tmp) <- sub(paste0("_", comparison), "", colnames(tmp))
    tmp
}
sig_tab <- significant_table(results, comparison = "liver_fat_cat22")

pandoc.table(
    sig_tab, style = "rmarkdown", split.table = Inf,
    caption = paste0("proteins with significant differences in ",
                     "concentrations between MASLD and control (none) participants."))

```

The full results table is shown below.

```{r glm-table-top-5, results = "asis"}
##
oorder <- results[order(results[, "p.adj_liver_fat_cat22"]), ]

pandoc.table(oorder[1:5, c(1, 4, 7, 10, 13, 17)], style = "rmarkdown",
             split.table = Inf, caption = "Top 5 proteins for type MASLD")

datatable(results) |>
    formatRound(grep("coef|p.value|p.adj|effect", colnames(results)),
                digits = 6)

```

There are `r nrow(sig_tab)` significantly associated proteins for SLD type
MASLD vs. none.


Significant metabolites adjusted for age and sex (top 5) are visualized in
boxplot in next figure. Points are colored according to sample sex.


```{r glm-prep for plotting}
#' prep for plotting
PLOTLY <- TRUE
ggplotly_or_not <- function(x, interactive = TRUE) {
    if (interactive) ggplotly(x)
    else print(x)
}
myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))

box_plot <- function(data, aes, analyte) {
    ggplot(data, aes) +
        geom_jitter(position = position_jitter(0.24),
                    aes(colour = Sex, alpha = 0.5 , size = (BMI/10))) +
        geom_violin(trim = FALSE, aes(alpha = 0.5)) +
        stat_summary(fun = median, geom = "point", shape = 23, size = 2) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) + labs(title = analyte) +
        theme_bw ()
}

scatter_plot <- function(data, aes, analyte) {
    ggplot(data, aes) +
        geom_jitter(position = position_jitter(0.24),
                    aes(colour = Sex, alpha = 0.5 , size = 0.1)) +
        stat_smooth(method = lm, se = TRUE) + #with standard error
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) + labs(title = analyte) +
        theme_bw ()
}

#' for boxplot func to work automatically
tmp_colnr <- which(colnames(gal_met_prot_data) == TARGET)
gal_met_prot_data$tmp_target <- gal_met_prot_data[, tmp_colnr]   #value of eg BMI

```


```{r glm-top1, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(oorder)[1]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]  #intensities of metabolite
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)

```

```{r glm-top2, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- oorder[2, 1]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r glm-top3, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- oorder[3, 1]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r glm-top4, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- oorder[4, 1]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r glm-top5, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- oorder[5, 1]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```


# Logistic model with naab13_cat - metabolites

Done for metabolites. FYI: here using lm function, but with factors for target,
so logistic regression was performed.

```{r}
gal_met_prot_data$liver_fat_cat1 <- as.factor(gal_met_prot_data$liver_fat_cat1)
summary(gal_met_prot_data$liver_fat_cat1)

gal_met_prot_data <- gal_met_prot_data[!is.na(gal_met_prot_data$liver_fat_cat2), ]
gal_met_prot_data$liver_fat_cat2 <- as.factor(gal_met_prot_data$liver_fat_cat2)
summary(gal_met_prot_data$liver_fat_cat2)

```

logistic models setup:

```{r define-lm-covariates-and-target}
print('covariates of the linear regression model:')
COVARIATES <- c("Sex", "Age_10")
print(COVARIATES)

print('target of the linear regression model:')
TARGET <- 'liver_fat_cat2'
print("liver_fat_cat2")

DATASET <- 'NAFLD'
print(paste0('final variables of the ', DATASET, ' dataset used:'))
#' variables are the output y in lm(y ~ .):
#print(metabolites)

print('nr of variables inspected:')
print(length(metabolites))

print('nr of individuals inspected:')
print(dim(gal_met_prot_data))

```

To identify metabolites related with the 2 (none, MASLD) of the 4 SLD types
(none, MASLD, MetALD, SLD), seperate linear regression models are fitted for all
metabolites separately (as a response) with sex, age and the (categorical)
FIB4 as a covariates.

Model coefficients and p-values are provided. Raw p-values are adjusted for
multiple hypothesis testing with the method from Benjamini & Hochberg.

```{r linear model, message=FALSE}
#' Define a function to fit the linear model:
#' - fit the linear model
#' - extract the coefficients and p-values skipping the intercept
#' - return the result as a one-row matrix, columns being coefficients
#'   and p-values
funForLinReg <- function(analyte, data) {
    y <- data[, analyte]
    data <- data[, c("Sex", "Age_10", TARGET)]
    res <- coef(summary(lm(y ~ ., data = data)))[-1, ]
    pvals <- matrix(c(res[, 1L], res[, 4L]), nrow = 1)
    rownames(pvals) <- analyte
    colnames(pvals) <- c(paste0("coef_", rownames(res)),
                         paste0("p-value_", rownames(res)))
    pvals
}

#' tests
#gal_met_prot_data[1:5, c("Sex", "Age", "FIB4")]
#gal_met_prot_data[1:5, metabolites[1]]
#data <- gal_met_prot_data
#analyte <- metabolites[1]
#funForLinReg(analyte, gal_met_prot_data)
#analyte <- metabolites[14]
#funForLinReg(analyte, gal_met_prot_data)

#' Fit linear models and extract coefficients and p-values on each protein
lm_output <- lapply(metabolites, FUN = funForLinReg, data = gal_met_prot_data) |>
    do.call(what = rbind)

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)

```

Next we calculate the effect sizes.

To obtain comparable coefficients, prior to linear modelling all the metabolites
are standardized to zero mean and unit variance (note that this does not
influence the linear models since they are affine equivariant). In such
autoscaled data, difference of 1 corresponds to a standard deviation of 1. This
parameter will be called "effect size" and will be use to evaluate group
differences.

```{r effectsize, message=FALSE}
#' Perform autoscaling for effect size (es) calculation
gal_met_prot_data_scaled <- gal_met_prot_data
gal_met_prot_data_scaled[, metabolites] <- gal_met_prot_data[, metabolites] |>
    as.matrix() |>
    scale(scale = TRUE)

#' fit linear models and extract the coefficients
es <- lapply(metabolites, FUN = funForLinReg, data = gal_met_prot_data_scaled) |>
    do.call(what = rbind)
es <- es[, grep("coef", colnames(es))]
colnames(es) <- sub("coef", "effectSize", colnames(es))
lm_output <- cbind(lm_output, es)
#head(lm_output)

```

To call a metabolites *significant* we consider, in addition to the
*statistical* significance levels also the magnitude of its coefficient,
relative to the coefficient of variation (CV) for that protein. The CV,
calculated on study-internal QC samples (pools of all samples) are a measure of
the technical noise specific for that protein in the present data set. Thus,
with this method we prioritize metabolites with a larger difference between the
observed (biological) difference and the (separately estimated) technical variance.

Below we add thus evaluate for each protein and comparison whether it
fulfills that criteria.

```{r define significance criteria}
#' define the different coefficients as comparisons
comps <- colnames(lm_output)[grep("coef", colnames(lm_output))]

#' define which metabolites are considered statistically significant
sign_p <- lm_output[, grep("p-adj", colnames(lm_output))] < 0.05
colnames(sign_p) <- sub("p-adj_", "significant_", colnames(sign_p))
#colSums(sign_p)

#' Convert the (absolute) difference in abundance into a difference in
#' percentage based on a (log2) coefficient.
diff_percentage <- function(x) {
    (2^abs(x) - 1) * 100
}
diff_perc <- apply(lm_output[, comps], MARGIN = 2, diff_percentage)
colnames(diff_perc) <- sub("coef", "diff_perc", colnames(diff_perc))

#' Define for which metabolites the difference in concentration is larger
#' than a certain proportion of the CV
cv_cut <- lapply(comps, function(z) {
    z <- sub("coef_", "", z)
    diff_perc[, paste0("diff_perc_", z)] > 50 *  ##after eval, set to 50
        met_ann[metabolites, "cv_qc_chris"]
}) |>
    do.call(what = cbind)
colnames(cv_cut) <- sub("coef_", "cv_cut_", comps)

#' Use statistical significance for non-categorical traits instead of bool
sign_cv <- sign_p & cv_cut
sign_cv[, "significant_Age_10"] <- sign_p[, "significant_Age_10"]

#' Combine with result from linear model converting to a data.frame to
#' avoid logicals being converted to numeric
lm_output <- data.frame(lm_output, sign_cv)

```

The number of metabolites called significant or with an adjusted p-value smaller
than 0.05 for each of the extracted coefficients are shown below. For a complete
comparison, the number of metabolites complying to either the adjusted p value
or either the CV-value, as single threshold criteria, are calculated and
summarized here as well.

```{r table-sig, results = "asis"}
#' merge to p-adj to cv significance summary
sig_tab <- data.frame(
    `p_adj < 0.05` = colSums(sign_p),
    `cv thresh` = colSums(cv_cut),
    significant = colSums(lm_output[, sub("coef_", "significant_", comps)]),
    check.names = FALSE)

pandoc.table(
    sig_tab, style = "rmarkdown",
    caption = paste0("Number of significant metabolites per tested covariate. ",
                     " *p_adj < 0.05*: number of statistically significant",
                     " metabolites.",
                     " *cv*: number of metabolites with percentage abundance",
                     " difference above the CV-based threshold.",
                     " *significant*: number of metabolites that fulfill ",
                     "the CV-based criteria in addition to the p-adj. threshold"))

```

Export results for inspection:

```{r export-results}
results <- data.frame(
    met_ann[rownames(lm_output), c("analyte_name", "analyte_class",
                                   "biochemical_name", "cv_qc_chris")],
    lm_output,
    diff_perc
)

#colnames(results)
#colnames(lm_output)

#' write to file for manual inspection,
#' inclusing inspection of the CV-value of each protein
handle <- file.path(RDATA_PATH, 'lm_output_results_metabolites.txt')
write.table(results, file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)

lm_output_met <- results

```

We visualize the associations in a heatmap. The effect size are added for the
associations, for the significant metabolites (min 1 significance to
covariate). also for the not significant associations, the effect size is shown.

```{r heatmap, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for metabolites with at least one significant association.", eval = FALSE}
#' Extract effect sizes
es_results <- results[, grep("effectSize_", colnames(results))] |>
    as.matrix()

#' Restrict to those significant for at least one of the groups (sex, age, SLD type)
sig_results <- results[, grep("significant_", colnames(results))] |>
    as.matrix()
es_results <- es_results[rowSums(sig_results) > 0, ]
print('nr of metabolites with >= 1 significant association:')
nrow(es_results)

#' hierarchical clustering
clust1 <- hclust(dist((es_results)), method = "ward.D")
clust2 <- hclust(dist(t(es_results)), method = "ward.D")

#' clean roundup nrs for plot
es_results[es_results == 0] <- NA
es_results <- round(es_results, 2)

#' info regarding FIB4
#print('nr of metabolites with a not-zero effect size in the FIB4 group:')
#sum(es_results[, 'effectSize_FIB4'] != 0)

#' in long format
es_results_long <- melt(as.matrix(es_results))

#' ggplot
plt <- es_results_long |>
    ggplot(aes(Var2, Var1, fill = value, label = value)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Effect Size") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 2.5) +
    theme_classic() +
    scale_x_discrete(
        expand = c(0, 0), limits = colnames(es_results)[clust2$order]) +
    scale_y_discrete(
        expand = c(0, 0), limits = rownames(es_results)[clust1$order]) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 10),
          axis.text.y = element_text(hjust = 1, size = 8),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.position = "none")

ggsave(paste0(IMAGE_PATH, "protein_SLDtypes_effect_sizes.png"),
       plot = plt, dpi = 600, width = 12,
       height = 12, units = "cm", scale = 2.5)
plt

```

In second plot, only the significant effect size labels were added to the
heat map. also, the dendograms after hierarchical clustering are shown.

```{r heatmap2, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for significant associations the value of the effect size is shown.", eval = FALSE}
#' keep just the effect sizes of the significant associations
es_results_sig <- es_results
es_results_sig[as.matrix(sig_results)[rownames(es_results), ] == 0] <- ""

#' info regarding FIB4 (check=ok)
print('nr of significant metabolites in the FIB4 group:')
sum(es_results_sig[, 'effectSize_liver_fat_cat22'] != "")

#' plot
colnames(es_results) <- sub("effectSize_", "", colnames(es_results))
pheatmap(es_results, display_numbers = es_results_sig,
         number_color = "#444444", fontsize_number = 7)

```

A considerable number of metabolites are significant both for MASLD and for sex,
in many cases even with the same sign.

We investigate the associations for MASLD, i.e. the difference in average
concentrations between participants of SLD types *MASLD* and *none*. We show the
results for that coefficient using a volcano plot, coloring significant
associations in red.

```{r volcano, FIG-volcano-FIB4, fig.path = IMAGE_PATH, fig.width = 8, fig.height = 8, fig.cap = "Volcano plot for the liver stiffness. Red colored points are those with an adjusted p-value smaller than 0.05."}
plot_volcano <- function(x, comparison = character(), ...) {
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    plot(X, -log10(Y), pch = 21,
         col = "#000000ce", bg = "#00000060",
         xlab = "coefficient", ylab = expression(-log[10](p[adjusted])),
         ...)
}
significant_points <- function(x, comparison = character(),
                               diff_perc = 2, col = "#E41A1CCE",
                               bg = "#E41A1C40", ...) {
    s <- x[, paste0("significant_", comparison)]
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    if (any(s))
        points(X[s], -log10(Y[s]), col = col, bg = bg, pch = 21)
}
plot_volcano(results, "liver_fat_cat22", main = "liver_fat_cat22")
grid()
abline(h = -log10(0.05), lty = 1, col = "grey")
significant_points(results, comparison = "liver_fat_cat22")

```

The (static) table with the significant metabolites is shown below:

```{r table-sig-sex, results = "asis", echo = FALSE}
significant_table <- function(x, comparison = character(),
                              columns = c("analyte_class", "cv_qc_chris")) {
    sigs <- x[, paste0("significant_", comparison)]
    columns <- c(columns, paste0(c("coef_", "p.adj_", "effectSize_"),
                                 comparison))
    tmp <- x[sigs, columns]
    tmp <- tmp[order(tmp[, paste0("p.adj_", comparison)]), ]
    colnames(tmp) <- sub(paste0("_", comparison), "", colnames(tmp))
    tmp
}
sig_tab <- significant_table(results, comparison = "liver_fat_cat22")

pandoc.table(
    sig_tab, style = "rmarkdown", split.table = Inf,
    caption = paste0("metabolites with significant differences in ",
                     "concentrations between MASLD and control (none) participants."))

```

The full results table is shown below.

```{r table-top-5, results = "asis"}
##
oorder <- results[order(results[, "p.adj_liver_fat_cat22"]), ]

pandoc.table(oorder[1:5, c(1, 4, 7, 10, 13, 17)], style = "rmarkdown",
             split.table = Inf, caption = "Top 5 metabolites for type MASLD")

datatable(results) |>
    formatRound(grep("coef|p.value|p.adj|effect", colnames(results)),
                digits = 6)

```

There are `r nrow(sig_tab)` significantly associated metabolites for SLD type
MASLD vs. none.


Significant metabolites adjusted for age and sex (top 5) are visualized in
boxplot in next figure. Points are colored according to sample sex.


```{r prep for plotting}
#' prep for plotting
PLOTLY <- TRUE
ggplotly_or_not <- function(x, interactive = TRUE) {
    if (interactive) ggplotly(x)
    else print(x)
}
myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))

box_plot <- function(data, aes, analyte) {
    ggplot(data, aes) +
        geom_jitter(position = position_jitter(0.24),
                    aes(colour = Sex, alpha = 0.5 , size = (BMI/10))) +
        geom_violin(trim = FALSE, aes(alpha = 0.5)) +
        stat_summary(fun = median, geom = "point", shape = 23, size = 2) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) + labs(title = analyte) +
        theme_bw ()
}

scatter_plot <- function(data, aes, analyte) {
    ggplot(data, aes) +
        geom_jitter(position = position_jitter(0.24),
                    aes(colour = Sex, alpha = 0.5 , size = 0.1)) +
        stat_smooth(method = lm, se = TRUE) + #with standard error
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) + labs(title = analyte) +
        theme_bw ()
}

#' for boxplot func to work automatically
tmp_colnr <- which(colnames(gal_met_prot_data) == TARGET)
gal_met_prot_data$tmp_target <- gal_met_prot_data[, tmp_colnr]   #value of eg BMI

```


```{r lm-top1, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(oorder)[1]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]  #intensities of metabolite
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)

```

```{r lm-top2, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- oorder[2, 1]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r lm-top3, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- oorder[3, 1]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r lm-top4, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- oorder[4, 1]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r lm-top5, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- oorder[5, 1]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```


# Conclusions

With (W) T2D: 103# total with 31 control + 38 grade 3 steatosis (balanced, small set size)

- 0 signif proteins
- 0 signif metabolites

Without (WO) T2D: 219# total with 110 control + 41 grade 3 steatosis (not balanced, ok set size)

- 7 signif proteins: GC, AFM, GC, AZGP1, SERPINC1, VTN, LRG1
- 12 signif metabolites: Glu, Tyr, C5, PC ae C34:3, C3, lysoPC a C18:1, Val, 
PC aa C42:1, Ala, Gly, lysoPC a C18:2

Question to Jo: how know because of the no-T2D, and not because of the 
inbalanced and small sample size? The original lm 
(steatosis 141 vs 79) is also inbalanced but less and bigger in sample size.

The siginificant variables more reflect a healthy vs unhealthy phenotype. as 
same for MASLD vs none, and control vs G3 steatosis. When the control group 
also has T2D comorbidity, no differneces anymore.

# Acknowledgments

Consulted sources used as basis for code:

- ~/nafld_proteo_metabolomics/NAFLD_metab_targeted_lm.Rmd
- ~?metabo_chris_untargeted/CHRIS_lm_pos.Rmd
- kora: CHRIS_KORA_tmetab_lm_binary.Rmd


# Session information

R packages used for the analysis:

```{r}
sessionInfo()
```

