---
title: "NAFLD_fibrosis_tmetab_lm"
author: "Marilyn De Graeve, Johannes Rainer, Christoph Grander"
affiliation: "Eurac Research, Bolzano, Italy"
date: "2025-01-27"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
---

**Modified**: `r file.info("NAFLD_fibrosis_tmetab_lm.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r style, message = FALSE, echo = FALSE, warning = FALSE, results = "asis"}
library("BiocStyle")
library("kableExtra")
library("knitr")
library("rmarkdown")
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now, units = "secs")
      # return a character string to show the time
      paste("    Time for this code chunk to run:", round(res,
        2), "seconds")
    }
  }
}))
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7, time_it = TRUE)

```

# Introduction

During this workflow, statistical analyis (linear models) are performed on the
targeted metabolomics data from the metabolic dysfunction-associated steatotic
liver disease (MASLD) cohort. MASLD, previously known as NAFLD is the most
chronic liver disease, affecting more than 30% of global population. The MAFLD
cohort is a subset of approximatley 353 # participants from the Cooperative
Health Research in South Tirol (CHRIS) study, which were requested to return for
a second resampling and additional phenotyping. The NAFLD cohort represent an
unhealty group with various degrees of steatotic (alcoholic) liver disease
classifications. See [nafld_data_overview.Rmd](nafld_data_overview.Rmd)
(respectively the rendered html report) for an overview of the data set and the
definition of the phenotype.

This document describes the statistical analysis using multiple linear models to
identify associations between metabolites and fibrosis, whitin participant with 
metabolic dysfunction-associated
steatotic liver disease (MASLD). The definition of control and MASLD types are:

- *none*: participants negative for steatotic liver disease.
- *MASLD*: (= Metabolic Dysfunction Associated Steatotic Liver Disease)
  participants with SLD, positive for cardiometabolic criteria but low alcohol
  consumption (below 40g/day). This includes also 8 participants without
  information on alcohol consumption. These participants did not provide the
  GA2LEN FFQ (hence missing values for quantified alcohol consumption), but
  according to alcohol-related information in other questionnaire they are not
  heavy drinkers and are thus added to this category.

The MetALD and SLD types are not included in this analysis. FYI:

- *MetALD*: (= MASLD and increased alcohol intake) participants with SLD,
  positive for cardiometabolic criteria but daily alcohol consumption > 40g/day.
- *SLD*: (= steatotic liver disease) participants with a SLD, but negative for
  cardiometabolic criteria. This includes potentially cryptogenic SLD, alcoholic
  liver disease (ALD), DILI or monogenic diseases.

See the `nafld_data_overview 1.html` report for further
information regarding the SLD sub-classification.

Regarding fibrosis, i.e.  the formation of excess fibrous connective tissue in 
an organ or tissue in a reparative or reactive process. In this scope: liver 
fibrosis. The variables containing this parameter is:

- *nascfib4*: fibrosis 4 index. Linear score calculated from standard clinical 
  parameters.

This parameter will be used as the target for a biomarker search in *MASLD* 
participants.


# Setup

## Directories

```{r directories, echo = FALSE}
#' General settings
filename <- "NAFLD_fibrosis_tmetab_lm"

#' Path to save images to; remove if exists.
IMAGE_PATH <- paste0("images/", filename, "/")
dir.create(IMAGE_PATH, recursive = TRUE)

#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)

#' Path to the data
#DATA_PATH <- '/home/mdegraeve/Documents/Files/Work_Eurac/Data/NAFLD' #FINDME!
DATA_PATH <- "."

```

## Packages

```{r packages, message=FALSE}
#' Load libraries
library(chrisUtils)
library(chrisData)
library(DT)
library(ggfortify)
library(ggplot2)
library(pander)
library(pheatmap)
library(plotly)
library(RColorBrewer)
library(readxl)
library(reshape2)
library(tidyfr)
library(vioplot)
library(writexl)

```


# Data import

The data from metabolites was converted into the TDF format and can hence be
loaded and imported using the `tidyfr` R package. Now we read the full data
which includes general information for each measurement.


```{r load_dataset, message = FALSE}
#' load TDF using tidyfr
#list_data_modules(path = DATA_PATH)
p180 <- data_module(name = "metabolomics_p180", version = "1.0.1.2",
                    path = DATA_PATH)
met_data <- data(p180)
met_ann <- labels(p180)
stopifnot(all(colnames(met_data) == rownames(met_ann)))
colnames(met_data) <- met_ann$description
rownames(met_ann) <- met_ann$description

#' define the columns with metabolite concentrations
metabolites <- rownames(met_ann)[met_ann$type != "categorical"]

#' problem in calculating lm below for
#'   'PC aa C30:2' #NOK, only 8 col!!
#'   'Histamine' NOK, only 8 col!!
#' because no values for each typeSLD, due to lot NA values
#' option 1: rm metabolites, not done
#metabolites <- metabolites[metabolites != 'PC aa C30:2']
#metabolites <- metabolites[metabolites != 'Histamine']
#' option2: retain only none and MASLD, see done later

#' select metabolites dataframe
met_data <- met_data[, metabolites]

#' Impute missing values using an uniform distribution, after gap filling
na_unidis <- function(z) {
    na <- is.na(z)
    if (any(na)) {
        min = min(z, na.rm = TRUE)
        z[na] <- runif(sum(na), min = min/2, max = min)
    }
    z
}

#' Row-wise impute missing values and add the data as a new assay
tmp <- apply(met_data, MARGIN = 1, na_unidis)
met_data <- t(tmp)

#' checks
#head(met_data)    #row samples, cols are metab name, matrix are absol conc
print('nr of  metabolites included:')
ncol(met_data)
print('included metabolites are:')
colnames(met_data)
#dim(met_data) #ok!

```

The loaded data contains the samples as rows, and the metabolite names as
colums. The measurements of the `met_data` object are the absolute
concentrations of metabolite per sample, in natural scale.

To ensure a normal distribution of the metabololits, we perform log2
transformation.

```{r log2-transformation, message = FALSE}
#' log2 scaling of absolute abundances metabolites
met_data <- log2(met_data)

```

Before proceeding with the data analysis we remove the POOL samples. In addition
we restrict to samples from the NAFLD study.

```{r subset to NAFLD set, message = FALSE, warning = FALSE}
met_data <- met_data[grep("^008", rownames(met_data)), ]

```

All available trait (phenotype) data was defined in the
[nafld_data_overview.Rmd](nafld_data_overview.Rmd), is loaded below and merged
with the log2 transformed metabolomics data of the NAFLD cohort.

```{r add phenotype dataa}
#' load xlsx file instead of below
handle <- 'masld_phenotypes.xlsx'
masld_phenotype_df <- read_excel(file.path(DATA_PATH, handle)) |>
    as.data.frame()
#head(masld_phenotype_df, 5)
#colnames(masld_phenotype_df)
#dim(masld_phenotype_df)

#' make sure sld_type is a factor with group 'none' as the baseline
masld_phenotype_df$sld_type <- factor(masld_phenotype_df$sld_type,
                                      levels = c('none', 'MASLD',
                                                 'MetALD', 'SLD'))
print('the sld types are:')
levels(masld_phenotype_df$sld_type)
print('nr of samples per sld types:')
table(masld_phenotype_df$sld_type)

#' merge with metabolomics data
met_data <- as.data.frame(met_data)
met_data$AID <- rownames(met_data)

met_data <- merge(met_data, masld_phenotype_df, by = 'AID')
#dim(met_data)
#head(met_data)

#' write to file for manuel inspection
#handle <- 'data/target_met_data.txt'
#write.table(met_data, file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)

```

Because for 2 metabolites, 'PC aa C30:2' and 'Histamine', only NAs are present
for the less populated `sld_type` groups (*MetALD* end *SLD*), only *none* and
*MASLD* types are considered further.

```{r keep only none and MASLD types}
#' because problem 2 metabolites only NAs in some sld types,
#' keep only none and MASLD
met_data <- met_data[met_data$sld_type %in% c('MASLD'), ]  #FINDME! 'none', 'MASLD'
met_data$sld_type <- droplevels(met_data$sld_type)
met_data$Sex <- factor(met_data$Sex)

#' Summary of available data:
summary(met_data[, c("sld_type", "Age", "Sex", "BMI")])

```

Create additional variable for age per 10y, because coefficients and effect
sizes correspond to one year difference, and the annual difference is
negligible. Dividing age with 10, difference in coefficients and effect sizes
would correspond to 10 years difference and it would be easier to define
meaningful cut-off.

```{r define-age10-variable, message = FALSE}
#' define age per 10y group
met_data$Age_10 <- met_data$Age / 10

```

Add the information from the [nafld_data_overview.Rmd](nafld_data_overview.Rmd) 
report regarding the fibrosis, using the parameter nascfib4: fibrosis 4 index. 
Linear score calculated from standard clinical parameters. Further called FIB4.

```{r load-liver-param}
#' from same excel loaded above, already added to met_data
#colnames(met_data)
summary(met_data[, c("FIB4", "T2D")])

```
FIB4 is numeric, continuous ranging between 0.3 - 7.9. A higher score is more 
liver fibrosis.


# Evaluation of the various variables in the data set

Overview of variable evaluation in the data set, prior to the lm regression:

- Violin plot of BMI of the data split by sex and SLD type. to evaluate whether
  there is a strong difference in BMI between MASLD and none control and, if so,
  if that variable also to the linear model.
- Violin plot for age.
- PCA of the data. 

## FIB4 distribution of data

Recap from overview report. 
"There seems to be only a small difference in average FIB4 values between male 
and female participants. A strong (positive) relationship is however visible 
between FIB4 and age. A slight negative relationship seems to be present between 
FIB4 and BMI."

```{r violin-plots, fig.height=8, fig.width=7}
#violin sex vs FIB4 
vioplot(split(met_data$FIB4,
              met_data$Sex),
        xlab = "Sex", ylab = "FIB4",
        main = "FIB4 distribution by sex")

#binned plot for visibility
met_data$FIB4_bin <- round(met_data$FIB4, 0)
summary(met_data$FIB4_bin)

#violin age vs FIB4_bin
vioplot(split(met_data$Age,
              met_data$FIB4_bin),
        xlab = "FIB4_bin", ylab = "Age",
        main = "FIB4_bin versus Age")

#violin BMI vs FIB4 
vioplot(split(met_data$BMI,
              met_data$FIB4_bin),
        xlab = "FIB4_bin", ylab = "BMI",
        main = "BMI_bin versus BMI")

#violin T2D vs FIB4 
vioplot(split(met_data$FIB4,
              met_data$T2D),
        xlab = "T2D", ylab = "FIB4",
        main = "FIB4 distribution by T2D")

```


## PCA score plot

PCA score plot of the study participants, collored by:

- BMI,
- Age,
- gender,
- T2D
- clinical phenotype (i.e. `FIB4` values).

Prior the PCA, data is normalized to 0 mean and standard deviation of
1 (autoscaling). Plot shows the scores and loadings of all subjects in the
space of the first two PCs, with 33% of total variances explained.
Fixed error for PCA calculation (NAs after scaling) by imputing above.

```{r pca, fig.height=5, fig.width=7}

#' scaling and PCA
scaledData <- scale(met_data[, metabolites], scale = TRUE)
scPCA <- svd(scaledData)
scPCA.scores <- scPCA$u %*% diag(scPCA$d)    ## scores
scPCA.loadings <- scPCA$v                    ## loadings
rownames(scPCA.loadings) <- metabolites
scPCA.variances <- round(100*((scPCA$d^2) / sum(scPCA$d^2)), 1)[1:10]

#' variance explained PCs
print('variance explained for PC1 + PC2:')
scPCA.variances[1] + scPCA.variances[2]

#' prep for plotting
PLOTLY <- TRUE
ggplotly_or_not <- function(x, interactive = TRUE) {
    if (interactive) ggplotly(x)
    else print(x)
}
myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))

#' plot with col BMI
BMI <- met_data$BMI
gg_1 <- autoplot(prcomp(scaledData), data = met_data) +
    geom_point(aes(colour = BMI), size = (BMI/10)) +
    scale_colour_gradientn(colours = myPalette(100)) +
    theme_bw()
ggplotly_or_not(gg_1, PLOTLY)

#' plot with col Age
Age <- met_data$Age
gg_1 <- autoplot(prcomp(scaledData), data = met_data) +
    geom_point(aes(colour = Age), size = (Age/25)) +
    scale_colour_gradientn(colours = myPalette(100)) +
    theme_bw()
ggplotly_or_not(gg_1, PLOTLY)

#' plot with col Sex
Sex <- met_data$Sex
gg_1 <- autoplot(prcomp(scaledData), data = met_data) +
    geom_point(aes(colour = Sex)) +
    scale_color_manual(name = Sex, values = c('pink', 'lightblue')) +
    theme_bw()
gg_1
#ggplotly_or_not(gg_1, PLOTLY)

#' plot with col T2D
T2D <- met_data$T2D
gg_1 <- autoplot(prcomp(scaledData), data = met_data) +
    geom_point(aes(colour = T2D)) +
    scale_color_manual(name = T2D, values = c('gold', 'purple')) +
    theme_bw()
gg_1
#ggplotly_or_not(gg_1, PLOTLY)

#' plot with col FIB4
FIB4 <- met_data$FIB4
gg_1 <- autoplot(prcomp(scaledData), data = met_data) +
    geom_point(aes(colour = FIB4), size = (FIB4)) +
    scale_colour_gradientn(colours = myPalette(100)) +
    theme_bw()
ggplotly_or_not(gg_1, PLOTLY)


#' plot with loading
gg_1 <- autoplot(prcomp(scaledData), data = met_data,
                 loadings = TRUE, loadings.colour = 'gray',
                 loadings.label = TRUE, loadings.label.size = 3,
                 loadings.label.colour ="darkblue", colour = NA) +
    geom_point(color = 'gray', size = 0.1) +
    theme_bw()
ggplotly_or_not(gg_1, PLOTLY)

```


# Linear model to evaluate (nasc)FIB4

linear models setup:

```{r define-lm-covariates-and-target}
print('covariates of the linear regression model:')
COVARIATES <- c("Sex", "Age_10")
print(COVARIATES)

print('target of the linear regression model:')
TARGET <- 'FIB4'
print("FIB4")

DATASET <- 'NAFLD'
print(paste0('final variables of the ', DATASET, ' dataset used:'))
#' variables are the output y in lm(y ~ .):
print(metabolites)

print('nr of variables inspected:')
print(length(metabolites))

print('nr of individuals inspected:')
print(dim(met_data))

```

To identify metabolites related with the 2 (none, MASLD) of the 4 SLD types
(none, MASLD, MetALD, SLD), seperate linear regression models are fitted for all
metabolites separately (as a response) with sex, age and the (categorical)
FIB4 as a covariates.

Model coefficients and p-values are provided. Raw p-values are adjusted for
multiple hypothesis testing with the method from Benjamini & Hochberg.

```{r linear model, message=FALSE}
#' Define a function to fit the linear model:
#' - fit the linear model
#' - extract the coefficients and p-values skipping the intercept
#' - return the result as a one-row matrix, columns being coefficients
#'   and p-values
funForLinReg <- function(analyte, data) {
    y <- data[, analyte]
    data <- data[, c("Sex", "Age_10", "FIB4")]
    res <- coef(summary(lm(y ~ ., data = data)))[-1, ]
    pvals <- matrix(c(res[, 1L], res[, 4L]), nrow = 1)
    rownames(pvals) <- analyte
    colnames(pvals) <- c(paste0("coef_", rownames(res)),
                         paste0("p-value_", rownames(res)))
    pvals
}

#' tests
#met_data[1:5, c("Sex", "Age", "FIB4")]
#met_data[1:5, metabolites[1]]
#data <- met_data
#analyte <- metabolites[1]
#funForLinReg(analyte, met_data)
#analyte <- metabolites[14]
#funForLinReg(analyte, met_data)

#' Fit linear models and extract coefficients and p-values on each metabolite
lm_output <- lapply(metabolites, FUN = funForLinReg, data = met_data) |>
    do.call(what = rbind)

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)

```

Next we calculate the effect sizes.

To obtain comparable coefficients, prior to linear modelling all the metabolites
are standardized to zero mean and unit variance (note that this does not
influence the linear models since they are affine equivariant). In such
autoscaled data, difference of 1 corresponds to a standard deviation of 1. This
parameter will be called "effect size" and will be use to evaluate group
differences.

```{r effectsize, message=FALSE}
#' Perform autoscaling for effect size (es) calculation
met_data_scaled <- met_data
met_data_scaled[, metabolites] <- met_data[, metabolites] |>
    as.matrix() |>
    scale(scale = TRUE)

#' fit linear models and extract the coefficients
es <- lapply(metabolites, FUN = funForLinReg, data = met_data_scaled) |>
    do.call(what = rbind)
es <- es[, grep("coef", colnames(es))]
colnames(es) <- sub("coef", "effectSize", colnames(es))
lm_output <- cbind(lm_output, es)
#head(lm_output)

```

To call a metabolites *significant* we consider, in addition to the
*statistical* significance levels also the magnitude of its coefficient,
relative to the coefficient of variation (CV) for that metabolite. The CV,
calculated on study-internal QC samples (pools of all samples) are a measure of
the technical noise specific for that metabolite in the present data set. Thus,
with this method we prioritize metabolites with a larger difference between the
observed (biological) difference and the (separately estimated) technical variance.

Below we add thus evaluate for each metabolite and comparison whether it
fulfills that criteria.

```{r define significance criteria}
#' define the different coefficients as comparisons
comps <- colnames(lm_output)[grep("coef", colnames(lm_output))]

#' define which metabolites are considered statistically significant
sign_p <- lm_output[, grep("p-adj", colnames(lm_output))] < 0.05
colnames(sign_p) <- sub("p-adj_", "significant_", colnames(sign_p))
#colSums(sign_p)

#' Convert the (absolute) difference in abundance into a difference in
#' percentage based on a (log2) coefficient.
diff_percentage <- function(x) {
    (2^abs(x) - 1) * 100
}
diff_perc <- apply(lm_output[, comps], MARGIN = 2, diff_percentage)
colnames(diff_perc) <- sub("coef", "diff_perc", colnames(diff_perc))

#' Define for which metabolites the difference in concentration is larger
#' than a certain proportion of the CV
cv_cut <- lapply(comps, function(z) {
    z <- sub("coef_", "", z)
    diff_perc[, paste0("diff_perc_", z)] > 50 *  #after eval, set to 50
        met_ann[metabolites, "cv_qc_chris"]
}) |>
    do.call(what = cbind)
colnames(cv_cut) <- sub("coef_", "cv_cut_", comps)

#' Use statistical significance for non-categorical traits instead of bool
sign_cv <- sign_p & cv_cut
sign_cv[, "significant_Age_10"] <- sign_p[, "significant_Age_10"]

#' Combine with result from linear model converting to a data.frame to
#' avoid logicals being converted to numeric
lm_output <- data.frame(lm_output, sign_cv)

```

The number of metabolites called significant or with an adjusted p-value smaller
than 0.05 for each of the extracted coefficients are shown below. For a complete
comparison, the number of metabolites complying to either the adjusted p value
or either the CV-value, as single threshold criteria, are calculated and
summarized here as well.

```{r table-sig-mets, results = "asis"}
#' merge to p-adj to cv significance summary
sig_tab <- data.frame(
    `p_adj < 0.05` = colSums(sign_p),
    `cv thresh` = colSums(cv_cut),
    significant = colSums(lm_output[, sub("coef_", "significant_", comps)]),
    check.names = FALSE)

pandoc.table(
    sig_tab, style = "rmarkdown",
    caption = paste0("Number of significant metabolites per tested covariate. ",
                     " *p_adj < 0.05*: number of statistically significant",
                     " metabolites.",
                     " *cv*: number of metabolites with percentage abundance",
                     " difference above the CV-based threshold.",
                     " *significant*: number of metabolites that fulfill ",
                     "the CV-based criteria in addition to the p-adj. threshold"))

```

Export results for inspection:

```{r export-results}
results <- data.frame(
    met_ann[rownames(lm_output), c("analyte_name", "analyte_class",
                                   "biochemical_name", "cv_qc_chris")],
    lm_output,
    diff_perc
)

#colnames(results)
#colnames(lm_output)

#' write to file for manual inspection,
#' inclusing inspection of the CV-value of each metabolite
#handle <- 'data/lm_output_results.txt'
#write.table(results, file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)

```

We visualize the associations in a heatmap. The effect size are added for the
associations, for the significant metabolites (min 1 significance to
covariate). also for the not significant associations, the effect size is shown.

```{r heatmap, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for metabolites with at least one significant association."}
#' Extract effect sizes
es_results <- results[, grep("effectSize_", colnames(results))] |>
    as.matrix()

#' Restrict to those significant for at least one of the groups (sex, age, SLD type)
sig_results <- results[, grep("significant_", colnames(results))] |>
    as.matrix()
es_results <- es_results[rowSums(sig_results) > 0, ]
print('nr of metabolites with >= 1 significant association:')
nrow(es_results)

#' hierarchical clustering
clust1 <- hclust(dist((es_results)), method = "ward.D")
clust2 <- hclust(dist(t(es_results)), method = "ward.D")

#' clean roundup nrs for plot
es_results[es_results == 0] <- NA
es_results <- round(es_results, 2)

#' info regarding FIB4
#print('nr of metabolites with a not-zero effect size in the FIB4 group:')
#sum(es_results[, 'effectSize_FIB4'] != 0) #FINDME!!!

#' in long format
es_results_long <- melt(as.matrix(es_results))

#' ggplot
plt <- es_results_long |>
    ggplot(aes(Var2, Var1, fill = value, label = value)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Effect Size") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 2.5) +
    theme_classic() +
    scale_x_discrete(
        expand = c(0, 0), limits = colnames(es_results)[clust2$order]) +
    scale_y_discrete(
        expand = c(0, 0), limits = rownames(es_results)[clust1$order]) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 10),
          axis.text.y = element_text(hjust = 1, size = 8),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.position = "none")

ggsave(paste0(IMAGE_PATH, "metabolite_SLDtypes_effect_sizes.png"),
       plot = plt, dpi = 600, width = 12,
       height = 12, units = "cm", scale = 2.5)
plt

```

In second plot, only the significant effect size labels were added to the
heat map. also, the dendograms after hierarchical clustering are shown.

```{r heatmap2, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for significant associations the value of the effect size is shown."}
#' keep just the effect sizes of the significant associations
es_results_sig <- es_results
es_results_sig[as.matrix(sig_results)[rownames(es_results), ] == 0] <- ""

#' info regarding FIB4 (check=ok)
print('nr of significant metabolites in the FIB4 group:')
sum(es_results_sig[, 'effectSize_FIB4'] != "") #FINDME

#' plot
colnames(es_results) <- sub("effectSize_", "", colnames(es_results))
pheatmap(es_results, display_numbers = es_results_sig,
         number_color = "#444444", fontsize_number = 7)

```

A considerable number of metabolites are significant both for MASLD and for sex,
in many cases even with the same sign.

We investigate the associations for MASLD, i.e. the difference in average
concentrations between participants of SLD types *MASLD* and *none*. We show the
results for that coefficient using a volcano plot, coloring significant
associations in red.

```{r volcano, FIG-volcano-FIB4, fig.path = IMAGE_PATH, fig.width = 8, fig.height = 8, fig.cap = "Volcano plot for the liver stiffness. Red colored points are those with an adjusted p-value smaller than 0.05."}
plot_volcano <- function(x, comparison = character(), ...) {
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    plot(X, -log10(Y), pch = 21,
         col = "#000000ce", bg = "#00000060",
         xlab = "coefficient", ylab = expression(-log[10](p[adjusted])),
         ...)
}
significant_points <- function(x, comparison = character(),
                               diff_perc = 2, col = "#E41A1CCE",
                               bg = "#E41A1C40", ...) {
    s <- x[, paste0("significant_", comparison)]
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    if (any(s))
        points(X[s], -log10(Y[s]), col = col, bg = bg, pch = 21)
}
plot_volcano(results, "FIB4", main = "MASLD vs none")
grid()
abline(h = -log10(0.05), lty = 1, col = "grey")
significant_points(results, comparison = "FIB4")

```

The (static) table with the significant metabolites is shown below:

```{r table-sig-sex, results = "asis", echo = FALSE}
significant_table <- function(x, comparison = character(),
                              columns = c("analyte_class", "cv_qc_chris")) {
    sigs <- x[, paste0("significant_", comparison)]
    columns <- c(columns, paste0(c("coef_", "p.adj_", "effectSize_"),
                                 comparison))
    tmp <- x[sigs, columns]
    tmp <- tmp[order(tmp[, paste0("p.adj_", comparison)]), ]
    colnames(tmp) <- sub(paste0("_", comparison), "", colnames(tmp))
    tmp
}
sig_tab <- significant_table(results, comparison = "FIB4")

pandoc.table(
    sig_tab, style = "rmarkdown", split.table = Inf,
    caption = paste0("Metabolites with significant differences in ",
                     "concentrations between MASLD and CTRL participants."))

```

The full results table is shown below.

```{r table-top-20, results = "asis"}
##
oorder <- results[order(results[, "p.adj_FIB4"]), ]

pandoc.table(oorder[1:5, c(1, 4, 7, 10, 13, 17)], style = "rmarkdown",
             split.table = Inf, caption = "Top 5 metabolites for type MASLD")

datatable(results) |>
    formatRound(grep("coef|p.value|p.adj|effect", colnames(results)),
                digits = 6)

```

There are `r nrow(sig_tab)` significantly associated metabolites for SLD type
MASLD vs. none.


Significant metabolites adjusted for age and sex (top 5) are visualized in
boxplot in next figure. Points are colored according to sample sex.


```{r prep for plotting}
#' prep for plotting
PLOTLY <- TRUE
ggplotly_or_not <- function(x, interactive = TRUE) {
    if (interactive) ggplotly(x)
    else print(x)
}
myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))

box_plot <- function(data, aes, analyte) {
    ggplot(data, aes) +
        geom_jitter(position = position_jitter(0.24),
                    aes(colour = Sex, alpha = 0.5 , size = (BMI/10))) +
        geom_violin(trim = FALSE, aes(alpha = 0.5)) +
        stat_summary(fun = median, geom = "point", shape = 23, size = 2) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) + labs(title = analyte) +
        theme_bw ()
}

scatter_plot <- function(data, aes, analyte) {
    ggplot(data, aes) +
        geom_jitter(position = position_jitter(0.24),
                    aes(colour = Sex, alpha = 0.5 , size = 0.1)) +
        stat_smooth(method = lm, se = TRUE) + #with standard error
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) + labs(title = analyte) +
        theme_bw ()
}

#' for boxplot func to work automatically
tmp_colnr <- which(colnames(met_data) == TARGET)
met_data$tmp_target <- met_data[, tmp_colnr]   #value of eg BMI

```


```{r lm-top1, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(oorder)[1]
met_data$top_analyte <- met_data[, top_analyte]  #intensities of metabolite
s_1 <- scatter_plot(met_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)

```

```{r lm-top2, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- oorder[2, 1]
met_data$top_analyte <- met_data[, top_analyte]
s_1 <- scatter_plot(met_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r lm-top3, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- oorder[3, 1]
met_data$top_analyte <- met_data[, top_analyte]
s_1 <- scatter_plot(met_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r lm-top4, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- oorder[4, 1]
met_data$top_analyte <- met_data[, top_analyte]
s_1 <- scatter_plot(met_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r lm-top5, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- oorder[5, 1]
met_data$top_analyte <- met_data[, top_analyte]
s_1 <- scatter_plot(met_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```


```{r save results for sensitivity analysis}
#' get coef values from both lm models
lm_original <- results$coef_FIB4
```


In addition, we also export the results table from the lm containing the m/z,
rt, metabolite feature ID, statistics (coef, es) and p-vlues (adj. p-value) for
subsequent analysis (eg pathway enrichment and analysis).

```{r export-lm-results, eval = FALSE}
#' add rownames as a column
results$AID <- rownames(results)

#' write results (original order) for further analysis
handle <- file.path(DATA_PATH, paste0(filename, '_results.txt'))
write.table(results, file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)

```


# Acknowledgments

Consulted sources used as basis for code:

- ~/nafld_proteo_metabolomics/NAFLD_metab_targeted_lm.Rmd
- ~?metabo_chris_untargeted/CHRIS_lm_pos.Rmd


# Session information

R packages used for the analysis:

```{r}
sessionInfo()
```


