---
title: "Overview and characterization of the NAFLD substudy"
author: "Johannes Rainer"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
bibliography: references.bib
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 10, fig.height = 7)
```

# Introduction

In this document we evaluate and analyze the clinical/phenotypic data available
for participants of the NAFLD sub-study of CHRIS. In addition, we
stratify/diagnose individuals following recent clinical guidelines.


# NAFLD data

Demographics of the NAFLD cohort and evaluation of available trait data.

```{r}
library(chrisData)
library(nafldData)
library(vioplot)

n_anns <- nafldTraitAnnotations()

#' personal data
n_perso <- nafldData("personal")
n_data <- n_perso[, c("AID", "aid_chris", "na_ager", "na_sex", "na_fwup")]
#' na_ager: rounded age
#' na_fwup: follow-up duration CHRIS->NAFLD

tmp <- nafldData("recruitment")
rownames(tmp) <- tmp$AID
n_data$case <- tmp[n_data$AID, "na_case"]

```

The data set is nicely balanced by sex:

```{r}
table(n_data$na_sex)
```

The age distribution of participants (split by sex) is shown below.

```{r, fig.cap = "Age distribution."}
vioplot(split(n_data$na_ager, n_data$na_sex), ylab = "age")
grid()
```

The distribution of time between participants' CHRIS baseline and NAFLD
participation is shown below.

```{r, fig.cap = "Distribution of the time span between CHRIS baseline and NAFLD."}
plot(density(n_data$na_fwup))
grid()
```

The time span between the two studies is thus between 2 to 5 years. We next
evaluate BMI and fat composition for the study participants, first split by
gender and then by case/control definition.

```{r}
#' BMI, fat etc.
n_anthro <- nafldData("anthro")
rownames(n_anthro) <- n_anthro$AID

n_data$BMI <- n_anthro[n_data$AID, "naan03a"]
n_data$body_fat <- n_anthro[n_data$AID, "naan04"]
n_data$visc_fat <- n_anthro[n_data$AID, "naan05"]

```

```{r, fig.cap = "Distribution of BMI and body fat per sex or *case*."}
par(mfrow = c(2, 3))
#' split by sex
vioplot(split(n_data$BMI, n_data$na_sex), main = "BMI", ylab = "BMI")
grid()
vioplot(split(n_data$body_fat, n_data$na_sex), main = "Body fat",
        ylab = "Body fat")
grid()
vioplot(split(n_data$visc_fat, n_data$na_sex), main = "Visceral fat",
        ylab = "Visceral fat")
grid()
#' split by case
vioplot(split(n_data$BMI, n_data$case), main = "BMI", ylab = "BMI")
grid()
vioplot(split(n_data$body_fat, n_data$case), main = "Body fat",
        ylab = "Body fat")
grid()
vioplot(split(n_data$visc_fat, n_data$case), main = "Visceral fat",
        ylab = "Visceral fat")
grid()

```

While the distribution of the BMI is balanced between female and male
participants, females have on average a much higher body fat percentage, while
males have on average a higher proportion of visceral fat. For cases and
controls (defined for the recruitment of CHRIS participants) on average higher
BMI, body fat and visceral fat can be observed for *cases*.

Note that there seems to be no dependency between age and BMI in the present
data set (data not shown).

We next evaluate the clinical examinations performed on all participants.

```{r}
#' Clinical examinations
tmp <- nafldData("clinical")
rownames(tmp) <- tmp$AID
```

We will use and evaluate the following parameters defined by a clinical
examination:

- *naab11*: liver stiffness (higher values are worse). Values for participants
  that have a *naab12* (liver stiffness IQR) > 30 should be either removed or
  taken with care.
- *naab12*: liver stiffness IQR: the IQR of repeated liver stiffness
  measurements. Values below 30 are considered OK.
- *naab13*: liver fat. Can be binarized at an absolute value of 288.
- *naab14*: subjective, visual evaluation of the liver fat. Categorical from 1
  to 3. Values 2 and 3 can be considered steatotic.
- *nascfib4*: fibrosis 4 index. Linear score calculated from standard clinical
  parameters.
- *nascmets*: metabolic syndrome, binary category.

```{r}
n_data$liver_stiffness <- tmp[n_data$AID, "naab11"]
n_data$liver_stiffness_iqr <- tmp[n_data$AID, "naab12"]
n_data$liver_fat <- tmp[n_data$AID, "naab13"]
n_data$visual_fat <- tmp[n_data$AID, "naab14"]
n_data$fib4 <- tmp[n_data$AID, "nascfib4"]
n_data$metsym <- tmp[n_data$AID, "nascmets"]
```

From the `r nrow(n_data)` participants,
`r sum(n_data$liver_stiffness_iqr >= 30, na.rm = TRUE)`
have a liver stiffness IQR `>= 30`.

At first we evaluate the distribution of the liver stiffness score in the data
set.

```{r, fig.cap = "Distribution of liver stiffness measurements. All values considered. Problematic values are highlighted in red in the lower panel.", fig.height = 10}
stiff_ok <- which(n_data$liver_stiffness_iqr < 30)
par(mfrow = c(2, 2))
vioplot(split(log2(n_data$liver_stiffness), n_data$na_sex),
        main = "Liver stiffness", ylab = expression(log[2](Liver~stiffness)))
grid()
vioplot(split(log2(n_data$liver_stiffness), n_data$case),
        main = "Liver stiffness", ylab = expression(log[2](Liver~stiffness)))
grid()
col <- rep("#ff0000", nrow(n_data))
col[stiff_ok] <- "#000000"
x <- n_data$na_ager
y <- log2(n_data$liver_stiffness)
plot(x, y, pch = 21, bg = paste0(col, "40"),
     col = paste0(col, "80"), main = "Liver stiffness",
     xlab = "age", ylab = expression(log[2](Liver~stiffness)))
grid()
abline(lm(y ~ x), lty = 1, col = "#00000080")
x <- n_data$BMI
plot(x, y, pch = 21, bg = paste0(col, "40"),
     col = paste0(col, "80"), main = "Liver stiffness",
     xlab = "BMI", ylab = expression(log[2](Liver~stiffness)))
grid()
abline(lm(y ~ x), lty = 1, col = "#00000080")
```

While there is a clear average difference for liver stiffness between male and
female as well as case and control cases, there seems to be no general age
dependency of the liver stiffness. Only for a subset of participants there seems
to be a (log linear) increase with age. In contrast, a (log-linear) relationship
between BMI and liver stiffness can be observed.

We next evaluate the distribution of liver fat values in the study cohort.

```{r, fig.cap = "Distribution of liver fat values in the study cohort.", fig.height = 10}
par(mfrow = c(2, 2))
vioplot(split(n_data$liver_fat, n_data$na_sex),
        main = "Liver fat", ylab = "Liver fat")
grid()
vioplot(split(n_data$liver_fat, n_data$case),
        main = "Liver fat", ylab = "Liver fat")
grid()
x <- n_data$na_ager
y <- n_data$liver_fat
plot(x, y, pch = 21, main = "Liver fat", xlab = "age", ylab = "Liver fat",
     bg = "#00000040", col = "#00000080")
grid()
abline(lm(y ~ x), lty = 1, col = "#00000080")
x <- n_data$BMI
plot(x, y, pch = 21, main = "Liver fat", xlab = "BMI", ylab = "Liver fat",
     bg = "#00000040", col = "#00000080")
grid()
abline(lm(y ~ x), lty = 1, col = "#00000080")
```

Similar to the liver stiffness, we see differences in average liver fat values
between male and female study participants as well as between cases and
controls, while there seems no age dependency. A strong relationship between BMI
and liver fat can be observed.

For completeness we also plot the visual fat categorization.

```{r, fig.cap = "Overview of visual fat categories.", fig.height = 10}
par(mfrow = c(2, 2))
split(n_data$visual_fat, n_data$na_sex) |>
    lapply(table) |>
    do.call(what = cbind) |>
    barplot()
split(n_data$visual_fat, n_data$case) |>
    lapply(table) |>
    do.call(what = cbind) |>
    barplot()
vioplot(split(n_data$na_ager, n_data$visual_fat), main = "Visual fat",
        ylab = "age")
grid()
vioplot(split(n_data$BMI, n_data$visual_fat), main = "Visual fat",
        ylab = "BMI")
grid()
```

A clear relationship can be seen for the visual fat evaluation with the cases as
well as BMI.

Next we evaluate the FIB4 score.

```{r, fig.cap = "Distribution of FIB4 scores.", fig.height = 10}
par(mfrow = c(2, 2))
vioplot(split(log2(n_data$fib4), n_data$na_sex),
        main = "FIB4", ylab = expression(log[2](FIB4)))
grid()
vioplot(split(log2(n_data$fib4), n_data$case),
        main = "FIB4", ylab = expression(log[2](FIB4)))
grid()
x <- n_data$na_ager
y <- log2(n_data$fib4)
plot(x, y, pch = 21, main = "FIB4", xlab = "age",
     ylab = expression(log[2](FIB4)),
     bg = "#00000040", col = "#00000080")
grid()
abline(lm(y ~ x), lty = 1, col = "#00000080")
x <- n_data$BMI
plot(x, y, pch = 21, main = "FIB4", xlab = "BMI",
     ylab = expression(log[2](FIB4)),
     bg = "#00000040", col = "#00000080")
grid()
abline(lm(y ~ x), lty = 1, col = "#00000080")
```

There seems to be only a small difference in average FIB4 values between male
and female participants or between cases and controls. A strong (positive)
relationship is however visible between FIB4 and age. A slight negative
relationship seems to be present between FIB4 and BMI.

```{r, fig.cap = "Overview of metabolic syndrome categories.", fig.height = 10}
par(mfrow = c(2, 2))
split(n_data$metsym, n_data$na_sex) |>
    lapply(table) |>
    do.call(what = cbind) |>
    barplot()
split(n_data$metsym, n_data$case) |>
    lapply(table) |>
    do.call(what = cbind) |>
    barplot()
vioplot(split(n_data$na_ager, n_data$metsym), main = "Metabolic syndrome",
        ylab = "age")
grid()
vioplot(split(n_data$BMI, n_data$metsym), main = "Metabolic syndrome",
        ylab = "BMI")
grid()
```

While no cases of metabolic syndrome seem to be balanced between male and female
participants and seem to be independent of the participants' ages, a strong
difference can be seen for cases and controls as well as for BMI.


# Classifying NAFLD/MASLD and diabetes

In this section we define different clinical phenotypes for study participants
based on the data collected in the study.


## Type 2 Diabetes

Type 2 Diabetes (T2B) is diagnosed for study participants using the information
and data from the study according to [@american_diabetes_association_2_2018]
[@resl_guidelines_2023]. The criteria are:

- HbA1C >= 6.5%
- OR Fasting glucose >= 126 mg/dl
- OR Anti-diabetic medication

We are loading in addition CHRIS baseline data to enable comparison of these
values in the baseline and the NAFLD study collection.

```{r, message = FALSE}
library(chrisData)
chris_med <- chrisData("drugs_summary")
rownames(chris_med) <- chris_med$AID
chris_med <- chris_med[n_data$aid_chris, ]

chris_gi <- chrisData("general")
rownames(chris_gi) <- chris_gi$AID
chris_gi <- chris_gi[n_data$aid_chris, ]

chris_int <- chrisData("interview")
rownames(chris_int) <- chris_int$AID
chris_int <- chris_int[n_data$aid_chris, ]

chris_lab <- chrisData("labdata")
rownames(chris_lab) <- chris_lab$AID
chris_lab <- chris_lab[n_data$aid_chris, ]

n_data$chris_dm <- chris_int$x0dm00 # self-reported diabetes diagnosis
n_data$chris_HbA1c <- chris_lab$x0lp06a # Glycated Hemoglobin
n_data$chris_glucose <- chris_lab$x0lp05 # fasting glucose
n_data$chris_dia_med <- chris_med$x0dd21 # diabetes medication
n_data$chris_sr_fasting <- chris_lab$x0bc12 # self-reported fasting
```

We next extract the same data from the NAFLD collection. We also update the
self-reported diabetes diagnosis considering all participants that reported
diabetes at baseline OR at NAFLD follow-up. Finally, we extract the information
on self-reported fasting and no drinking.

```{r}
med <- nafldData("Drug")
rownames(med) <- med$AID
med <- med[n_data$AID, ]

int <- nafldData("interview")
rownames(int) <- int$AID
int <- int[n_data$AID, ]

lab <- nafldData("Labdata")
rownames(lab) <- lab$AID
lab <- lab[n_data$AID, ]

#' self-reported diabetes considering also baseline
n_data$self_dm <- n_data$chris_dm
n_data$self_dm[which(int$nadm00 == "Yes")] <- "Yes"
n_data$HbA1c <- lab$nalp06a
n_data$glucose <- lab$nalp05
n_data$dia_med <- med$nadd21

#' self-reported fasting.
n_data$sr_fasting <- lab$nabc12
n_data$sr_no_drinking <- lab$nabc13
```

Glucose levels can be affected by the (missing) fasting status. Thus we next
evaluate glucose levels in participants with or without self-reported fasting
(and no drinking).

```{r, fig.cap = "Glucose concentrations for self-reported fasting (or no drinking) status.", fig.width = 12, fig.height = 5}
library(beeswarm)

par(mfrow = c(1, 2))
beeswarm(split(n_data$glucose, n_data$sr_fasting), pch = 21,
         bg = "#00000020", main = "Self-reported fasting",
         ylab = "Glucose")
grid()
abline(h = 126)
beeswarm(split(n_data$glucose, n_data$sr_no_drinking), pch = 21,
         bg = "#00000020", main = "Self-reported no drinking",
         ylab = "Glucose")
grid()
abline(h = 126)
```

Drinking did not affect the glucose levels, while for some participant that did
not fast the glucose levels are higher than the cut-off of 126 used in diabetes
diagnosis. Thus, we below only consider glucose levels for participants that did
not eat (self-reported) prior to the blood drawing.

And we are next diagnosing the NAFLD study participants based on the clinical
and medication information (ignoring the self-reported diagnosis).

```{r}
#' Diagnosing in NAFLD
t2d <- rep(FALSE, nrow(n_data))

#' Anti-diabetic medication: nadd21
t2d[which(n_data$dia_med == "Yes")] <- TRUE

#' Glycated hemoglobin %: nalp06a
t2d[which(n_data$HbA1c >= 6.5)] <- TRUE

#' fasting glucose: nalp05 + fasting
t2d[which(n_data$glucose >= 126 & n_data$sr_fasting == "Yes")] <- TRUE

#' finally define those without clear information.
no_lab <- is.na(n_data$HbA1c)
t2d[is.na(n_data$dia_med) & no_lab] <- NA

n_data$t2d <- t2d
```

The same criteria are applied to CHRIS baseline samples.

```{r}
#' Diagnosing in CHRIS
t2d <- rep(FALSE, nrow(n_data))

#' Anti-diabetic medication: nadd21
t2d[which(n_data$chris_dia_med == "Yes")] <- TRUE

#' Glycated hemoglobin %: nalp06a
t2d[which(n_data$chris_HbA1c >= 6.5)] <- TRUE

#' fasting glucose: nalp05
t2d[which(n_data$chris_glucose >= 126 &
          n_data$chris_sr_fasting == "Yes")] <- TRUE

n_data$chris_t2d <- t2d
```

Note that for two participants no glycated hemoglobin measurement was
available: one of them was (in CHRIS baseline) diabetic and one was selected as
control (i.e., not diabetic). According to the available medication information
in the NAFLD substudy these two participants were still diagnosed the same
(i.e. one as diabetic and one not).

```{r, results = "asis"}
tmp <- data.frame(
    count = as.integer(table(n_data$t2d)),
    females = vapply(split(n_data$na_sex, n_data$t2d),
                     function(x) sum(x == "Female"), integer(1)),
    age = vapply(split(n_data$na_ager, n_data$t2d), mean, numeric(1)),
    BMI = vapply(split(n_data$BMI, n_data$t2d), mean, numeric(1), na.rm = TRUE)
)
library(pander)
pandoc.table(
    tmp, style = "rmarkdown",
    caption = paste0("Overview of NAFLD participants with and without type 2",
                     " diabetes (diagnosed based on laboratory parameters ",
                     "and medication)."))

```

Below we compare the total number of diagnosed T2D in baseline and NAFLD as well
as the self-reported diabetes status.

```{r, results = "asis"}
data.frame(status = c("No", "Yes"),
           NAFLD_T2D = as.integer(table(n_data$t2d)),
           CHRIS_T2D = as.integer(table(n_data$chris_t2d)),
           NAFLD_SR_DM = as.integer(table(n_data$self_dm)),
           CHRIS_SR_DM = as.integer(table(n_data$chris_dm))) |>
    pandoc.table(style = "rmarkdown",
                 caption = paste0("Diabetes cases in CHRIS baseline and ",
                                  "NAFLD, diagnosed and self-reported (SR)."))
```

Surprisingly, the number of cases are higher in CHRIS baseline than in the NAFLD
studies, both diagnosed and self-reported. The higher proportion of T2D
diagnosed participants in CHRIS baseline seems to be mostly driven by glycated
hemoglobin measurements (see tables below). Also interestingly, self-reported
diabetes is much higher than diagnosed (type 2) diabetes in NAFLD.

```{r, results = "asis"}
data.frame(status = c("No", "Yes"),
           medication = as.integer(table(n_data$dia_med == "Yes")),
           HbA1c = as.integer(table(n_data$HbA1c >= 6.5)),
           glucose = as.integer(table(n_data$glucose >= 126 &
                                      n_data$sr_fasting == "Yes"))) |>
    pandoc.table(style = "rmarkdown",
                 caption = paste0("Diagnostic criteria in NAFLD."))
```

```{r, results = "asis"}
data.frame(status = c("No", "Yes"),
           medication = as.integer(table(n_data$chris_dia_med == "Yes")),
           HbA1c = as.integer(table(n_data$chris_HbA1c >= 6.5)),
           glucose = as.integer(table(n_data$chris_glucose >= 126 &
                                      n_data$chris_sr_fasting == "Yes"))) |>
    pandoc.table(style = "rmarkdown",
                 caption = paste0("Diagnostic criteria in CHRIS."))
```

Below we evaluate the data for participants for which the diagnosed T2B differs
between baseline and NAFLD collection.

```{r, results = "asis"}
sel <- n_data$t2d != n_data$chris_t2d
tmp <- n_data[sel, ]
tmp[, c("chris_dm", "self_dm", "chris_t2d", "t2d", "chris_HbA1c", "HbA1c",
        "chris_glucose", "glucose")] |>
    pandoc.table(style = "rmarkdown", split.table = Inf)
```

All these participants were diagnosed with T2D based on CHRIS baseline data, but
not with data from the NAFLD collection. For all these participants the percent
glycated hemoglobin values are below the cut-off of 6.5% in NAFLD while they do
have higher levels in CHRIS baseline. Note that all HbA1c measurements in
baseline, for the NAFLD participants were done on a Tosoh Automated
Glycohemoglobin Analyzer HLC-723 G8 (TOSOH CORPORATION SHIBA-KOEN FIRST BLDG)
while the instrument was changed on 2015-05-15 to a ARKRAY ADAMS A1c HA-8180V
instrument (hence all measurements in the NAFLD study were performed on that
instrument).

Below we compare the HbA1c measurements from the two instruments to evaluate
whether there was an (apparent) systematic shift.

```{r, fig.cap = "HbA1c concentrations (%) quantified using the old (FALSE) and new (TRUE) instrument."}
tmp <- chrisData("lab")
rownames(tmp) <- tmp$AID
tmp_gi <- chrisData("general")
rownames(tmp_gi) <- tmp_gi$AID
tmp_gi <- tmp_gi[rownames(tmp), ]
hba1c <- tmp$x0lp06a
inst_cat <- tmp_gi$x0_examd > as.Date("2015-05-15")

vioplot(split(hba1c, inst_cat), ylab = "HbA1c [%]",
        main = "Glycohemoglobin before and after 2015-05-15")
grid()
abline(h = 6.5)
```

The average percentage of HbA1c quantified on the *old* instrument was higher
than on the new instrument.

The number of participants (in CHRIS baseline) with an HbA1c concentration
higher than 6.5% per instrument are shown below.

```{r, results = "asis"}
after <- hba1c[inst_cat]
before <- hba1c[!inst_cat]

#' Test mean scaling...
#' before <- before * mean(after, na.rm = TRUE) / mean(before, na.rm = TRUE)

data.frame(instrument = c("A", "B"),
           total_count = c(length(before), length(after)),
           higher_6.5 = c(sum(before > 6.4, na.rm = TRUE),
                          sum(after > 6.4, na.rm = TRUE))) |>
    pandoc.table(
        style = "rmarkdown",
        caption = paste0("Number of measurements performed with the old (A), ",
                         "or new (B) instrument and number of participants ",
                         "with a HbA1c > 6.5%."))

vioplot(list(before, after))
grid()
```

Indeed, it looks like we might have a problem here! About twice as many
participants have elevated HbA1c levels on the old instrument.

- [X] DONE: check instruments between baseline and NAFLD study - was there a
      change that could explain differences in HbA1c?
- [ ] TODO [Essi]: check if there was difference in lifestyle or nutrition
      between baseline and NAFLD.
- [ ] TODO: reach out to Martin, what is his opinion on that?
- [ ] TODO: discuss with Christoph, what is his opinion on that?
- [ ] TODO: how to normalize the values.

We next evaluate distribution of age, BMI, sex etc for participants with and
without T2D.

```{r, fig.cap = "Age distribution for female and male study participants with and without T2D."}
f <- paste0(n_data$na_sex, n_data$t2d)
vioplot(split(n_data$na_ager, f), main = "Age/T2D", ylab = "age [years]")
grid()
```

Age distribution is similar for female or male study participants with or
without T2D. We next evaluate the distribution of BMI values for T2D and control
(again stratified by sex).

```{r, fig.cap = "Distribution of BMI values for female and male study participants with and without T2D."}
vioplot(split(n_data$BMI, f), main = "BMI/T2D", ylab = "BMI")
grid()
```

Participants with T2D have a consistently higher BMI, both in female and male
participants. We next evaluate the liver stiffness measurements (removing
measurements that don't fulfill the quality requirements).

```{r, fig.cap = "Distribution of liver stiffness values for female and male study participants with and without T2D."}
lstiff <- n_data$liver_stiffness
lstiff[n_data$liver_stiffness_iqr > 30] <- NA
vioplot(split(log2(lstiff), f), main = "liver stiffness/T2D",
        ylab = expression(log[2]~liver~stiffness))
grid()
```

Participants with T2D have slightly higher liver stiffness measurements than
those without. This might however be a consequence of their higher BMI.

```{r, fig.cap = "Distribution of visceral fat values for female and male study participants with and without T2D."}
vioplot(split(n_data$visc_fat, f), main = "Visceral fat/T2D",
        ylab = expression(visceral~fat))
grid()
```

Again, participants with T2D have on average higher visceral fat.

```{r, fig.cap = "Distribution of FIB4 values for female and male study participants with and without T2D."}
vioplot(split(n_data$fib4, f), main = "FIB4/T2D",
        ylab = expression(FIB4))
grid()
```

FIB4 values are higher for participants without T2D.


# Classification of steatosis in MASLD/NAFLD

Clinical diagnosis of metabolic dysfunction associated steatotic liver disease
(MASLD [@rinella_multisociety_2024], formerly NAFLD
[@european_association_for_the_study_of_the_liver_easl_easl-easd-easo_2016])
is based on a range of parameters:

1) Presence of steatotic liver disease: sonographic evaluation, steatose score
   of 2 - 3.

2) Presence of cardiometabolic risk factors/criteria (at least one out of 5):
   a) BMI >= 25 OR waist circumference > 94 (M) 80 (F) cm.
   b) Fasting glucose >= 100 mg/dl OR HbA1c >= 5.7% OR type 2 diabetes OR
      treatment for type 2 diabetes.
   c) Blood pressure >= 130/85 mmHg OR specific antihypertensive drug treatment.
   d) Triglycerides >= 150 mg/dl OR lipid lowering treatment.
   e) HDL <= 40 (M) <= 50 (F) mg/dl OR lipid lowering treatment.

3) Rule out alcoholic liver disease. Alcohol consumption > 40g/day (one
   *standard drink* is 10g alcohol).

This diagnosis includes also metabolic syndrome (MetS) which is diagnosed based
on the presence of 3 risk factors:

- (must have): central obesity (WC >= 94 (M) >= 80 (F) cm.
- (two out of four):
  - TG >= 150 mg/dl OR specific treatment for this lipid abnormality (**d**).
  - HDL < 40 (M) 50 (F) mg/dl OR specific treatment for this lipid abnormality
    (**e**).
  - systolic BP >= 130 or diastolic >= 85 mmHG OR treatment for hypertension
    (**c**).
  - fasting glucose >= 100 mg/dl OR diagnosed T2D (**b**).

Below we perform the diagnosis based on the above criteria.

Presence of steatotic liver disease.

```{r}
#' steatose score: naab14, visual scoring of liver fat.
n_data$SLD <- n_data$visual_fat > 1
summary(n_data$SLD)
```

**a)** BMI >= 25 or waist circumference > 94 (M) 80 (F) cm.

```{r}
#' Waist circumference
n_data$waist_circumference <- n_anthro[n_data$AID, "naan12"]
#' Classify
n_data$cardiometabolic_a <-
    n_data$BMI >= 25 |
    (n_data$waist_circumference >= 94 & n_data$na_sex == "Male") |
    (n_data$waist_circumference >= 80 & n_data$na_sex == "Female")
```

A quite large proportion of participants fulfill this criteria - also maybe not
unexpected based on the selection criteria for study participants.

**b)** Fasting glucose >= 100 mg/dl OR HbA1c >= 5.7% OR type 2 diabetes OR
treatment for type 2 diabetes (already included in the diagnosis of T2D
performed in the previous section).

```{r}
#' Classify
n_data$cardiometabolic_b <-
    n_data$glucose >= 100 |
    n_data$HbA1c > 5.7 |
    n_data$t2d
```

**c)** Blood pressure >= 130/85 mmHg OR specific antihypertensive drug
treatment.

```{r}
#' Add blood pressure measurements.
n_data$systolic_bp <- n_anthro[n_data$AID, "nabp01"] # systolic BP [mm/Hg]
n_data$diastolic_bp <- n_anthro[n_data$AID, "nabp02"] # diastolic BP [mm/Hg]
n_data$systolic_bp[n_data$systolic_bp < 0] <- NA_real_
n_data$diastolic_bp[n_data$diastolic_bp < 0] <- NA_real_
#' Add blood pressure lowering therapy
n_data$pb_med <- med$nadd20 # blood pressure lowering agents
#' Classify
n_data$cardiometabolic_c <-
    n_data$systolic_bp >= 130 |
    n_data$diastolic_bp >= 85 |
    n_data$pb_med == "Yes"
```

**d)** Triglycerides >= 150 mg/dl OR lipid lowering treatment.

```{r}
#' triglycerides [mg/dl]
n_data$triglycerides <- lab[n_data$AID, "nalp19"]
n_data$lm_med <- med$nadd22 # lipid modifying agents
n_data$ms_med <- med$nadd23 # metabolic syndrome medication
#' Classify
n_data$cardiometabolic_d <-
    n_data$triglycerides >= 150 |
    n_data$lm_med == "Yes"
```

**e)** HDL <= 40 (M) <= 50 (F) mg/dl OR lipid lowering treatment.

```{r}
#' HDL [mg/dl]: nalp17
n_data$HDL <- lab[n_data$AID, "nalp17"]
#' Classify
n_data$cardiometabolic_e <-
    (n_data$HDL <= 40 & n_data$na_sex == "Male") |
    (n_data$HDL <= 50 & n_data$na_sex == "Female") |
    n_data$lm_med == "Yes"
```

Summary of the numbers of numbers of cases for each sub-criteria is provided
below.

```{r, results = "asis"}
cardiom <- n_data[, c("cardiometabolic_a", "cardiometabolic_b",
                      "cardiometabolic_c", "cardiometabolic_d",
                      "cardiometabolic_e")]
lapply(cardiom, summary) |>
    do.call(what = rbind) |>
    pandoc.table(
        style = "rmarkdown",
        caption = paste0("Number of cases for each cardiometabolic sub-criteria"))

```

Most participants scored positive in subcategories **c** (hypertension) and
**a** (BMI and waist circumference). We visually inspect the overlap between the
sub-criteria also with an upset plot.

```{r, caption = "Overview and overlap of participants scoring positive in the individual cardiometabolic sub-criteria."}
library(UpSetR)
cardiom <- as.data.frame(lapply(cardiom, as.integer))
upset(cardiom, order.by = "degree")
```

A high number of participants score positive in all sub-criteria.

At last we combine the scores from the sub-criteria to diagnose MASLD based on
the criteria 1) and 2) from the definition. Criteria 3), related to alcohol
consumption will be evaluated later.

1) Presence of steatotic liver disease: sonographic evaluation, steatose score
   of 2 - 3.

2) Presence of cardiometabolic risk factors/criteria (at least one out of 5):
   a) BMI >= 25 OR waist circumference > 94 (M) 80 (F) cm.
   b) Fasting glucose >= 100 mg/dl OR HbA1c >= 5.7% OR type 2 diabetes OR
      treatment for type 2 diabetes.
   c) Blood pressure >= 130/85 mmHg OR specific antihypertensive drug treatment.
   d) Triglycerides >= 150 mg/dl OR lipid lowering treatment.
   e) HDL <= 40 (M) <= 50 (F) mg/dl OR lipid lowering treatment.

```{r}
#' MASLD:
#' - steatotic liver disease
#' - one out of 5 of a), b), c), d) and e)
n_data$MASLD <-
    n_data$SLD &
    rowSums(n_data[, c("cardiometabolic_a", "cardiometabolic_b",
                       "cardiometabolic_c", "cardiometabolic_d",
                       "cardiometabolic_e")], na.rm =TRUE) > 0
summary(n_data$MASLD)
```

The MASLD positive/negative cases are thus almost balanced in the present
study. Also, maybe interesting, there were in total `r sum(n_data$SLD)`
individuals scored positive for steatotic liver disease, but 2 of them did not
fulfill the additional criteria for MASLD.

Again, we next evaluate distribution of age, BMI, sex etc for participants with
and without MASLD.

```{r, fig.cap = "Age distribution for female and male study participants with and without MASLD."}
f <- paste0(n_data$na_sex, n_data$MASLD)
vioplot(split(n_data$na_ager, f), main = "Age/MASLD", ylab = "age [years]")
grid()
```

Age distribution is about similar for female participants. For males, MASLD
negative have an on average slightly higher age. We next evaluate the
distribution of BMI values for MASLD positive/negative cases (again stratified
by sex).

```{r, fig.cap = "Distribution of BMI values for female and male study participants with and without MASLD."}
vioplot(split(n_data$BMI, f), main = "BMI/MASLD", ylab = "BMI")
grid()
```

Participants with MASLD have a consistently higher BMI, both in female and male
participants. We next evaluate the liver stiffness measurements (removing
measurements that don't fulfill the quality requirements).

```{r, fig.cap = "Distribution of liver stiffness values for female and male study participants with and without MASLD."}
lstiff <- n_data$liver_stiffness
lstiff[n_data$liver_stiffness_iqr > 30] <- NA
vioplot(split(log2(lstiff), f), main = "liver stiffness/MASLD",
        ylab = expression(log[2]~liver~stiffness))
grid()
```

Participants with MASLD have slightly higher liver stiffness measurements than
those without. This might however be a consequence of their higher BMI.

```{r, fig.cap = "Distribution of visceral fat values for female and male study participants with and without MASLD."}
vioplot(split(n_data$visc_fat, f), main = "Visceral fat/MASLD",
        ylab = expression(visceral~fat))
grid()
```

Again, participants with MASLD have on average higher visceral fat.

```{r, fig.cap = "Distribution of FIB4 values for female and male study participants with and without MASLD."}
vioplot(split(n_data$fib4, f), main = "FIB4/MASLD",
        ylab = expression(FIB4))
grid()
```

FIB4 values are higher for participants without MASLD.

At last we evaluate also the T2D cases with and without MASLD.

```{r}
table(n_data[, c("t2d", "MASLD")])
```

A higher proportion of participants with T2D score in addition also positive for
MASLD.


# Classification of Metabolic Syndrome

Metabolic syndrome (MetS) is diagnosed based on the presence of 3 risk factors
(and partially overlaps with the MASLD definition above):

- (must have): central obesity (WC >= 94 (M) >= 80 (F) cm.
- (two out of four):
  - TG >= 150 mg/dl OR specific treatment for this lipid abnormality (**d**).
  - HDL < 40 (M) 50 (F) mg/dl OR specific treatment for this lipid abnormality
    (**e**).
  - systolic BP >= 130 or diastolic >= 85 mmHG OR treatment for hypertension
    (**c**).
  - fasting glucose >= 100 mg/dl OR diagnosed T2D (**b**).

```{r}
n_data$MetS <-
    ((n_data$waist_circumference >= 94 & n_data$na_sex == "Male") |
     (n_data$waist_circumference >= 80 & n_data$na_sex == "Female")) &
    rowSums(n_data[, c("cardiometabolic_b", "cardiometabolic_c",
                       "cardiometabolic_d", "cardiometabolic_e")],
            na.rm =TRUE) > 1
```

The overlap between participants with metabolic syndrome and MASLD are shown
below.

```{r}
table(n_data[, c("MetS", "MASLD")])
```

For the biggest part, individuals with MetS are also positive for MASLD. On the
other side, also a quite large proportion of individuals are classified to have
MASLD but are not positive for MetS. Most T2D positive cases score also positive
for MetS (see below).

```{r}
table(n_data[, c("t2d", "MetS")])
```

Again, we next evaluate distribution of age, BMI, sex etc for participants with
and without MetS.

```{r, fig.cap = "Age distribution for female and male study participants with and without MetS."}
f <- paste0(n_data$na_sex, n_data$MetS)
vioplot(split(n_data$na_ager, f), main = "Age/MetS", ylab = "age [years]")
grid()
```

For females the average age of MetS positive cases is higher while for male
participant the opposite effect can be observed. We next evaluate the
distribution of BMI values for MetS and control (again stratified by sex).

```{r, fig.cap = "Distribution of BMI values for female and male study participants with and without MetS."}
vioplot(split(n_data$BMI, f), main = "BMI/MetS", ylab = "BMI")
grid()
```

Participants with MetS have a consistently higher BMI, both in female and male
participants. We next evaluate the liver stiffness measurements (removing
measurements that don't fulfill the quality requirements).

```{r, fig.cap = "Distribution of liver stiffness values for female and male study participants with and without MetS."}
lstiff <- n_data$liver_stiffness
lstiff[n_data$liver_stiffness_iqr > 30] <- NA
vioplot(split(log2(lstiff), f), main = "liver stiffness/MetS",
        ylab = expression(log[2]~liver~stiffness))
grid()
```

Participants with MetS have slightly higher liver stiffness measurements than
those without. This might however be a consequence of their higher BMI.

```{r, fig.cap = "Distribution of visceral fat values for female and male study participants with and without MetS."}
tmp <- split(n_data$visc_fat, f)
tmp <- lapply(tmp, function(z) z[!is.na(z)])
tmp <- tmp[lengths(tmp) > 0]
vioplot(tmp, main = "Visceral fat/MetS",
        ylab = expression(visceral~fat))
grid()
```

Again, participants with MetS have on average higher visceral fat.

```{r, fig.cap = "Distribution of FIB4 values for female and male study participants with and without MetS."}
vioplot(split(n_data$fib4, f), main = "FIB4/MetS",
        ylab = expression(FIB4))
grid()
```

FIB4 values are higher for participants without MetS.



# NAFLD omics data overview


Get the IDs of NAFLD participants and their respective CHRIS AID.

```{r, message = FALSE}
library(chrisData)

gi <- chrisData("general")

nafld <- gi[which(gi$x0_nafld == "Yes"), ]
```

Participation year (in CHRIS baseline) for the `r nrow(nafld)` NAFLD study
participants.

```{r}
barplot(table(nafld$x0_examy))
```

We determine the number of participants with available CHRIS targeted
metabolomics as well as CHRIS ScanningSWATH and Somalogic proteomics data from
the baseline and NAFLD collections.

```{r, message = FALSE}
library(tidyfr)
soma <- data_module(name = "chris_somalogic", version = "1.0.3.2", path = ".")
scan <- data_module(name = "chris_scanningswath_proteins", version = "1.0.0.2",
                    path = ".")
p180 <- data_module(name = "metabolomics_p180", version = "1.0.1.2", path = ".")

soma_aid <- rownames(data(soma))
scan_aid <- rownames(data(scan))
p180_aid <- rownames(data(p180))
```

```{r, results = "asis"}
library(pander)
tmp <- data.frame(baseline = c(sum(nafld$AID %in% p180_aid),
                               sum(nafld$AID %in% scan_aid),
                               sum(nafld$AID %in% soma_aid)),
                  NAFLD = c(sum(nafld$aid_nafld %in% p180_aid),
                            sum(nafld$aid_nafld %in% scan_aid),
                            sum(nafld$aid_nafld %in% soma_aid)))
rownames(tmp) <- c("targeted metabolomics",
                   "ScanningSWATH proteomics",
                   "SomaLogic proteomics")
pandoc.table(tmp, style = "rmarkdown",
             caption = paste0("Number of NAFLD study participants (n = ",
                              nrow(nafld), ") with available omics data"))
```

Note: these numbers don't include any of the drop outs up to version 2 (July
2023).


```{r}
do <- read.table(system.file("txt", "drop-outs.txt", package = "chrisData"),
                 sep = "\t", header = TRUE, colClasses = "character")
any(do$aid %in% nafld$AID)
any(do$aid %in% nafld$aid_nafld)
```


## Metabolomics data

We next evaluate metabolite concentrations in baseline and NAFLD. We first
calculate correlations for metabolites (for each metabolite correlating
measurements in baseline to those from NAFLD).

```{r}
p180_data <- data(p180)
p180_labels <- labels(p180)
metabolites <- p180_labels$description
metabolites <- metabolites[-grep("assay flags", metabolites)]
colnames(p180_data) <- p180_labels$description

a <- as.matrix(p180_data[n_data$aid_chris, metabolites])
b <- as.matrix(p180_data[n_data$AID, metabolites])

cors <- vapply(seq_len(ncol(a)), function(i) {
    cor(a[, i], b[, i], use = "pairwise.complete.obs",
        method = "spearman")
}, numeric(1))
```

The table below lists the correlation coefficients (between values from baseline
and NAFLD) for all metabolites. Metabolites are ordered by magnitude of the
correlation coefficient.

```{r}
tmp <- data.frame(p180_labels[match(metabolites, p180_labels$description),
                              c("analyte_name", "analyte_class")],
                  rho = cors)
pandoc.table(tmp[order(tmp$rho, decreasing = TRUE), ], style = "rmarkdown")
```

Some metabolites yield consistent values while others show much lower
*reproducibility*.

```{r, fig.cap = "Correlation coefficients for metabolites (ordered by magnitude)."}
plot(sort(tmp$rho), pch = 21, col = "#00000080", bg = "#00000040",
     ylab = "rho", xlab = "rank", main = "Metabolites")
grid()
```

We plot next the data for the best and worst correlating metabolite.

```{r, fig.cap = "Correlations between metabolite abundances in CHRIS baseline and NAFLD for the metabolite with the highest and lowest correlation. Correlations calculated for each metabolite on pairs of measurements for the same participant."}
par(mfrow = c(1, 2))

i <- which.max(tmp$rho)
plot(log2(a[, i]), log2(b[, i]),
     main = colnames(a)[i],
     xlab = expression(log[2]~baseline),
     ylab = expression(log[2]~NAFLD), pch = 21,
     col = "#00000080", bg = "#00000040")
grid()
abline(0, 1)

i <- which.min(abs(tmp$rho))
plot(log2(a[, i]), log2(b[, i]),
     main = colnames(a)[i],
     xlab = expression(log[2]~baseline),
     ylab = expression(log[2]~NAFLD), pch = 21,
     col = "#00000080", bg = "#00000040")
grid()
abline(0, 1)
```

Metabolite C14:2-OH showed also previously differences between measurements for
the first 5000 and second 2000 participants.

Next we calculate correlations for participants (correlate measurements for
each participant between baseline and NAFLD).

```{r}
cors <- vapply(seq_len(nrow(a)), function(i) {
    cor(a[i, ], b[i, ], use = "pairwise.complete.obs",
        method = "spearman")
}, numeric(1))

tmp <- data.frame(AID = rownames(a), rho = cors)
```

```{r, fig.cap = "Correlation coefficients for pairwise correlations of all metabolite measurements from baseline and NAFLD per participant."}
plot(sort(tmp$rho), pch = 21, col = "#00000080", bg = "#00000040",
     ylab = "rho", xlab = "rank", main = "Participants")
grid()
```

Correlation of values between baseline and NAFLD are highly correlated. As
examples we plot below the values for the best and worst correlating
participant.

```{r}
par(mfrow = c(1, 2))

i <- which.max(tmp$rho)
plot(log2(a[i, ]), log2(b[i, ]),
     xlab = expression(log[2]~baseline),
     ylab = expression(log[2]~NAFLD), pch = 21,
     col = "#00000080", bg = "#00000040")
grid()
abline(0, 1)

i <- which.min(tmp$rho)
plot(log2(a[i, ]), log2(b[i, ]),
     xlab = expression(log[2]~baseline),
     ylab = expression(log[2]~NAFLD), pch = 21,
     col = "#00000080", bg = "#00000040")
grid()
abline(0, 1)

```

Values for the first participant are nicely correlated. For the second,
correlation is present, but there seems to be a systematic shift in some values
(eventually e.g. values from FIA measurement).


## Proteomics data

We evaluate also correlations for the available proteomics data.

```{r}
swath_data <- data(scan)
swath_labels <- labels(scan)
proteins <- swath_labels$description
proteins <- proteins[swath_labels$Genes != ""]
colnames(swath_data) <- swath_labels$description

a <- as.matrix(swath_data[n_data$aid_chris, proteins])
b <- as.matrix(swath_data[n_data$AID, proteins])

cors <- vapply(seq_len(ncol(a)), function(i) {
    cor(a[, i], b[, i], use = "pairwise.complete.obs",
        method = "spearman")
}, numeric(1))
```

The table below lists the correlation coefficients (between values from baseline
and NAFLD) for all proteins. Proteins are ordered by magnitude of the
correlation coefficient.

```{r}
tmp <- data.frame(
    Genes = swath_labels[match(proteins, swath_labels$description),
                         c("Genes")],
    rho = cors)
rownames(tmp) <- proteins
pandoc.table(tmp[order(tmp$rho, decreasing = TRUE), ], style = "rmarkdown")
```

High correlations can be observed for many proteins.

```{r, fig.cap = "Correlation coefficients for proteins (ordered by magnitude)."}
plot(sort(tmp$rho), pch = 21, col = "#00000080", bg = "#00000040",
     ylab = "rho", xlab = "rank", main = "Proteins")
grid()
```

We plot next the data for the best and worst correlating protein.

```{r, fig.cap = "Correlations between protein abundances in CHRIS baseline and NAFLD for the metabolite with the highest and lowest correlation. Correlations calculated for each protein on pairs of measurements for the same participant."}
par(mfrow = c(1, 2))

i <- which.max(tmp$rho)
plot(log2(a[, i]), log2(b[, i]),
     main = tmp[i, "Genes"],
     xlab = expression(log[2]~baseline),
     ylab = expression(log[2]~NAFLD), pch = 21,
     col = "#00000080", bg = "#00000040")
grid()
abline(0, 1)

i <- which.min(abs(tmp$rho))
plot(log2(a[, i]), log2(b[, i]),
     main = tmp[i, "Genes"],
     xlab = expression(log[2]~baseline),
     ylab = expression(log[2]~NAFLD), pch = 21,
     col = "#00000080", bg = "#00000040")
grid()
abline(0, 1)
```

Next we calculate correlations for participants (correlate measurements for
each participant between baseline and NAFLD).

```{r}
cors <- vapply(seq_len(nrow(a)), function(i) {
    cor(a[i, ], b[i, ], use = "pairwise.complete.obs",
        method = "spearman")
}, numeric(1))

tmp <- data.frame(AID = rownames(a), rho = cors)
```

Note that for the in total `r nrow(tmp)` participants we have only for
`r sum(!is.na(tmp$rho))` measurements both in the baseline and NAFLD collection
(mostly because measurements are missing in baseline).

```{r, fig.cap = "Correlation coefficients for pairwise correlations of all protein measurements from baseline and NAFLD per participant."}
plot(sort(tmp$rho), pch = 21, col = "#00000080", bg = "#00000040",
     ylab = "rho", xlab = "rank", main = "Participants")
grid()
```

Correlation of values between baseline and NAFLD are highly correlated. As
examples we plot below the values for the best and worst correlating
participant.

```{r}
par(mfrow = c(1, 2))

i <- which.max(tmp$rho)
plot(log2(a[i, ]), log2(b[i, ]),
     xlab = expression(log[2]~baseline),
     ylab = expression(log[2]~NAFLD), pch = 21,
     col = "#00000080", bg = "#00000040")
grid()
abline(0, 1)

i <- which.min(tmp$rho)
plot(log2(a[i, ]), log2(b[i, ]),
     xlab = expression(log[2]~baseline),
     ylab = expression(log[2]~NAFLD), pch = 21,
     col = "#00000080", bg = "#00000040")
grid()
abline(0, 1)

```


# Session Information

```{r}
sessionInfo()
```

# References