---
title: "NAFLD_fibrosis_prot_lm"
author: Marilyn De Graeve, Johannes Rainer, Christoph Grander"
affiliation: "Eurac Research, Bolzano, Italy"
date: "2025-01-27"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
---

**Modified**: `r file.info("NAFLD_fibrosis_prot_lm.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r style, message = FALSE, echo = FALSE, warning = FALSE, results = "asis"}
library("BiocStyle")
library("kableExtra")
library("knitr")
library("rmarkdown")
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now, units = "secs")
      # return a character string to show the time
      paste("    Time for this code chunk to run:", round(res,
        2), "seconds")
    }
  }
}))
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7, time_it = TRUE)

```

# Introduction

During this workflow, statistical analyis (linear models) are performed on the
relatively quantified Scanning SWATH proteomics data (148 proteins) from the
metabolic dysfunction-associated steatotic liver disease (MASLD) cohort. MASLD,
previously known as NAFLD is the most chronic liver disease, affecting more than
30% of global population. The MAFLD cohort is a subset of approximatley 353 #
participants from the Cooperative Health Research in South Tirol (CHRIS) study,
which were requested to return for a second resampling and additional
phenotyping. The NAFLD cohort represent an unhealty group with various degrees
of steatotic (alcoholic) liver disease classifications. See
[nafld_data_overview.Rmd](nafld_data_overview.Rmd) (respectively the rendered
html report) for an overview of the data set and the definition of the
phenotype.

This document describes the statistical analysis using multiple linear models to
identify associations between proteins and metabolic dysfunction-associated
steatotic liver disease (MASLD). The definition of control and MASLD types are:

- *none*: participants negative for steatotic liver disease.
- *MASLD*: (= Metabolic Dysfunction Associated Steatotic Liver Disease)
  participants with SLD, positive for cardiometabolic criteria but low alcohol
  consumption (below 40g/day). This includes also 8 participants without
  information on alcohol consumption. These participants did not provide the
  GA2LEN FFQ (hence missing values for quantified alcohol consumption), but
  according to alcohol-related information in other questionnaire they are not
  heavy drinkers and are thus added to this category.

The MetALD and SLD types are not included in this analysis. FYI:

- *MetALD*: (= MASLD and increased alcohol intake) participants with SLD,
  positive for cardiometabolic criteria but daily alcohol consumption > 40g/day.
- *SLD*: (= steatotic liver disease) participants with a SLD, but negative for
  cardiometabolic criteria. This includes potentially cryptogenic SLD, alcoholic
  liver disease (ALD), DILI or monogenic diseases.

See the `nafld_data_overview 1.html` report for further
information regarding the SLD sub-classification.

Regarding fibrosis, i.e.  the formation of excess fibrous connective tissue in 
an organ or tissue in a reparative or reactive process. In this scope: liver 
fibrosis. The variables containing this parameter is:

- *nascfib4*: fibrosis 4 index. Linear score calculated from standard clinical 
  parameters.

This parameter will be used as the target for a biomarker search in *MASLD* 
participants.


# Setup

## Directories

```{r directories, echo = FALSE}
#' General settings
filename <- " NAFLD_fibrosis_prot_lm"

#' Path to save images to; remove if exists.
IMAGE_PATH <- paste0("images/", filename, "/")
dir.create(IMAGE_PATH, recursive = TRUE)

#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)

#' Path to the data
#DATA_PATH <- '/home/mdegraeve/Documents/Files/Work_Eurac/Data/NAFLD' #FINDME!
DATA_PATH <- "."

```

## Packages

```{r packages, message=FALSE}
#' Load libraries
library(chrisUtils)
library(chrisData)
library(DT)
library(ggfortify)
library(ggplot2)
library(pander)
library(pheatmap)
library(plotly)
library(RColorBrewer)
library(readxl)
library(reshape2)
library(tidyfr)
library(vioplot)
library(writexl)

```


# Data import

The data from proteins was converted into the TDF format and can hence be
loaded and imported using the `tidyfr` R package. Now we read the full data
which includes general information for each measurement.


```{r load_dataset, message = FALSE}
#' load TDF using tidyfr
#list_data_modules(path = DATA_PATH)
md1 <- data_module(name = "chris_scanningswath_proteins", version = "1.0.0.2",
                    path = DATA_PATH)
prot_data <- data(md1)
prot_ann <- labels(md1)
stopifnot(all(colnames(prot_data) == rownames(prot_ann)))
colnames(prot_data) <- prot_ann$description
rownames(prot_ann) <- prot_ann$description

#' define the columns with protein concentrations
proteins <- rownames(prot_ann)[prot_ann$type != "categorical"]
proteins <- rownames(prot_ann)[prot_ann$Genes != ""]

#' problem in calculating lm below for
#'   'PC aa C30:2' #NOK, only 8 col!!
#'   'Histamine' NOK, only 8 col!!
#' because no values for each typeSLD, due to lot NA values
#' option 1: rm proteins, not done
#proteins <- proteins[proteins != 'PC aa C30:2']
#proteins <- proteins[proteins != 'Histamine']
#' option2: retain only none and MASLD, see done later

#' select proteins dataframe
prot_data <- prot_data[, proteins]
#head(prot_data)    #row samples, cols are protein name, matrix are absol conc
print('nr of  proteins included:')
ncol(prot_data)
print('included proteins are:')
colnames(prot_data)
#dim(prot_data)
#head(prot_data)

```

The loaded data contains the samples as rows, and the protein names as
colums. The measurements of the `prot_data` object are the relative
concentrations of protein per sample. No data transformation is performed, a
normal distribution of the proteomics is assumed.

Before proceeding with the data analysis we remove the POOL samples. In addition
we restrict to samples from the NAFLD study.

```{r subset to NAFLD set, message = FALSE, warning = FALSE}
prot_data <- prot_data[grep("^008", rownames(prot_data)), ]

```

All available trait (phenotype) data was defined in the
[nafld_data_overview.Rmd](nafld_data_overview.Rmd), is loaded below and merged
with the log2 transformed proteomics data of the NAFLD cohort.

```{r add phenotype dataa}
#' load xlsx file instead of below
handle <- 'masld_phenotypes.xlsx'
masld_phenotype_df <- read_excel(file.path(DATA_PATH, handle)) |>
    as.data.frame()
#head(masld_phenotype_df, 5)
#colnames(masld_phenotype_df)
#dim(masld_phenotype_df)

#' make sure sld_type is a factor with group 'none' as the baseline
masld_phenotype_df$sld_type <- factor(masld_phenotype_df$sld_type,
                                      levels = c('none', 'MASLD',
                                                 'MetALD', 'SLD'))
print('the sld types are:')
levels(masld_phenotype_df$sld_type)
print('nr of samples per sld types:')
table(masld_phenotype_df$sld_type)

#' merge with proteomics data
prot_data$AID <- rownames(prot_data)
#dim(prot_data)
prot_data <- merge(prot_data, masld_phenotype_df, by = 'AID')
#dim(prot_data)

#' write to file for manuel inspection
handle <- file.path(RDATA_PATH, 'prot_data.txt')
write.table(prot_data, file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)

```

Because for 2 proteins, 'PC aa C30:2' and 'Histamine', only NAs are present
for the less populated `sld_type` groups (*MetALD* end *SLD*), only *none* and
*MASLD* types are considered further.

```{r keep only none and MASLD types}
#' because problem 2 proteins only NAs in some sld types,
#' keep only none and MASLD
prot_data <- prot_data[prot_data$sld_type %in% c('MASLD'), ] #'none', 
prot_data$sld_type <- droplevels(prot_data$sld_type)
prot_data$Sex <- factor(prot_data$Sex)

#' Summary of available data:
summary(prot_data[, c("sld_type", "Age", "Sex", "BMI")])

```

Create additional variable for age per 10y, because coefficients and effect
sizes correspond to one year difference, and the anual difference is
neglectible. Dividing age with 10, difference in coefficents and effect sizes
would correspond to 10 years difference and it would be easier to define
meaningful cut-off.

```{r define-age10-variable, message = FALSE}
#' define age per 10y group
prot_data$Age_10 <- prot_data$Age / 10

```

```{r load-liver-param}
#' from same excel loaded above, already added to prot_data
#colnames(prot_data)
summary(prot_data[, c("FIB4", "T2D")])

```

FIB4 is numeric, continuous ranging between 0.3 - 7.9. A higher score is more 
liver fibrosis.


# Evaluation of the various variables in the data set

Overview of variable evaluation in the data set, prior to the lm regression:

- Violin plot of BMI of the data split by sex and SLD type. to evaluate whether
  there is a strong difference in BMI between MASLD and none control and, if so,
  if that variable also to the linear model.
- Violin plot for age.
- PCA of the data.

## BMI distribution of data

Recap from overview report. 
"There seems to be only a small difference in average FIB4 values between male 
and female participants. A strong (positive) relationship is however visible 
between FIB4 and age. A slight negative relationship seems to be present between 
FIB4 and BMI."

```{r violin-plots, fig.height=8, fig.width=7}
#violin sex vs FIB4 
vioplot(split(prot_data$FIB4,
              prot_data$Sex),
        xlab = "Sex", ylab = "FIB4",
        main = "FIB4 distribution by sex")

#binned plot for visibility
prot_data$FIB4_bin <- round(prot_data$FIB4, 0)
summary(prot_data$FIB4_bin)

#violin age vs FIB4_bin
vioplot(split(prot_data$Age,
              prot_data$FIB4_bin),
        xlab = "FIB4_bin", ylab = "Age",
        main = "FIB4_bin versus Age")

#violin BMI vs FIB4 
vioplot(split(prot_data$BMI,
              prot_data$FIB4_bin),
        xlab = "FIB4_bin", ylab = "BMI",
        main = "BMI_bin versus BMI")

#violin T2D vs FIB4 
vioplot(split(prot_data$FIB4,
              prot_data$T2D),
        xlab = "T2D", ylab = "FIB4",
        main = "FIB4 distribution by T2D")

```


## PCA score plot

PCA score plot of the study participants, collored by:

- BMI,
- Age,
- gender,
- T2D,
- clinical phenotype (i.e. `FIB4` values).

Prior the PCA, data is normalized to 0 mean and standard deviation of
1 (autoscaling). Plot shows the scores and loadings of all subjects in the
space of the first two PCs, with 14% of total variances explained.


```{r pca, fig.height=7, fig.width=7}

#' scaling and PCA
scaledData <- scale(prot_data[, proteins], scale = TRUE)
scPCA <- svd(scaledData)
scPCA.scores <- scPCA$u %*% diag(scPCA$d)    ## scores
scPCA.loadings <- scPCA$v                    ## loadings
rownames(scPCA.loadings) <- proteins
scPCA.variances <- round(100*((scPCA$d^2) / sum(scPCA$d^2)), 1)[1:10]

#' variance explained PCs
print('variance explained for PC1 + PC2:')
scPCA.variances[1] + scPCA.variances[2]

#' prep for plotting
PLOTLY <- TRUE
ggplotly_or_not <- function(x, interactive = TRUE) {
    if (interactive) ggplotly(x)
    else print(x)
}
myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))

#' plot with col BMI
BMI <- prot_data$BMI
gg_1 <- autoplot(prcomp(scaledData), data = prot_data) +
    geom_point(aes(colour = BMI), size = (BMI/10)) +
    scale_colour_gradientn(colours = myPalette(100)) +
    theme_bw()
ggplotly_or_not(gg_1, PLOTLY)

#' plot with col Age
Age <- prot_data$Age
gg_1 <- autoplot(prcomp(scaledData), data = prot_data) +
    geom_point(aes(colour = Age), size = (Age/25)) +
    scale_colour_gradientn(colours = myPalette(100)) +
    theme_bw()
ggplotly_or_not(gg_1, PLOTLY)

#' plot with col Sex
Sex <- prot_data$Sex
gg_1 <- autoplot(prcomp(scaledData), data = prot_data) +
    geom_point(aes(colour = Sex)) +
    scale_color_manual(name = Sex, values = c('pink', 'lightblue')) +
    theme_bw()
gg_1
#ggplotly_or_not(gg_1, PLOTLY)

#' plot with col T2D
T2D <- prot_data$T2D
gg_1 <- autoplot(prcomp(scaledData), data = prot_data) +
    geom_point(aes(colour = T2D)) +
    scale_color_manual(name = T2D, values = c('gold', 'purple')) +
    theme_bw()
gg_1
#ggplotly_or_not(gg_1, PLOTLY)

#' plot with col FIB4
FIB4 <- prot_data$FIB4
gg_1 <- autoplot(prcomp(scaledData), data = prot_data) +
    geom_point(aes(colour = FIB4), size = (FIB4)) +
    scale_colour_gradientn(colours = myPalette(100)) +
    theme_bw()
ggplotly_or_not(gg_1, PLOTLY)

#' plot with loading
gg_1 <- autoplot(prcomp(scaledData), data = prot_data,
                 loadings = TRUE, loadings.colour = 'gray',
                 loadings.label = TRUE, loadings.label.size = 3,
                 loadings.label.colour ="darkblue", colour = NA) +
    geom_point(color = 'gray', size = 0.1) +
    theme_bw()
ggplotly_or_not(gg_1, PLOTLY)

```

None of the variables BMI, age, sex, sld_type seem to affect a seperation in the 
PCA score plots. 


# Linear model to evaluate SLD types - original

linear models setup:

```{r define-lm-covariates-and-target}
print('covariates of the linear regression model:')
COVARIATES <- c("Sex", "Age_10")
print(COVARIATES)

print('target of the linear regression model:')
TARGET <- 'FIB4'
print("FIB4")

DATASET <- 'NAFLD'
print(paste0('final variables of the ', DATASET, ' dataset used:'))
#' variables are the output y in lm(y ~ .):
print(proteins)

print('nr of variables inspected:')
print(length(proteins))

print('nr of individuals inspected:')
print(dim(prot_data))

```

To identify proteins related with the 2 (none, MASLD) of the 4 SLD types
(none, MASLD, MetALD, SLD), seperate linear regression models are fitted for all
proteins separately (as a response) with sex, age and the (categorical)
FIB4 as a covariates.

Model coefficients and p-values are provided. Raw p-values are adjusted for
multiple hypothesis testing with the method from Benjamini & Hochberg.

```{r linear model, message=FALSE}
#' Define a function to fit the linear model:
#' - fit the linear model
#' - extract the coefficients and p-values skipping the intercept
#' - return the result as a one-row matrix, columns being coefficients
#'   and p-values
funForLinReg <- function(analyte, data) {
    y <- data[, analyte]
    data <- data[, c("Sex", "Age_10", "FIB4")]
    res <- coef(summary(lm(y ~ ., data = data)))[-1, ]
    pvals <- matrix(c(res[, 1L], res[, 4L]), nrow = 1)
    rownames(pvals) <- analyte
    colnames(pvals) <- c(paste0("coef_", rownames(res)),
                         paste0("p-value_", rownames(res)))
    pvals
}

#' tests
#prot_data[1:5, c("Sex", "Age", "FIB4")]
#prot_data[1:5, proteins[1]]
#data <- prot_data
#analyte <- proteins[1]
#funForLinReg(analyte, prot_data)
#analyte <- proteins[14]
#funForLinReg(analyte, prot_data)

#' Fit linear models and extract coefficients and p-values on each protein
lm_output <- lapply(proteins, FUN = funForLinReg, data = prot_data) |>
    do.call(what = rbind)

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)

```

Next we calculate the effect sizes.

To obtain comparable coefficients, prior to linear modelling all the proteins
are standardized to zero mean and unit variance (note that this does not
influence the linear models since they are affine equivariant). In such
autoscaled data, difference of 1 corresponds to a standard deviation of 1. This
parameter will be called "effect size" and will be use to evaluate group
differences.

```{r effectsize, message=FALSE}
#' Perform autoscaling for effect size (es) calculation
prot_data_scaled <- prot_data
prot_data_scaled[, proteins] <- prot_data[, proteins] |>
    as.matrix() |>
    scale(scale = TRUE)

#' fit linear models and extract the coefficients
es <- lapply(proteins, FUN = funForLinReg, data = prot_data_scaled) |>
    do.call(what = rbind)
es <- es[, grep("coef", colnames(es))]
colnames(es) <- sub("coef", "effectSize", colnames(es))
lm_output <- cbind(lm_output, es)
#head(lm_output)

```

To call a proteins *significant* we consider, in addition to the
*statistical* significance levels also the magnitude of its coefficient,
relative to the coefficient of variation (CV) for that protein. The CV,
calculated on study-internal QC samples (pools of all samples) are a measure of
the technical noise specific for that protein in the present data set. Thus,
with this method we prioritize proteins with a larger difference between the
observed (biological) difference and the (separately estimated) technical variance.

Below we add thus evaluate for each protein and comparison whether it
fulfills that criteria.

```{r define significance criteria}
#' define the different coefficients as comparisons
comps <- colnames(lm_output)[grep("coef", colnames(lm_output))]

#' define which proteins are considered statistically significant
sign_p <- lm_output[, grep("p-adj", colnames(lm_output))] < 0.05
colnames(sign_p) <- sub("p-adj_", "significant_", colnames(sign_p))
#colSums(sign_p)

#' Convert the (absolute) difference in abundance into a difference in
#' percentage based on a (log2) coefficient.
diff_percentage <- function(x) {
    (2^abs(x) - 1) * 100
}
diff_perc <- apply(lm_output[, comps], MARGIN = 2, diff_percentage)
colnames(diff_perc) <- sub("coef", "diff_perc", colnames(diff_perc))

#' Define for which proteins the difference in concentration is larger
#' than a certain proportion of the CV
cv_cut <- lapply(comps, function(z) {
    z <- sub("coef_", "", z)
    diff_perc[, paste0("diff_perc_", z)] > 50 *  ##after eval, set to 50
        prot_ann[proteins, "cv_qc_chris"]
}) |>
    do.call(what = cbind)
colnames(cv_cut) <- sub("coef_", "cv_cut_", comps)

#' Use statistical significance for non-categorical traits instead of bool
sign_cv <- sign_p & cv_cut
sign_cv[, "significant_Age_10"] <- sign_p[, "significant_Age_10"]

#' Combine with result from linear model converting to a data.frame to
#' avoid logicals being converted to numeric
lm_output <- data.frame(lm_output, sign_cv)

```

The number of proteins called significant or with an adjusted p-value smaller
than 0.05 for each of the extracted coefficients are shown below. For a complete
comparison, the number of proteins complying to either the adjusted p value
or either the CV-value, as single threshold criteria, are calculated and
summarized here as well.

```{r table-sig, results = "asis"}
#' merge to p-adj to cv significance summary
sig_tab <- data.frame(
    `p_adj < 0.05` = colSums(sign_p),
    `cv thresh` = colSums(cv_cut),
    significant = colSums(lm_output[, sub("coef_", "significant_", comps)]),
    check.names = FALSE)

pandoc.table(
    sig_tab, style = "rmarkdown",
    caption = paste0("Number of significant proteins per tested covariate. ",
                     " *p_adj < 0.05*: number of statistically significant",
                     " proteins.",
                     " *cv*: number of proteins with percentage abundance",
                     " difference above the CV-based threshold.",
                     " *significant*: number of proteins that fulfill ",
                     "the CV-based criteria in addition to the p-adj. threshold"))

```

Export results for inspection:

```{r export-results}
results <- data.frame(
    prot_ann[rownames(lm_output), c("description", "long_description", "PeptideNN",
                         "PeptideNames", "Protein.Ids", "Protein.Names",
                         "Genes", "Modified.Sequence", "Stripped.Sequence",
                         "Precursor.Id", "cv_qc_chris")],
    lm_output,
    diff_perc
)

#colnames(results)
#colnames(lm_output)

#' write to file for manual inspection,
#' inclusing inspection of the CV-value of each protein
handle <- file.path(RDATA_PATH, 'lm_output_results.txt')
write.table(results, file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)

```

We visualize the associations in a heatmap. The effect size are added for the
associations, for the significant proteins (min 1 significance to
covariate). also for the not significant associations, the effect size is shown.

```{r heatmap, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for proteins with at least one significant association."}
#' Extract effect sizes
es_results <- results[, grep("effectSize_", colnames(results))] |>
    as.matrix()

#' Restrict to those significant for at least one of the groups (sex, age, SLD type)
sig_results <- results[, grep("significant_", colnames(results))] |>
    as.matrix()
es_results <- es_results[rowSums(sig_results) > 0, ]
print('nr of proteins with >= 1 significant association:')
nrow(es_results)

#' hierarchical clustering
clust1 <- hclust(dist((es_results)), method = "ward.D")
clust2 <- hclust(dist(t(es_results)), method = "ward.D")

#' clean roundup nrs for plot
es_results[es_results == 0] <- NA
es_results <- round(es_results, 2)

#' info regarding FIB4
#print('nr of proteins with a not-zero effect size in the FIB4 group:')
#sum(es_results[, 'effectSize_FIB4'] != 0)

#' in long format
es_results_long <- melt(as.matrix(es_results))

#' ggplot
plt <- es_results_long |>
    ggplot(aes(Var2, Var1, fill = value, label = value)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Effect Size") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 2.5) +
    theme_classic() +
    scale_x_discrete(
        expand = c(0, 0), limits = colnames(es_results)[clust2$order]) +
    scale_y_discrete(
        expand = c(0, 0), limits = rownames(es_results)[clust1$order]) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 10),
          axis.text.y = element_text(hjust = 1, size = 8),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.position = "none")

ggsave(paste0(IMAGE_PATH, "protein_SLDtypes_effect_sizes.png"),
       plot = plt, dpi = 600, width = 12,
       height = 12, units = "cm", scale = 2.5)
plt

```

In second plot, only the significant effect size labels were added to the
heat map. also, the dendograms after hierarchical clustering are shown.

```{r heatmap2, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for significant associations the value of the effect size is shown."}
#' keep just the effect sizes of the significant associations
es_results_sig <- es_results
es_results_sig[as.matrix(sig_results)[rownames(es_results), ] == 0] <- ""

#' info regarding FIB4 (check=ok)
print('nr of significant proteins in the FIB4 group:')
sum(es_results_sig[, 'effectSize_FIB4'] != "")

#' plot
colnames(es_results) <- sub("effectSize_", "", colnames(es_results))
pheatmap(es_results, display_numbers = es_results_sig,
         number_color = "#444444", fontsize_number = 7)

```

A considerable number of proteins are significant both for MASLD and for sex,
in many cases even with the same sign.

We investigate the associations for MASLD, i.e. the difference in average
concentrations between participants of SLD types *MASLD* and *none*. We show the
results for that coefficient using a volcano plot, coloring significant
associations in red.

```{r volcano, FIG-volcano-FIB4, fig.path = IMAGE_PATH, fig.width = 8, fig.height = 8, fig.cap = "Volcano plot for the liver stiffness. Red colored points are those with an adjusted p-value smaller than 0.05."}
plot_volcano <- function(x, comparison = character(), ...) {
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    plot(X, -log10(Y), pch = 21,
         col = "#000000ce", bg = "#00000060",
         xlab = "coefficient", ylab = expression(-log[10](p[adjusted])),
         ...)
}
significant_points <- function(x, comparison = character(),
                               diff_perc = 2, col = "#E41A1CCE",
                               bg = "#E41A1C40", ...) {
    s <- x[, paste0("significant_", comparison)]
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    if (any(s))
        points(X[s], -log10(Y[s]), col = col, bg = bg, pch = 21)
}
plot_volcano(results, "FIB4", main = "MASLD vs none")
grid()
abline(h = -log10(0.05), lty = 1, col = "grey")
significant_points(results, comparison = "FIB4")

```

The (static) table with the significant proteins is shown below:

```{r table-sig-sex, results = "asis", echo = FALSE}
significant_table <- function(x, comparison = character(),
                              columns = c("Genes", "description", "cv_qc_chris")) {
    sigs <- x[, paste0("significant_", comparison)]
    columns <- c(columns, paste0(c("coef_", "p.adj_", "effectSize_"),
                                 comparison))
    tmp <- x[sigs, columns]
    tmp <- tmp[order(tmp[, paste0("p.adj_", comparison)]), ]
    colnames(tmp) <- sub(paste0("_", comparison), "", colnames(tmp))
    tmp
}
sig_tab <- significant_table(results, comparison = "FIB4")

pandoc.table(
    sig_tab, style = "rmarkdown", split.table = Inf,
    caption = paste0("proteins with significant differences in ",
                     "concentrations between MASLD and control (none) participants."))

```

The full results table is shown below.

```{r table-top-5, results = "asis"}
##
oorder <- results[order(results[, "p.adj_FIB4"]), ]

pandoc.table(oorder[1:5, c(1, 4, 7, 10, 13, 17)], style = "rmarkdown",
             split.table = Inf, caption = "Top 5 proteins for type MASLD")

datatable(results) |>
    formatRound(grep("coef|p.value|p.adj|effect", colnames(results)),
                digits = 6)

```

There are `r nrow(sig_tab)` significantly associated proteins for SLD type
MASLD vs. none.


Significant metabolites adjusted for age and sex (top 5) are visualized in
boxplot in next figure. Points are colored according to sample sex.


```{r prep for plotting}
#' prep for plotting
PLOTLY <- TRUE
ggplotly_or_not <- function(x, interactive = TRUE) {
    if (interactive) ggplotly(x)
    else print(x)
}
myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))

box_plot <- function(data, aes, analyte) {
    ggplot(data, aes) +
        geom_jitter(position = position_jitter(0.24),
                    aes(colour = Sex, alpha = 0.5 , size = (BMI/10))) +
        geom_violin(trim = FALSE, aes(alpha = 0.5)) +
        stat_summary(fun = median, geom = "point", shape = 23, size = 2) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) + labs(title = analyte) +
        theme_bw ()
}

scatter_plot <- function(data, aes, analyte) {
    ggplot(data, aes) +
        geom_jitter(position = position_jitter(0.24),
                    aes(colour = Sex, alpha = 0.5 , size = 0.1)) +
        stat_smooth(method = lm, se = TRUE) + #with standard error
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) + labs(title = analyte) +
        theme_bw ()
}

#' for boxplot func to work automatically
tmp_colnr <- which(colnames(prot_data) == TARGET)
prot_data$tmp_target <- prot_data[, tmp_colnr]   #value of eg BMI

```


```{r lm-top1, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(oorder)[1]
prot_data$top_analyte <- prot_data[, top_analyte]  #intensities of metabolite
s_1 <- scatter_plot(prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)

```

```{r lm-top2, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- oorder[2, 1]
prot_data$top_analyte <- prot_data[, top_analyte]
s_1 <- scatter_plot(prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r lm-top3, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- oorder[3, 1]
prot_data$top_analyte <- prot_data[, top_analyte]
s_1 <- scatter_plot(prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r lm-top4, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- oorder[4, 1]
prot_data$top_analyte <- prot_data[, top_analyte]
s_1 <- scatter_plot(prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r lm-top5, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- oorder[5, 1]
prot_data$top_analyte <- prot_data[, top_analyte]
s_1 <- scatter_plot(prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```


```{r save results for sensitivity analysis}
#' get coef values from both lm models
lm_original <- results$coef_FIB4
```


In addition, we also export the results table from the lm containing the m/z,
rt, metabolite feature ID, statistics (coef, es) and p-vlues (adj. p-value) for
subsequent analysis (eg pathway enrichment and analysis).

```{r export-lm-results, eval = FALSE}
#' add rownames as a column
results$AID <- rownames(results)

#' write results (original order) for further analysis
handle <- file.path(DATA_PATH, paste0(filename, '_results.txt'))
write.table(results, file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)

```


# Acknowledgments

Consulted sources used as basis for code:

- ~/nafld_proteo_metabolomics/NAFLD_metab_targeted_lm.Rmd
- ~?metabo_chris_untargeted/CHRIS_lm_pos.Rmd


# Session information

R packages used for the analysis:

```{r}
sessionInfo()
```


