---
title: "NAFLD_MASLD_sensititvity_analysis"
author: "Marilyn De Graeve, Johannes Rainer, Christoph Grander"
affiliation: "Eurac Research, Bolzano, Italy"
date: "2025-02-10"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
---

**Modified**: `r file.info("NAFLD_sensititvity_analysis.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r style, message = FALSE, echo = FALSE, warning = FALSE, results = "asis"}
library("BiocStyle")
library("kableExtra")
library("knitr")
library("rmarkdown")
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now, units = "secs")
      # return a character string to show the time
      paste("    Time for this code chunk to run:", round(res,
        2), "seconds")
    }
  }
}))
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7, time_it = TRUE)

```


# Introduction

During this workflow, sensitivity analyses are performed to see if there are
potential unique metabolites/proteins between the lm in the MASLD cohort vs in
the none (control, healthy) cohort of the NAFLD dataset.

For this, the results are loaded of the saved lm results file for fibrosis and
steatosis, each time run and renamed according to its subcohort (MASLD, none).


# Setup

## Directories

```{r directories, echo = FALSE}
#' General settings
filename <- " NAFLD_MASLD_sensititvity_analysis"

#' Path to the data
#FINDME!!  see manual below

```

## Packages

```{r packages, message=FALSE}
#' Load libraries
library(ggplot2)
library(pander)
library(plotly)

```


# 1 Fibrosis - tprot

## load lm results

```{r}
#' get coef values from both lm models
handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/ NAFLD_fibrosis_prot_lm/ NAFLD_fibrosis_prot_lm_results-masld.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_masld <- results$coef_FIB4

handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/ NAFLD_fibrosis_prot_lm/ NAFLD_fibrosis_prot_lm_results-none.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_none <- results$coef_FIB4

rownames(results) <- results$AID

```


## Sensitivity analysis

The coeficients (MASLD vs none) from the linear model to evaluate SLD
types - original is compared to the linear model to evaluate SLD types - with
BMI, using a sensitivity analysis.

```{r}
#' merge coef values from both lm models above
sensit <-data.frame(featureid = rownames(results),
                  x = lm_masld,
                  y = lm_none)
#head(sensit)

#' prep for plotting
PLOTLY <- TRUE
ggplotly_or_not <- function(x, interactive = TRUE) {
    if (interactive) ggplotly(x)
    else print(x)
}

#' scatterplot with coeff vales
scatter_plot <- function(data) {
    ggplot(data, aes(x = x, y = y)) +
        geom_point(aes(x,y), colour = "gray65") +
        stat_smooth(method = lm, se = TRUE) + #with standard error
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) +
        labs(title = 'Fibrosis tprot coefficients between datasets:',
             x = 'lm model - MASLD',
             y = 'lm model - none') +
        theme_bw ()
}

s_1 <- scatter_plot(sensit)
ggplotly_or_not(s_1, PLOTLY)

```

```{r, eval = FALSE}
#' FC for see diff between 2 datasets
sensit$FC <- abs(sensit$x + 0.00000001) / abs(sensit$y + 0.00000001)
sensit$log2FC <- log2(abs(sensit$FC))
# 0.5 and > 2 intersting, NO too many
#summary(sensit)
#dim(sensit)
hist(sensit$log2FC)
#sensit <- sensit[sensit$log2FC < 0.5 | sensit$log2FC > 2.0, ]  #94 of 148 left, maybe more diff?
#sensit <- sensit[sensit$FC > 25, ] #5 left
sensit <- sensit[sensit$log2FC < -5 | sensit$log2FC > +5, ]  #10 left

#' table with highest dev coef
pandoc.table(
    sensit, style = "rmarkdown", split.table = Inf,
    caption = paste0("Coeficients of lm models using proteins ",
                     "between MASLD and control (none) participants",
                     ", only log2FC < -5 and log2FC > +5"))

```


## Table with differences

```{r, results = "asis"}
#add difference column
sensit$diff <- abs(sensit$x - sensit$y)
sensit <- sensit[order(sensit$diff, decreasing = TRUE), ]

pandoc.table(sensit[1:20, ], style = "rmarkdown",
             split.table = Inf, caption = "Top 20 different variable coeficients between lm model with- and without medication")

```


# 2 Fibrosis - tmetab

## load lm results

```{r}
#' get coef values from both lm models
handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_fibrosis_tmetab_lm/NAFLD_fibrosis_tmetab_lm_results-masld.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_masld <- results$coef_FIB4

handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_fibrosis_tmetab_lm/NAFLD_fibrosis_tmetab_lm_results-none.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_none <- results$coef_FIB4

rownames(results) <- results$AID

```


## Sensitivity analysis

The coeficients (MASLD vs none) from the linear model to evaluate SLD
types - original is compared to the linear model to evaluate SLD types - with
BMI, using a sensitivity analysis.

```{r}
#' merge coef values from both lm models above
sensit <-data.frame(featureid = rownames(results),
                  x = lm_masld,
                  y = lm_none)
#head(sensit)

#' scatterplot with coeff vales
scatter_plot <- function(data) {
    ggplot(data, aes(x = x, y = y)) +
        geom_point(aes(x,y), colour = "gray65") +
        stat_smooth(method = lm, se = TRUE) + #with standard error
        #geom_text(x = -Inf, y = Inf, label = equation, parse = TRUE, vjust = 1.5, hjust = -0.1) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) +
        labs(title = 'Fibrosis tmetab coefficients between datasets:',
             x = 'lm model - MASLD',
             y = 'lm model - none') +
        theme_bw ()
}

s_1 <- scatter_plot(sensit)
ggplotly_or_not(s_1, PLOTLY)

```

```{r, eval = FALSE}
#' FC for see diff between 2 datasets
sensit$FC <- abs(sensit$x + 0.00000001) / abs(sensit$y + 0.00000001)
sensit$log2FC <- log2(abs(sensit$FC))
# 0.5 and > 2 intersting, NO too many
#summary(sensit)
#dim(sensit)
hist(sensit$log2FC)
#sensit <- sensit[sensit$log2FC < 0.5 | sensit$log2FC > 2.0, ]  #94 of 148 left, maybe more diff?
#sensit <- sensit[sensit$FC > 25, ] #5 left
sensit <- sensit[sensit$log2FC < -5 | sensit$log2FC > +5, ]  #10 left

#' table with highest dev coef
pandoc.table(
    sensit, style = "rmarkdown", split.table = Inf,
    caption = paste0("Coeficients of lm models using proteins ",
                     "between MASLD and control (none) participants",
                     ", only log2FC < -5 and log2FC > +5"))

```


## Table with differences

```{r, results = "asis"}
#add difference column
sensit$diff <- abs(sensit$x - sensit$y)
sensit <- sensit[order(sensit$diff, decreasing = TRUE), ]

pandoc.table(sensit[1:20, ], style = "rmarkdown",
             split.table = Inf, caption = "Top 20 different variable coeficients between lm model with- and without medication")

```


# 3 Steatosis - tprot

## load lm results

```{r}
#' get coef values from both lm models
handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/ NAFLD_steatosis_prot_lm/ NAFLD_steatosis_prot_lm_results-masld.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_masld <- results$coef_liver_stiffness

handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/ NAFLD_steatosis_prot_lm/ NAFLD_steatosis_prot_lm_results-none.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_none <- results$coef_liver_stiffness

rownames(results) <- results$AID

```


## Sensitivity analysis

The coeficients (MASLD vs none) from the linear model to evaluate SLD
types - original is compared to the linear model to evaluate SLD types - with
BMI, using a sensitivity analysis.

```{r}
#' merge coef values from both lm models above
sensit <-data.frame(featureid = rownames(results),
                  x = lm_masld,
                  y = lm_none)
#head(sensit)

#' scatterplot with coeff vales
scatter_plot <- function(data) {
    ggplot(data, aes(x = x, y = y)) +
        geom_point(aes(x,y), colour = "gray65") +
        stat_smooth(method = lm, se = TRUE) + #with standard error 
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) +
        labs(title = 'Steatosis tprot coefficients between datasets:',
             x = 'lm model - MASLD',
             y = 'lm model - none') +
        theme_bw ()
}

s_1 <- scatter_plot(sensit)
ggplotly_or_not(s_1, PLOTLY)

```

```{r, eval = FALSE}
#' FC for see diff between 2 datasets
sensit$FC <- abs(sensit$x + 0.00000001) / abs(sensit$y + 0.00000001)
sensit$log2FC <- log2(abs(sensit$FC))
# 0.5 and > 2 intersting, NO too many
#summary(sensit)
#dim(sensit)
hist(sensit$log2FC)
#sensit <- sensit[sensit$log2FC < 0.5 | sensit$log2FC > 2.0, ]  #94 of 148 left, maybe more diff?
#sensit <- sensit[sensit$FC > 25, ] #5 left
sensit <- sensit[sensit$log2FC < -5 | sensit$log2FC > +5, ]  #10 left

#' table with highest dev coef
pandoc.table(
    sensit, style = "rmarkdown", split.table = Inf,
    caption = paste0("Coeficients of lm models using proteins ",
                     "between MASLD and control (none) participants",
                     ", only log2FC < -5 and log2FC > +5"))

```


## Table with differences

```{r, results = "asis"}
#add difference column
sensit$diff <- abs(sensit$x - sensit$y)
sensit <- sensit[order(sensit$diff, decreasing = TRUE), ]

pandoc.table(sensit[1:20, ], style = "rmarkdown",
             split.table = Inf, caption = "Top 20 different variable coeficients between lm model with- and without medication")

```


# 4 Steatosis - tmetab

## load lm results

```{r}
#' get coef values from both lm models
handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_steatosis_tmetab_lm/NAFLD_steatosis_tmetab_lm_results-masld.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_masld <- results$coef_liver_stiffness

handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_steatosis_tmetab_lm/NAFLD_steatosis_tmetab_lm_results-none.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_none <- results$coef_liver_stiffness

rownames(results) <- results$AID

```


## Sensitivity analysis

The coeficients (MASLD vs none) from the linear model to evaluate SLD
types - original is compared to the linear model to evaluate SLD types - with
BMI, using a sensitivity analysis.

```{r}
#' merge coef values from both lm models above
sensit <-data.frame(featureid = rownames(results),
                  x = lm_masld,
                  y = lm_none)
#head(sensit)

#' scatterplot with coeff vales
scatter_plot <- function(data) {
    ggplot(data, aes(x = x, y = y)) +
        geom_point(aes(x,y), colour = "gray65") +
        stat_smooth(method = lm, se = TRUE) + #with standard error
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) +
        labs(title = 'Steatosis tmetab coefficients between datasets:',
             x = 'lm model - MASLD',
             y = 'lm model - none') +
        theme_bw ()
}

s_1 <- scatter_plot(sensit)
ggplotly_or_not(s_1, PLOTLY)

```

```{r, eval = FALSE}
#' FC for see diff between 2 datasets
sensit$FC <- abs(sensit$x + 0.00000001) / abs(sensit$y + 0.00000001)
sensit$log2FC <- log2(abs(sensit$FC))
# 0.5 and > 2 intersting, NO too many
#summary(sensit)
#dim(sensit)
hist(sensit$log2FC)
#sensit <- sensit[sensit$log2FC < 0.5 | sensit$log2FC > 2.0, ]  #94 of 148 left, maybe more diff?
#sensit <- sensit[sensit$FC > 25, ] #5 left
sensit <- sensit[sensit$log2FC < -5 | sensit$log2FC > +5, ]  #10 left

#' table with highest dev coef
pandoc.table(
    sensit, style = "rmarkdown", split.table = Inf,
    caption = paste0("Coeficients of lm models using proteins ",
                     "between MASLD and control (none) participants",
                     ", only log2FC < -5 and log2FC > +5"))

```


## Table with differences

```{r, results = "asis"}
#add difference column
sensit$diff <- abs(sensit$x - sensit$y)
sensit <- sensit[order(sensit$diff, decreasing = TRUE), ]

pandoc.table(sensit[1:20, ], style = "rmarkdown",
             split.table = Inf, caption = "Top 20 different variable coeficients between lm model with- and without medication")

```


# 5 Steatosis - tprot (naab13)

## load lm results

```{r}
#' get coef values from both lm models
handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/ NAFLD_steatosis_prot_lm/ NAFLD_steatosis_prot_lm_results-naab13-masld.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_masld <- results$coef_liver_fat

handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/ NAFLD_steatosis_prot_lm/ NAFLD_steatosis_prot_lm_results-naab13-none.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_none <- results$coef_liver_fat

rownames(results) <- results$AID

```


## Sensitivity analysis

The coeficients (MASLD vs none) from the linear model to evaluate SLD
types - original is compared to the linear model to evaluate SLD types - with
BMI, using a sensitivity analysis.

```{r}
#' merge coef values from both lm models above
sensit <-data.frame(featureid = rownames(results),
                  x = lm_masld,
                  y = lm_none)
#head(sensit)

#' scatterplot with coeff vales
scatter_plot <- function(data) {
    ggplot(data, aes(x = x, y = y)) +
        geom_point(aes(x,y), colour = "gray65") +
        stat_smooth(method = lm, se = TRUE) + #with standard error 
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) +
        labs(title = 'Steatosis naab13 tprot coefficients between datasets:',
             x = 'lm model - MASLD',
             y = 'lm model - none') +
        theme_bw ()
}

s_1 <- scatter_plot(sensit)
ggplotly_or_not(s_1, PLOTLY)

```

```{r, eval = FALSE}
#' FC for see diff between 2 datasets
sensit$FC <- abs(sensit$x + 0.00000001) / abs(sensit$y + 0.00000001)
sensit$log2FC <- log2(abs(sensit$FC))
# 0.5 and > 2 intersting, NO too many
#summary(sensit)
#dim(sensit)
hist(sensit$log2FC)
#sensit <- sensit[sensit$log2FC < 0.5 | sensit$log2FC > 2.0, ]  #94 of 148 left, maybe more diff?
#sensit <- sensit[sensit$FC > 25, ] #5 left
sensit <- sensit[sensit$log2FC < -5 | sensit$log2FC > +5, ]  #10 left

#' table with highest dev coef
pandoc.table(
    sensit, style = "rmarkdown", split.table = Inf,
    caption = paste0("Coeficients of lm models using proteins ",
                     "between MASLD and control (none) participants",
                     ", only log2FC < -5 and log2FC > +5"))

```


## Table with differences

```{r, results = "asis"}
#add difference column
sensit$diff <- abs(sensit$x - sensit$y)
sensit <- sensit[order(sensit$diff, decreasing = TRUE), ]

pandoc.table(sensit[1:20, ], style = "rmarkdown",
             split.table = Inf, caption = "Top 20 different variable coeficients between lm model with- and without medication")

```



# 6 Steatosis - tmetab (naab13)

## load lm results

```{r}
#' get coef values from both lm models
handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_steatosis_tmetab_lm/NAFLD_steatosis_tmetab_lm_results-naab13-masld.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_masld <- results$coef_liver_fat

handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_steatosis_tmetab_lm/NAFLD_steatosis_tmetab_lm_results-naab13-none.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_none <- results$coef_liver_fat

rownames(results) <- results$AID

```


## Sensitivity analysis

The coeficients (MASLD vs none) from the linear model to evaluate SLD
types - original is compared to the linear model to evaluate SLD types - with
BMI, using a sensitivity analysis.

```{r}
#' merge coef values from both lm models above
sensit <-data.frame(featureid = rownames(results),
                  x = lm_masld,
                  y = lm_none)
#head(sensit)

#' scatterplot with coeff vales
scatter_plot <- function(data) {
    ggplot(data, aes(x = x, y = y)) +
        geom_point(aes(x,y), colour = "gray65") +
        stat_smooth(method = lm, se = TRUE) + #with standard error
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) +
        labs(title = 'Steatosis naab13 tmetab coefficients between datasets:',
             x = 'lm model - MASLD',
             y = 'lm model - none') +
        theme_bw ()
}

s_1 <- scatter_plot(sensit)
ggplotly_or_not(s_1, PLOTLY)

```

```{r, eval = FALSE}
#' FC for see diff between 2 datasets
sensit$FC <- abs(sensit$x + 0.00000001) / abs(sensit$y + 0.00000001)
sensit$log2FC <- log2(abs(sensit$FC))
# 0.5 and > 2 intersting, NO too many
#summary(sensit)
#dim(sensit)
hist(sensit$log2FC)
#sensit <- sensit[sensit$log2FC < 0.5 | sensit$log2FC > 2.0, ]  #94 of 148 left, maybe more diff?
#sensit <- sensit[sensit$FC > 25, ] #5 left
sensit <- sensit[sensit$log2FC < -5 | sensit$log2FC > +5, ]  #10 left

#' table with highest dev coef
pandoc.table(
    sensit, style = "rmarkdown", split.table = Inf,
    caption = paste0("Coeficients of lm models using proteins ",
                     "between MASLD and control (none) participants",
                     ", only log2FC < -5 and log2FC > +5"))

```


## Table with differences

```{r, results = "asis"}
#add difference column
sensit$diff <- abs(sensit$x - sensit$y)
sensit <- sensit[order(sensit$diff, decreasing = TRUE), ]

pandoc.table(sensit[1:20, ], style = "rmarkdown",
             split.table = Inf, caption = "Top 20 different variable coeficients between lm model with- and without medication")

```


# Acknowledgments

- NAFLD_metab_targeted_lm
- NAFLD_fibrosis_prot_lm, etc.


# Session information

R packages used for the analysis:

```{r}
sessionInfo()
```


