---
title: "Exploratory data analysis of the NAFLD Scanning SWATH proteomics data set"
author: "Nikola Dordevic, Essi Hantikainen, Johannes Rainer"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
#bibliography: references.bib
---

**Modified**: `r file.info("NAFLD_scanningswath_proteomics_exploratory_analysis.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r biocstyle, echo = FALSE, results = "asis", message = FALSE}
#' rmarkdown format settings
library(BiocStyle)
library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8, fig.height = 7)
BiocStyle::markdown()
```



```{r settings, echo = FALSE}
#' General settings
filename <- "NAFLD_scanningswath_proteomics_exploratory_analysis"
#' Path to save images to
IMAGE_PATH <- paste0("images/", filename, "/")
if (file.exists(IMAGE_PATH))
    unlink(IMAGE_PATH, recursive = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE)
#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)
```


# Introduction

In this document we perform an unsupervised *exploratory* data analysis of the
Scanning SWATH proteomics data set of the NAFLD study to evaluate and identify
any relations between proteins and/or participants.

# Data import

Below we load all required packages for the analysis and retrieve the data.

```{r libraries}
library(CompMetaboTools)
library(tidyfr)
library(chrisUtils)
library(chrisData)
library(nafldData)
library(RColorBrewer)
library(pander)
library(pheatmap)
library(DT)
library(ggfortify)
library(ggplot2)
library(reshape2)
library(plotly)
library(rrcov)
library(WGCNA)
library(flashClust)
options(rgl.useNULL = TRUE)
library(rgl)
library(writexl)
```


We next load the data the TDFF data module.

```{r, message = FALSE}
scan <- data_module(name = "chris_scanningswath_proteins", version = "1.0.0.2",
                    path = ".")
protSS_data <- data(scan)
protSS_ann <- labels(scan)
stopifnot(all(colnames(protSS_data) == rownames(protSS_ann)))
colnames(protSS_data) <- protSS_ann$description
rownames(protSS_ann) <- protSS_ann$description
## define the columns with peptide concentrations
proteins <- rownames(protSS_ann)[protSS_ann$Genes != ""]
protSS_data <- protSS_data[, proteins]
```

Scaninng SWATH data is already log2 transformed!
Before proceeding with the data analysis we restrict to samples from the NAFLD
study.

```{r keep-NAFLDs, message = FALSE, warning = FALSE}
protSS_data <- protSS_data[grep("^008", rownames(protSS_data)), ]
```

# General data overview

In this dataset `r ncol(protSS_data)` proteins measured from
`r nrow(protSS_data)` samples collected from the NAFLD study are evaluated. Note
that all proteins are log2 transformed.

Distributions of all 148 metabolites are visualized in histograms.

```{r Distributions}
PlotHist <- function(AnlName) {
    Anl <- protSS_data[, AnlName]
    plot(density(Anl, na.rm = T), col = 2, main = AnlName,
                                               xlab = "Conc.")
    }
anlts <- colnames(protSS_data)
splits <- rep(seq_len(ceiling(length(anlts) / 12)), each = 12)
splits <- splits[1:length(anlts)]
anlts <- split(anlts, splits)
for (i in seq_along(anlts)) {
    anlts_i <- anlts[[i]]
    par(mfrow = c(3, 4))
    for (anlt in anlts_i)
        PlotHist(anlt)
}
```


Demographics of the NAFLD cohort and evaluation of available trait data.

```{r}
n_anns <- nafldTraitAnnotations()

#' personal data
n_perso <- nafldData("personal")
rownames(n_perso) <- n_perso$AID
##
protSS_data$AID <- n_perso[rownames(protSS_data), "AID"]
protSS_data$AID_CHRIS <- n_perso[rownames(protSS_data), "aid_chris"]
protSS_data$Age <- n_perso[rownames(protSS_data), "na_ager"]
protSS_data$Sex <- n_perso[rownames(protSS_data), "na_sex"]
protSS_data$fwup <- n_perso[rownames(protSS_data), "na_fwup"]


#' na_ager: rounded age
#' na_fwup: follow-up duration CHRIS->NAFLD

tmp <- nafldData("recruitment")
rownames(tmp) <- tmp$AID
protSS_data$case <- tmp[rownames(protSS_data), "na_case"]
```

The time span between the two studies is thus between 2 to 5 years. We next
add additional anthropometric traits to the data.

```{r}
#' BMI, fat etc.
n_anthro <- nafldData("anthro")
rownames(n_anthro) <- n_anthro$AID

protSS_data$BMI <- n_anthro[rownames(protSS_data), "naan03"]
protSS_data$body_fat <- n_anthro[rownames(protSS_data), "naan04"]
protSS_data$visc_fat <- n_anthro[rownames(protSS_data), "naan05"]
```

We next add the clinical examinations performed on all participants.

```{r}
#' Clinical examinations
tmp <- nafldData("clinical")
rownames(tmp) <- tmp$AID
```

We will use and evaluate the following parameters defined by a clinical
examination:

- *naab11*: liver stiffness (higher values are worse). Values for participants
  that have a *naab12* (liver stiffness IQR) > 30 should be either removed or
  taken with care.
- *naab12*: liver stiffness IQR: the IQR of repeated liver stiffness
  measurements. Values below 30 are considered OK.
- *naab13*: liver fat. Can be binarized at an absolute value of 288.
- *naab14*: subjective, visual evaluation of the liver fat. Categorical from 1
  to 3. Values 2 and 3 can be considered steatotic.
- *nascfib4*: fibrosis 4 index. Linear score calculated from standard clinical
  parameters.
- *nascmets*: metabolic syndrome, binary category.

```{r}
protSS_data$liver_stiffness <- tmp[rownames(protSS_data), "naab11"]
protSS_data$liver_stiffness_iqr <- tmp[rownames(protSS_data), "naab12"]
protSS_data$liver_fat <- tmp[rownames(protSS_data), "naab13"]
protSS_data$visual_fat <- tmp[rownames(protSS_data), "naab14"]
protSS_data$fib4 <- tmp[rownames(protSS_data), "nascfib4"]
protSS_data$metsym <- tmp[rownames(protSS_data), "nascmets"]
```


In next table, general overview for all metabolites are presented (mean,
standard deviation, coefficients of variation of QC samples):

```{r}
Table_1 <- data.frame(
    Genes = protSS_ann[proteins,"Genes"],
    Description = protSS_ann[proteins,"long_description"],
    Mean = colMeans(protSS_data[,proteins], na.rm = TRUE),
    SD = sapply(protSS_data[,proteins], sd, na.rm = TRUE),
    qc_cv = protSS_ann[proteins, "cv_qc_chris"])
    #CV_QC = rowData(chris_plasma_proteins)$cv_qc)
datatable(Table_1) |>
formatRound(columns = seq(3, ncol(Table_1)), digits = 3)
```

There are no missing values!



# Outlier detection

Outlier detection was performed using robust PCA as described by Hubert et al.
2005 and implemented in the rrcov package for R
[@todorovObjectOrientedFrameworkRobust2009].
It is important to be aware of possible outliers, because they can affect the
results of some possible future analyses (e.g. some parametric multivariate
methods like PLS is highly sensitive to them).
In addition, they can provide us insight into some possible technical issues
during measurements or they can detect samples with some interestin biological
properties (with high or low abundances of some metabolites).
In next figure, diagnostic plot of the dataset based on two robust principal
components is presented:

```{r, PCA_Hubert, fig.path = IMAGE_PATH, fig.width=12, fig.height=10}
## outlier identification:
par(mfrow=c(1,1))
pca_Hubert <- PcaHubert(protSS_data[,proteins])
plot(pca_Hubert)
```

Looking at the plot there are no clear outliers.


# General unsupervised grouping of samples (PCA)

In next figures, scores and loadings from the PCA are presented considering all
proteins.
Prior the PCA, data is normalized to 0 mean and standard deviation of 1
(z-score or autoscaling).
Plot shows the scores and loadings of all subjects in the space of the first 3
PCs, with almost 18.4% of total variances explained.

In next PCA, samples are colored acordin sample sex:

```{r}
##
scaledData <- scale(protSS_data[,proteins], scale = TRUE)
scPCA <- svd(scaledData)
scPCA.scores <- scPCA$u %*% diag(scPCA$d)    ## scores
scPCA.loadings <- scPCA$v                    ## loadings
scPCA.variances <- round(100*((scPCA$d^2)/sum(scPCA$d^2)),1)[1:10]
rownames(scPCA.loadings) <- proteins

plot3d(scPCA.scores[,1], scPCA.scores[,2], scPCA.scores[,3], xlab=paste("PC 1(",
       scPCA.variances[1], "%)", sep = ""), ylab = paste("PC 2(",
       scPCA.variances[2], "%)", sep = ""), zlab = paste("PC 3(",
       scPCA.variances[3], "%)", sep = ""), pch = 19,
       col = as.integer(protSS_data$Sex))
#text3d(39*scPCA.loadings[,1:3], texts=rownames(scPCA.loadings), col="darkgrey")
#coords <- NULL
#for (i in 1:nrow(scPCA.loadings)) {
#  coords <- rbind(coords, rbind(c(0,0,0),39*scPCA.loadings[i,1:3]))}
#lines3d(coords, col="darkgrey", lwd=1)
rglwidget()
```

There is overlap between males and females, but small trend can be observed in
PC1 vs PC2!


In next PCA, study samples are colored by case/control definition:

```{r}

plot3d(scPCA.scores[,1], scPCA.scores[,2], scPCA.scores[,3], xlab=paste("PC 1(",
       scPCA.variances[1], "%)", sep = ""), ylab = paste("PC 2(",
       scPCA.variances[2], "%)", sep = ""), zlab = paste("PC 3(",
       scPCA.variances[3], "%)", sep = ""), pch = 19,
       col = as.integer(protSS_data$case))
rglwidget()
```

Also here, there is moderate shift beetween 2 groups!


# General clustering of proteins

In this section we aim to identify proteins with highly correlated (or anti
correlated) concentrations to potentially get some more detailed insides into
the dataset. For that, Pearson correlation coefficients between all proteins are
computed and visualized.  To detect groups with correlated proteins which might
belong from the same class, correlation network analysis is performed.


## Correlation analyses

Pearson's correlation coefficients between all proteins are computed, grouped
and visualized:

```{r, corr-plot-features, fig.path = IMAGE_PATH, fig.width = 12, fig.height = 10}
cormat <- cor(protSS_data[,proteins], method = "pearson",
              use = "pairwise.complete.obs")
reorder_cormat <- function(cormat){
    ## Use correlation between variables as distance
    dd <- as.dist((1-cormat)/2)
    hc <- hclust(dd)
    cormat <- cormat[hc$order, hc$order]
}
## Get upper triangle of the correlation matrix
get_upper_tri <- function(cormat){
    cormat[lower.tri(cormat)]<- NA
    return(cormat)
}
# Reorder the correlation matrix
cormat <- reorder_cormat(cormat)
upper_tri <- get_upper_tri(cormat)
# Melt the correlation matrix
melted_cormat <- melt(upper_tri, na.rm = TRUE)
# Create a ggheatmap
ggpHMap <- ggplot(melted_cormat, aes(Var2, Var1, fill = value)) +
geom_tile(color = "white") + scale_fill_gradient2(low = "blue", high = "red",
mid = "white", midpoint = 0, limit = c(-1,1), space = "Lab",
name="Pearson\nCorrelation") + theme_minimal() +
theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 10, hjust = 1),
axis.text.y = element_text(vjust = 1, size = 10, hjust = 1),
axis.title.x = element_blank(), axis.title.y = element_blank(),
panel.grid.major = element_blank(), panel.border = element_blank(),
panel.background = element_blank(), axis.ticks = element_blank(),
legend.justification = c(1, 0), legend.position = c(0.6, 0.7),
legend.direction = "horizontal") + coord_fixed()
ggplotly(ggpHMap)
```

There is a large number of proteins and figure is not very clear. Please zoom in
the part of figure which you want to analyse further.
All correlation coefficients are summariezed in table and ordered according
the Pearson correlation coefficients:

```{r, message = FALSE}
CorTable <- melted_cormat[order(melted_cormat[,3],decreasing=TRUE),]
colnames(CorTable) <- c("Prot. 1", "Prot. 2", "Corr")
## exclude the results between the same metabolites
CorTable_2 <- CorTable[-c(which(CorTable[, 1] == CorTable[,2])),]
datatable(CorTable_2, caption = "Pearson correlations between all proteins") %>%
formatRound(columns = "Corr", digits = 4)
```

In addition we cluster the proteins based on their pairwise correlation
coefficient.

```{r corr-heatmap-features, fig.path = IMAGE_PATH, fig.width = 16, fig.height = 14}
cormat2 <- cormat
rownames(cormat2) <- protSS_ann[rownames(cormat),"long_description"]
pheatmap(cormat2)
```


## Correlation network analysis

To identify groups of highly corraleted proteins, correlation network analysis
is performed. Weighted gene coexpression network analysis (WGCNA) implemented in
the WGCNA R package is used for determining groups (modules) of higlhy
correlated proteins
[@langfelderWGCNAPackageWeighted2008][@langfelderFastFunctionsRobust2012].
Modules are identified using unsupervised clustering Dynamic Branch Cut method
with a minimal number of 2 proteins per module.
Note that modules are randomly assigned colors, while unassigned proteins are
grouped in the grey module.
Results are visualized with hierarchical clustering dendrogram and correlations
between all proteins grouped together are computed.

In next figure, there is a hierarchical cluster tree of the 148 proteins,
color-coded according to the identified modules (groups):

```{r WGCNA_figure, fig.path = IMAGE_PATH, results='hide', message=FALSE, warning=FALSE, fig.width=32, fig.height=20, out.width='1600px'}
## WGCNA
# Choose a set of soft thresholding powers
powers <- c(1:20) # in practice this should include powers up to 20.
# choose power based on SFT criterion
sft <- pickSoftThreshold(protSS_data[,proteins], powerVector = powers,
                         networkType = "unsigned")
## https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/faq.html
## point 6: low connectivity; soft-thresholding power = 6
adjacencyA1 <- adjacency(protSS_data[,proteins], power = 6,
                         type = "unsigned")
diag(adjacencyA1) <- 0
dissTOMA1 <- 1 - TOMsimilarity(adjacencyA1, TOMType = "unsigned")
colnames(dissTOMA1) <- colnames(adjacencyA1)
biomTreeA1 <- flashClust(as.dist(dissTOMA1), method = "average")
mColorh <- NULL
tree <- cutreeHybrid(dendro = biomTreeA1, pamStage = FALSE, minClusterSize = 2,
                     cutHeight = 0.99, deepSplit = 3, distM = dissTOMA1)
mColorh <- cbind(mColorh,labels2colors(tree$labels))
plotDendroAndColors(biomTreeA1, mColorh, paste("dpSplt =", 3),
                    main = "Hierarchical cluster",dendroLabels = NULL)
```

If parameters are in the same and long branch, means that they are correlated.
Because of large number of proteins with a long names figure is not clear, thus
tables are provided.
Numbers of the proteins associatet to each module are presented here:

```{r}
# To take out module information
GroupVar <- cbind(proteins, mColorh)
orderedGV <- GroupVar[order(GroupVar[, 2]), ]
## number of variables per module:
sort(table(orderedGV[, 2]))
```

We next create tables for all groups of correlated proteins. Shown are the
long_description and the mean correlation of the protein against all proteins
within the cluster (including itself).

```{r, results = "asis"}
clusters <- split(GroupVar[, 1], GroupVar[, 2])
clusters <- clusters[order(lengths(clusters))]

clust_table <- function(x) {
    cormat_x <- cormat[x, x]
    data.frame(
        Class  = protSS_ann[x,"long_description"],
        mean_cor = rowMeans(cormat_x))
}

for (cluster in clusters) {
    pandoc.table(clust_table(cluster), style = "rmarkdown", split.table = "Inf")
}
```

Most of the clusters with highly correlated proteins make biologically
sense.

In next table, identified correlated proteins per module with correlation
coefficients are presented:

```{r}
##  Take module membership info and check correlations per module!
corPerModul <- function (Data, colour) {
colourModule <- round(cor(Data[,orderedGV[which(orderedGV[,2]==colour),1]],
                          use = "pairwise.complete.obs"), 4)
colourModule[lower.tri(colourModule)] <- NA
diag(colourModule) <- NA
colourModule = as.data.frame(as.table(colourModule))
colourModule = na.omit(colourModule)
colourModule[,"Module"] <- colour
return(colourModule)
}
cluster_colors <- unique(mColorh)
cluster_colors <- cluster_colors[cluster_colors != "grey"]
CorrPerModule <- do.call(rbind.data.frame,
                         lapply(cluster_colors, function(z)
                             corPerModul(protSS_data[,proteins], z)))
colnames(CorrPerModule)[3] <- "Corr"
datatable(CorrPerModule, caption = "Correlations per module")
```

Seems that there are moderate to strong correlations between proteins which
belongs to the same module (group).


# Session information

R packages used for the analysis:

```{r}
sessionInfo()
```

# References
