---
title: "NAFLD_T2D_sensititvity_analysis"
author: "Marilyn De Graeve, Johannes Rainer, Christoph Grander"
affiliation: "Eurac Research, Bolzano, Italy"
date: "2025-02-10"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
---

**Modified**: `r file.info("NAFLD_T2D_sensititvity_analysis.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r style, message = FALSE, echo = FALSE, warning = FALSE, results = "asis"}
library("BiocStyle")
library("kableExtra")
library("knitr")
library("rmarkdown")
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now, units = "secs")
      # return a character string to show the time
      paste("    Time for this code chunk to run:", round(res,
        2), "seconds")
    }
  }
}))
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7, time_it = TRUE)

```


# Introduction

During this workflow, sensitivity analyses are performed to investigate
influence and potential confounding of T2D on the steatosis results:
coefficients for steatosis were compared between the standard linear model and
a model accounting for T2D. 

For this, the original results are loaded of the saved lm results file for
fibrosis and steatosis The lms with T2D as added covariate are calculated there. 
For this, Rdata is loaded here. Loaded (wo T2D in covariates) and calculated
models (with T2D in covariates) are afterwards compared here.

In addition, steatosis associations were performed
separately on participants with (n = 103) or without T2D (n = 219). FINDME!!

# Setup

## Directories

```{r directories, echo = FALSE}
#' General settings
filename <- " NAFLD_T2D_sensititvity_analysis"

#' Path to the data
#No, manualy done below

```

## Packages

```{r packages, message=FALSE}
#' Load libraries
library(ggplot2)
library(pander)
library(plotly)

```


# 1 Fibrosis - tprot

## load lm results (wo T2D in covariates)

```{r}
#' get coef values from lm model
handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_fibrosis_explore/lm_output_results_prot.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_none <- results$coef_liver_stiffness_cat23

handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_fibrosis_explore/lm_output_results_prot_t2d.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_T2D <- results$coef_liver_stiffness_cat23

#rownames(results) <- results$Genes

```

## Sensitivity analysis

The coeficients (MASLD vs none) from the linear model to evaluate SLD
types - original is compared to the linear model to evaluate SLD types - with
BMI, using a sensitivity analysis.

```{r}
#' merge coef values from both lm models above
sensit <- data.frame(featureid = rownames(results),
                  x = lm_T2D,
                  y = lm_none)
#head(sensit)

#' prep for plotting
PLOTLY <- TRUE
ggplotly_or_not <- function(x, interactive = TRUE) {
    if (interactive) ggplotly(x)
    else print(x)
}

#' scatterplot with coeff vales
scatter_plot <- function(data) {
    ggplot(data, aes(x = x, y = y)) +
        geom_point(aes(x,y), colour = "gray65") +
        stat_smooth(method = lm, se = TRUE) + #with standard error
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) +
        labs(title = 'Fibrosis tprot coefficients between datasets:',
             x = 'lm model - MASLD',
             y = 'lm model - none') +
        theme_bw ()
}

s_1 <- scatter_plot(sensit)
ggplotly_or_not(s_1, PLOTLY)

```

```{r, eval = FALSE}
#' FC for see diff between 2 datasets
sensit$FC <- abs(sensit$x + 0.00000001) / abs(sensit$y + 0.00000001)
sensit$log2FC <- log2(abs(sensit$FC))
# 0.5 and > 2 intersting, NO too many
#summary(sensit)
#dim(sensit)
hist(sensit$log2FC)
#sensit <- sensit[sensit$log2FC < 0.5 | sensit$log2FC > 2.0, ]  #94 of 148 left, maybe more diff?
#sensit <- sensit[sensit$FC > 25, ] #5 left
sensit <- sensit[sensit$log2FC < -5 | sensit$log2FC > +5, ]  #10 left

#' table with highest dev coef
pandoc.table(
    sensit, style = "rmarkdown", split.table = Inf,
    caption = paste0("Coeficients of lm models using proteins ",
                     "between MASLD and control (none) participants",
                     ", only log2FC < -5 and log2FC > +5"))

```


## Table with differences

```{r, results = "asis"}
#add difference column
sensit$diff <- abs(sensit$x - sensit$y)
sensit <- sensit[order(sensit$diff, decreasing = TRUE), ]

pandoc.table(sensit[1:20, ], style = "rmarkdown",
             split.table = Inf, caption = "Top 20 different variable coeficients between lm model with- and without T2D as covariates")

```


# 2 Fibrosis - tmetab

## load lm results

```{r}
#' get coef values from both lm models
handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_fibrosis_explore/lm_output_results_metab.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_none <- results$coef_liver_stiffness_cat23

handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_fibrosis_explore/lm_output_results_metab_t2d.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_T2D <- results$coef_liver_stiffness_cat23

#rownames(results) <- results$analyte_name

```


## Sensitivity analysis

The coeficients (MASLD vs none) from the linear model to evaluate SLD
types - original is compared to the linear model to evaluate SLD types - with
BMI, using a sensitivity analysis.

```{r}
#' merge coef values from both lm models above
sensit <- data.frame(featureid = rownames(results),
                  x = lm_T2D,
                  y = lm_none)
#head(sensit)

#' scatterplot with coeff vales
scatter_plot <- function(data) {
    ggplot(data, aes(x = x, y = y)) +
        geom_point(aes(x,y), colour = "gray65") +
        stat_smooth(method = lm, se = TRUE) + #with standard error
        #geom_text(x = -Inf, y = Inf, label = equation, parse = TRUE, vjust = 1.5, hjust = -0.1) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) +
        labs(title = 'Fibrosis tmetab coefficients between datasets:',
             x = 'lm model - MASLD',
             y = 'lm model - none') +
        theme_bw ()
}

s_1 <- scatter_plot(sensit)
ggplotly_or_not(s_1, PLOTLY)

```

```{r, eval = FALSE}
#' FC for see diff between 2 datasets
sensit$FC <- abs(sensit$x + 0.00000001) / abs(sensit$y + 0.00000001)
sensit$log2FC <- log2(abs(sensit$FC))
# 0.5 and > 2 intersting, NO too many
#summary(sensit)
#dim(sensit)
hist(sensit$log2FC)
#sensit <- sensit[sensit$log2FC < 0.5 | sensit$log2FC > 2.0, ]  #94 of 148 left, maybe more diff?
#sensit <- sensit[sensit$FC > 25, ] #5 left
sensit <- sensit[sensit$log2FC < -5 | sensit$log2FC > +5, ]  #10 left

#' table with highest dev coef
pandoc.table(
    sensit, style = "rmarkdown", split.table = Inf,
    caption = paste0("Coeficients of lm models using proteins ",
                     "between MASLD and control (none) participants",
                     ", only log2FC < -5 and log2FC > +5"))

```


## Table with differences

```{r, results = "asis"}
#add difference column
sensit$diff <- abs(sensit$x - sensit$y)
sensit <- sensit[order(sensit$diff, decreasing = TRUE), ]

pandoc.table(sensit[1:20, ], style = "rmarkdown",
             split.table = Inf, caption = "Top 20 different variable coeficients between lm model with- and without T2D as covariates")

```


# 3 Steatosis - tprot

## load lm results

```{r}
#' get coef values from both lm models
handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_steatosis_explore/lm_output_results_proteins.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_T2D <- results$coef_liver_fat_cat22

handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_steatosis_explore/lm_output_results_proteins_t2d.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_none <- results$coef_liver_fat_cat22

#rownames(results) <- results$Genes

```


## Sensitivity analysis

The coeficients (MASLD vs none) from the linear model to evaluate SLD
types - original is compared to the linear model to evaluate SLD types - with
BMI, using a sensitivity analysis.

```{r}
#' merge coef values from both lm models above
sensit <- data.frame(featureid = rownames(results),
                  x = lm_T2D,
                  y = lm_none)
#head(sensit)

#' scatterplot with coeff vales
scatter_plot <- function(data) {
    ggplot(data, aes(x = x, y = y)) +
        geom_point(aes(x,y), colour = "gray65") +
        stat_smooth(method = lm, se = TRUE) + #with standard error 
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) +
        labs(title = 'Steatosis tprot coefficients between datasets:',
             x = 'lm model - MASLD',
             y = 'lm model - none') +
        theme_bw ()
}

s_1 <- scatter_plot(sensit)
ggplotly_or_not(s_1, PLOTLY)

```

```{r, eval = FALSE}
#' FC for see diff between 2 datasets
sensit$FC <- abs(sensit$x + 0.00000001) / abs(sensit$y + 0.00000001)
sensit$log2FC <- log2(abs(sensit$FC))
# 0.5 and > 2 intersting, NO too many
#summary(sensit)
#dim(sensit)
hist(sensit$log2FC)
#sensit <- sensit[sensit$log2FC < 0.5 | sensit$log2FC > 2.0, ]  #94 of 148 left, maybe more diff?
#sensit <- sensit[sensit$FC > 25, ] #5 left
sensit <- sensit[sensit$log2FC < -5 | sensit$log2FC > +5, ]  #10 left

#' table with highest dev coef
pandoc.table(
    sensit, style = "rmarkdown", split.table = Inf,
    caption = paste0("Coeficients of lm models using proteins ",
                     "between MASLD and control (none) participants",
                     ", only log2FC < -5 and log2FC > +5"))

```


## Table with differences

```{r, results = "asis"}
#add difference column
sensit$diff <- abs(sensit$x - sensit$y)
sensit <- sensit[order(sensit$diff, decreasing = TRUE), ]

pandoc.table(sensit[1:20, ], style = "rmarkdown",
             split.table = Inf, caption = "Top 20 different variable coeficients between lm model with- and without T2D as covariates")

```


# 4 Steatosis - tmetab

## load lm results

```{r}
#' get coef values from both lm models
handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_steatosis_explore/lm_output_results_metabolites.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_T2D <- results$coef_liver_fat_cat22

handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_steatosis_explore/lm_output_results_metabolites_t2d.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_none <- results$coef_liver_fat_cat22

#rownames(results) <- results$analyte_name

```


## Sensitivity analysis

The coeficients (MASLD vs none) from the linear model to evaluate SLD
types - original is compared to the linear model to evaluate SLD types - with
BMI, using a sensitivity analysis.

```{r}
#' merge coef values from both lm models above
sensit <- data.frame(featureid = rownames(results),
                  x = lm_T2D,
                  y = lm_none)
#head(sensit)

#' scatterplot with coeff vales
scatter_plot <- function(data) {
    ggplot(data, aes(x = x, y = y)) +
        geom_point(aes(x,y), colour = "gray65") +
        stat_smooth(method = lm, se = TRUE) + #with standard error
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) +
        labs(title = 'Steatosis tmetab coefficients between datasets:',
             x = 'lm model - MASLD',
             y = 'lm model - none') +
        theme_bw ()
}

s_1 <- scatter_plot(sensit)
ggplotly_or_not(s_1, PLOTLY)

```

```{r, eval = FALSE}
#' FC for see diff between 2 datasets
sensit$FC <- abs(sensit$x + 0.00000001) / abs(sensit$y + 0.00000001)
sensit$log2FC <- log2(abs(sensit$FC))
# 0.5 and > 2 intersting, NO too many
#summary(sensit)
#dim(sensit)
hist(sensit$log2FC)
#sensit <- sensit[sensit$log2FC < 0.5 | sensit$log2FC > 2.0, ]  #94 of 148 left, maybe more diff?
#sensit <- sensit[sensit$FC > 25, ] #5 left
sensit <- sensit[sensit$log2FC < -5 | sensit$log2FC > +5, ]  #10 left

#' table with highest dev coef
pandoc.table(
    sensit, style = "rmarkdown", split.table = Inf,
    caption = paste0("Coeficients of lm models using proteins ",
                     "between MASLD and control (none) participants",
                     ", only log2FC < -5 and log2FC > +5"))

```


## Table with differences

```{r, results = "asis"}
#add difference column
sensit$diff <- abs(sensit$x - sensit$y)
sensit <- sensit[order(sensit$diff, decreasing = TRUE), ]

pandoc.table(sensit[1:20, ], style = "rmarkdown",
             split.table = Inf, caption = "Top 20 different variable coeficients between lm model with- and without T2D as covariates")

```


# 5 Steatosis T2Dsplit w vs wo - tprot

## load lm results

```{r}
#' get coef values from both lm models
handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_steatosis_T2Dsplit/lm_output_results_proteinsW.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_T2D <- results$coef_liver_fat_cat22

handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_steatosis_T2Dsplit/lm_output_results_proteinsWO.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_none <- results$coef_liver_fat_cat22

#rownames(results) <- results$Genes

```


## Sensitivity analysis

The coeficients (MASLD vs none) from the linear model to evaluate SLD
types - original is compared to the linear model to evaluate SLD types - with
BMI, using a sensitivity analysis.

```{r}
#' merge coef values from both lm models above
sensit <- data.frame(featureid = rownames(results),
                  x = lm_T2D,
                  y = lm_none)
#head(sensit)

#' scatterplot with coeff vales
scatter_plot <- function(data) {
    ggplot(data, aes(x = x, y = y)) +
        geom_point(aes(x,y), colour = "gray65") +
        stat_smooth(method = lm, se = TRUE) + #with standard error 
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) +
        labs(title = 'Steatosis tprot coefficients between datasets:',
             x = 'lm model - MASLD',
             y = 'lm model - none') +
        theme_bw ()
}

s_1 <- scatter_plot(sensit)
ggplotly_or_not(s_1, PLOTLY)

```

```{r, eval = FALSE}
#' FC for see diff between 2 datasets
sensit$FC <- abs(sensit$x + 0.00000001) / abs(sensit$y + 0.00000001)
sensit$log2FC <- log2(abs(sensit$FC))
# 0.5 and > 2 intersting, NO too many
#summary(sensit)
#dim(sensit)
hist(sensit$log2FC)
#sensit <- sensit[sensit$log2FC < 0.5 | sensit$log2FC > 2.0, ]  #94 of 148 left, maybe more diff?
#sensit <- sensit[sensit$FC > 25, ] #5 left
sensit <- sensit[sensit$log2FC < -5 | sensit$log2FC > +5, ]  #10 left

#' table with highest dev coef
pandoc.table(
    sensit, style = "rmarkdown", split.table = Inf,
    caption = paste0("Coeficients of lm models using proteins ",
                     "between MASLD and control (none) participants",
                     ", only log2FC < -5 and log2FC > +5"))

```


## Table with differences

```{r, results = "asis"}
#add difference column
sensit$diff <- abs(sensit$x - sensit$y)
sensit <- sensit[order(sensit$diff, decreasing = TRUE), ]

pandoc.table(sensit[1:20, ], style = "rmarkdown",
             split.table = Inf, caption = "Top 20 different variable coeficients between lm model of ppl with and ppl wo T2D")

```


# 6 Steatosis T2Dsplit w vs wo - tmetab

## load lm results

```{r}
#' get coef values from both lm models
handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_steatosis_T2Dsplit/lm_output_results_metabolitesW.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_T2D <- results$coef_liver_fat_cat22

handle <- "/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_steatosis_T2Dsplit/lm_output_results_metabolitesWO.txt"
results <- read.csv(handle, header = TRUE, sep = "\t")
dim(results)
lm_none <- results$coef_liver_fat_cat22

#rownames(results) <- results$analyte_name

```


## Sensitivity analysis

The coeficients (MASLD vs none) from the linear model to evaluate SLD
types - original is compared to the linear model to evaluate SLD types - with
BMI, using a sensitivity analysis.

```{r}
#' merge coef values from both lm models above
sensit <- data.frame(featureid = rownames(results),
                  x = lm_T2D,
                  y = lm_none)
#head(sensit)

#' scatterplot with coeff vales
scatter_plot <- function(data) {
    ggplot(data, aes(x = x, y = y)) +
        geom_point(aes(x,y), colour = "gray65") +
        stat_smooth(method = lm, se = TRUE) + #with standard error
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) +
        labs(title = 'Steatosis tmetab coefficients between datasets:',
             x = 'lm model - MASLD',
             y = 'lm model - none') +
        theme_bw ()
}

s_1 <- scatter_plot(sensit)
ggplotly_or_not(s_1, PLOTLY)

```

```{r, eval = FALSE}
#' FC for see diff between 2 datasets
sensit$FC <- abs(sensit$x + 0.00000001) / abs(sensit$y + 0.00000001)
sensit$log2FC <- log2(abs(sensit$FC))
# 0.5 and > 2 intersting, NO too many
#summary(sensit)
#dim(sensit)
hist(sensit$log2FC)
#sensit <- sensit[sensit$log2FC < 0.5 | sensit$log2FC > 2.0, ]  #94 of 148 left, maybe more diff?
#sensit <- sensit[sensit$FC > 25, ] #5 left
sensit <- sensit[sensit$log2FC < -5 | sensit$log2FC > +5, ]  #10 left

#' table with highest dev coef
pandoc.table(
    sensit, style = "rmarkdown", split.table = Inf,
    caption = paste0("Coeficients of lm models using proteins ",
                     "between MASLD and control (none) participants",
                     ", only log2FC < -5 and log2FC > +5"))

```


## Table with differences

```{r, results = "asis"}
#add difference column
sensit$diff <- abs(sensit$x - sensit$y)
sensit <- sensit[order(sensit$diff, decreasing = TRUE), ]

pandoc.table(sensit[1:20, ], style = "rmarkdown",
             split.table = Inf, caption = "Top 20 different variable coeficients between lm model of ppl with and ppl wo T2D")

```


# Acknowledgments

- NAFLD_metab_targeted_lm
- NAFLD_fibrosis_prot_lm, etc.


# Session information

R packages used for the analysis:

```{r}
sessionInfo()
```


