---
title: "Mummichog annotation on NAFLD umetab (pos) after lm with liver_fat_cat2"
author: "Marilyn De Graeve, Johannes Rainer"
affiliation: "Eurac Research, Bolzano, Italy"
date: "2024-12-13"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
---

**Modified**: `r file.info("NAFLD_umetabpos_mummichog_annotations.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r configuration, include = TRUE, cached = FALSE}

#' introduction
STUDY <- "NAFLD cohort"
TARGET <- "liver_fat_cat2"
DATASET <- "untargeted metabolomics, positive mode"

#' settings
FILENAME <- "NAFLD_umetabpos_naab13cat2"
LOCATION <- 'local' #'server' OR 'local', change in chunk directories if needed

#' target variables
#print('target of the linear regression model:')
TARGET_COLNAME <- 'liver_fat_cat2'
#print('more concrete defined target of the linear regression model:')
TARGET_LMNAME <- 'liver_fat_cat22'

```


```{r style, echo = FALSE, results = "hide", message = FALSE}
#' rmarkdown format settings
library("BiocStyle")
library("kableExtra")
library("knitr")
library("rmarkdown")
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7, time_it = TRUE)
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now, units = "secs")
      # return a character string to show the time
      paste("    Time for this code chunk to run:", round(res,
        2), "seconds")
    }
  }
}))
BiocStyle::markdown()

```

# Introduction

In this document, the Mummichog algorithm from *todo!!!* are performed on the
significant metabolite list on the targeted metabolomics data after statistical
analyis (linear models or t-test todo!!!) from the CHRIS obesity project.

As POC, we will use the lm results table from:
(metabo_chris_untargeted/CHRIS_lm_pos.html)[metabo_chris_untargeted/CHRIS_lm_pos.html].
There, the exploratory statistical analyis (linear models) are performed on the
untargeted metabolomics data from the CHRIS study (baseline); using the
continious variable "BMI" as a target of interest.

The (static) table with the significant metabolites only contained 11
significant variables, maybe not enough to leverage the power of the Mummichog
algoritm. ??todo with testdata first!!! => NO, input is full list, so also give
the not-significant w their values. 

This document will be TEMPLATE, to be used for different datasets - target
significant tables.


# Setup

## Directories

```{r directories}
#' General settings
filename <- "Mummichog_annotations"

#' Path to the data
DATA_PATH <- '/home/mdegraeve/Documents/Files/Work_Eurac/Projects/biochem_pathway_analysis/data' #FINDME!
#DATA_OUTPUT_PATH <- "data"

```

## Packages

```{r packages}


```

## functions

```{r functions}


```

# Steps

The following steps are performed:

- [x] in the metabo_chris_untargeted/CHRIS_lm_pos.html: write table with the ALL
  (both significant and non-significant) metabolites with mz, rt, id, p-value (=
  adj.p lm) and statistic (= coef lm) after performing lm with `r TARGET` as
  target, using the `R DATASET` dataset.
- [x] rename colnames so c(m/z, retention_time, p_value, statistic,
  CompoundID_from_user, ...)
- [ ] run mummichog 
  - [x] (now via bash, 
  - [ ] future w spectripy)
- [x] load and cleanup results into 1 df.
- [ ] add results as putative IDs into the sumExp object and save as Rdata.


# Load and rename results from lm

load the results from lm analysis. write table with the ALL
(both significant and non-significant) metabolites with mz, rt, id, p-value (=
  adj.p lm) and statistic (= coef lm) after performing lm with `r TARGET` as
  target, using the `R DATASET` dataset.

```{r load-results-lm}

#' example data: 
#    - dataset: umetab pos chris (3820#, come from umExp_chris_filtered_norm.RData) 
#    - target: bmi (float, continious)
#    - covariates: sex, age 
#' after lm: 11 metab signif eg FT01399

#' load data lm-results
handle <- file.path(DATA_PATH, 'lm_output_results_metab.txt')
lm_output_results_metab <- read.csv(handle, header = TRUE, sep="\t")  |>
    as.data.frame()
lm_output_results_metab[1:10, 1:10]
dim(lm_output_results_metab)
```

rename colnames to c(m/z, retention_time, p_value, statistic,
CompoundID_from_user, ...) so it fits the input in mummichog, write the table.

```{r make-write-mummichog-input}
#' make new df with correct names for mummichog
df <- NULL
df$m.z  <- lm_output_results_metab$mzmed
df$retention_time  <- lm_output_results_metab$rtmed
df$p_value  <- lm_output_results_metab[, paste0('p.adj_', TARGET_LMNAME)]
df$statistic  <-lm_output_results_metab[, paste0('coef_', TARGET_LMNAME)]
df$CompoundID_from_user <- lm_output_results_metab$AID
df <- as.data.frame(df)
head(df)

#TODO m/z ipv m.z ...!!!! ok?


#' write the mummichog input table
handle <- file.path(DATA_PATH, 'mummichog_input.txt')
write.table(df, file = handle, sep = "\t", row.names = FALSE, col.names = TRUE)

```


# Run mummichog algoritm 

## Run mummichog algoritm using Bash

Before trying to integrate the mummichog algorithm using spectripy, we first
want to run it automatically from this script. it is run locally, on a Linux
computer with the Mummichog installed according to the instructions from
'https://github.com/shuzhao-li/mummichog'.

After install using pip in the linux terminal, it is run using commands:

- cd Documents/Files/Work_Eurac/Projects/biochem_pathway_analysis/
- python3 -m mummichog.main -f testdata0710.txt -o t2

the results are saved in the same folder, as the input. you will get a folder
with various output. in the first instance, i.e. for the putative annotation of
untargeted metabolic features, we are interested in 1 tabular output. See
section [Results import] below.

run the command using R (calls bash), for linux:

```{r run-mummichog-from-bash}

#TODO!!!
#command <- 'python3.10 -m mummichog.main -f ~/Documents/Files/Work_Eurac/Projects/biochem_pathway_analysis/testdata0710.txt -o t2' #test
#command <- 'bash ~/Documents/Files/Work_Eurac/Code/Sh_scripts/mummichog2.sh'
command <- 'python3.10 -m mummichog.main -f ~/Documents/Files/Work_Eurac/Projects/biochem_pathway_analysis/data/mummichog_input.txt -o t2'

command
system(command)

```

This will be either called in the spectripy environment, if possible. will try
that next. therefor, no need to go too much into win/linux differences (see old
code marilyn if needed).


# Results from mummichog (pasted from html report)

For the NAFLD untargeted metabolomics - pos mode lm model with MAFLD sld type, 
pasted from the mummichog html report:


Mummichog Report: t2
mummichog version: 2.7.0, using metabolic model MFN_1.10.4.
User input data

User supplied 10882 features as reference list.We are using 2700 features (p < 0.604157) as significant list. The feature level data are shown in the Manhattan plots below.
Top pathways

The reference feature list is mapped to 532 EmpiricalCompounds, significant features to 155 EmpiricalCompounds. In the table below, pathway_size is number of detected EmpiricalCompounds for each pathway; overlap_size is number of significant EmpiricalCompounds. Empirical p-values are estimated by permutation test. Details on EmpiricalCompounds are in 'tables/ListOfEmpiricalCompounds.tsv'.
Pathways 	overlap_size	pathway_size	p-value 	overlap_EmpiricalCompounds
Alanine and Aspartate Metabolism 	7	11	0.00563	E13,E421,E46,E210,E14,E924,E274
Ubiquinone Biosynthesis 	4	5	0.00815	E442,E924,E913,E1001
Lysine metabolism 	11	22	0.00874	E13,E345,E277,E177,E262,E179,E178,E14,E1001,E924,E265
Prostaglandin formation from arachidonate 	10	20	0.01244	E1055,E616,E951,E1020,E440,E497,E943,E495,E557,E924
Vitamin B3 (nicotinate and nicotinamide) metabolism 	6	10	0.01311	E13,E210,E14,E1001,E924,E274
Phosphatidylinositol phosphate metabolism 	8	15	0.01311	E914,E705,E1247,E448,E450,E457,E916,E876
N-Glycan Degradation 	6	10	0.01311	E448,E644,E643,E450,E457,E352
Vitamin A (retinol) metabolism 	5	8	0.01714	E735,E827,E1247,E924,E975
Glycine, serine, alanine and threonine metabolism 	16	38	0.01807	E924,E134,E13,E140,E143,E463,E14,E138,E464,E173,E139,E1001,E135,E209,E650,E192
Nitrogen metabolism 	4	6	0.02185	E14,E210,E13,E274
Histidine metabolism 	9	19	0.02378	E13,E152,E14,E153,E653,E1001,E924,E732,E650
Valine, leucine and isoleucine degradation 	9	19	0.02378	E924,E13,E140,E185,E14,E138,E9,E139,E135
Urea cycle/amino group metabolism 	17	42	0.02437	E130,E416,E13,E177,E262,E46,E210,E115,E443,E179,E178,E14,E229,E1001,E209,E650,E192
Keratan sulfate degradation 	3	4	0.02764	E643,E352,E644
Butanoate metabolism 	7	14	0.0284	E924,E13,E140,E14,E138,E139,E135
Purine metabolism 	12	28	0.02891	E740,E65,E302,E13,E210,E14,E788,E488,E213,E924,E650,E274
Fatty acid oxidation 	2	2	0.03033	E913,E1271
Omega-3 fatty acid metabolism 	2	2	0.03033	E924,E817
Arginine and Proline Metabolism 	11	26	0.03739	E130,E13,E177,E262,E171,E46,E210,E179,E178,E14,E274
Arachidonic acid metabolism 	6	12	0.03865	E13,E951,E943,E14,E964,E924
Caffeine metabolism 	4	7	0.04983	E445,E449,E450,E448
Top modules

Network modules may capture concerted metabolite activities that are missed by predefined pathways. Use pull-down menus to visualize modules and the activity network, which tries to combine the results of pathway/module analyses. The visualization requires interenet connection.
Showing
Node size Visual style
module_1, p=0.04307, 87 metabolites
D-Proline, L-Proline, L-4-Hydroxyglutamate semialdehyde, AMP, L-2-Amino-3-oxobutanoic acid; L-2-Amino-3-oxobutanoate; L-2-Amino-acetoacetate; (S)-2-Amino-3-oxobutanoic acid, L-Threonine, L-Valine, L-Glutamate, 5-Oxoproline, L-1-Pyrroline-3-hydroxy-5-carboxylate, L-Leucine, 4-Methyl-2-oxopentanoate, kinetensin 1-3, L-Isoleucine, L-Aspartate, L-Tryptophan, 3-(4-Hydroxyphenyl)pyruvate, 3-Methyl-2-oxopentanoate; 2-Oxo-3-methylvalerate, L-Citrulline; 2-Amino-5-ureidovaleric acid; Citrulline, N-Carbamoyl-L-aspartate, L-Glutamine, N-Acetyl-L-aspartate, Ornithine, N2-Acetyl-L-ornithine, Ornithine; 2,5-Diaminovaleric acid; 2,5-Diaminopentanoic acid; 2,5-Diaminopentanoate, 6-Amino-2-oxohexanoate; 2-Oxo-6-aminocaproate, L-Lysine, L-2-Aminoadipate 6-semialdehyde; 2-Aminoadipate 6-semialdehyde, L-2-Aminoadipate; L-alpha-Aminoadipate; L-alpha-Aminoadipic acid; L-2-Aminoadipic acid; L-2-Aminohexanedioate, 5-Phospho-beta-D-ribosylamine, 5-Formiminotetrahydrofolate; 5-Formimidoyltetrahydrofolate, leukotriene D5, N-Formyl-L-glutamate, Dihydropteroate; 7,8-Dihydropteroate, Tetrahydrofolyl-[Glu](n); Tetrahydrofolyl-[Glu](n+1); THF-polyglutamate; Tetrahydropteroyl-[gamma-Glu]n; Tetrahydropteroyl-[gamma-Glu]n+1, 5,6,7,8-Tetrahydrofolate, Xanthine, Inosine, Guanosine, Urate, Octanoic acid; Caprylic acid; Octylic acid, 5-Hydroxyisourate, D-Glucose, myo-Inositol, Melibiose, Lactose, Epimelibiose, Maltose, 1-alpha-D-Galactosyl-myo-inositol; 1-O-alpha-D-Galactosyl-D-myo-inositol; Galactinol, Sucrose, D-Mannose, Trehalose, 3-beta-D-Galactosyl-sn-glycerol; Galactosylglycerol, beta-D-Glucose, alpha-D-Glucose, beta-D-Fructose; beta-Fruit sugar; beta-D-arabino-Hexulose; beta-Levulose; Fructose, Isomaltose; Brachiose, Phosphatidic acid (phospholipids), N-Acetyl-D-glucosamine, 2-Hydroxy-3-(4-hydroxyphenyl)propenoate; 4-Hydroxy-enol-phenylpyruvate, D-Tryptophan, N-Acetyl-D-mannosamine; 2-Acetamido-2-deoxy-D-mannose, Thymidine, Uridine, dGMP, Prostaglandin A2; PGA2; Medullin, Prostaglandin E2, Prostaglandin C2; PGC2, Prostaglandin J2; PGJ2, Prostaglandin D2, Prostaglandin B2; PGB2, 6-(alpha-D-Glucosaminyl)-1D-myo-inositol, 11-dehydrothromboxane B2, 3alpha,7alpha,12alpha-Trihydroxy-5beta-cholestanoate, pentadecanoyl Coenzyme A, 13'-carboxy-alpha-tocopherol, , 13'-carboxy-gama-tocotrienol, (5Z,13E)-11alpha-Hydroxy-9,15-dioxoprost-13-enoate; (5Z)-(15S)-11alpha-Hydroxy-9,15-dioxoprosta-13-enoate, (5Z)-(15S)-11alpha-Hydroxy-9,15-dioxoprostanoate, Prostaglandin F2alpha; (5Z,13E)-(15S)-9alpha,11alpha,15-Trihydroxyprosta-5,13-dienoate; (+)-Prostaglandin F2a; 7-[3,5-Dihydroxy-2-(3-hydroxy-1-octenyl)cyclopentyl]-5-heptenoic acid; 9a,11a-PGF2; Amoglandin; Cyclosin; Dinoprost; Enzaprost; Enzaprost F; Panacelan; PGF2a; 9,11,15-Trihydroxy-(5Z,9a,11a,13E,15S)-prosta-5,13-dien-1-oic acid; Prostaglandin F2a; Prostin F 2 alpha; Protamodin; U 14583, Prostaglandin H2, 11-epi-Prostaglandin F2alpha; 11-epi-Prostaglandin F2a; 11-epi-PGF2alpha; 11-epi-PGF2a, Alprostadil; Prostaglandin E1; (13E)-(15S)-11alpha,15-Dihydroxy-9-oxoprost-13-enoate, (13E)-11alpha-Hydroxy-9,15-dioxoprost-13-enoate, Prostaglandin I2; (5Z,13E)-(15S)-6,9alpha-Epoxy-11alpha,15-dihydroxyprosta-5,13-dienoate; Prostacyclin; PGI2; Epoprostenol, Thromboxane A2; (5Z,13E)-(15S)-9alpha,11alpha-Epoxy-15-hydroxythromboxa-5,13-dienoate; (5Z,9alpha,11alpha,13E,15S)-9,11-Epoxy-15-hydroxythromboxa-5,13-dien-1-oic acid
EmpiricalCompounds favored by above analyses

An EmpiricalCompound is a computational unit for a tentative metabolite. It can group multiple ions, and be any of the isobaric/isomeric species. This table contains EmpiricalCompounds prioritized by pathway/module analysis.
EmpiricalCompound	Input m/z 	Retention time	ion	mz_diff	Statistic 	Significant
E788 	Best guess: C00387, Guanosine	C00387
	284.099148677354 	159.906221932334	M+H[1+]	0.0002	-0.08	no
	306.081033078351 	159.859620122684	M+Na[1+]	0.0001	-0.56	yes
E274 	Best guess: C00819, D-Glutamine	C00064, C05100, C00819
	169.058421834848 	175.584664990003	M+Na[1+]	0.0	-0.07	yes
	147.07714663581 	175.767827420671	M+H[1+]	0.0008	-0.05	yes
	148.079397135472 	175.77525914372	M(C13)+H[1+]	-0.0004	-0.04	no
E1271 	Best guess: ptdcacoa, pentadecanoyl Coenzyme A	ptdcacoa
	987.297366391483 	157.593298308763	M[1+]	-0.0	-0.33	yes
	988.300151362901 	157.457743742745	M+H[1+]	-0.005	-0.16	yes
	989.301392052679 	157.44211747464	M(C13)+H[1+]	-0.0072	-0.42	yes
E1159 	Best guess: CE7087, leukotriene D5	CE7087
	493.241140570286 	31.3497652946349	M[1+]	0.0044	-0.19	yes
	494.245150239307 	31.1978105551189	M+H[1+]	0.0006	0.0	no
E463 	Best guess: CE4968, isovalerylglycine	CE4970, CE4968
	181.071817352089 	35.7767213729386	M+Na[1+]	0.0008	0.54	yes
E464 	Best guess: CE4968, isovalerylglycine	CE4970, CE4968
	181.071893935303 	48.395716891956	M+Na[1+]	0.0009	0.77	yes
E83 	Best guess: C00525, D-Tryptophan	C00078, C00525
	205.097357202632 	151.125579809398	M+H[1+]	0.0002	0.08	yes
	206.100391669395 	151.584971894894	M(C13)+H[1+]	-0.0002	0.09	yes
	227.079219757571 	151.638186830096	M+Na[1+]	0.0	0.05	no
E277 	Best guess: C00047, L-Lysine	C00047
	169.094276546959 	186.749748975364	M+Na[1+]	-0.0005	0.09	no
	147.11317977405 	186.663332904751	M+H[1+]	0.0004	0.07	yes
	148.115911464333 	186.676263723386	M(C13)+H[1+]	-0.0003	0.06	no
E951 	Best guess: C14782, 11,12,15-THETA	C04843, C04741, C05959, C14814, C00639, C14782
	355.246812754374 	26.4828196127864	M+H[1+]	-0.0011	-0.12	yes
	423.236261966704 	26.7705082081122	M+HCOONa[1+]	0.001	0.04	no
E959 	Best guess: CE5787, kinetensin 1-3	CE5787
	382.229266736866 	46.2882791830158	M+Na[1+]	-0.0007	0.19	yes
E964 	Best guess: C14773, 8,9-DHET	C14774, C14775, C14772, C14773
	361.235110719539 	28.4093513503427	M+Na[1+]	0.0001	-0.03	no
	407.240831734866 	26.2537253231465	M+HCOONa[1+]	0.0005	0.14	yes
E557 	Best guess: C05963, 	C05963
	197.115137498282 	29.205428175067	M+H+Na[2+]	-0.0009	-0.22	no
	393.227197673302 	26.2659931314194	M+Na[1+]	0.0024	0.14	yes
E185 	Best guess: C03465, 3-Methyl-2-oxopentanoate	C00233, C06102, C03465
	199.058394448999 	28.5983334515759	M+HCOONa[1+]	0.0007	0.15	yes
	131.070444236393 	28.9561817678334	M+H[1+]	0.0002	0.08	no
E817 	Best guess: C16300, Stearidonic acid	C16300
	299.197511395329 	28.1538661302838	M+Na[1+]	-0.0007	0.25	yes
E1055 	Best guess: CE4980, Prostaglandin PGE2 glyceryl ester	CE4980
	427.265865360397 	26.2659931314194	M+H[1+]	-0.0032	0.08	yes
E302 	Best guess: C00385, Xanthine	C00385
	153.040635389064 	136.835601204689	M+H[1+]	-0.0	0.21	yes
E488 	Best guess: C11821, 5-Hydroxyisourate	C11821
	185.03055983449 	140.168844037611	M+H[1+]	-0.0	0.19	yes
	186.032803429111 	140.598836704487	M(C13)+H[1+]	-0.0012	-0.16	no
	207.012405662767 	140.874345026774	M+Na[1+]	-0.0002	0.05	no
E905 	Best guess: C00921, Dihydropteroate	C00921
	337.10531714664 	53.4866201835127	M+Na[1+]	0.0033	0.32	yes
E975 	Best guess: CE2961, 4-hydroxy-all-trans-retinyl acetate	CE2961
	367.227296190073 	28.5137755636823	M+Na[1+]	0.0029	-0.2	yes
E416 	Best guess: C00437, N2-Acetyl-L-ornithine	C00437
	175.108136486149 	167.735437364001	M+H[1+]	0.0005	0.01	no
	197.089817424914 	168.061432313048	M+Na[1+]	0.0001	-0.24	yes
E914 	Best guess: C05402, Melibiose	C00208, C00252, C00243, C01083, C00089, C01235, C05400, C05402
	342.113599195123 	162.860424626754	M[1+]	-0.0021	0.09	no
	343.12415051353 	167.866271582616	M+H[1+]	0.0007	0.25	yes
	365.106018541642 	167.78469939128	M+Na[1+]	0.0005	0.24	yes
E827 	Best guess: CE5013, 	CE5932, C02075, C06428, CE5013
	325.216105993762 	28.7004468294589	M+Na[1+]	0.0023	0.12	yes
E115 	Best guess: C00791, 	C00791
	136.047906253237 	166.019343990759	M+Na[1+]	-0.0003	0.32	yes
E705 	Best guess: C04299, D-myo-Inositol 1,2-cyclic phosphate	C04299
	243.027912158688 	207.286715425103	M+H[1+]	0.0014	-0.07	no
	244.03053453726 	207.256185483421	M(C13)+H[1+]	0.0007	-0.11	yes
E913 	Best guess: C00399, Ubiquinone	C00399
	341.171532876928 	32.9526989584018	M+Na[1+]	-0.0009	0.21	yes
	387.180755392711 	30.7682192018288	M+HCOONa[1+]	0.003	0.13	no
E495 	Best guess: CE1447, 11-dehydrothromboxane B2	CE5926, CE5930, CE2056, CE5534, CE5533, CE1447
	368.220037884916 	37.0349444783669	M+H[1+]	0.0007	0.14	yes
E497 	Best guess: CE1447, 11-dehydrothromboxane B2	CE5926, CE5930, CE2056, CE5534, CE5533, CE1447
	390.198554968854 	136.566127425071	M+Na[1+]	-0.0028	0.36	yes
E709 	Best guess: C00214, Thymidine	C00214
	265.081209040718 	129.401222712167	M+Na[1+]	0.0016	-0.38	yes
E192 	Best guess: C00300, Creatine	C00300
	154.058683137714 	165.985827414505	M+Na[1+]	-0.0001	0.39	yes
	132.077077955487 	165.875646360483	M+H[1+]	0.0003	0.14	yes
	133.079574927692 	165.881223095189	M(C13)+H[1+]	-0.0006	0.32	yes
E421 	Best guess: C01042, N-Acetyl-L-aspartate	C01045, C01042
	176.056298120512 	169.327142201084	M+H[1+]	0.0009	-0.1	no
	198.038224312001 	169.460765142265	M+Na[1+]	0.0008	-0.29	yes
E916 	Best guess: C15658, 6-(alpha-D-Glucosaminyl)-1D-myo-inositol	C15658
	342.139577181708 	187.169469108707	M+H[1+]	0.0001	0.26	yes
E130 	Best guess: C00148, L-Proline	CE4788, C00763, C00148
	116.071284919705 	189.644756681073	M+H[1+]	0.0007	0.0	no
	117.074698392449 	189.665850435534	M(C13)+H[1+]	0.0007	0.13	yes
E616 	Best guess: CE5730, 	CE5730
	409.254803811515 	26.4844670729491	M+H[1+]	-0.0037	0.15	yes
	477.245954048195 	26.5913439820278	M+HCOONa[1+]	0.0001	0.06	no
E432 	Best guess: C00438, N-Carbamoyl-L-aspartate	C00438
	178.054117210875 	159.205870495265	M(C13)+H[1+]	0.0001	-0.26	yes
	199.033243878992 	159.388305807604	M+Na[1+]	0.0007	-0.15	no
E135 	Best guess: C00183, L-Valine	C00719, C00431, C00183
	118.086804647355 	32.9362964155388	M+H[1+]	0.0005	0.38	yes
E209 	Best guess: C01602, Ornithine	C00077, C00515, C01602
	133.097773966659 	189.657109088885	M+H[1+]	0.0006	0.07	no
	134.100535435361 	189.665177541712	M(C13)+H[1+]	-0.0	0.09	yes
E924 	Best guess: C00362, dGMP	C00020, C00362
	347.065031186628 	175.77098476885	M[1+]	0.0025	-0.16	yes
	348.066998039671 	175.679483748155	M+H[1+]	-0.0034	-0.13	yes
E724 	Best guess: C00299, Uridine	C00299
	267.058842433434 	152.434294127664	M+Na[1+]	0.0001	-0.09	yes
E134 	Best guess: C03508, L-2-Amino-3-oxobutanoic acid	C00441, C03508
	118.049746600734 	52.221062708748	M+H[1+]	-0.0001	-0.68	yes
E1086 	Best guess: CE7072, 13'-carboxy-gama-tocotrienol	CE7072
	462.273718176345 	44.5203155431044	M+Na[1+]	-0.0004	0.36	yes
E140 	Best guess: C00183, L-Valine	C00719, C00431, C00183
	186.07516756858 	168.208323462175	M+HCOONa[1+]	0.0015	-0.16	yes
	140.068422504314 	167.895746409619	M+Na[1+]	0.0001	-0.08	no
E210 	Best guess: C00402, D-Aspartate	C00049, C00402
	134.045133649988 	178.13491741794	M+H[1+]	0.0004	0.13	yes
	135.048067520858 	178.38724570132	M(C13)+H[1+]	-0.0001	0.02	no
E440 	Best guess: CE5537, 	CE5928, CE5925, C05953, C05957, C05954, C05955, CE5535, CE5537
	335.222907660932 	45.3857366020265	M+H[1+]	0.0012	0.09	yes
E138 	Best guess: C00183, L-Valine	C00719, C00431, C00183
	118.086986533199 	156.325662772725	M+H[1+]	0.0007	-0.15	yes
	140.068674270398 	157.005328859031	M+Na[1+]	0.0004	-0.09	yes
E139 	Best guess: C00183, L-Valine	C00719, C00431, C00183
	118.087085786408 	162.545776621999	M+H[1+]	0.0008	0.08	yes
	119.090009346851 	162.367934974042	M(C13)+H[1+]	0.0003	0.01	no
E213 	Best guess: C00294, Inosine	C00294
	269.088267405768 	142.609612316691	M+H[1+]	0.0002	0.01	no
	270.091176060444 	142.789472665335	M(C13)+H[1+]	-0.0003	-0.25	yes
	291.07024352127 	142.966507264613	M+Na[1+]	0.0002	-0.03	no
E1001 	Best guess: C00021, S-Adenosyl-L-homocysteine	C00021
	384.120179196921 	167.12341231574	M[1+]	-0.0009	0.04	no
	385.127238511038 	167.250608939691	M+H[1+]	-0.0016	-0.06	no
	386.131115707301 	166.956281029636	M(C13)+H[1+]	-0.0012	-0.13	yes
	407.107984462153 	166.903479951716	M+Na[1+]	-0.0029	-0.12	no
E732 	Best guess: C00664, 5-Formiminotetrahydrofolate	C00664
	473.184949015672 	187.381259387352	M+H[1+]	-0.0042	0.23	yes
	473.185934623436 	181.324869460584	M+H[1+]	-0.0032	0.59	yes
E933 	Best guess: C04707, (5Z,13E)-11alpha-Hydroxy-9,15-dioxoprost-13-enoate	C04707, C06439
	351.213811984449 	35.4420398477052	M+H[1+]	-0.0028	0.26	yes
E735 	Best guess: C11061, retinoyl glucuronide	C11061
	250.118475071993 	27.046577397134	M+H+Na[2+]	-0.0003	-0.13	no
	477.245954048195 	26.5913439820278	M+H[1+]	-0.0023	0.06	no
	478.250504993974 	26.400778920538	M(C13)+H[1+]	-0.0012	0.29	yes
E740 	Best guess: C03090, 5-Phospho-beta-D-ribosylamine	C03090
	252.02554472671 	141.038031941081	M+Na[1+]	0.0012	-1.0	yes
E1103 	Best guess: C04722, 3alpha,7alpha,12alpha-Trihydroxy-5beta-cholestanoate	C04722
	473.324748111999 	138.237949520824	M+Na[1+]	0.001	-0.43	yes
E46 	Best guess: C00327, L-Citrulline	C00327
	176.103446232011 	178.627029922609	M+H[1+]	0.0005	-0.1	yes
	177.105896431171 	178.592407820409	M(C13)+H[1+]	-0.0005	-0.13	yes
	198.085122749498 	178.626196942728	M+Na[1+]	0.0001	-0.14	yes
E143 	Best guess: C00188, L-Threonine	C05519, C00188
	120.066290033161 	168.835003077402	M+H[1+]	0.0008	-0.03	no
	121.069309907678 	168.792187754212	M(C13)+H[1+]	0.0004	-0.03	no
	142.047694857258 	168.914330475562	M+Na[1+]	0.0002	-0.21	yes
E443 	Best guess: C01586, Hippurate	CE5536, C01586
	180.065504638412 	35.4004172163323	M+H[1+]	0.0	-0.37	yes
E442 	Best guess: C01179, 3-(4-Hydroxyphenyl)pyruvate	C05350, C01197, C01179
	180.042090938812 	128.466715841144	M[1+]	0.0003	0.24	no
	203.031653738617 	128.542209228999	M+Na[1+]	0.0001	-0.51	yes
E9 	Best guess: C00407, L-Isoleucine	C00123, C00407
	154.083826186935 	149.216792267092	M+Na[1+]	-0.0001	0.21	yes
	132.102514777352 	149.119199747165	M+H[1+]	0.0006	0.18	yes
	133.10538154778 	148.235535180792	M(C13)+H[1+]	0.0001	0.15	yes
E229 	Best guess: C02735, Phenylethanolamine	C00483, C01183, C02735
	137.082248303144 	36.6779886741168	M[1+]	-0.0013	-0.26	yes
	138.091851433665 	34.1326207541589	M+H[1+]	0.0005	-0.08	no
E445 	Best guess: C07480, 	CE5541, C13747, C07480
	180.065504638412 	35.4004172163323	M[1+]	0.0014	-0.37	yes
	181.071817352089 	35.7767213729386	M+H[1+]	-0.0002	0.54	yes
	182.075189814211 	35.7414999247126	M(C13)+H[1+]	-0.0002	0.49	yes
E345 	Best guess: C00956, L-2-Aminoadipate	C00956
	162.0762122647 	169.673156820422	M+H[1+]	0.0001	0.21	yes
E448 	Best guess: C00159, D-Mannose	C13747, C07480, C00031, C00221, C02336, C00267, C00962, C00795, C00137, C00984, C00159
	181.071893935303 	48.395716891956	M+H[1+]	0.0012	0.77	yes
	182.074863871563 	48.4397586734721	M(C13)+H[1+]	0.0008	0.49	yes
E152 	Best guess: C05827, 3-Methylimidazole acetaldehyde	C05827
	125.071908137835 	34.3395345695163	M+H[1+]	0.0009	0.15	yes
E449 	Best guess: C07480, 	C13747, C07480
	203.052595526013 	154.234821007971	M+Na[1+]	-0.0014	0.01	no
	249.06116640893 	151.605063344907	M+HCOONa[1+]	0.0018	0.08	yes
E153 	Best guess: C05827, 3-Methylimidazole acetaldehyde	C05827
	125.072176177696 	57.466084965214	M+H[1+]	0.0012	0.26	yes
E450 	Best guess: C00159, D-Mannose	C13747, C07480, C00031, C00221, C02336, C00267, C00962, C00795, C00137, C00984, C00159
	203.052933801202 	166.215699669738	M+Na[1+]	0.0003	0.11	yes
E352 	Best guess: C01019, 6-Deoxy-L-galactose	C01019
	165.076356713199 	128.907481863291	M+H[1+]	0.0006	-0.4	yes
E1119 	Best guess: CE5843, 13'-carboxy-alpha-tocopherol	CE5843
	460.358369776795 	32.9589117432608	M+H[1+]	0.0037	-0.18	yes
E644 	Best guess: C00140, N-Acetyl-D-glucosamine	C01074, C00645, C00140
	244.079340211545 	167.231472492246	M+Na[1+]	0.0002	0.11	yes
E14 	Best guess: C05938, L-4-Hydroxyglutamate semialdehyde	C00217, C00025, C05938
	148.060799740294 	169.340766074335	M+H[1+]	0.0003	0.32	yes
	149.06369420469 	169.343572290557	M(C13)+H[1+]	-0.0002	0.37	yes
E643 	Best guess: C00140, N-Acetyl-D-glucosamine	C01074, C00645, C00140
	244.077864278451 	63.1414111868012	M+Na[1+]	-0.0013	0.01	no
	244.077405685064 	73.4649063592259	M+Na[1+]	-0.0018	-0.36	yes
E457 	Best guess: C00159, D-Mannose	C00031, C00221, C02336, C00267, C00962, C00795, C00137, C00984, C00159
	181.071817352089 	35.7767213729386	M+H[1+]	0.0011	0.54	yes
	182.075189814211 	35.7414999247126	M(C13)+H[1+]	0.0011	0.49	yes
E941 	Best guess: C03541, Tetrahydrofolyl-[Glu](n)	C03541
	726.248446642947 	170.104977343275	M+Na[1+]	0.003	-0.25	yes
E13 	Best guess: C05938, L-4-Hydroxyglutamate semialdehyde	C00217, C00025, C05938
	148.060374999736 	79.1985434835427	M+H[1+]	-0.0001	-0.21	yes
E171 	Best guess: C04282, 1-Pyrroline-4-hydroxy-2-carboxylate	C01879, C04281, C04282
	152.031900666236 	175.563251387819	M+Na[1+]	0.0	-0.11	yes
	198.038224312001 	169.460765142265	M+HCOONa[1+]	0.0009	-0.29	yes
	130.050533235302 	175.794675266091	M+H[1+]	0.0007	-0.03	no
	130.050326538094 	169.258016891767	M+H[1+]	0.0005	0.28	yes
	131.053423804078 	175.772996569593	M(C13)+H[1+]	0.0001	-0.08	yes
	131.052881275153 	169.351244113728	M(C13)+H[1+]	-0.0004	0.46	yes
E1020 	Best guess: C13856, 2-Arachidonoylglycerol	C13856
	401.263510729393 	46.2541299515766	M+Na[1+]	-0.0028	0.18	yes
E1247 	Best guess: C00416, Phosphatidic acid (phospholipids)	C00416
	723.488616656648 	155.98888842267	M+Na[1+]	-0.005	-0.16	yes
	769.504977894454 	157.293541814095	M+HCOONa[1+]	0.006	-0.07	no
E943 	Best guess: C00427, Prostaglandin H2	C00696, C04853, C06315, C06314, C00584, C04654, C04671, C01312, C02198, C00427
	353.228921623428 	33.0699719762715	M+H[1+]	-0.0034	0.17	yes
E173 	Best guess: CE1554, N-acetyl-L-alanine	CE1554
	153.040635389064 	136.835601204689	M+Na[1+]	0.0009	0.21	yes
E653 	Best guess: C00386, Carnosine	C06486, C00386
	249.097128076742 	33.2577263383329	M+Na[1+]	0.0013	0.21	yes
E650 	Best guess: C00101, 5,6,7,8-Tetrahydrofolate	C00101
	224.095808761035 	178.610537395497	M(C13)+2H[2+]	0.0013	0.02	no
	468.162575518918 	177.66750080005	M+Na[1+]	0.0023	0.47	yes
E876 	Best guess: C05981, phosphatidylinositol-3,4,5-trisphosphate 	C05981
	644.981077752684 	164.702534540136	M+H[1+]	0.0028	-0.34	yes
	712.962169179343 	164.298936665285	M+HCOONa[1+]	-0.0035	-0.2	no
E261 	Best guess: C06423, Octanoic acid	C06423
	167.10415138422 	184.789992918668	M+Na[1+]	-0.0001	0.16	yes
E65 	Best guess: C00366, Urate	C00366
	169.036215622006 	167.167239834218	M+H[1+]	0.0006	0.21	yes
	191.01736359876 	166.302828104537	M+Na[1+]	-0.0002	-0.09	no
E177 	Best guess: C00408, L-Pipecolate	C05936, C00408
	130.086145291386 	35.4874165339146	M+H[1+]	-0.0001	0.31	yes
E262 	Best guess: C04076, L-2-Aminoadipate 6-semialdehyde	C03239, C02946, C04076
	168.064087372536 	162.778132740707	M+Na[1+]	0.0009	-0.04	no
	146.081082471696 	167.536097002276	M+H[1+]	-0.0001	-0.06	no
	146.081594949794 	162.378967156258	M+H[1+]	0.0004	-0.31	yes
E179 	Best guess: C00408, L-Pipecolate	C05936, C00408
	130.086550332681 	186.728312901591	M+H[1+]	0.0003	0.11	yes
	131.089604989269 	186.744739676169	M(C13)+H[1+]	-0.0001	-0.01	no
E178 	Best guess: C00408, L-Pipecolate	C05936, C00408
	130.086480734306 	153.346487770778	M+H[1+]	0.0002	-0.39	yes
E780 	Best guess: C05401, 3-beta-D-Galactosyl-sn-glycerol	C05401
	277.089645271842 	170.158152692971	M+Na[1+]	0.0002	-0.16	yes
E265 	Best guess: C05543, 3-Dehydroxycarnitine	C05543
	146.117642585952 	68.2175524262869	M+H[1+]	0.0001	-0.17	yes
	147.120586642635 	68.5224626835855	M(C13)+H[1+]	-0.0004	-0.08	no
More data

Full data tables from pathway analysis, module analysis and activity network are stored in the 'tables/' directory. The tsv files can be imported into a spreadsheet program, e.g. MS Excel. Under the 'figures/network_modules/' directory are files intended for Cytoscape (cytoscape.org) visualization. Please refer to Cytoscape's guides for details. Details of this run are recorded in file mummichog.log.txt.
Mummichog algorithms are described in Li et al. Predicting Network Activity from High Throughput Metabolomics. PLoS Computational Biology (2013); doi:10.1371/journal.pcbi.1003123. This software is provided as is, without warranty of any kind. Visit http://mummichog.org for updates.


# Results mummichog import

After running the algoritm (bash script on pc with py/mummichoc installed), 
todo into spectripy... !!!,
the following table os imported:

- userInput_to_EmpiricalCompounds.tsv

The first file is a copy of the data user gave to the algoritm, with the rowxxx identifiers needed as first column.
The second file gives the hypothetical annotations (kegg ID and comon name used in KEGG) for the retrieved metabolites.

please note that only metabolites which are present in KEGG can be retrieved using this approach.

TODO!!! path, filename to auto select latest or in config?? or see git if name output cn change name/dir??

```{r load-mummichig-result-table}
#' load mummichog result files
#' TODO rename using bash folder
#DIR <- "1732118659.9126198.t2"   #test
DIR <- '1734448517.6233082.t2'    #umetab-bmi-umetabpos-lm
DIR <- '1737464806.4199815.t2'    #nafld-masld-umetabpos-lm
DATA_OUTPUT_PATH <- paste0('/home/mdegraeve/Documents/Files/Work_Eurac/Projects/biochem_pathway_analysis/', DIR, '/tables')


#' userInput_to_EmpiricalCompounds
handle <- file.path(DATA_OUTPUT_PATH, "userInput_to_EmpiricalCompounds.tsv")
userInput_to_EmpiricalCompounds <- read.csv(handle, header = TRUE, sep="\t")  |>
    as.data.frame()
userInput_to_EmpiricalCompounds[1:10, 1:10]

```

info regarding results file:

- adducts = str_row_ion
- keggid = compounds, also sometimes CE id from bigg)
- name = compound_names

See remarks regarding the paradoxical results in the CHRIS/example rmd report.
See here the final selected clean up of the mummichog results.


```{r cleanup-table-unique-rowids}
#' checks
#' total rowids (inlc duplic)
nrow(userInput_to_EmpiricalCompounds)
#' nr of non-duplic
length(userInput_to_EmpiricalCompounds$input_row[!duplicated(as.list(userInput_to_EmpiricalCompounds$input_row))])
#' nr of duplic
length(unique(userInput_to_EmpiricalCompounds$input_row[duplicated(as.list(userInput_to_EmpiricalCompounds$input_row))]))
#' needed length if correct merged == unique
length(unique(as.list(userInput_to_EmpiricalCompounds$input_row)))
#' nr of exactly 1 time occur in df
length(subset(userInput_to_EmpiricalCompounds$input_row, !(duplicated(userInput_to_EmpiricalCompounds$input_row) | duplicated(userInput_to_EmpiricalCompounds$input_row, fromLast = TRUE))))


#' names of the replicate rowids to be merged
doubles <- userInput_to_EmpiricalCompounds$input_row[duplicated(as.list(userInput_to_EmpiricalCompounds$input_row))] #double hits, to merge into 1 rowid
doubles <- unique(doubles)

#' loop to make subdf with solved doubles
df_all <- NULL
i <- 1
for(i in 1:length(doubles)){
    #print(doubles[i])
        
    double_df <- userInput_to_EmpiricalCompounds[userInput_to_EmpiricalCompounds$input_row == doubles[i], ]

    #' add to empty df
    df <- NULL
    df$input_row <- double_df$input_row[1]
    df$EID <- paste(c(unlist(double_df$EID)),collapse = ';')
    df$str_row_ion <- paste(c(unlist(double_df$str_row_ion)),collapse = ';')
    df$compounds <- paste(c(unlist(double_df$compounds)),collapse = ';')
    df$compound_names  <- paste(c(unlist(double_df$compound_names)),collapse = ';')
    df$input_row.1 <- double_df$input_row[1]
    df$m.z  <- paste(c(unlist(double_df$m.z)),collapse = ';')
    df$retention_time  <- paste(c(unlist(double_df$retention_time)),collapse = ';')
    df$p_value  <- paste(c(unlist(double_df$p_value)),collapse = ';')
    df$statistic  <- paste(c(unlist(double_df$statistic)),collapse = ';')
    df$CompoundID_from_user <- paste(c(unlist(double_df$CompoundID_from_user)),collapse = ';')
    df <- as.data.frame(df)

    df_all <- rbind(df_all, df) 
}

#' checks
dim(df_all)
df_all[, 1:3]
colnames(df_all)
length(unique(as.list(df_all$input_row)))

#' subset original mummichog results table with rowid nr exactly 1 time occur in df
df_unique <- subset(userInput_to_EmpiricalCompounds, !(duplicated(userInput_to_EmpiricalCompounds$input_row) | duplicated(userInput_to_EmpiricalCompounds$input_row, fromLast = TRUE)))
dim(df_unique)
colnames(df_unique)

#' merge the non-duplic and the solved into 1 df again
clean_df <- rbind(df_unique, df_all)
clean_df <- clean_df[order(clean_df$input_row),]
dim(clean_df)
rownames(clean_df) <- clean_df$input_row

#' write for testing
handle <- file.path(DATA_PATH, 'cleanup_mummichog_output.txt')
write.table(clean_df, file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)

#fyi, top metab from lm FT01399 == glutamate according to mummichog

```

todo!!!! see in clean df still some rowIDs merged in the untargeted metab df!!!


The computational 'EmpiricalCompound' and other information for a tentative
metabolite is now linked to the feature ID from the untargeted metabolomics
dataset. The candidate metabolite can group multiple ions, and be any of the
isobaric/isomeric species. See the original Mummichog 'results.html' for the
visualisation and for the EmpiricalCompounds prioritized by
pathway/module analysis.  

TODO: how add these to umetab dataset? using summExp object add description
maybe? FINDME!!! (todo after lcms semi quant untarg added, see there how).


# Subset the significant metabolites from the lm, if contain annotation

The input data is loaded in the `lm_output_results_metab` object. After 
subsetting to keep only the significant untargeted metabolites, There are 13 
significantly associated metabolites for SLD type MASLD vs. none.

the significant metabolites are then retrieved in the cleaned up table with
putative annotations is in the `clean_df` object.

```{r}
#subset lm output with only the significant
#dim(lm_output_results_metab)
#colnames(lm_output_results_metab)
lm_signif <- lm_output_results_metab[lm_output_results_metab[, paste0('significant_', TARGET_LMNAME)], ]  
print('nr of significant metabolites:')
dim(lm_signif) 
print('significant metabolites from lm, both annotated and not-annotated:')
lm_signif$AID

#keep only the annotated by subsetting clean_df
lm_signif_with_annotation <- clean_df[clean_df$CompoundID_from_user %in% lm_signif$AID, ]
#head(lm_signif_with_annotation)
print('putative annotated significant metabolites:')
lm_signif_with_annotation$compound_names

#all info together of significant:
lm_signif_with_annotation

```

[1] "N-Acetylmethionine"                                                                                                                                                                               
[2] "L-Aspartate 4-semialdehyde; Aspartate beta-semialdehyde; L-Aspartic 4-semialdehyde L-2-Amino-3-oxobutanoic acid; L-2-Amino-3-oxobutanoate; L-2-Amino-acetoacetate; (S)-2-Amino-3-oxobutanoic acid"
[3] "5-Oxoproline L-1-Pyrroline-3-hydroxy-5-carboxylate 1-Pyrroline-4-hydroxy-2-carboxylate"                                                                                                           
[4] "L-Leucine$L-Isoleucine"                                                                                                                                                                           
[5] "D-Glutamate; D-Glutamic acid; D-Glutaminic acid; D-2-Aminoglutaric acid L-Glutamate L-4-Hydroxyglutamate semialdehyde"                                                                            
[6] "D-Glutamate; D-Glutamic acid; D-Glutaminic acid; D-2-Aminoglutaric acid L-Glutamate L-4-Hydroxyglutamate semialdehyde"     

# Acknowlegments

# Perspectives

https://www.metaboanalyst.ca/resources/vignettes/Functional_Analysis_global_metabolomics.html
found the metaboAnalystR tutorial, with code in R


# Session information

```{r}
sessionInfo()

```
