---
title: "duplicated_nafld"
author:
  - name: "Philippine Louail"
affiliation: "Eurac Research, Bolzano, Italy"
output: rmdformats::html_clean   
date: "2023-10-16"
---

```{r style, message = FALSE, echo = FALSE, warning = FALSE, results = "asis"}
library("BiocStyle")
library("knitr")
library("rmarkdown")
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7)
```

# Packages

```{r warning=FALSE, include=FALSE}
library(MsExperiment)
library(RSQLite)
library(xcms)
library(Spectra)
library(RColorBrewer)
library(pander)
library(readxl)
library(MetaboCoreUtils)
library(pheatmap)
library(MsBackendSql)
```



```{r}
#' Load the spectra data from the database
sps <- Spectra("data/NAFLD_untargeted.sqlite", drv = SQLite(),
               source = MsBackendOfflineSql())

pd <- as.data.frame(read_xlsx("data/nafld_untargeted_metabolomics.xlsx"))

#' Subset to positive polarity
pd <- pd[pd$polarity == "POS", ]
batches <- c("20200204", "20201026", "20200901", "20210922")
pd <- pd[pd$batch == batches,] ## specifically for analysis of the duplicated samples 


#' Subset the Spectra to the respective files
sps <- sps[basename(sps$dataOrigin) %in% pd$file]
sps$file <- basename(sps$dataOrigin)

#' Create MsExperiment and fill with data
nafld <- MsExperiment()
sampleData(nafld) <- DataFrame(pd)
spectra(nafld) <- sps

#' Link spectra to samples
nafld <- linkSampleData(nafld, with = "sampleData.file = spectra.file")
```

```{r define-colors, include=FALSE}
#' Define colors for the groups.
col_phenotype <- brewer.pal(4, "Set1")[c(2, 1, 4, 3)]
names(col_phenotype) <- c("study", "POOL", "QC", "BLANK")

col_sample <- col_phenotype[sampleData(nafld)$sample_type] 
```

```{r}
#' Set up parallel processing using 3 cores
if (.Platform$OS.type == "unix") {
    register(bpstart(MulticoreParam(3)))
} else {
    register(bpstart(SnowParam(3)))
}
```


Get some data overview to be sure everything is fine (it was not before for example)
```{r some healthy stuff}
#' check number of spectra per file
fromFile(nafld) |> 
    table()

#' List number of spectra per MS level
spectra(nafld) |>
    msLevel() |>
    table()

#' Get the distribution of peak counts per file
spectra(nafld) |>
    lengths() |>
    split(fromFile(nafld)) |>
    lapply(quantile)
```



```{r Plot bpc before filtering, echo=FALSE}
#' First extract and plot bpc for full data
bpc <- chromatogram(nafld, aggregationFun = "max")

#' Plot
png(paste("bpc duplicated samples.png"))
plot(bpc, col = paste0(col_phenotype[bpc$sample_type]), main = "BPC")
grid()
legend("topright", col = col_phenotype,
       legend = names(col_phenotype), lty = 1)
dev.off()

#' Filter for ret time 
nafld <- filterRt(nafld, c(0, 350))

#' Plot again
bpc <- chromatogram(nafld, aggregationFun = "max")
png(paste("bpc duplicated samples_after_filtering.png"))
plot(bpc, col = paste0(col_phenotype[bpc$sample_type]), main = "BPC after filtering")
grid()
legend("topright", col = col_phenotype,
       legend = names(col_phenotype), lty = 1)
dev.off()

```
## Internal standards

change this so that chromatogram part outside of loop
```{r EIC extract for internal standard }
#' get the list 
intern_standard <- read.delim("C:/Users/plouail/Documents/GitHub/lcms-standards/data/internal_standards.txt") 
intern_standard <- intern_standard[-4, ] ## remove succinic acid because no POS

#'generate calcualte formula
intern_standard[, "mz"] <- mass2mz(intern_standard$formula, adduct = "[M+H]+")[, 1]
intern_standard[3, "mz"] <- mass2mz(intern_standard[3,"formula"], adduct = "[M+Na]+")[, 1]


##' Extract the EICs no RT 
#' Expand the m/z range by 0.01 on both sides
intern_standard$mzmin <- intern_standard$mz - 0.005
intern_standard$mzmax <- intern_standard$mz + 0.005
intern_standard$rtmin <- 0
intern_standard$rtmax <- 350

eics <- chromatogram(nafld, mz = as.matrix(intern_standard[, 14:15]), rt = as.matrix(intern_standard[, 16:17])) 

for (i in seq_len(nrow(intern_standard))) {
    x <- intern_standard[i, c(2,3,4,6)]
    png(paste("png/internal_standard/chromatogram/chromatogram", x$abbreviation, ".png"))
    plot(eics[i], main = paste("chromatogram_all_sample_for_", x$abbreviation), 
         col = paste0(col_phenotype[eics$sample_type], 80))
    grid()
    legend("topright", col = col_phenotype,
           legend = names(col_phenotype), lty = 1)
    abline(v = x$RT, col = "red", )
    dev.off()
}

##'  extract EIC with RT

intern_standard[,"rtmin"] <- intern_standard[, "RT"] - 10
intern_standard[,"rtmax"] <- intern_standard[, "RT"] + 10
eics <- chromatogram(nafld, mz = as.matrix(intern_standard[, 14:15]), rt = as.matrix(intern_standard[, 16:17]))
for (i in seq_len(nrow(intern_standard))) {
    x <- intern_standard[i, c(2,3,4,6)]
    #'plot it add expected retention time
    png(paste("png/internal_standard/EIC/EIC", x$abbreviation, ".png"))
    plot(eics[i], main = paste("EIC_all_sample_for_", x$abbreviation), 
         col = paste0(col_phenotype[eics$sample_type], 80))
    grid()
    legend("topright", col = col_phenotype,
           legend = names(col_phenotype), lty = 1)
    abline(v = x$RT, col = "red", )
    dev.off()
}

#' what does it mean exactly ? what about the multiple peaks ones ? do on ALL sample ?
```

## Standards

```{r full standard}
#' get the list 
standard <- read.delim("C:/Users/plouail/Documents/GitHub/lcms-standards/data/standards_dilution.txt")

#' select for pos standard
x <- standard[!is.na(standard$POS),]
#' calculate mass 
x$mass <- calculateMass(x$formula)

#' Calculate charge 
for (i in seq_len(nrow(x))){
    xi <- x[i,]
    x[i, "mz"] <- mass2mz(xi$mass, adduct = xi$POS)[, 1]
}

##' Extract the EICs no RT 

#' Expand the m/z range by 0.01 on both sides
x$mzmin <- x$mz - 0.005
x$mzmax <- x$mz + 0.005
x$rtmin <- 0
x$rtmax <- 350

#' extract chromatogram
eics <- chromatogram(nafld, mz = as.matrix(x[, 18:19]), 
                     rt = as.matrix(x[, 20:21]))

#' loop through standard list and extract chromatogram
for (i in seq_len(nrow(x))) {
    #'plot it add expected retention time
    png(paste("png/standard/chromatogram/chromatogram", x$abbreviation, ".png"))
    plot(eics[i], main = paste("chromatogram_all_sample_for_", x$abbreviation), 
         col = paste0(col_phenotype[eics$sample_type], 80))
    grid()
    legend("topright", col = col_phenotype,
           legend = names(col_phenotype), lty = 1)
    abline(v = x$RT, col = "red", )
    dev.off()
}
```

We therefore have an data set of `r length(nafld)` samples for a total of `r length(spectra(nafld))` spectra. 


The retention time range for the entire data set is: `r range(rtime(spectra(nafld)))`

```{r heatmap}
#' Heatmap from total ion chromatogram 
tic <- chromatogram(nafld, aggregationFun = "sum") |> 
  bin(binSize = 2)

ticmap <- do.call(cbind, lapply(tic, intensity)) |>
  cor() |> 
  pheatmap()
```

```{r bps}
#' plot some  only the POOL 
pool_sample <- which(sampleData(nafld)$sample_type == "POOL")
pool_sample <- nafld[pool_sample]

bps <- spectra(pool_sample) |>
  combineSpectra(f = fromFile(pool_sample), intensityFun = max, ppm = 5)  

par(mfrow = c(2,1), mar = c(4, 4.5, 1, 0.5)) # again not sure what to look for here 
plotSpectraOverlay(bps)
plotSpectra(bps)[5:8]
plotSpectra(bps)[9:10]
```

# Preprocessing
```{r}

```

## Chromatographic peak detection

finding cwp parameter 

- peakwidth: observing previous EIC, expecting between 2 to 20s 
_ ppm: discuss with johannes bc hard to find good parameter.. 
```{r ppm parameter }

#' choose a nice compound here 
cystine_mz <- calculateMass("C6H12N2O4S2") |> 
  mass2mz("[M+H]+")
  
cystine_mz <- cystine_mz[1, 1]

#' Restrict the data to signal from Cystine
cst <- nafld[1L] |>
  spectra() |>
  filterRt(rt = c(200, 220)) |>
  filterMzRange(mz = cystine_mz + c(-0.01, 0.01))

lengths(cst)

#' Calculate the difference in m/z values between scans
mz_diff <- cst |>
    mz() |>
    unlist() |>
    diff() |>
    abs()

#' Express it in ppm
range(mz_diff * 1e6 / mean(unlist(mz(cst))))
```

```{r}
#' extract test ion 
#' for cystine 
eic_cystine <- chromatogram(nafld, rt = c(200, 235),
                           mz = cystine_mz + c(-0.05, 0.05))
#' histidine

his_mz <- calculateMass("C6H9N3O2") |> 
  mass2mz("[M+H]+")
  
his_mz <- his_mz[1, 1]

eic_his <- chromatogram(nafld, rt = c(170, 210),
                           mz = his_mz + c(-0.05, 0.05))
```


```{r warning=FALSE}
param <- CentWaveParam(peakwidth = c(2, 20), ppm = 40, integrate = 2)

#' test for some compounds:
#' cystine
cystine_test <- findChromPeaks(eic_cystine, param = param)
head(chromPeaks(cystine_test))

#' histidine
his_test <- findChromPeaks(eic_his, param = param)
head(chromPeaks(his_test))
```

```{r echo=FALSE}
#' Plot test chromatogram
par(mfrow = c(1, 2))
plot(cystine_test, main = "Cystine", col = paste0(col_phenotype[eic_cystine$sample_type], 80))
grid()

plot(his_test, main = "Histidine", col = paste0(col_phenotype[eic_his$sample_type], 80))
grid()
legend("topright", col = col_phenotype,
       legend = names(col_phenotype), lty = 1)
```

In my opinion looks pretty good, should apply to all data: 

```{r}
nafld <- findChromPeaks(nafld, param = param)

#' test for ions of interest 
eic_cystine <- chromatogram(nafld, rt = c(200, 220),
                           mz = cystine_mz + c(-0.05, 0.05),
                           aggregationFun = "max")

eic_his <- chromatogram(nafld, rt = c(170, 205),
                           mz = his_mz + c(-0.05, 0.05),
                           aggregationFun = "max")

par(mfrow = c(1, 2))
plot(eic_cystine, main = "Cystine", col = paste0(col_phenotype[eic_cystine$sample_type], 80))
grid()
plot(eic_his, main = "Histidine", col = paste0(col_phenotype[eic_his$sample_type], 80))
grid()
legend("topright", col = col_phenotype,
       legend = names(col_phenotype), lty = 1)

```

Look at the nb of peak per files
```{r}
#' Count peaks per file
chromPeaks(nafld)[, "sample"] |>
    table()

#' sample 8 has lower amount than rest 
```

### Refine chromatographic peaks 


test our dataset:
that doesnt really work tbh
```{r}
#' Select 4 random peaks
mz_rt <- data.frame(row.names=c("Area 1", "Area 2", "Area 3", "Area 4"), 
                    rtmin = c(137.291, 161.5420, 161.5420, 162.3770),
                    rtmax = c(148.965, 167.4010, 166.8430, 167.1200), 
                    mzmin = c(251.1842, 939.8238, 992.8133, 910.8906),
                    mzmax = c(251.1881, 939.9147, 992.9208, 910.9350)) |> 
  as.matrix()


#' Expand the rt range by 10 seconds on both sides
mz_rt[, "rtmin"] <- mz_rt[, "rtmin"] - 5
mz_rt[, "rtmax"] <- mz_rt[, "rtmax"] + 5
#' Expand the m/z range by 0.005 on both sides
mz_rt[, "mzmin"] <- mz_rt[, "mzmin"] - 0.001
mz_rt[, "mzmax"] <- mz_rt[, "mzmax"] + 0.001


#' Extract the EICs
eics <- chromatogram(nafld, mz = mz_rt[, 3:4], rt = mz_rt[, 1:2])
#' Plot the EICs
plot(eics)
```


Choose parameter: 

- expandRt: switch it to 5 ? or keep 2.5 
- 
- 

let's see what happens
```{r test merging}
#' set up the parameter 
param <- MergeNeighboringPeaksParam(expandRt = 5, expandMz = 0.001,
                                   minProp = 3/4) 

#' Perform the peak refinement on the EICs
eics <- refineChromPeaks(eics, param = param)
plot(eics)
```
- DO we need to show for this dataset, can barely see anything


```{r}
#' apply to all dataset 
nafld <- refineChromPeaks(nafld, param = param)
```
"Reduced from 715317 to 643063 chromatographic peaks."



# Session information

```{r}
sessionInfo()
```