---
title: "Steatosis associated metabolites, lipids and proteins - with PDI index as covariate"
author: "Marilyn De Graeve, Johannes Rainer, Christoph Grander"
affiliation: "Eurac Research, Bolzano, Italy"
date: "2025-02-27"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
---

**Modified**: `r file.info("NAFLD_steatosis_foodindex_influence.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r style, message = FALSE, echo = FALSE, warning = FALSE, results = "asis"}
library("BiocStyle")
library("kableExtra")
library("knitr")
library("rmarkdown")
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now, units = "secs")
      # return a character string to show the time
      paste("    Time for this code chunk to run:", round(res,
        2), "seconds")
    }
  }
}))
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7, time_it = TRUE)
set.seed(1)
```

# Introduction

During this workflow, exploratory analyis (Spearman correlations) are performed
on the relatively quantified Scanning SWATH proteomics data (148 proteins) from
the metabolic dysfunction-associated steatotic liver disease (MASLD)
cohort. MASLD, previously known as NAFLD is the most chronic liver disease,
affecting more than 30% of global population. The MAFLD cohort is a subset of
approximatley 353 # participants from the Cooperative Health Research in South
Tyrol (CHRIS) study, which were requested to return for a second resampling and
additional phenotyping. The NAFLD cohort represent an unhealty group with
various degrees of steatotic (non-alcoholic) liver disease classifications. See
[nafld_data_overview.Rmd](nafld_data_overview.Rmd) (respectively the rendered
html report) for an overview of the data set and the definition of the
phenotype.

This documents describes the association analysis between circulating
metabolites, lipids and proteins with liver steatosis.

See the `nafld_data_overview 1.html` report for further
information regarding the SLD sub-classification.

Regarding steatosis, i.e. Abnormal retention of fat (lipids) within a cell or
organ. In this scope: liver steatosis.
The variables containing this parameter is:

Presence of steatotic liver disease (SLD): sonographic evaluation, steatose
score of 2 – 3.

- *naab13*: liver fat. Can be binarized at an absolute value of 288. "liver_fat"
- *naab14*: subjective, visual evaluation of the liver fat. Categorical from 1
  to 3. Values 2 and 3 can be considered steatotic. "visual_fat"

This parameter will be used as the target for a biomarker search in *MASLD*
participants.


# Setup

## Directories

```{r directories, echo = FALSE}
#' General settings
filename <- "NAFLD_steatosis_foodindex_influence"

#' Path to save images to; remove if exists.
IMAGE_PATH <- paste0("images/", filename, "/")
dir.create(IMAGE_PATH, recursive = TRUE)

#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)

#' Path to the data
DATA_PATH <- '/home/mdegraeve/Documents/Files/Work_Eurac/Data/NAFLD' #FINDME!
#DATA_PATH <- "."


DATASET <- 'CHRIS-NAFLD, none and MASLD'

```

## Packages

```{r packages, message=FALSE}
#' Load libraries
library(chrisUtils)
library(chrisData)
library(DT)
library(ggfortify)
library(ggplot2)
library(pander)
library(pheatmap)
library(plotly)
library(RColorBrewer)
library(readxl)
library(reshape2)
library(tidyfr)
library(vioplot)
library(writexl)

```


# Data import

## Circulating proteins (tprot) dataset (Scanning SWATH)

The data from proteins was converted into the TDF format and can hence be
loaded and imported using the `tidyfr` R package. Now we read the full data
which includes general information for each measurement.

```{r load_prot_dataset, message = FALSE}
#' load TDF using tidyfr
#list_data_modules(path = DATA_PATH)
md1 <- data_module(name = "chris_scanningswath_proteins", version = "1.0.0.3",
                   path = DATA_PATH)
prot_data <- data(md1)
prot_ann <- labels(md1)
stopifnot(all(colnames(prot_data) == rownames(prot_ann)))
colnames(prot_data) <- prot_ann$description
rownames(prot_ann) <- prot_ann$description

#' define the columns with protein concentrations
proteins <- rownames(prot_ann)[prot_ann$type != "categorical"]
proteins <- rownames(prot_ann)[prot_ann$Genes != ""]

#' select proteins dataframe
prot_data <- prot_data[, proteins]

```

The loaded data contains the samples as rows, and the protein names as
colums. The measurements of the `prot_data` object are the relative
concentrations of protein per sample. No data transformation is performed as the
abundances are already reported in log2 scale, a normal distribution of the
proteomics is assumed.

Before proceeding with the data analysis we remove the POOL samples. In addition
we restrict to samples from the NAFLD study. The number of samples and proteins
are:

```{r subset to NAFLD set, message = FALSE, warning = FALSE}
prot_data <- prot_data[grep("^008", rownames(prot_data)), ]
dim(prot_data)
```


## Targeted metabolomics (tmetab) dataset (Biocrates p180 kit)

The data from metabolites was converted into the TDF format and can hence be
loaded and imported using the `tidyfr` R package. Now we read the full data
which includes general information for each measurement.

```{r load_tmetb_dataset, message = FALSE}
#' load TDF using tidyfr
#list_data_modules(path = DATA_PATH)
p180 <- data_module(name = "metabolomics_p180", version = "1.0.1.3",
                    path = DATA_PATH)
met_data <- data(p180)
met_ann <- labels(p180)
stopifnot(all(colnames(met_data) == rownames(met_ann)))
colnames(met_data) <- met_ann$description
rownames(met_ann) <- met_ann$description

#' define the columns with metabolite concentrations
metabolites <- rownames(met_ann)[met_ann$type != "categorical"]

#' select metabolites dataframe
met_data <- met_data[, metabolites]
#' Remove the individuals with more than 80% missing values
rem <- apply(met_data, MARGIN = 1, function(z) {
    sum(is.na(z)) > length(z) * 0.8
})
met_data <- met_data[!rem, ]


#' Impute missing values using an uniform distribution, after gap filling
#na_unidis <- function(z) {
#    na <- is.na(z)
#    if (any(na)) {
#        min = min(z, na.rm = TRUE)
#        z[na] <- runif(sum(na), min = min/2, max = min)
#    }
#    z
#}

#' Row-wise impute missing values and add the data as a new assay
#tmp <- apply(met_data, MARGIN = 1, na_unidis)
#met_data <- t(tmp)
```

The loaded data contains the samples as rows, and the metabolite names as
colums. The measurements of the `met_data` object are the absolute
concentrations of metabolite per sample, in natural scale.

To ensure a normal distribution of the metabololits, we perform log2
transformation.

```{r m log2-transformation, message = FALSE}
#' log2 scaling of absolute abundances metabolites
met_data <- log2(met_data)
```

Before proceeding with the data analysis we remove the POOL samples. In addition
we restrict to samples from the NAFLD study. The number of samples and
metabolites are listed below.

```{r m subset to NAFLD set, message = FALSE, warning = FALSE}
met_data <- met_data[grep("^008", rownames(met_data)), ]
dim(met_data)
```


## Diet GA2LEN data

Dietary information, i.e. GA2LEN food frequency questionaires, from the MASLD
study were loaded and included in the analysis.

DO NOT RUN next 2 chuncks, we use the newly created dietary indices from Essi
(10/07/2025). 

```{r diet-data, eval = FALSE}
#' 1. load the metadata regarding the GA2LEN FFQs
library(nafldData)
#nafldTraitAnnotations()|> head()

galan <- nafldData("FFQ")
rownames(galan) <- galan$AID

#variables, subset:
#naff001 - naff229 are galan FFQ, and the rest = not!

galan <- galan[, 1:230] #incl AID
ffq_items <- colnames(galan)[2:230]

#' change to numerical scores instead of FFQ answers
#' solve using category encoder, in R see:
#' https://www.r-bloggers.com/2020/02/a-guide-to-encoding-categorical-features-using-r/

#' select categor variables
categor_variables <- NULL
for (x in ffq_items){
  if (class(galan[[x]]) == 'factor' | class(galan[[x]]) == 'character'){  #factor and character
    categor_variables <- c(categor_variables, x)
  }
}

#' using ordinal encoding on categorical vars
encode_ordinal <- function(x) {
  if (is.factor(x)) {
    order <- levels(x)   #retain defined factor level order, wo NA but next step ignored so fine
    x <- as.numeric(factor(x, levels = order, exclude = NA)) #as numeric, keep NA in dataset
  } else if (is.character(x)) {
    print(x) #*
  }
  x
}

#* text info: NOT
    #naff246 - naff249b
    #naff260 - naff263a

#' testing
summary(galan$naff001)

#' apply encoding on categor vars
galan[, categor_variables] <- lapply(galan[, categor_variables], encode_ordinal)
dim(galan)
```

Corresponding fake CV-values (no 'pool' samples) for diet. Too many AIDs, but
ok, since the filtering on included rownames is done in the lm.

```{r kora-cv, eval = FALSE}
#' 1. load the metadata regarding the GA2LEN FFQs
library(nafldData)
#nafldTraitAnnotations()|> head()

nafldAnn <- nafldTraitAnnotations()
kora_ann <- nafldAnn[ffq_items, ]
rownames(kora_ann) <- ffq_items
#head(kora_ann)

#' TEMP, just accept all features now as having good cv
cv_values <- 0
#' save in the met_ann as column
kora_ann$cv_qc_chris <- cv_values
```

The newly created dietary indices from Essi are saved under 
NAFLD_dietary_indices_data_Marilyn. And note in the
lm: "you should also correct your models for total energy intake (just have it
as covariate in the model, nanu006a)". This was added in the section of dietary
lms.

```{r NAFLD-indices}
#' load the NAFLD_dietary_indices_data
handle <- file.path(DATA_PATH, 'GALAN', 'NAFLD_dietary_indices_data_Marilyn.txt')
galan <- read.table(handle, header = TRUE, sep = "\t")  |>
    as.data.frame()

#'galan is with , instead of decimal point so change
comma_to_point <- function(x) {
  x <- scan(text = x, dec = ",", sep = ".")
  x
}
galan <- apply(galan, MARGIN = 2, comma_to_point) |>
    as.data.frame()
rownames(galan) <- paste0('00', as.character(galan$AID))
#dim(galan)
#colnames(galan)
#galan$AID[1:5]
#head(galan)

#' replace the aids columns with 00 start and as a string
galan$AID <- rownames(galan)

#' variables
ffq_items <- colnames(galan)[2:ncol(galan)] #excl AID col
dim(galan)
```

Also add the 4 macros and add to the `ffq_items`.

```{r NAFLD_food_macros}
library(nafldData)
#need these one
#nanu003 (protein), nanu004 (total fat), nanu005 (total carbs), nanu019b (saturated fatty acids), nanu023b (monounsaturated fatty acids), nanu025b (polyunsaturated fatty acids), nanu028b (trans fatty acids)

ffq <- nafldData("FFQ")
rownames(ffq) <- ffq$AID

galan$protein <- ffq[rownames(galan), 'nanu003']
galan$total_fat <- ffq[rownames(galan), 'nanu004']
galan$total_carbs <- ffq[rownames(galan), 'nanu005']
galan$saturated_fatty_acids <- ffq[rownames(galan), 'nanu019b']
galan$monounsaturated_fatty_acids <- ffq[rownames(galan), 'nanu023b']
galan$polyunsaturated_fatty_acids <- ffq[rownames(galan), 'nanu025b']
galan$trans_fatty_acids <- ffq[rownames(galan), 'nanu028b']

#' variables
ffq_items <- colnames(galan)[2:ncol(galan)] #excl AID col
dim(galan)
```


## Metadata (excel)

All available trait (phenotype) data was defined in the
[nafld_data_overview.Rmd](nafld_data_overview.Rmd), is loaded below and merged
with the log2 transformed proteomics data of the NAFLD cohort.

```{r add phenotype dataa}
#' load xlsx file instead of below
handle <- 'masld_phenotypes.xlsx'
masld_phenotype_df <- read_excel(file.path(DATA_PATH, handle)) |>
    as.data.frame()
rownames(masld_phenotype_df) <- masld_phenotype_df$AID
#head(masld_phenotype_df, 5)
metadata <- colnames(masld_phenotype_df)[3:ncol(masld_phenotype_df)] #info wo AIDs
#dim(masld_phenotype_df)

#' make sure sld_type is a factor with group 'none' as the baseline
masld_phenotype_df$sld_type <- factor(masld_phenotype_df$sld_type,
                                      levels = c('none', 'MASLD',
                                                 'MetALD', 'SLD'))
print('the sld types are:')
levels(masld_phenotype_df$sld_type)
print('nr of samples per sld types:')
table(masld_phenotype_df$sld_type)
```


# Merge tmetab, tprot, kora and metadata

Merging all data tables. Note that restrict the data to participants for which
we have proteomics **and** metabolomics data. This also ensures that any
potential study drop-outs are no longer considered in the data (as their
proteomics or metabolomics data was removed).

There were less AIDs present in the dietary data, so we restrict with the data
to participants for which we have proteomics **and** metabolomics **and** 
dietary data. ??? TODO findme!!!

```{r merge-dfs}
#AIDs <- intersect(rownames(prot_data), rownames(met_data))
AIDs1 <- intersect(rownames(prot_data), rownames(met_data))
AIDs2 <- intersect(rownames(masld_phenotype_df), rownames(galan))
AIDs <- intersect(AIDs1, AIDs2)

#' Subset all datasets to the same IDs
prot_data <- prot_data[AIDs, ]
met_data <- met_data[AIDs, ]
masld_phenotype_df <- masld_phenotype_df[AIDs, ]
galan <- galan[AIDs, ]

#' merge with proteomics data
prot_data$AID <- rownames(prot_data)
prot_data <- merge(prot_data, masld_phenotype_df, by = 'AID')

#' merge all sets
met_data <- as.data.frame(met_data)
met_data$AID <- rownames(met_data)
met_prot_data <- merge(prot_data, met_data, by = 'AID')

#' merge with metabolomics data only
met_data <- merge(met_data, masld_phenotype_df, by = 'AID')

#' merge all sets, galan too
met_prot_data <- as.data.frame(met_prot_data)
rownames(met_prot_data) <- met_prot_data$AID
gal_met_prot_data <- merge(met_prot_data, galan, by = 'AID') #galan has only 001 AIDs
rownames(gal_met_prot_data) <- gal_met_prot_data$AID
#gal_met_prot_data[1:5, 1:5]
dim(gal_met_prot_data)

#' write to file for manuel inspection
handle <- 'data/gal_met_prot_data.txt'
write.table(gal_met_prot_data, file = handle, sep = "\t",
            row.names = FALSE, col.names = TRUE)
```

All data sets are subset to `r length(AIDs)` study participants.


# Clean data and restrict to relevant participants

The data is further processed and cleaned to remove unreliable
measurements. Subsequently the data is restricted to relevant participants. We
thus next:

- remove unreliable liver fat measurements
- exclude participants with MetALD and SLD to ensure a more homogeneous data set
- remove participants with missing values in relevant variables
- define variables and phenotype(s) for the analysis

Add the information from the [nafld_data_overview.Rmd](nafld_data_overview.Rmd)
report regarding the steatosis, using the parameter naab11
(higher values are worse). Values for participants that have a naab12
(liver stiffness IQR) > 30 should be either removed or taken with care. Below we
remove these unreliable measurements.

```{r load-liver-param2}
#' remove samples with  naab12 (liver stiffness IQR) > 30
gal_met_prot_data$liver_stiffness[
                      which(gal_met_prot_data$liver_stiffnes_iqr > 30)] <- NA
```

We next remove the MetALD and SLD participants.

```{r keep only none and MASLD types}
#' because problem 2 proteins only NAs in some sld types,
#' keep only none and MASLD
gal_met_prot_data <- gal_met_prot_data[gal_met_prot_data$sld_type %in%
                                       c('none', 'MASLD'), ] #
gal_met_prot_data$sld_type <- droplevels(gal_met_prot_data$sld_type)
gal_met_prot_data$sld_type <- factor(gal_met_prot_data$sld_type)
```

We next remove participants with missing values for relevant variables:
- liver fat measurement

```{r}
#' Define variables for which we don't allow missing values
no_na_var <- c("liver_fat")
has_na <- lapply(gal_met_prot_data[, no_na_var, drop = FALSE], is.na) |>
    do.call(what = cbind)
#' Remove rows (participants) with a missing value in any of the above
#' variables
gal_met_prot_data <- gal_met_prot_data[rowSums(has_na) == 0, , drop = FALSE]
```

Remove fibrosis outlier values: **CURRENTLY NOT DONE**.
Previous outlier removal removed a participant with a FIB4 score of 7.8. This
participant was originally removed. The second outlier removal excluded
a participant with a liver_stiffness value of 34. Both participants are still
present, as fibrosis is not the phenotype of interest anymore.

We next define the categorical variables from numeric measurements:

Add a categoric liver_stiffness category too, using both the definition of
Christoph (7, 15 cutoffs) and in literature (7, 12.5 kPa cutoffs).

```{r liver-stiffness-category}
#' naab11 according to Christoph
gal_met_prot_data$liver_stiffness_cat1 <- 2
gal_met_prot_data$liver_stiffness_cat1[
                      is.na(gal_met_prot_data$liver_stiffness)] <- NA
gal_met_prot_data$liver_stiffness_cat1[
                      which(gal_met_prot_data$liver_stiffness < 7)] <- 1
gal_met_prot_data$liver_stiffness_cat1[
                      which(gal_met_prot_data$liver_stiffness > 15)] <- 3
gal_met_prot_data$liver_stiffness_cat1 <-
    factor(gal_met_prot_data$liver_stiffness_cat1)

#' naab11 according to literature
gal_met_prot_data$liver_stiffness_cat2 <- 2
gal_met_prot_data$liver_stiffness_cat2[
                      is.na(gal_met_prot_data$liver_stiffness)] <- NA
gal_met_prot_data$liver_stiffness_cat2[
                      which(gal_met_prot_data$liver_stiffness < 7)] <- 1     #mild or absent fibrosis
gal_met_prot_data$liver_stiffness_cat2[
                      which(gal_met_prot_data$liver_stiffness > 12.5)] <- 3  #cirrhosis
gal_met_prot_data$liver_stiffness_cat2 <-
    factor(gal_met_prot_data$liver_stiffness_cat2)
#' appearantly differnet cutoffs, depending what background...
#' https://www.mskcc.org/cancer-care/patient-education/understanding-your-fibroscan-results
```

Add a categoric liver_fat (i.e. steatosis) category, using the following
cutoffs:

- "normal": A score below 238 dB/m means the amount of fatty change in your
  liver is not higher than normal.
- grade steatosis S1: 238 to 260 dB/m
- grade steatosis S2: 260 to 290 dB/m
- grade steatosis S3: 290 to 400 dB/m

Depending on the distribution, look only at the extreme normal - steatosis.

```{r liver-fat-category}
#' naab13 make 4 groups
gal_met_prot_data$liver_fat_cat1 <- NA
gal_met_prot_data$liver_fat_cat1[which(gal_met_prot_data$liver_fat < 238)] <- 1 #normal
gal_met_prot_data$liver_fat_cat1[which(gal_met_prot_data$liver_fat >= 238)] <- 2
gal_met_prot_data$liver_fat_cat1[which(gal_met_prot_data$liver_fat > 260)] <- 3
gal_met_prot_data$liver_fat_cat1[which(gal_met_prot_data$liver_fat > 290)] <- 4
gal_met_prot_data$liver_fat_cat1 <- factor(gal_met_prot_data$liver_fat_cat1)

#' naab13 make 2 groups
gal_met_prot_data$liver_fat_cat2 <- NA
gal_met_prot_data$liver_fat_cat2[which(gal_met_prot_data$liver_fat < 238)] <- 1 #normal
gal_met_prot_data$liver_fat_cat2[which(gal_met_prot_data$liver_fat > 290)] <- 2  #steatosis, high S3

#' appearantly differnet cutoffs, depending what background...
#' https://www.mskcc.org/cancer-care/patient-education/understanding-your-fibroscan-results
```

At last we ensure categorical variables are encoded as a factor. Also, we create
additional variable for age per 10y, because coefficients and effect sizes
correspond to one year difference, and the anual difference is
neglectible. Dividing age with 10, difference in coefficents and effect sizes
would correspond to 10 years difference and it would be easier to define
meaningful cut-off. Finally, we add a visc_subcut_fat variable on the dataset,
comprising the ratio of visceral to subcutaneous fat thickness.


```{r define-age10-variable, message = FALSE}
#' Convert categorical variables to factors
col_cat <- c("case", "self_DM", "med_DM", "fasting", "no_drinking",
             "med_BP", "med_lip", "med_MS", "T2D", "Sex", "MASLD")
for (i in col_cat)
    gal_met_prot_data[, i] <- as.factor(gal_met_prot_data[, i])

#' define age per 10y group
gal_met_prot_data$Age_10 <- gal_met_prot_data$Age / 10
gal_met_prot_data$visc_subcut_fat <- gal_met_prot_data$visc_fat /
    gal_met_prot_data$body_fat
```


## Summary of the final data set

```{r}
summary(gal_met_prot_data[, c("Age", "Sex", "T2D", "MASLD", "FIB4", "BMI",
                              "liver_stiffness", "liver_fat", "liver_fat_cat1",
                              "visc_subcut_fat")])
```

The variables that are not allowed to conatin NAs are: sex, age and liver_fat.
Participants with NA in BMI, and those 2 with extreme liver stiffness/FIB4
scores are not removed.


# Association analyses

We next perform the association analyses between food items, metabolites and
proteins with the liver fat categories.


## Association between circulating proteins and liver fat (steatosis)

Model setup with "estim_daily_nutr_intake" and "pdi_score" in addition:
as estim is all coef=0, rm and keep only pdi_score. this can be compared with 
the model in "NAFLD_steatosis_associations.html" 

```{r define-lm--covariates-and-target}
print('covariates of the linear regression model:')
COVARIATES <- c("Sex", "Age_10", "pdi_score")
print(COVARIATES)

print('target of the linear regression model:')
TARGET <- 'liver_fat_cat1'
print(TARGET)

DATASET <- 'NAFLD'
print(paste0('final variables of the ', DATASET, ' dataset used:'))
#' variables are the output y in lm(y ~ .):
#print(proteins)

print('nr of variables inspected:')
print(length(proteins))

print('nr of individuals inspected:')
print(nrow(gal_met_prot_data))

```

To identify proteins associated with liver steatosis, seperate linear regression
models are fitted for all proteins separately using their (log2) abundance as
response variable with sex, age and the (categorical) steatosis as a covariates.

Model coefficients and p-values are provided. Raw p-values are adjusted for
multiple hypothesis testing with the method from Benjamini & Hochberg.

Function to use.

```{r linear model-diet with CIs, message=FALSE}

#' Define a function to fit the linear model:
#' - fit the linear model
#' - extract the coefficients and p-values skipping the intercept
#' - return the result as a one-row matrix, columns being coefficients
#'   and p-values
funForLinReg <- function(analyte, data) {
    y <- data[, analyte]
    data <- data[, c(COVARIATES, TARGET)]
    res <- coef(summary(lm(y ~ ., data = data)))[-1, ]
    pvals <- matrix(c(res[, 1L], res[, 4L]), nrow = 1)
    rownames(pvals) <- analyte
    colnames(pvals) <- c(paste0("coef_", rownames(res)),
                         paste0("p-value_", rownames(res)))
    pvals
}

#' tests
#gal_met_prot_data[1:5, c("Sex", "Age", "FIB4")]
#gal_met_prot_data[1:5, ffq_items[1]]
#data <- gal_met_prot_data
#analyte <- ffq_items[1]
#funForLinReg(analyte, gal_met_prot_data)
#analyte <- ffq_items[14]
#funForLinReg(analyte, gal_met_prot_data)
#for (analyte in ffq_items) {
#  print(analyte)
#  funForLinReg(analyte, gal_met_prot_data)
#}

```

run the LM.

```{r linear-model-prot}
#' Fit linear models and extract coefficients and p-values on each protein
lm_output <- lapply(proteins, FUN = funForLinReg, data = gal_met_prot_data) |>
    do.call(what = rbind)

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)
```

Next we calculate the effect sizes.

To obtain comparable coefficients, prior to linear modelling all the
proteins
are standardized to zero mean and unit variance (note that this does not
influence the linear models since they are affine equivariant). In such
autoscaled data, difference of 1 corresponds to a standard deviation of 1. This
parameter will be called "effect size" and will be use to evaluate group
differences.

```{r lm--effectsize, message=FALSE}
#' Perform autoscaling for effect size (es) calculation
gal_met_prot_data_scaled <- gal_met_prot_data
gal_met_prot_data_scaled[, proteins] <- gal_met_prot_data[, proteins] |>
    as.matrix() |>
    scale(scale = TRUE)

#' fit linear models and extract the coefficients
es <- lapply(proteins, FUN = funForLinReg, data = gal_met_prot_data_scaled) |>
    do.call(what = rbind)
es <- es[, grep("coef", colnames(es))]
colnames(es) <- sub("coef", "effectSize", colnames(es))
lm_output <- cbind(lm_output, es)
```

To call a proteins *significant* we consider, in addition to the *statistical*
significance levels also the magnitude of its coefficient, relative to the
coefficient of variation (CV) for that protein. The CV, calculated on
study-internal QC samples (pools of all samples) are a measure of the technical
noise specific for that protein in the present data set. Thus, with this method
we prioritize proteins with a larger difference between the observed
(biological) difference and the (separately estimated) technical variance.

Below we add thus evaluate for each protein and comparison whether it
fulfills that criteria.

```{r lm--define significance criteria}
#' define the different coefficients as comparisons
comps <- colnames(lm_output)[grep("coef", colnames(lm_output))]

#' define which proteins are considered statistically significant
sign_p <- lm_output[, grep("p-adj", colnames(lm_output))] < 0.05
colnames(sign_p) <- sub("p-adj_", "significant_", colnames(sign_p))
#colSums(sign_p)

#' Convert the (absolute) difference in abundance into a difference in
#' percentage based on a (log2) coefficient.
diff_percentage <- function(x) {
    (2^abs(x) - 1) * 100
}
diff_perc <- apply(lm_output[, comps], MARGIN = 2, diff_percentage)
colnames(diff_perc) <- sub("coef", "diff_perc", colnames(diff_perc))

#' Define for which proteins the difference in concentration is larger
#' than a certain proportion of the CV
cv_cut <- lapply(comps, function(z) {
    z <- sub("coef_", "", z)
    diff_perc[, paste0("diff_perc_", z)] > 50 *  ##after eval, set to 50
        prot_ann[proteins, "cv_qc_chris"]
}) |>
    do.call(what = cbind)
colnames(cv_cut) <- sub("coef_", "cv_cut_", comps)

#' Use statistical significance for non-categorical traits instead of bool
sign_cv <- sign_p & cv_cut
sign_cv[, "significant_Age_10"] <- sign_p[, "significant_Age_10"]

#' Combine with result from linear model converting to a data.frame to
#' avoid logicals being converted to numeric
lm_output <- data.frame(lm_output, sign_cv)
```

The number of proteins called significant or with an adjusted p-value smaller
than 0.05 for each of the extracted coefficients are shown below. For a complete
comparison, the number of proteins complying to either the adjusted p value
or either the CV-value, as single threshold criteria, are calculated and
summarized here as well.

```{r lm--table-sig, results = "asis"}
#' merge to p-adj to cv significance summary
sig_tab <- data.frame(
    `p_adj < 0.05` = colSums(sign_p),
    `cv thresh` = colSums(cv_cut),
    significant = colSums(lm_output[, sub("coef_", "significant_", comps)]),
    check.names = FALSE)

pandoc.table(
    sig_tab, style = "rmarkdown",
    caption = paste0("Number of significant proteins per tested covariate. ",
                     " *p_adj < 0.05*: number of statistically significant",
                     " proteins.",
                     " *cv*: number of proteins with percentage abundance",
                     " difference above the CV-based threshold.",
                     " *significant*: number of proteins that fulfill ",
                     "the CV-based criteria in addition to the p-adj. threshold"))
```

Export results for inspection:

```{r lm--export-results}
results_prot <- data.frame(
    prot_ann[rownames(lm_output),
             c("description", "long_description", "PeptideNN",
               "PeptideNames", "Protein.Ids", "Protein.Names",
               "Genes", "Modified.Sequence", "Stripped.Sequence",
               "Precursor.Id", "cv_qc_chris")],
    lm_output,
    diff_perc
)

#' inclusing inspection of the CV-value of each protein
handle <- file.path(RDATA_PATH, 'lm_output_results_proteins.txt')
write.table(results_prot, file=handle, sep ="\t",
            row.names = FALSE, col.names = TRUE)
```

The full results are shown below.

```{r, eval = TRUE}
keep <- grepl("description|Genes|coef|p.adj", colnames(results_prot))
tmp <- results_prot[, keep]
datatable(tmp) |>
    formatRound(columns = which(vapply(tmp, is.numeric, logical(1))),
                digits = 3)
```

We visualize the associations in a heatmap. The effect size are added for the
associations, for the significant proteins (min 1 significance to
covariate). also for the not significant associations, the effect size is shown.

```{r prot-lm--heatmap, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for proteins with at least one significant association."}
#' Extract effect sizes
es_results <- results_prot[, grep("effectSize_", colnames(results_prot))] |>
    as.matrix()

#' Restrict to those significant for at least one of the groups (sex, age, SLD type)
sig_results <- results_prot[, grep("significant_", colnames(results_prot))] |>
    as.matrix()
es_results <- es_results[rowSums(sig_results) > 0, ]
print('nr of proteins with >= 1 significant association:')
nrow(es_results)

#' hierarchical clustering
clust1 <- hclust(dist((es_results)), method = "ward.D")
clust2 <- hclust(dist(t(es_results)), method = "ward.D")

#' clean roundup nrs for plot
es_results[es_results == 0] <- NA
es_results <- round(es_results, 2)

#' info regarding FIB4
#print('nr of proteins with a not-zero effect size in the FIB4 group:')
#sum(es_results[, 'effectSize_FIB4'] != 0)

#' in long format
es_results_long <- melt(as.matrix(es_results))

#' ggplot
plt <- es_results_long |>
    ggplot(aes(Var2, Var1, fill = value, label = value)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Effect Size") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 2.5) +
    theme_classic() +
    scale_x_discrete(
        expand = c(0, 0), limits = colnames(es_results)[clust2$order]) +
    scale_y_discrete(
        expand = c(0, 0), limits = rownames(es_results)[clust1$order]) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 10),
          axis.text.y = element_text(hjust = 1, size = 8),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.position = "none")

ggsave(paste0(IMAGE_PATH, "protein_SLDtypes_effect_sizes.png"),
       plot = plt, dpi = 600, width = 12,
       height = 12, units = "cm", scale = 2.5)
plt

```

In second plot, only the significant effect size labels were added to the
heat map. also, the dendograms after hierarchical clustering are shown.

```{r prot-lm--heatmap2, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for significant associations the value of the effect size is shown."}
#' keep just the effect sizes of the significant associations
es_results_sig <- es_results
es_results_sig[as.matrix(sig_results)[rownames(es_results), ] == 0] <- ""

#' plot
colnames(es_results) <- sub("effectSize_", "", colnames(es_results))
pheatmap(es_results, display_numbers = es_results_sig,
         number_color = "#444444", fontsize_number = 7)

```

As expected, the strongest effects are visible in the comparison between
steatosis grade 4 and 1.

In the third heat map, we plot the coefficients of molecules with >= 1
significant association in the comparisons of naab13_cat1:

```{r proteins-heatmap3, fig.path = IMAGE_PATH, res = 150, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for significant associations the value of the effect size is shown."}

#' Extract coefficients
coef_results_prot <- results_prot[, grep("coef_", colnames(results_prot))] |>
    as.matrix()

#' drop sex, age
coef_results_prot <- coef_results_prot[, c("coef_liver_fat_cat12", "coef_liver_fat_cat13", "coef_liver_fat_cat14")] |>
    as.matrix()

#' Restrict to those significant for at least one of the groups (sex, age, SLD type)
sig_results_prot <- results_prot[, c("significant_liver_fat_cat12", "significant_liver_fat_cat13", "significant_liver_fat_cat14")] |>
    as.matrix()
coef_results_prot <- coef_results_prot[rowSums(sig_results_prot) > 0, ]
print('nr of proteins with >= 1 significant association:')
nrow(coef_results_prot)

#' hierarchical clustering
clust1 <- hclust(dist((coef_results_prot)), method = "ward.D")
clust2 <- hclust(dist(t(coef_results_prot)), method = "ward.D")

#' clean roundup nrs for plot
coef_results_prot[coef_results_prot == 0] <- NA
coef_results_prot <- round(coef_results_prot, 2)

#' in long format
coef_results_prot_long <- melt(as.matrix(coef_results_prot))

#' ggplot
plt <- coef_results_prot_long |>
    ggplot(aes(Var2, Var1, fill = value, label = value)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Coefficient") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 3) +
    theme_classic() +
    #scale_x_discrete(
    #    expand = c(0, 0), limits = colnames(coef_results_prot)[clust2$order]) +
    scale_y_discrete(
        expand = c(0, 0), limits = rownames(coef_results_prot)[clust1$order]) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 5),
          axis.text.y = element_text(hjust = 1, size = 10),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.position = "none")

ggsave(paste0(IMAGE_PATH, "protein_naab13cat1_HMcoef.png"),
       plot = plt, dpi = 600, width = 12,
       height = 12, units = "cm", scale = 2.5)
plt


#' keep just the Coefficients of the significant associations
coef_results_prot_sig <- coef_results_prot
coef_results_prot_sig[as.matrix(sig_results_prot)[rownames(coef_results_prot), ] == 0] <- ""

#' plot
#colnames(coef_results_prot) <- sub("Coefficient_", "", colnames(coef_results_prot))
#pheatmap(coef_results_prot, display_numbers = es_results_prot_sig,
#         number_color = "#444444", fontsize_number = 7,
#         cluster_cols = FALSE)

```

We next create volcano plots showing the results for the individual steatosis
categories.

```{r prot-FIG-volcano-steatosis, fig.path = IMAGE_PATH, fig.width = 15, fig.height = 5, fig.cap = "Volcano plots for the protein associations with the various steatosis categories (compared to category 1)."}
plot_volcano <- function(x, comparison = character(), ...) {
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    plot(X, -log10(Y), pch = 21,
         col = "#000000ce", bg = "#00000060",
         xlab = "coefficient", ylab = expression(-log[10](p[adjusted])),
         ...)
}
significant_points <- function(x, comparison = character(),
                               diff_perc = 2, col = "#E41A1CCE",
                               bg = "#E41A1C40", ...) {
    s <- x[, paste0("significant_", comparison)]
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    if (any(s))
        points(X[s], -log10(Y[s]), col = col, bg = bg, pch = 21)
}

par(mfrow = c(1, 3))
xl <- range(results_prot[, c("coef_liver_fat_cat12", "coef_liver_fat_cat13",
                             "coef_liver_fat_cat14")])
yl <- c(0, -log10(min(results_prot[, c("p.adj_liver_fat_cat12",
                                       "p.adj_liver_fat_cat13",
                                       "p.adj_liver_fat_cat14")])))
sigs <- results_prot$significant_liver_fat_cat12 |
    results_prot$significant_liver_fat_cat13 |
    results_prot$significant_liver_fat_cat14

plot_volcano(results_prot, "liver_fat_cat12", main = "Category 2",
             xlim = xl, ylim = yl)
points(results_prot$coef_liver_fat_cat12[sigs],
       -log10(results_prot$p.adj_liver_fat_cat12[sigs]),
       pch= 21, col = "#E41A1CCE", bg = "#E41A1C40")
grid()
abline(h = -log10(0.05), lty = 2, col = "grey")
plot_volcano(results_prot, "liver_fat_cat13", main = "Category 3",
             xlim = xl, ylim = yl)
points(results_prot$coef_liver_fat_cat13[sigs],
       -log10(results_prot$p.adj_liver_fat_cat13[sigs]),
       pch= 21, col = "#E41A1CCE", bg = "#E41A1C40")
grid()
abline(h = -log10(0.05), lty = 2, col = "grey")
plot_volcano(results_prot, "liver_fat_cat14", main = "Category 4",
             xlim = xl, ylim = yl)
points(results_prot$coef_liver_fat_cat14[sigs],
       -log10(results_prot$p.adj_liver_fat_cat14[sigs]),
       pch= 21, col = "#E41A1CCE", bg = "#E41A1C40")
grid()
abline(h = -log10(0.05), lty = 2, col = "grey")

```

Difference of abundance increases with the steatosis grade. Also, the
significant proteins for steatosis grade 4 show a trend of difference already in
steatosis grade 2 and 3 participants.

Proteins with significant different abundance for at least one steatosis grade
class are shown below.

```{r lm--table-sig-sex, results = "asis", echo = FALSE}
tmp <- results_prot[sigs, c("Genes",
                       "coef_SexMale", "p.adj_SexMale",
                       "coef_Age_10", "p.adj_Age_10",
                       "coef_liver_fat_cat12", "p.adj_liver_fat_cat12",
                       "coef_liver_fat_cat13", "p.adj_liver_fat_cat13",
                       "coef_liver_fat_cat14", "p.adj_liver_fat_cat14")]
tmp <- tmp[order(tmp$p.adj_liver_fat_cat14), ]

pandoc.table(
    tmp, style = "rmarkdown", split.table = Inf,
    caption = paste0("Protiens significantly associated with at least ",
                     "on steatosis category"))
```

Significant proteins associated with steatosis (top 5 strongest associated
proteins) are visualized below. Points are colored according to sample sex.

```{r lm--prep for plotting}
#' prep for plotting
PLOTLY <- FALSE
ggplotly_or_not <- function(x, interactive = TRUE) {
    if (interactive) ggplotly(x)
    else print(x)
}
myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))

box_plot <- function(data, aes, analyte) {
    ggplot(data, aes) +
        geom_jitter(position = position_jitter(0.24),
                    aes(colour = Sex, alpha = 0.5 , size = (BMI/10))) +
        geom_violin(trim = FALSE, aes(alpha = 0.5)) +
        stat_summary(fun = median, geom = "point", shape = 23, size = 2) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) + labs(title = analyte) +
        theme_bw ()
}

scatter_plot <- function(data, aes, analyte) {
    ggplot(data, aes) +
        geom_jitter(position = position_jitter(0.24),
                    aes(colour = Sex, alpha = 0.5 , size = 0.1)) +
        stat_smooth(method = lm, se = TRUE) + #with standard error
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) + labs(title = analyte) +
        theme_bw ()
}

#' for boxplot func to work automatically
tmp_colnr <- which(colnames(gal_met_prot_data) == TARGET)
gal_met_prot_data$tmp_target <- gal_met_prot_data[, tmp_colnr]   #value of eg BMI
```

```{r prot-lm-top1, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[1]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]  #intensities of metabolite
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r prot-lm-top2, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[2]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r prot-lm-top3, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[3]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r prot-lm-top4, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[4]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r prot-lm-top5, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[5]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

### Summary

- 20 instead of 24 proteins are significantly changed between contrl-grade 3 
steatosis, when adding PDI_index as covariate. 
- 0 metabolites are significant if looking at pdi score


### Compare LM with and without PDI covariate - scatter plot

read the coeficients from model with and without

```{r}
#' load results from y ~ Sex + Age_10 + pdi_score + liver_fat_cat1
handle <- 'data/RData/NAFLD_steatosis_foodindex_influence/lm_output_results_proteins.txt'
coef_with_prot <- read.table(handle, header = TRUE, sep = "\t")  |>
    as.data.frame()
#head(coef_with_prot)

#' load results from y ~ Sex + Age_10 + liver_fat_cat1
handle <- 'data/RData/NAFLD_steatosis_associations/lm_output_results_proteins.txt'
coef_without_prot <- read.table(handle, header = TRUE, sep = "\t")  |>
    as.data.frame()
#head(coef_without_prot)

```

plot scatter

```{r}
#' scatter
plot(coef_without_prot$coef_liver_fat_cat14, coef_with_prot$coef_liver_fat_cat14,
     xlab = expression(coef),
     ylab = expression(coef[+PDI]), main = "Steatosis grade 3 vs normal",
     pch = 21, col = "#00000080", bg = "#00000040")

```

diff proteins are: lost: "Gelsolin" "Complement factor I"            
 "Corticosteroid-binding globulin" "Ficolin-3"    
and won new signif: /.
 
=> Those were not discussed in paper, as not >1 metabolite altered in within the
same pathway.

```{r}

#' find the ones that are unique to coef without pdi (48)
setdiff(coef_without_prot[coef_without_prot$significant_liver_fat_cat14, ]$long_description,
       coef_with_prot[coef_with_prot$significant_liver_fat_cat14, ]$long_description)

#' find the ones that are unique to coef without pdi (48)
setdiff(coef_with_prot[coef_with_prot$significant_liver_fat_cat14, ]$long_description,
       coef_without_prot[coef_without_prot$significant_liver_fat_cat14, ]$long_description)

```

check: 24 - 4 + 0 = 20


### Compare LM with and without PDI covariate

compare the with and without interaction LMs from above. Use the log likelihood
ratio test.

```{r log-likelihood-ratio-test-prot}
library(lmtest)

#' for dataset SWATH proteins
variables <- proteins

#' loop over all the metabolites
pvals <- c()
vars <- c()

for (i in variables){
  #print(i)

  #analyte
  analyte <- i

  #data
  data <- gal_met_prot_data
  data <- data[complete.cases(data[, analyte]), ]
  y <- data[, analyte]
  data <- data[, c(COVARIATES,
                   TARGET)]
  data <- Filter(function(x) length(unique(x)) > 1, data)

  #lm with
  lm_with <- lm(y ~ Sex + Age_10 + pdi_score + liver_fat_cat1,
                          data = data, na.action = na.omit)
  #lm without
  lm_without <- lm(y ~ Sex + Age_10 + liver_fat_cat1,
                          data = data, na.action = na.omit)

  #' info: https://www.statology.org/likelihood-ratio-test-in-r/
  #' h0: both models fit data equally well
  #' if p < 0.05, reject null hypothesis (aka a significant improvement in fit over the 1st model)
  #perform likelihood ratio test for differences in models
  llr <- lrtest(lm_with, lm_without)

  # Extract p-value
  p_value <- llr$`Pr(>Chisq)`[2]
  #print(p_value)  #warning: is rounded up to 0 is very small

  #' save info
  vars <- c(vars, i)
  pvals <- c(pvals, p_value)
}

llr_summary <- NULL
llr_summary$variables <- vars
llr_summary$pvals <- pvals
llr_summary <- as.data.frame(llr_summary)

```

For each p-value in the llr, if p-value > 0.05, we keep the null hypothesis. 

Thus, we would conclude that the model with the PDI covariate in the predictors 
offers EQUAL fit over the model without the PDI covariate in the predictors.

print the results for PDI and all proteins/metabolites with signif differ. 
Note MDG: this is ignoring FDR? so...

```{r summary-llr-test-prot}
#' All pvals
#pandoc.table(llr_summary, style = "rmarkdown", split.table = Inf,
#             caption = "P-values from likelihood ratio test")

#' which protein/metab is p-value < 0.05, so does not behave same?
tmp <- llr_summary[llr_summary$pvals < 0.05, ]
tmp
nrow(tmp)

```

for these 9, differs:

          variables       pvals
6            P02649 0.038029009
13           P07360 0.015030745
19           P01031 0.012112569
21  P06396;P06396-2 0.011919936
34           P00751 0.019058778
64           P08603 0.008878359
98           P22352 0.018512294
99           P06396 0.005681851
101          P05156 0.032024852


## Association between serum metabolites and liver fat (steatosis)

Perform multiple linear regression analyses for each metabolite to identify
those with significant differences in abundances between steatosis categories.

Model setup with "estim_daily_nutr_intake" and "pdi_score" in addition:
as estim is all coef=0, rm and keep only pdi_score. this can be compared with 
the model in "NAFLD_steatosis_associations.html" 

```{r define-lm-covariates-and-target}
print('covariates of the linear regression model:')
COVARIATES <- c("Sex", "Age_10", "pdi_score")
print(COVARIATES)

print('target of the linear regression model:')
TARGET <- 'liver_fat_cat1'
print(TARGET)

DATASET <- 'NAFLD'
print(paste0('final variables of the ', DATASET, ' dataset used:'))

print('nr of variables inspected:')
print(length(metabolites))

print('nr of individuals inspected:')
print(nrow(gal_met_prot_data))
```

Linear regression models are fitted for all metabolites separately (as a
response) with sex, age and the (categorical) steatosis category as a
covariates.

Model coefficients and p-values are provided. Raw p-values are adjusted for
multiple hypothesis testing with the method from Benjamini & Hochberg.

```{r linear model, message=FALSE}

#' Fit linear models and extract coefficients and p-values on each protein
lm_output <- lapply(metabolites, FUN = funForLinReg,
                    data = gal_met_prot_data) |>
    do.call(what = rbind)

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)
```

Next we calculate the effect sizes.

To obtain comparable coefficients, prior to linear modelling all the metabolites
are standardized to zero mean and unit variance (note that this does not
influence the linear models since they are affine equivariant). In such
autoscaled data, difference of 1 corresponds to a standard deviation of 1. This
parameter will be called "effect size" and will be use to evaluate group
differences.

```{r effectsize, message=FALSE}
#' Perform autoscaling for effect size (es) calculation
gal_met_prot_data_scaled <- gal_met_prot_data
gal_met_prot_data_scaled[, metabolites] <- gal_met_prot_data[, metabolites] |>
    as.matrix() |>
    scale(scale = TRUE)

#' fit linear models and extract the coefficients
es <- lapply(metabolites, FUN = funForLinReg,
             data = gal_met_prot_data_scaled) |>
    do.call(what = rbind)
es <- es[, grep("coef", colnames(es))]
colnames(es) <- sub("coef", "effectSize", colnames(es))
lm_output <- cbind(lm_output, es)
```

To call a metabolites *significant* we consider, in addition to the
*statistical* significance levels also the magnitude of its coefficient,
relative to the coefficient of variation (CV) for that protein. The CV,
calculated on study-internal QC samples (pools of all samples) are a measure of
the technical noise specific for that protein in the present data set. Thus,
with this method we prioritize metabolites with a larger difference between the
observed (biological) difference and the (separately estimated) technical
variance.

Below we add thus evaluate for each protein and comparison whether it
fulfills that criteria.

```{r define significance criteria}
#' define the different coefficients as comparisons
comps <- colnames(lm_output)[grep("coef", colnames(lm_output))]

#' define which metabolites are considered statistically significant
sign_p <- lm_output[, grep("p-adj", colnames(lm_output))] < 0.05
colnames(sign_p) <- sub("p-adj_", "significant_", colnames(sign_p))
#colSums(sign_p)

#' Convert the (absolute) difference in abundance into a difference in
#' percentage based on a (log2) coefficient.
diff_percentage <- function(x) {
    (2^abs(x) - 1) * 100
}
diff_perc <- apply(lm_output[, comps], MARGIN = 2, diff_percentage)
colnames(diff_perc) <- sub("coef", "diff_perc", colnames(diff_perc))

#' Define for which metabolites the difference in concentration is larger
#' than a certain proportion of the CV
cv_cut <- lapply(comps, function(z) {
    z <- sub("coef_", "", z)
    diff_perc[, paste0("diff_perc_", z)] > 50 *  ##after eval, set to 50
        met_ann[metabolites, "cv_qc_chris"]
}) |>
    do.call(what = cbind)
colnames(cv_cut) <- sub("coef_", "cv_cut_", comps)

#' Use statistical significance for non-categorical traits instead of bool
sign_cv <- sign_p & cv_cut
sign_cv[, "significant_Age_10"] <- sign_p[, "significant_Age_10"]

#' Combine with result from linear model converting to a data.frame to
#' avoid logicals being converted to numeric
lm_output <- data.frame(lm_output, sign_cv)
```

The number of metabolites called significant or with an adjusted p-value smaller
than 0.05 for each of the extracted coefficients are shown below. For a complete
comparison, the number of metabolites complying to either the adjusted p value
or either the CV-value, as single threshold criteria, are calculated and
summarized here as well.

```{r table-sig, results = "asis"}
#' merge to p-adj to cv significance summary
sig_tab <- data.frame(
    `p_adj < 0.05` = colSums(sign_p),
    `cv thresh` = colSums(cv_cut),
    significant = colSums(lm_output[, sub("coef_", "significant_", comps)]),
    check.names = FALSE)

pandoc.table(
    sig_tab, style = "rmarkdown",
    caption = paste0("Number of significant metabolites per tested covariate. ",
                     " *p_adj < 0.05*: number of statistically significant",
                     " metabolites.",
                     " *cv*: number of metabolites with percentage abundance",
                     " difference above the CV-based threshold.",
                     " *significant*: number of metabolites that fulfill ",
                     "the CV-based criteria in addition to the p-adj. threshold"))
```

Export results for inspection:

```{r export-results}
results_met <- data.frame(
    met_ann[rownames(lm_output), c("analyte_name", "analyte_class",
                                   "biochemical_name", "cv_qc_chris")],
    lm_output,
    diff_perc
)

#' write to file for manual inspection,
#' inclusing inspection of the CV-value of each protein
handle <- file.path(RDATA_PATH, 'lm_output_results_metabolites.txt')
write.table(results_met, file=handle, sep ="\t",
            row.names = FALSE, col.names = TRUE)
```

The full results are shown below.

```{r, eval = TRUE}
keep <- grepl("analyte_name|coef|p.adj", colnames(results_met))
tmp <- results_met[, keep]
datatable(tmp) |>
    formatRound(columns = which(vapply(tmp, is.numeric, logical(1))),
                digits = 3)
```

We visualize the associations in a heatmap. The effect size are added for the
associations, for the significant metabolites (min 1 significance to
covariate). also for the not significant associations, the effect size is shown.

```{r met-heatmap, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for metabolites with at least one significant association."}
#' Extract effect sizes
es_results <- results_met[, grep("effectSize_", colnames(results_met))] |>
    as.matrix()

#' Restrict to those significant for at least one of the groups (sex, age, SLD type)
sig_results <- results_met[, grep("significant_", colnames(results_met))] |>
    as.matrix()
es_results <- es_results[rowSums(sig_results) > 0, ]
print('nr of metabolites with >= 1 significant association:')
nrow(es_results)

#' hierarchical clustering
clust1 <- hclust(dist((es_results)), method = "ward.D")
clust2 <- hclust(dist(t(es_results)), method = "ward.D")

#' clean roundup nrs for plot
es_results[es_results == 0] <- NA
es_results <- round(es_results, 2)

#' info regarding FIB4
#print('nr of metabolites with a not-zero effect size in the FIB4 group:')
#sum(es_results[, 'effectSize_FIB4'] != 0)

#' in long format
es_results_long <- melt(as.matrix(es_results))

#' ggplot
plt <- es_results_long |>
    ggplot(aes(Var2, Var1, fill = value, label = value)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Effect Size") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 2.5) +
    theme_classic() +
    scale_x_discrete(
        expand = c(0, 0), limits = colnames(es_results)[clust2$order]) +
    scale_y_discrete(
        expand = c(0, 0), limits = rownames(es_results)[clust1$order]) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 10),
          axis.text.y = element_text(hjust = 1, size = 8),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.position = "none")

ggsave(paste0(IMAGE_PATH, "protein_SLDtypes_effect_sizes.png"),
       plot = plt, dpi = 600, width = 12,
       height = 12, units = "cm", scale = 2.5)
plt

```

In second plot, only the significant effect size labels were added to the
heat map. also, the dendograms after hierarchical clustering are shown.

```{r met-heatmap2, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for significant associations the value of the effect size is shown."}
#' keep just the effect sizes of the significant associations
es_results_sig <- es_results
es_results_sig[as.matrix(sig_results)[rownames(es_results), ] == 0] <- ""

#' plot
colnames(es_results) <- sub("effectSize_", "", colnames(es_results))
pheatmap(es_results, display_numbers = es_results_sig,
         number_color = "#444444", fontsize_number = 7)
```

In the third heat map, we plot the coefficients of molecules with >= 1
significant association in the comparisons of naab13_cat1:

```{r metabolites-heatmap3, fig.path = IMAGE_PATH, res = 150, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for significant associations the value of the effect size is shown."}

#' Extract coefficients
coef_results_met <- results_met[, grep("coef_", colnames(results_met))] |>
    as.matrix()

#' drop sex, age
coef_results_met <- coef_results_met[, c("coef_liver_fat_cat12", "coef_liver_fat_cat13", "coef_liver_fat_cat14")] |>
    as.matrix()

#' Restrict to those significant for at least one of the groups (sex, age, SLD type)
sig_results_met <- results_met[, c("significant_liver_fat_cat12", "significant_liver_fat_cat13", "significant_liver_fat_cat14")] |>
    as.matrix()
coef_results_met <- coef_results_met[rowSums(sig_results_met) > 0, ]
print('nr of proteins with >= 1 significant association:')
nrow(coef_results_met)

#' hierarchical clustering
clust1 <- hclust(dist((coef_results_met)), method = "ward.D")
clust2 <- hclust(dist(t(coef_results_met)), method = "ward.D")

#' clean roundup nrs for plot
coef_results_met[coef_results_met == 0] <- NA
coef_results_met <- round(coef_results_met, 2)

#' in long format
coef_results_met_long <- melt(as.matrix(coef_results_met))

#' ggplot
plt <- coef_results_met_long |>
    ggplot(aes(Var2, Var1, fill = value, label = value)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Coefficient") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 1.5) +
    theme_classic() +
    #scale_x_discrete(
    #    expand = c(0, 0), limits = colnames(coef_results_met)[clust2$order]) +
    scale_y_discrete(
        expand = c(0, 0), limits = rownames(coef_results_met)[clust1$order]) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 5),
          axis.text.y = element_text(hjust = 1, size = 5),
          plot.margin = margin(t = 5, l = 145, r = 145, b = 5), #top left right bottom 
          legend.position = "none")

ggsave(paste0(IMAGE_PATH, "metab_naab13cat1_HMcoef.png"),
       plot = plt, dpi = 600, width = 12,
       height = 12, units = "cm", scale = 2.5)
plt


#' keep just the Coefficients of the significant associations
coef_results_met_sig <- coef_results_met
coef_results_met_sig[as.matrix(sig_results_met)[rownames(coef_results_met), ] == 0] <- ""

#' plot
#colnames(coef_results_met) <- sub("Coefficient_", "", colnames(coef_results_met))
#pheatmap(coef_results_met, display_numbers = es_results_met_sig,
#         number_color = "#444444", fontsize_number = 7,
#         cluster_cols = FALSE)

```

We investigate the associations for MASLD, i.e. the difference in average
concentrations between participants of SLD types *MASLD* and *none*. We show the
results for that coefficient using a volcano plot, coloring significant
associations in red.

```{r met-FIG-volcano-steatosis, fig.path = IMAGE_PATH, fig.width = 15, fig.height = 5, fig.cap = "Volcano plots for associations of metabolites with steatosis categories."}
par(mfrow = c(1, 3))
xl <- range(results_met[, c("coef_liver_fat_cat12", "coef_liver_fat_cat13",
                            "coef_liver_fat_cat14")])
yl <- c(0, -log10(min(results_met[, c("p.adj_liver_fat_cat12",
                                      "p.adj_liver_fat_cat13",
                                      "p.adj_liver_fat_cat14")])))
sigs <- results_met$significant_liver_fat_cat12 |
    results_met$significant_liver_fat_cat13 |
    results_met$significant_liver_fat_cat14

plot_volcano(results_met, "liver_fat_cat12", main = "Category 2",
             xlim = xl, ylim = yl)
points(results_met$coef_liver_fat_cat12[sigs],
       -log10(results_met$p.adj_liver_fat_cat12[sigs]),
       pch= 21, col = "#E41A1CCE", bg = "#E41A1C40")
grid()
abline(h = -log10(0.05), lty = 2, col = "grey")
plot_volcano(results_met, "liver_fat_cat13", main = "Category 3",
             xlim = xl, ylim = yl)
points(results_met$coef_liver_fat_cat13[sigs],
       -log10(results_met$p.adj_liver_fat_cat13[sigs]),
       pch= 21, col = "#E41A1CCE", bg = "#E41A1C40")
grid()
abline(h = -log10(0.05), lty = 2, col = "grey")
plot_volcano(results_met, "liver_fat_cat14", main = "Category 4",
             xlim = xl, ylim = yl)
points(results_met$coef_liver_fat_cat14[sigs],
       -log10(results_met$p.adj_liver_fat_cat14[sigs]),
       pch= 21, col = "#E41A1CCE", bg = "#E41A1C40")
grid()
abline(h = -log10(0.05), lty = 2, col = "grey")
```

Similar to the proteomics analysis, significance of difference in abundance
increases. For metabolites we see however a large number of significant
metabolites for the comparison between steatosis grade 3 and 1.

Metabolites with significant different abundance for at least one steatosis
grade class are shown below.

```{r, results = "asis", echo = FALSE}
tmp <- results_met[sigs, c("analyte_name",
                           "coef_SexMale", "p.adj_SexMale",
                           "coef_Age_10", "p.adj_Age_10",
                           "coef_liver_fat_cat12", "p.adj_liver_fat_cat12",
                           "coef_liver_fat_cat13", "p.adj_liver_fat_cat13",
                           "coef_liver_fat_cat14", "p.adj_liver_fat_cat14")]
tmp <- tmp[order(tmp$p.adj_liver_fat_cat14), ]

pandoc.table(
    tmp, style = "rmarkdown", split.table = Inf,
    caption = paste0("Metabolites significantly associated with at least ",
                     "on steatosis category"))
```

Significant metabolites associated with steatosis (top 5 strongest associated
metabolites) are visualized below. Points are colored according to sample sex.

```{r met-lm-top1, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[1]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]  #intensities of metabolite
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r met-lm-top2, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[2]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r met-lm-top3, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[3]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r met-lm-top4, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[4]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r met-lm-top5, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[5]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

### Summary

- 45 instead of 48 metabolites are significantly changed between contrl-grade 3 
steatosis, when adding PDI_index as covariate. 
- 0 metabolites are significant if looking at pdi score


### Compare LM with and without PDI covariate - scatter plot

read the coeficients from model with and without

```{r}
#' load results from y ~ Sex + Age_10 + pdi_score + liver_fat_cat1
handle <- 'data/RData/NAFLD_steatosis_foodindex_influence/lm_output_results_metabolites.txt'
coef_with_metab <- read.table(handle, header = TRUE, sep = "\t")  |>
    as.data.frame()
#head(coef_with_metab)

#' load results from y ~ Sex + Age_10 + liver_fat_cat1
handle <- 'data/RData/NAFLD_steatosis_associations/lm_output_results_metabolites.txt'
coef_without_metab <- read.table(handle, header = TRUE, sep = "\t")  |>
    as.data.frame()
#head(coef_without_metab)

```

plot scatter

```{r}
#' scatter
plot(coef_without_metab$coef_liver_fat_cat14, coef_with_metab$coef_liver_fat_cat14,
     xlab = expression(coef),
     ylab = expression(coef[+PDI]), main = "Steatosis grade 3 vs normal",
     pch = 21, col = "#00000080", bg = "#00000040")

```

diff metabolites are:  lost "PC aa C42:2"   "PC ae C38:5"   "SM (OH) C22:2" "SM C20:2" 
and won new signif: "PC aa C38:3".

=> PCs and SMs are discussed, in SI. for PCs, are enough other singif (24#), 
BUT for SMs. story changes:

"The sphingolipids, two sphingomyelins (SM C16:0, SM C24:1) and one hydroxylated 
sphingomyelin (SM(OH) C22:2) were less abundant in participants with grade 3 steatosis."

FUCK: txt 50# with 3 SMs and fig2 (48#) NOT THE SAME!!!

```{r}

#' find the ones that are unique to coef without pdi (48)
setdiff(coef_without_metab[coef_without_metab$significant_liver_fat_cat14, ]$analyte_name,
        coef_with_metab[coef_with_metab$significant_liver_fat_cat14, ]$analyte_name)

#' find the ones that are unique to coef without pdi (48)
setdiff(coef_with_metab[coef_with_metab$significant_liver_fat_cat14, ]$analyte_name,
        coef_without_metab[coef_without_metab$significant_liver_fat_cat14, ]$analyte_name)

```

check: 48 - 4 + 1 = 45


### Compare LM with and without PDI covariate - LLR

compare the with and without pdi covariate LMs from above. Use the log
likelihood ratio test.

```{r log-likelihood-ratio-test-metab}
library(lmtest)

#' for dataset p180 metabolites
variables <- metabolites

#' loop over all the metabolites
pvals <- c()
vars <- c()

for (i in variables){
  #print(i)

  #analyte
  analyte <- i

  #data
  data <- gal_met_prot_data
  data <- data[complete.cases(data[, analyte]), ]
  y <- data[, analyte]
  data <- data[, c(COVARIATES,
                   TARGET)]
  data <- Filter(function(x) length(unique(x)) > 1, data)

  #lm with
  lm_with <- lm(y ~ Sex + Age_10 + pdi_score + liver_fat_cat1,
                          data = data, na.action = na.omit)
  #lm without
  lm_without <- lm(y ~ Sex + Age_10 + liver_fat_cat1,
                          data = data, na.action = na.omit)

  #' info: https://www.statology.org/likelihood-ratio-test-in-r/
  #' h0: both models fit data equally well
  #' if p < 0.05, reject null hypothesis (aka a significant improvement in fit over the 1st model)
  #perform likelihood ratio test for differences in models
  llr <- lrtest(lm_with, lm_without)

  # Extract p-value
  p_value <- llr$`Pr(>Chisq)`[2]
  #print(p_value)  #warning: is rounded up to 0 is very small

  #' save info
  vars <- c(vars, i)
  pvals <- c(pvals, p_value)
}

llr_summary <- NULL
llr_summary$variables <- vars
llr_summary$pvals <- pvals
llr_summary <- as.data.frame(llr_summary)

```

For each p-value in the llr, if p-value > 0.05, we keep the null hypothesis. 

Thus, we would conclude that the model with the PDI covariate in the predictors 
offers EQUAL fit over the model without the PDI covariate in the predictors.

print the results for PDI and all proteins/metabolites that changed signif.
Note MDG: this is ignoring FDR? so ...

```{r summary-llr-test}
#' all pvals
#pandoc.table(llr_summary, style = "rmarkdown", split.table = Inf,
#             caption = "P-values from likelihood ratio test")

#' which protein/metab is p-value < 0.05, so does not behave same?
tmp <- llr_summary[llr_summary$pvals < 0.05, ]
tmp
nrow(tmp)

```

7              Cit 0.046226781
10             Gln 0.015896425
54           C18:1 0.008621638
71              H1 0.007293453
73  lysoPC a C16:0 0.024924469
76  lysoPC a C18:0 0.002862579
78  lysoPC a C18:2 0.018601506
81  lysoPC a C24:0 0.003209068
84  lysoPC a C28:0 0.033155920
96     PC aa C34:2 0.038986821
101    PC aa C36:2 0.018815404
102    PC aa C36:3 0.044569882
119    PC aa C42:1 0.019636742
120    PC aa C42:2 0.014375154
126    PC ae C30:2 0.044524886
131    PC ae C34:2 0.037502543
132    PC ae C34:3 0.023542256
133    PC ae C36:0 0.036211698
135    PC ae C36:2 0.030447679
136    PC ae C36:3 0.045107979
144    PC ae C38:5 0.014668528
146    PC ae C40:1 0.047323642
150    PC ae C40:5 0.010551782
152    PC ae C42:0 0.028190909
153    PC ae C42:1 0.047619847
154    PC ae C42:2 0.041163006
157    PC ae C42:5 0.008117250
161    PC ae C44:6 0.013920563
167       SM C16:0 0.019591646
168       SM C16:1 0.003010586


# Nice scatter plot for SI paper

We merge the 2 scatter plots to put in paper, or at least to share with Essi in
the discussion.

Export for the paper SI Table S2B:

```{r FIG-scatter_PDI, fig.path = IMAGE_PATH, fig.width = 15, fig.height = 7, fig.cap = "Scatter plots for the associations with and without adding PDI as covariate."}
#' 2 plots horizontally
par(mfrow = c(1, 2))

#' fixed limits
xl <- range(coef_without_prot$coef_liver_fat_cat14,
            coef_with_prot$coef_liver_fat_cat14,
            coef_without_metab$coef_liver_fat_cat14,
            coef_with_metab$coef_liver_fat_cat14)

#' scatter with grid, abline
plot(coef_without_prot$coef_liver_fat_cat14, coef_with_prot$coef_liver_fat_cat14,
     xlab = expression(Coefficient),
     ylab = expression(Coefficient_with_PDI),
     main = "Steatosis grade 3 vs normal - Proteins",
     xlim = xl, ylim = xl,
     pch = 21, col = "#00000080", bg = "#00000040")
grid()
abline(a = 0, b = 1, lty = 2, col = "grey") #y = bx + a
plot(coef_without_metab$coef_liver_fat_cat14, coef_with_metab$coef_liver_fat_cat14,
     xlab = expression(Coefficient),
     ylab = expression(Coefficient_with_PDI),
     main = "Steatosis grade 3 vs normal - Metabolites and lipids",
     xlim = xl, ylim = xl,
     pch = 21, col = "#00000080", bg = "#00000040")
grid()
abline(a = 0, b = 1, lty = 2, col = "grey")

```


# Acknowledgments

Consulted sources used as basis for code:

- ~/nafld_proteo_metabolomics/NAFLD_metab_targeted_lm.Rmd
- ~?metabo_chris_untargeted/CHRIS_lm_pos.Rmd
- kora: CHRIS_KORA_tmetab_lm_binary.Rmd


# Session information

R packages used for the analysis:

```{r}
sessionInfo()
```
