---
title: "NAFLD associated Scanning SWATH proteins"
author: "Nikola Dordevic, Johannes Rainer"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
#bibliography: references.bib
---

**Modified**: `r file.info("NAFLD_related_proteins.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r biocstyle, echo = FALSE, results = "hide", message = FALSE}
#' rmarkdown format settings
library(BiocStyle)
library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8, fig.height = 7)
BiocStyle::markdown()
```

```{r settings}
#' General settings
filename <- "NAFLD_related_proteins"
#' Path to save images to
IMAGE_PATH <- paste0("images/", filename, "/")
if (file.exists(IMAGE_PATH))
    unlink(IMAGE_PATH, recursive = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE)
#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)
chris_path <- "."
```


# Introduction

This document describes the analysis to identify Scanning SWATH proteins that 
are related to liver stiffness.


# Data import

Below we load all required packages for the analysis and retrieve the data.

```{r libraries}
library(chrisUtils)
library(chrisData)
library(nafldData)
library(RColorBrewer)
library(pander)
library(pheatmap)
library(DT)
library(ggfortify)
library(ggplot2)
library(ggpubr)
library(reshape2)
library(plotly)
options(rgl.useNULL = TRUE)
library(rgl)
library(readxl)
library(writexl)
library(tidyfr)
#library(atc)
```

The data from Scanning Swath was converted into the TDF format and can hence be
loaded and imported using the `tidyfr` R package. Now we read the full data
which includes general information for each measurement along with the
quantified protein levels.


```{r, message = FALSE}
##
scan <- data_module(name = "chris_scanningswath_proteins", version = "1.0.0.1",
                    path = ".")
protSS_data <- data(scan)
protSS_ann <- labels(scan)
stopifnot(all(colnames(protSS_data) == rownames(protSS_ann)))
colnames(protSS_data) <- protSS_ann$description
rownames(protSS_ann) <- protSS_ann$description
## define the columns with peptide concentrations
proteins <- rownames(protSS_ann)[protSS_ann$Genes != ""]
protSS_data <- protSS_data[, proteins]
```

Scaninng SWATH data is already log2 transformed!
Before proceeding with the data analysis we restrict to samples from the NAFLD 
study.

```{r keep-NAFLDs, message = FALSE, warning = FALSE}
protSS_data <- protSS_data[grep("^008", rownames(protSS_data)), ]
```

Demographics of the NAFLD cohort and evaluation of available trait data.

```{r}
n_anns <- nafldTraitAnnotations()
#' personal data
n_perso <- nafldData("personal")
rownames(n_perso) <- n_perso$AID
##                                          
protSS_data$AID <- n_perso[rownames(protSS_data), "AID"]
protSS_data$AID_CHRIS <- n_perso[rownames(protSS_data), "aid_chris"]
protSS_data$Age <- n_perso[rownames(protSS_data), "na_ager"]
protSS_data$Sex <- n_perso[rownames(protSS_data), "na_sex"]
protSS_data$fwup <- n_perso[rownames(protSS_data), "na_fwup"]
#' na_ager: rounded age
#' na_fwup: follow-up duration CHRIS->NAFLD
tmp <- nafldData("recruitment")
rownames(tmp) <- tmp$AID
protSS_data$case <- tmp[rownames(protSS_data), "na_case"]
```

The time span between the two studies is thus between 2 to 5 years. We next
evaluate BMI and fat composition for the study participants, first split by
gender and then by case/control definition.

```{r}
#' BMI, fat etc.
n_anthro <- nafldData("anthro")
rownames(n_anthro) <- n_anthro$AID

protSS_data$BMI <- n_anthro[rownames(protSS_data), "naan03"]
protSS_data$body_fat <- n_anthro[rownames(protSS_data), "naan04"]
protSS_data$visc_fat <- n_anthro[rownames(protSS_data), "naan05"]
```

We next evaluate the clinical examinations performed on all participants.

```{r}
#' Clinical examinations
tmp <- nafldData("clinical")
rownames(tmp) <- tmp$AID
```

We will use and evaluate the following parameters defined by a clinical
examination:

- *naab11*: liver stiffness (higher values are worse). Values for participants
  that have a *naab12* (liver stiffness IQR) > 30 should be either removed or
  taken with care.
- *naab12*: liver stiffness IQR: the IQR of repeated liver stiffness
  measurements. Values below 30 are considered OK.
- *naab13*: liver fat. Can be binarized at an absolute value of 288.
- *naab14*: subjective, visual evaluation of the liver fat. Categorical from 1
  to 3. Values 2 and 3 can be considered steatotic.
- *nascfib4*: fibrosis 4 index. Linear score calculated from standard clinical
  parameters.
- *nascmets*: metabolic syndrome, binary category.

```{r}
protSS_data$liver_stiffness <- tmp[rownames(protSS_data), "naab11"]
protSS_data$liver_stiffness_iqr <- tmp[rownames(protSS_data), "naab12"]
protSS_data$liver_fat <- tmp[rownames(protSS_data), "naab13"]
protSS_data$visual_fat <- tmp[rownames(protSS_data), "naab14"]
protSS_data$fib4 <- tmp[rownames(protSS_data), "nascfib4"]
protSS_data$metsym <- tmp[rownames(protSS_data), "nascmets"]
```

We exclude  measurements/participants that have a value >= 30 for variable 
"naab12" (which  is the liver stiffness IQR). The liver stiffness values for 
these are expected to be of bad quality.

```{r}
protSS_data <- protSS_data[-which(protSS_data$liver_stiffness_iqr >= 30),]
```

# General data overview and visualization
 
In resulting dataset, there are `r length(proteins)` proteins measured from`
r nrow(protSS_data)` samples in total and general overview of the trait is 
summarized below:

```{r, message=FALSE}
Overview <- summary(
        protSS_data[, c("Sex", "Age", "BMI", "case", "body_fat", "visc_fat", 
                        "liver_stiffness" , "liver_stiffness_iqr", "liver_fat",     
                        "visual_fat", "fib4", "metsym")])
print(Overview)
```

                        
Pearson correlation between liver stiffness and BMI:

```{r}
cor(protSS_data$liver_stiffness, protSS_data$BMI, use = "complete.obs")
```

Spearman correlation between liver stiffness and BMI:

```{r}
cor(protSS_data$liver_stiffness, protSS_data$BMI, method = "spearman", 
    use = "complete.obs")
```
 
In next figure, BMI vs. liver stiffness is visualized in scatter plot. Points 
are colored acording case-control info and shaped with sample sex:

```{r, BP_PVCbyBMI_fig2, fig.path = IMAGE_PATH, fig.width=10, fig.height=6}
BMIbyliver_stiffness <- ggplot(protSS_data, aes(x = liver_stiffness, y = BMI, 
color = case, shape = Sex)) + geom_point() + 
theme(axis.text.x = element_text(angle = 90, vjust = 0.8, size = 8),
axis.title.y = element_blank()) + labs(title = "BMI Vs. liver stiffness") +
theme_bw ()
ggplotly(BMIbyliver_stiffness)
```


Histogram of liver stiffness is visualized as a raw score and after log and log2 
transformation:

```{r, hist_fig, fig.path = IMAGE_PATH, fig.width=10, fig.height=4}
##
par(mfrow = c(1, 3))
hist(protSS_data$liver_stiffness, main ="raw score")
hist(log(protSS_data$liver_stiffness), main ="Log transformed")
hist(log2(protSS_data$liver_stiffness), main ="Log2 transformed")
```


# Linear regression analysis

To identify proteins related with liver stiffness, linear regression models 
are fitted for all proteins separately (as a response) and sex, age and liver 
stiffness as a covariates. 
Model coefficients and p-values are provided. Raw p-values are adjusted for 
multiple hypothesis testing with the method from Benjamini & Hochberg.

   
```{r, message=FALSE}
## make a function for LM
FunForLinReg <- function(AnalyteName, Data) {
    Analyte <- Data[, AnalyteName]
    LMReg <- lm(Analyte ~ Sex + Age + liver_stiffness,
                data = Data)
    res <- coef(summary(LMReg))[-1, ]
    p.Values <- data.frame(c(res[, 1],
                             res[, 4]))
    names(p.Values) <- AnalyteName
    rownames(p.Values) <- c(paste0("coef_", rownames(res)),
                            paste0("p-value_", rownames(res)))
    return(p.Values)
}
## do it in one run

LMRegtests <- do.call(cbind.data.frame,
                      lapply(proteins,
                             function(x) FunForLinReg(x, protSS_data)))
Test_output <- t(LMRegtests)
## Adjusting for multiple hypothesis testing
adjp <- apply(Test_output[, grep("p-value", colnames(Test_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
Test_output <- cbind(Test_output, adjp)
```

To obtain comparable coefficients, prior to linear modelling all the proteins
are standardized to zero mean and unit variance (note that this does not
influence the linear models since they are affine equivariant).
In autoscaled data, difference of 1 corresponds to standard deviation of 1.
This parameter will be called "effect size" and will be use to evaluate group
differences.

```{r, message=FALSE}
## perform data normalization (autoscaling)
protSS_data_AS <- protSS_data
protSS_data_AS[,proteins] <- scale(protSS_data_AS[,proteins], scale = TRUE)
## do it in one run
LMRegtests_AS <- do.call(cbind.data.frame,
                      lapply(proteins,
                             function(x) FunForLinReg(x, protSS_data_AS)))
Test_output_AS <- t(LMRegtests_AS)
EffectSize <- Test_output_AS[,grep("coef",colnames(Test_output_AS))]
colnames(EffectSize) <- sub("coef", "effect_size", colnames(EffectSize))
```

To call a proteins *significant* we require that it has, in addition to an
adjusted p-value smaller than 0.05, also a difference in abundances (expressed
in percentage) which is at least once as high as the coefficient of variation
for that proteine. The coefficient of variation was calculated on the QC CHRIS
Pool samples hence it represents the technical variability that was observed for
each protein.

Below we add thus evaluate for each protein and comparison whether it fulfills
that criteria.

```{r define-significant}
#' Calculate the (absolute) difference in abundance expressed as a percentage
#' based on a (log2) coefficient.
diff_percentage <- function(x) {
    (2^abs(x) - 1) * 100
}
comps <- colnames(Test_output)[grep("coef", colnames(Test_output))]
#' restrict to categorical variables
comps <- comps[comps != "coef_Age"]
diff_perc <- apply(Test_output[, comps],
                   MARGIN = 2, diff_percentage) 
colnames(diff_perc) <- sub("coef", "diff_perc", colnames(diff_perc))   
#' define which metabolites are considered significant
sigs <- lapply(comps, function(z) {
    z <- sub("coef_", "", z)
    Test_output[, paste0("p-adj_", z)] < 0.05 &
        diff_perc[, paste0("diff_perc_", z)] > 
                                       100 * protSS_ann[proteins, "cv_qc_chris"]
})
sigs <- do.call(cbind, sigs)
colnames(sigs) <- sub("coef", "significant", comps)
#' Add the significant for numeric variable
sigs <- cbind(sigs, significant_Age = Test_output[, "p-adj_Age"] < 0.05)
sigs[,"significant_liver_stiffness"] <- Test_output[, "p-adj_liver_stiffness"] < 0.05
```

The number of protein called significant or with an adjusted p-value smaller
than 0.05 for each of the extracted coefficients are shown below.

```{r table-sig-mets, results = "asis"}
sig_tab <- apply(Test_output[, grep("adj", colnames(Test_output))], 2,
                 function(z) sum(z < 0.05))
names(sig_tab) <- sub("p-adj_", "", names(sig_tab))
sig_tab <- data.frame(
    `p_adj < 0.05` = sig_tab,
    significant = colSums(sigs[, paste0("significant_", names(sig_tab))]),
    check.names = FALSE)
pandoc.table(data.frame(no_sign = sig_tab, check.names = FALSE),
             style = "rmarkdown",
             caption = "Number of significant proteins for each comparison.")
```

Export results:

```{r export-results}
results <- data.frame(
    protSS_ann[rownames(Test_output), c("Genes", "long_description")],
    Test_output,
    EffectSize,
    diff_perc,
    sigs)
```

Volcano plot for the liver stiffness. Red colored points are those with an 
adjusted p-value smaller than 0.05.


```{r FIG-volcano-sex, fig.path = IMAGE_PATH, fig.width = 8, fig.height = 8, fig.cap = "Volcano plot for the liver stiffness. Red colored points are those with an adjusted p-value smaller than 0.05."}
plot_volcano <- function(x, comparison = character(), ...) {
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    plot(X, -log10(Y), pch = 21,
         col = "#000000ce", bg = "#00000060",
         xlab = "coefficient", ylab = expression(-log[10](p[adjusted])),
         ...)
}
significant_points <- function(x, comparison = character(),
                               diff_perc = 2, col = "#E41A1CCE",
                               bg = "#E41A1C40", ...) {
    sig <- x[, paste0("significant_", comparison)]
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    if (any(sig))
        points(X[sig], -log10(Y[sig]), col = col, bg = bg, pch = 21)
}
plot_volcano(results, "liver_stiffness", main = "liver stiffness")
abline(h = -log10(0.05), lty = 2, col = "grey")
significant_points(results, comparison = "liver_stiffness")
```

The (static) table with the significant proteins is shown below:

```{r table-sig-sex, results = "asis", echo = FALSE}
significant_table <- function(x, comparison = character(),
                              columns = c("Genes", "long_description")) {
 sigs <- x[, paste0("significant_", comparison)]
 columns <- c(columns, paste0(c("coef_", "p.adj_", "effect_size_"), comparison))
 tmp <- x[sigs, columns]
 tmp <- tmp[order(tmp[, paste0("p.adj_", comparison)]), ]
 colnames(tmp) <- sub(paste0("_", comparison), "", colnames(tmp))
 tmp
}
sig_tab <- significant_table(results, "liver_stiffness")
## save significant
Signif_AllAges <- rownames(sig_tab)
##
pandoc.table(sig_tab[, !colnames(sig_tab) %in% c("Genes", "long_description")],
             style = "rmarkdown", split.table = Inf,
             caption = "Significant protein for liver stiffness")
```

Show in table with the results for all proteins from the test ordered with the 
tests p-values and effect size:

```{r table-top-20, results = "asis"}
##
Order <- results[order(results[, "p.value_liver_stiffness"]), ]
#pandoc.table(Order[1:5, c(2, 6, 14, 22, 45)], style = "rmarkdown", 
#             split.table = Inf, caption = "Top 5 metabolites for PVC")
datatable(Order[,c(1, 2, 5, 8, 11, 14, 18)]) %>%
          formatRound(columns=c(3:6), digits = 6)
```


There are six significantly associated proteins for liver stiffness!
Significant proteins (top 3) are visualized in boxplot in next figure. Points 
are colored according to sample sex.

```{r, f1, fig.path = IMAGE_PATH, fig.width=10, fig.height=6}
Sp_f1 <- ggplot(protSS_data, aes(x = liver_stiffness, y = `P02774;P02774-3`, 
color = case, shape = Sex)) + geom_point() + 
theme(axis.text.x = element_text(angle = 90, vjust = 0.8, size = 8),
axis.title.y = element_blank()) + theme_bw ()
ggplotly(Sp_f1)
```

```{r, f2, fig.path = IMAGE_PATH, fig.width=10, fig.height=6}
Sp_f2 <- ggplot(protSS_data, aes(x = liver_stiffness, y = P04004, 
color = case, shape = Sex)) + geom_point() + 
theme(axis.text.x = element_text(angle = 90, vjust = 0.8, size = 8),
axis.title.y = element_blank()) + theme_bw ()
ggplotly(Sp_f2)
```
   
```{r, f3, fig.path = IMAGE_PATH, fig.width=10, fig.height=6}
Sp_f3 <- ggplot(protSS_data, aes(x = liver_stiffness, y = O95445, 
color = case, shape = Sex)) + geom_point() + 
theme(axis.text.x = element_text(angle = 90, vjust = 0.8, size = 8),
axis.title.y = element_blank()) + theme_bw ()
ggplotly(Sp_f3)
```


# Conclusion


# Session information

R packages used for the analysis:

```{r}
sessionInfo()
```

# References
