---
title: "NAFLD"
author: "Johannes Rainer"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 10, fig.height = 7)
```

# NAFLD data

Demographics of the NAFLD cohort and evaluation of available trait data.

```{r}
library(chrisData)
library(nafldData)
library(vioplot)

n_anns <- nafldTraitAnnotations()

#' personal data
n_perso <- nafldData("personal")
n_data <- n_perso[, c("AID", "aid_chris", "na_ager", "na_sex", "na_fwup")]
#' na_ager: rounded age
#' na_fwup: follow-up duration CHRIS->NAFLD

tmp <- nafldData("recruitment")
rownames(tmp) <- tmp$AID
n_data$case <- tmp[n_data$AID, "na_case"]

```

The data set is nicely balanced by sex:

```{r}
table(n_data$na_sex)
```

The age distribution of participants (split by sex) is shown below.

```{r, fig.cap = "Age distribution."}
vioplot(split(n_data$na_ager, n_data$na_sex), ylab = "age")
grid()
```

The distribution of time between participants' CHRIS baseline and NAFLD
participation is shown below.

```{r, fig.cap = "Distribution of the time span between CHRIS baseline and NAFLD."}
plot(density(n_data$na_fwup))
grid()
```

The time span between the two studies is thus between 2 to 5 years. We next
evaluate BMI and fat composition for the study participants, first split by
gender and then by case/control definition.

```{r}
#' BMI, fat etc.
n_anthro <- nafldData("anthro")
rownames(n_anthro) <- n_anthro$AID

n_data$BMI <- n_anthro[n_data$AID, "naan03"]
n_data$body_fat <- n_anthro[n_data$AID, "naan04"]
n_data$visc_fat <- n_anthro[n_data$AID, "naan05"]

```

```{r, fig.cap = "Distribution of BMI and body fat per sex or *case*."}
par(mfrow = c(2, 3))
#' split by sex
vioplot(split(n_data$BMI, n_data$na_sex), main = "BMI", ylab = "BMI")
grid()
vioplot(split(n_data$body_fat, n_data$na_sex), main = "Body fat",
        ylab = "Body fat")
grid()
vioplot(split(n_data$visc_fat, n_data$na_sex), main = "Visceral fat",
        ylab = "Visceral fat")
grid()
#' split by case
vioplot(split(n_data$BMI, n_data$case), main = "BMI", ylab = "BMI")
grid()
vioplot(split(n_data$body_fat, n_data$case), main = "Body fat",
        ylab = "Body fat")
grid()
vioplot(split(n_data$visc_fat, n_data$case), main = "Visceral fat",
        ylab = "Visceral fat")
grid()

```

While the distribution of the BMI is balanced between female and male
participants, females have on average a much higher body fat percentage, while
males have on average a higher proportion of visceral fat. For cases and
controls (defined for the recruitment of CHRIS participants) on average higher
BMI, body fat and visceral fat can be observed for *cases*.

Note that there seems to be no dependency between age and BMI in the present
data set (data not shown).

We next evaluate the clinical examinations performed on all participants.

```{r}
#' Clinical examinations
tmp <- nafldData("clinical")
rownames(tmp) <- tmp$AID
```

We will use and evaluate the following parameters defined by a clinical
examination:

- *naab11*: liver stiffness (higher values are worse). Values for participants
  that have a *naab12* (liver stiffness IQR) > 30 should be either removed or
  taken with care.
- *naab12*: liver stiffness IQR: the IQR of repeated liver stiffness
  measurements. Values below 30 are considered OK.
- *naab13*: liver fat. Can be binarized at an absolute value of 288.
- *naab14*: subjective, visual evaluation of the liver fat. Categorical from 1
  to 3. Values 2 and 3 can be considered steatotic.
- *nascfib4*: fibrosis 4 index. Linear score calculated from standard clinical
  parameters.
- *nascmets*: metabolic syndrome, binary category.

```{r}
n_data$liver_stiffness <- tmp[n_data$AID, "naab11"]
n_data$liver_stiffness_iqr <- tmp[n_data$AID, "naab12"]
n_data$liver_fat <- tmp[n_data$AID, "naab13"]
n_data$visual_fat <- tmp[n_data$AID, "naab14"]
n_data$fib4 <- tmp[n_data$AID, "nascfib4"]
n_data$metsym <- tmp[n_data$AID, "nascmets"]
```

From the `r nrow(n_data)` participants, 
`r sum(n_data$liver_stiffness_iqr >= 30, na.rm = TRUE)`
have a liver stiffness IQR `>= 30`.

At first we evaluate the distribution of the liver stiffness score in the data
set.

```{r, fig.cap = "Distribution of liver stiffness measurements. All values considered. Problematic values are highlighted in red in the lower panel.", fig.height = 10}
stiff_ok <- which(n_data$liver_stiffness_iqr < 30)
par(mfrow = c(2, 2))
vioplot(split(log2(n_data$liver_stiffness), n_data$na_sex),
        main = "Liver stiffness", ylab = expression(log[2](Liver~stiffness)))
grid()
vioplot(split(log2(n_data$liver_stiffness), n_data$case),
        main = "Liver stiffness", ylab = expression(log[2](Liver~stiffness)))
grid()
col <- rep("#ff0000", nrow(n_data))
col[stiff_ok] <- "#000000"
x <- n_data$na_ager
y <- log2(n_data$liver_stiffness)
plot(x, y, pch = 21, bg = paste0(col, "40"),
     col = paste0(col, "80"), main = "Liver stiffness",
     xlab = "age", ylab = expression(log[2](Liver~stiffness)))
grid()
abline(lm(y ~ x), lty = 1, col = "#00000080")
x <- n_data$BMI
plot(x, y, pch = 21, bg = paste0(col, "40"),
     col = paste0(col, "80"), main = "Liver stiffness",
     xlab = "BMI", ylab = expression(log[2](Liver~stiffness)))
grid()
abline(lm(y ~ x), lty = 1, col = "#00000080")
```

While there is a clear average difference for liver stiffness between male and
female as well as case and control cases, there seems to be no general age
dependency of the liver stiffness. Only for a subset of participants there seems
to be a (log linear) increase with age. In contrast, a (log-linear) relationship
between BMI and liver stiffness can be observed.

We next evaluate the distribution of liver fat values in the study cohort.

```{r, fig.cap = "Distribution of liver fat values in the study cohort.", fig.height = 10}
par(mfrow = c(2, 2))
vioplot(split(n_data$liver_fat, n_data$na_sex),
        main = "Liver fat", ylab = "Liver fat")
grid()
vioplot(split(n_data$liver_fat, n_data$case),
        main = "Liver fat", ylab = "Liver fat")
grid()
x <- n_data$na_ager
y <- n_data$liver_fat
plot(x, y, pch = 21, main = "Liver fat", xlab = "age", ylab = "Liver fat",
     bg = "#00000040", col = "#00000080")
grid()
abline(lm(y ~ x), lty = 1, col = "#00000080")
x <- n_data$BMI
plot(x, y, pch = 21, main = "Liver fat", xlab = "BMI", ylab = "Liver fat",
     bg = "#00000040", col = "#00000080")
grid()
abline(lm(y ~ x), lty = 1, col = "#00000080")
```

Similar to the liver stiffness, we see differences in average liver fat values
between male and female study participants as well as between cases and
controls, while there seems no age dependency. A strong relationship between BMI
and liver fat can be observed.

For completeness we also plot the visual fat categorization.

```{r, fig.cap = "Overview of visual fat categories.", fig.height = 10}
par(mfrow = c(2, 2))
split(n_data$visual_fat, n_data$na_sex) |>
    lapply(table) |>
    do.call(what = cbind) |>
    barplot()
split(n_data$visual_fat, n_data$case) |>
    lapply(table) |>
    do.call(what = cbind) |>
    barplot()
vioplot(split(n_data$na_ager, n_data$visual_fat), main = "Visual fat",
        ylab = "age")
grid()
vioplot(split(n_data$BMI, n_data$visual_fat), main = "Visual fat",
        ylab = "BMI")
grid()
```

A clear relationship can be seen for the visual fat evaluation with the cases as
well as BMI.

Next we evaluate the FIB4 score.

```{r, fig.cap = "Distribution of FIB4 scores.", fig.height = 10}
par(mfrow = c(2, 2))
vioplot(split(log2(n_data$fib4), n_data$na_sex),
        main = "FIB4", ylab = expression(log[2](FIB4)))
grid()
vioplot(split(log2(n_data$fib4), n_data$case),
        main = "FIB4", ylab = expression(log[2](FIB4)))
grid()
x <- n_data$na_ager
y <- log2(n_data$fib4)
plot(x, y, pch = 21, main = "FIB4", xlab = "age",
     ylab = expression(log[2](FIB4)),
     bg = "#00000040", col = "#00000080")
grid()
abline(lm(y ~ x), lty = 1, col = "#00000080")
x <- n_data$BMI
plot(x, y, pch = 21, main = "FIB4", xlab = "BMI",
     ylab = expression(log[2](FIB4)),
     bg = "#00000040", col = "#00000080")
grid()
abline(lm(y ~ x), lty = 1, col = "#00000080")
```

There seems to be only a small difference in average FIB4 values between male
and female participants or between cases and controls. A strong (positive)
relationship is however visible between FIB4 and age. A slight negative
relationship seems to be present between FIB4 and BMI.

```{r, fig.cap = "Overview of metabolic syndrome categories.", fig.height = 10}
par(mfrow = c(2, 2))
split(n_data$metsym, n_data$na_sex) |>
    lapply(table) |>
    do.call(what = cbind) |>
    barplot()
split(n_data$metsym, n_data$case) |>
    lapply(table) |>
    do.call(what = cbind) |>
    barplot()
vioplot(split(n_data$na_ager, n_data$metsym), main = "Metabolic syndrome",
        ylab = "age")
grid()
vioplot(split(n_data$BMI, n_data$metsym), main = "Metabolic syndrome",
        ylab = "BMI")
grid()
```

While no cases of metabolic syndrome seem to be balanced between male and female
participants and seem to be independent of the participants' ages, a strong
difference can be seen for cases and controls as well as for BMI.

```{r}
#' Drug usage. only categorical.
```

What still needs to be done:

- Diabetes diagnosis (using the definition from the shared document).


# NAFLD omics data overview

Get the IDs of NAFLD participants and their respective CHRIS AID.


```{r, message = FALSE}
library(chrisData)

gi <- chrisData("general")

nafld <- gi[which(gi$x0_nafld == "Yes"), ]
```

Participation year (in CHRIS baseline) for the `r nrow(nafld)` NAFLD study
participants.

```{r}
barplot(table(nafld$x0_examy))
```

We determine the number of participants with available CHRIS targeted
metabolomics as well as CHRIS ScanningSWATH and Somalogic proteomics data from
the baseline and NAFLD collections.

```{r, message = FALSE}
library(tidyfr)
soma <- data_module(name = "chris_somalogic", version = "1.0.3.2", path = ".")
scan <- data_module(name = "chris_scanningswath_proteins", version = "1.0.0.2",
                    path = ".")
p180 <- data_module(name = "metabolomics_p180", version = "1.0.1.2", path = ".")

soma_aid <- rownames(data(soma))
scan_aid <- rownames(data(scan))
p180_aid <- rownames(data(p180))
```

```{r, results = "asis"}
library(pander)
tmp <- data.frame(baseline = c(sum(nafld$AID %in% p180_aid),
                               sum(nafld$AID %in% scan_aid),
                               sum(nafld$AID %in% soma_aid)),
                  NAFLD = c(sum(nafld$aid_nafld %in% p180_aid),
                            sum(nafld$aid_nafld %in% scan_aid),
                            sum(nafld$aid_nafld %in% soma_aid)))
rownames(tmp) <- c("targeted metabolomics",
                   "ScanningSWATH proteomics",
                   "SomaLogic proteomics")
pandoc.table(tmp, style = "rmarkdown",
             caption = paste0("Number of NAFLD study participants (n = ",
                              nrow(nafld), ") with available omics data"))
```

Note: these numbers don't include any of the drop outs up to version 2 (July
2023).


```{r}
do <- read.table(system.file("txt", "drop-outs.txt", package = "chrisData"),
                 sep = "\t", header = TRUE, colClasses = "character")
any(do$aid %in% nafld$AID)
any(do$aid %in% nafld$aid_nafld)
```
