---
title: "NAFLD preprocessing pos"
author: "Philippine Louail"
date: "2023-10-05"
output: html_document
---

```{r style, message = FALSE, echo = FALSE, warning = FALSE, results = "asis"}
library("BiocStyle")
library("knitr")
library("rmarkdown")
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7)
```

# Packages

```{r}
library(MsExperiment)
library(RSQLite)
library(xcms)
library(Spectra)
library(RColorBrewer)
library(pander)
library(readxl)
library(MetaboCoreUtils)
library(pheatmap)
library(MsBackendSql)
```


# Data import 

Create a `MsExperiment` representing the data
```{r}
#' Load the spectra data from the database
sps <- Spectra("data/NAFLD_untargeted.sqlite", drv = SQLite(),
               source = MsBackendOfflineSql())

pd <- as.data.frame(read_xlsx("data/nafld_untargeted_metabolomics.xlsx"))

#' Subset to positive polarity
pd <- pd[pd$polarity == "POS", ]
#' remove samples here ? 

#' Subset the Spectra to the respective files
sps <- sps[basename(sps$dataOrigin) %in% pd$file]
sps$file <- basename(sps$dataOrigin)

#' Create MsExperiment and fill with data
nafld_pos <- MsExperiment()
sampleData(nafld_pos) <- DataFrame(pd)
spectra(nafld_pos) <- sps

#' Link spectra to samples
nafld_pos <- linkSampleData(nafld_pos, with = "sampleData.file = spectra.file")
```

```{r}
#' Set up parallel processing using 3 cores
if (.Platform$OS.type == "unix") {
    register(bpstart(MulticoreParam(3)))
} else {
    register(bpstart(SnowParam(3)))
}
```
s
# Data organisation

The experimental data is now a `MsExperiment` object:
```{r}
nafld_pos
```

There is `r length(sampleData(nafld_pos))` samples in this data set.

```{r phenodata, echo=FALSE, results='asis'}
sampleData(nafld_pos)[,-c(1,3,5,7)] |> 
    as.data.frame() |>
    head() |>
    pandoc.table(style = "rmarkdown", caption = "Samples from the data set.")

which(table(fromFile(nafld_pos)) <1000) ## one weird sample

nafld_pos <- nafld_pos[table(fromFile(nafld_pos)) > 1000] ## removing that one weird sample. 

```


```{r define-colors, include=FALSE}
#' Define colors for the groups.
col_phenotype <- brewer.pal(4, "Set1")[c(2, 1, 4, 3)]
names(col_phenotype) <- c("study", "POOL", "QC", "BLANK")

col_sample <- col_phenotype[sampleData(nafld_pos)$sample_type] 
```


We therefore have an data set of `r length(nafld_pos)` samples for a total of `r length(spectra(nafld_pos))` spectra. 


The retention time range for the entire data set is: `r range(rtime(spectra(nafld_pos)))`

# Visualisation 

BPS ?
BPC 
```{r bpc and ret time filtering }
#' First extract and plot bpc for full data
bpc <- chromatogram(nafld_pos, aggregationFun = "max")


#' Look into this weird area to see if usefull, also check some spectrum there to see the m/z, if superior to 500 prob just remove
data <- filterRt(nafld_pos, c(350, 380))
data_chrom <- chromatogram(data, aggregationFun = "max")
plot(data_chrom, col = paste0(col_phenotype[data_chrom$sample_type]), main = "check for 300 to 400 sec")
grid()
legend("topright", col = col_phenotype,
       legend = names(col_phenotype), lty = 1)

#' check some spectrum
data_spec <- spectra(data) |> filterRt(c(365,366))
plotSpectra(data_spec[1:4])

#' seems like the spectra are all below/around 200 mz

max <- intensity(data_spec) |> 
    max()|>
    which.max()

max

weird_sample <- dataOrigin(data_spec[465])

#' lets see his bpc 

table_sample <- sampleData(data)
which(table_sample$sample == "0080243672")

chromatogram(data[which(table_sample$sample == "0080243672")], aggregationFun = "max") |> 
    plot()
#' so 2 run for same sample, apparently there was ome issues, let's look if there's more like that
duplicate <- sampleData(nafld_pos)$sample
vector_of_duplcited <- which(duplicated(duplicate))
sample_duplicated <- duplicate[vector_of_duplcited]

sample_duplicated <- sample_duplicated[grep("^008", sample_duplicated)]
sample_duplicated
#' so all of these are duplicate 
#' let's jsut do preprocessing only on this part of dataset, 
#' subset for these plates (4 in total) - see johannes paper for plte to extract
batches <- c("20200204", "20201026", "20200901", "20210922")
#' look for the duplicated_nafld.Rmd file to see analysis of them
#' 
#' what do i do with these now ?



#'for now keep, look again after peak detection/refinement, 
#'if nothing then remove everything after 350s

```


```{r Plot bpc before filtering, echo=FALSE}
plot(bpc, col = paste0(col_phenotype[bpc$sample_type]), main = "BPC") #changed the code here idk if that is a  big problem
grid()
legend("topright", col = col_phenotype,
       legend = names(col_phenotype), lty = 1)
```

should I reduce ret time ?

# Visualize internal standard in all samples

```{r EIC extract for internal standard }

#' get the list 
intern_standard <- read.delim("C:/Users/plouail/Documents/GitHub/lcms-standards/data/internal_standards.txt")
View(intern_standard)
##'for one standard
x <- intern_standard[1, c(1,3,4,6)]
x$mz <- mass2mz(x$formula, adduct = x$POS)

#'chrom extraction for it's specific m/z range but full rt
chrom <- chromatogram(nafld_pos, mz = x$mz + c(-0.001, 0.001), 
                      aggregationFun = "max")

#'plot it add expected retention time: line(v =)
png(paste("chromatogram", x$name, ".png"))
plot(chrom, main = paste("chromatogram_all_sample_for_", x$name), col = paste0(col_phenotype[chrom$sample_type], 80))
grid()
abline(v = x$RT, col = "red", )
dev.off()

#'export into png and check it
#'
#'make it loop through internal standard list

#'make EIC now adapting for rt expected 
#' 
 

```


# Session information

```{r}
sessionInfo()
```

