---
title: "Mummichog annotation on NAFLD umetab (pos) after lm with liver_fat_cat2"
author: "Marilyn De Graeve, Johannes Rainer"
affiliation: "Eurac Research, Bolzano, Italy"
date: "2024-12-13"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
---

**Modified**: `r file.info("NAFLD_naab13_mummichog.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r configuration, include = TRUE, cached = FALSE}

#' introduction
STUDY <- "NAFLD cohort"
TARGET <- "liver_fat_cat2"
DATASET <- "untargeted metabolomics, positive mode"

#' settings
FILENAME <- "NAFLD_umetab_naab13cat2_mummichog"
LOCATION <- 'local' #'server' OR 'local', change in chunk directories if needed

#' target variables
#print('target of the linear regression model:')
TARGET_COLNAME <- 'liver_fat_cat2'
#print('more concrete defined target of the linear regression model:')
TARGET_LMNAME <- 'liver_fat'

```


```{r style, echo = FALSE, results = "hide", message = FALSE}
#' rmarkdown format settings
library("BiocStyle")
library("kableExtra")
library("knitr")
library("rmarkdown")
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7, time_it = TRUE)
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now, units = "secs")
      # return a character string to show the time
      paste("    Time for this code chunk to run:", round(res,
        2), "seconds")
    }
  }
}))
BiocStyle::markdown()

```

# Introduction

In this document, the Mummichog algorithm from *todo!!!* are performed on the
significant metabolite list on the targeted metabolomics data after statistical
analyis.

We use the logistic regression analysis results table from:
(nafld_proteo_metabolomics/NAFLD_metab_untargeted_pos_lm.html)[nafld_proteo_metabolomics/NAFLD_metab_untargeted_pos_lm.html].
There, the exploratory statistical analyis (linear models) are performed on the
untargeted metabolomics data from the CHRIS-NAFLD study; using the
catagoris variable "liver_fat_cat" as a target of interest. Other covariates
were sex and age. 

The (static) table with the significant metabolites contained 8
significant variables, to use the Mummichog algoritm, input is full list, so
also give the not-significant feature their adj p-values. 


# Setup

## Directories

```{r directories}
#' General settings
filename <- "NAFLD_umetab_naab13cat2_mummichog"

#' Path to the data
DATA_PATH <- '/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_metab_untargeted_pos_lm/' #FINDME!      /home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_metab_untargeted_pos_lm/
#DATA_OUTPUT_PATH <- "data"

```

## Packages

```{r packages}


```

## functions

```{r functions}


```

# Steps

The following steps are performed:

- [x] in the metabo_chris_untargeted/CHRIS_lm_pos.html: write table with the ALL
  (both significant and non-significant) metabolites with mz, rt, id, p-value (=
  adj.p lm) and statistic (= coef lm) after performing lm with `r TARGET` as
  target, using the `R DATASET` dataset.
- [x] rename colnames so c(m/z, retention_time, p_value, statistic,
  CompoundID_from_user, ...)
- [x] run mummichog 
  - [x] now via bash, 
  - [/] future w spectripy
- [x] load and cleanup results into 1 df.
- [/] add results as putative IDs into the sumExp object and save as Rdata.


# Load and rename results from lm

load the results from lm analysis. write table with the ALL
(both significant and non-significant) metabolites with mz, rt, id, p-value (=
  adj.p lm) and statistic (= coef lm) after performing lm with `r TARGET` as
  target, using the `R DATASET` dataset.

```{r load-results-lm}

#' load data lm-results
handle <- file.path(DATA_PATH, 'lm_output_results_umetab_liver_fat_cat2.txt')
lm_output_results_metab <- read.csv(handle, header = TRUE, sep = "\t")  |>
    as.data.frame()
lm_output_results_metab[1:10, 1:10]
dim(lm_output_results_metab)

```

rename colnames to c(m/z, retention_time, p_value, statistic,
CompoundID_from_user, ...) so it fits the input in mummichog, write the table.

```{r make-write-mummichog-input}
#' make new df with correct names for mummichog
df <- NULL
df$m.z  <- lm_output_results_metab$mzmed
df$retention_time  <- lm_output_results_metab$rtmed
df$p_value  <- lm_output_results_metab[, paste0('p.adj_', TARGET_LMNAME)]
df$statistic  <- lm_output_results_metab[, paste0('coef_', TARGET_LMNAME)]
df$CompoundID_from_user <- lm_output_results_metab$AID
df <- as.data.frame(df)
head(df)

#fyi: m/z ipv m.z, ok? yes


#' write the mummichog input table
handle <- file.path(DATA_PATH, 'mummichog_input.txt')
write.table(df, file = handle, sep = "\t", row.names = FALSE, col.names = TRUE)

```


# Run mummichog algoritm 

## Run mummichog algoritm using Bash

Before trying to integrate the mummichog algorithm using spectripy, we first
want to run it automatically from this script. it is run locally, on a Linux
computer with the Mummichog installed according to the instructions from
'https://github.com/shuzhao-li/mummichog'.

After install using pip in the linux terminal, it is run using commands:

- cd Documents/Files/Work_Eurac/Projects/biochem_pathway_analysis/
- python3 -m mummichog.main -f testdata0710.txt -o t2

the results are saved in the same folder, as the input. you will get a folder
with various output. in the first instance, i.e. for the putative annotation of
untargeted metabolic features, we are interested in 1 tabular output. See
section [Results import] below.

run the command using R (calls bash), for linux:

```{r run-mummichog-from-bash}

#TODO!!!
#command <- 'python3.10 -m mummichog.main -f ~/Documents/Files/Work_Eurac/Projects/biochem_pathway_analysis/testdata0710.txt -o t2' #test
#command <- 'bash ~/Documents/Files/Work_Eurac/Code/Sh_scripts/mummichog2.sh'
#command <- 'python3.10 -m mummichog.main -f ~/Documents/Files/Work_Eurac/Projects/biochem_pathway_analysis/data/mummichog_input.txt -o t2'
command <- 'python3.12 -m mummichog.main -f ~/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/data/RData/NAFLD_metab_untargeted_pos_lm/mummichog_input.txt -o t2' 

command
system(command)

```

This will be either called in the spectripy environment, if possible. will try
that next. therefor, no need to go too much into win/linux differences (see old
code marilyn if needed).


# Results from mummichog (pasted from html report)

For the NAFLD untargeted metabolomics - pos mode lm model with MAFLD sld type, 
pasted from the mummichog html report:


Mummichog Report: t2
mummichog version: 2.7.0, using metabolic model MFN_1.10.4.
User input data

User supplied 10882 features as reference list.We are using 78 features (p < 0.200000) as significant list. The feature level data are shown in the Manhattan plots below.
Top pathways

The reference feature list is mapped to 532 EmpiricalCompounds, significant features to 12 EmpiricalCompounds. In the table below, pathway_size is number of detected EmpiricalCompounds for each pathway; overlap_size is number of significant EmpiricalCompounds. Empirical p-values are estimated by permutation test. Details on EmpiricalCompounds are in 'tables/ListOfEmpiricalCompounds.tsv'.
Pathways 	overlap_size	pathway_size	p-value 	overlap_EmpiricalCompounds
Purine metabolism 	4	28	0.00143	E213,E488,E274,E14
Nitrogen metabolism 	2	6	0.00302	E274,E14
Glutathione Metabolism 	2	9	0.00672	E171,E14
Arginine and Proline Metabolism 	3	26	0.00723	E171,E274,E14
Vitamin B3 (nicotinate and nicotinamide) metabolism 	2	10	0.00781	E274,E14
Glutamate metabolism 	2	10	0.00781	E274,E14
Alanine and Aspartate Metabolism 	2	11	0.01042	E274,E14
Aminosugars metabolism 	2	16	0.01983	E274,E14
Valine, leucine and isoleucine degradation 	2	19	0.02412	E9,E14
Carbon fixation 	1	3	0.02571	E565
Lysine metabolism 	2	22	0.03042	E515,E14
Aspartate and asparagine metabolism 	3	49	0.03176	E171,E274,E14
Methionine and cysteine metabolism 	2	26	0.03832	E14,E606
Pyrimidine metabolism 	2	29	0.04571	E274,E14
Top modules

Network modules may capture concerted metabolite activities that are missed by predefined pathways. Use pull-down menus to visualize modules and the activity network, which tries to combine the results of pathway/module analyses. The visualization requires interenet connection.
Showing
Node size Visual style
module_1, p=0.00446, 5 metabolites
5-Oxoproline, L-Glutamate, L-Leucine, L-Isoleucine, L-Glutamine
EmpiricalCompounds favored by above analyses

An EmpiricalCompound is a computational unit for a tentative metabolite. It can group multiple ions, and be any of the isobaric/isomeric species. This table contains EmpiricalCompounds prioritized by pathway/module analysis.
EmpiricalCompound	Input m/z 	Retention time	ion	mz_diff	Statistic 	Significant
E171 	Best guess: C04282, 1-Pyrroline-4-hydroxy-2-carboxylate	C01879, C04281, C04282
	152.031900666236 	175.563251387819	M+Na[1+]	0.0	-0.02	no
	198.038224312001 	169.460765142265	M+HCOONa[1+]	0.0009	-0.21	no
	130.050533235302 	175.794675266091	M+H[1+]	0.0007	-0.07	no
	130.050326538094 	169.258016891767	M+H[1+]	0.0005	0.41	yes
	131.053423804078 	175.772996569593	M(C13)+H[1+]	0.0001	-0.13	yes
	131.052881275153 	169.351244113728	M(C13)+H[1+]	-0.0004	0.67	yes
E565 	Best guess: C00279, D-Erythrose 4-phosphate	C00279
	201.015802875844 	33.2074381743636	M+H[1+]	-0.0001	0.04	no
	222.997777237245 	33.1988098525901	M+Na[1+]	-0.0001	0.76	yes
E9 	Best guess: C00407, L-Isoleucine	C00123, C00407
	154.083826186935 	149.216792267092	M+Na[1+]	-0.0001	0.22	no
	132.102514777352 	149.119199747165	M+H[1+]	0.0006	0.18	no
	133.10538154778 	148.235535180792	M(C13)+H[1+]	0.0001	0.13	yes
E606 	Best guess: C02712, N-Acetylmethionine	C02712
	214.051329273015 	206.998387554432	M+Na[1+]	0.0005	-0.75	yes
E488 	Best guess: C11821, 5-Hydroxyisourate	C11821
	185.03055983449 	140.168844037611	M+H[1+]	-0.0	0.35	yes
	186.032803429111 	140.598836704487	M(C13)+H[1+]	-0.0012	0.3	no
	207.012405662767 	140.874345026774	M+Na[1+]	-0.0002	0.39	no
E515 	Best guess: C03793, N6,N6,N6-Trimethyl-L-lysine	C03793
	211.143887812461 	44.0575279018813	M+Na[1+]	0.0021	0.21	yes
	257.149628503066 	45.19503364425	M+HCOONa[1+]	0.0025	0.19	no
E213 	Best guess: C00294, Inosine	C00294
	269.088267405768 	142.609612316691	M+H[1+]	0.0002	-0.46	no
	270.091176060444 	142.789472665335	M(C13)+H[1+]	-0.0003	-0.61	yes
	291.07024352127 	142.966507264613	M+Na[1+]	0.0002	-0.41	no
E274 	Best guess: C00819, D-Glutamine	C00064, C05100, C00819
	169.058421834848 	175.584664990003	M+Na[1+]	0.0	-0.05	no
	147.07714663581 	175.767827420671	M+H[1+]	0.0008	-0.1	yes
	148.079397135472 	175.77525914372	M(C13)+H[1+]	-0.0004	-0.05	no
E14 	Best guess: C05938, L-4-Hydroxyglutamate semialdehyde	C00217, C00025, C05938
	148.060799740294 	169.340766074335	M+H[1+]	0.0003	0.36	yes
	149.06369420469 	169.343572290557	M(C13)+H[1+]	-0.0002	0.41	yes
More data

Full data tables from pathway analysis, module analysis and activity network are stored in the 'tables/' directory. The tsv files can be imported into a spreadsheet program, e.g. MS Excel. Under the 'figures/network_modules/' directory are files intended for Cytoscape (cytoscape.org) visualization. Please refer to Cytoscape's guides for details. Details of this run are recorded in file mummichog.log.txt.
Mummichog algorithms are described in Li et al. Predicting Network Activity from High Throughput Metabolomics. PLoS Computational Biology (2013); doi:10.1371/journal.pcbi.1003123. This software is provided as is, without warranty of any kind. Visit http://mummichog.org for updates.


# Results mummichog import

After running the algoritm (bash script on pc with py/mummichoc installed), 
todo into spectripy... !!!,
the following table os imported:

- userInput_to_EmpiricalCompounds.tsv

The first file is a copy of the data user gave to the algoritm, with the rowxxx identifiers needed as first column.
The second file gives the hypothetical annotations (kegg ID and comon name used in KEGG) for the retrieved metabolites.

please note that only metabolites which are present in KEGG can be retrieved using this approach.

TODO!!! path, filename to auto select latest or in config?? or see git if name output cn change name/dir??

```{r load-mummichig-result-table}
#' load mummichog result files
#' TODO rename using bash folder
#DIR <- "1732118659.9126198.t2"   #test
DIR <- '1734448517.6233082.t2'    #umetab-bmi-umetabpos-lm
DIR <- '1737464806.4199815.t2'    #nafld-masld-umetabpos-lm aka
DIR <- '1737464806.4199815.t2'    #nafld-umetab_MASLD_lm
DIR <- '1747388849.322413.t2'     #nafld_umetab_naab13cat2_lm
DATA_OUTPUT_PATH <- paste0('/home/mdegraeve/Documents/Files/Work_Eurac/Projects/nafld_proteo_metabolomics/', DIR, '/tables')
#was saved in the working directory in VScode!


#' userInput_to_EmpiricalCompounds
handle <- file.path(DATA_OUTPUT_PATH, "userInput_to_EmpiricalCompounds.tsv")
userInput_to_EmpiricalCompounds <- read.csv(handle, header = TRUE, sep="\t")  |>
    as.data.frame()
userInput_to_EmpiricalCompounds[1:10, 1:10]

```

info regarding results file:

- adducts = str_row_ion
- keggid = compounds, also sometimes CE id from bigg)
- name = compound_names

See remarks regarding the paradoxical results in the CHRIS/example rmd report.
See here the final selected clean up of the mummichog results.


```{r cleanup-table-unique-rowids}
#' checks
#' total rowids (inlc duplic)
nrow(userInput_to_EmpiricalCompounds)
#' nr of non-duplic
length(userInput_to_EmpiricalCompounds$input_row[!duplicated(as.list(userInput_to_EmpiricalCompounds$input_row))])
#' nr of duplic
length(unique(userInput_to_EmpiricalCompounds$input_row[duplicated(as.list(userInput_to_EmpiricalCompounds$input_row))]))
#' needed length if correct merged == unique
length(unique(as.list(userInput_to_EmpiricalCompounds$input_row)))
#' nr of exactly 1 time occur in df
length(subset(userInput_to_EmpiricalCompounds$input_row, !(duplicated(userInput_to_EmpiricalCompounds$input_row) | duplicated(userInput_to_EmpiricalCompounds$input_row, fromLast = TRUE))))


#' names of the replicate rowids to be merged
doubles <- userInput_to_EmpiricalCompounds$input_row[duplicated(as.list(userInput_to_EmpiricalCompounds$input_row))] #double hits, to merge into 1 rowid
doubles <- unique(doubles)

#' loop to make subdf with solved doubles
df_all <- NULL
i <- 1
for(i in 1:length(doubles)){
    #print(doubles[i])
        
    double_df <- userInput_to_EmpiricalCompounds[userInput_to_EmpiricalCompounds$input_row == doubles[i], ]

    #' add to empty df
    df <- NULL
    df$input_row <- double_df$input_row[1]
    df$EID <- paste(c(unlist(double_df$EID)),collapse = ';')
    df$str_row_ion <- paste(c(unlist(double_df$str_row_ion)),collapse = ';')
    df$compounds <- paste(c(unlist(double_df$compounds)),collapse = ';')
    df$compound_names  <- paste(c(unlist(double_df$compound_names)),collapse = ';')
    df$input_row.1 <- double_df$input_row[1]
    df$m.z  <- paste(c(unlist(double_df$m.z)),collapse = ';')
    df$retention_time  <- paste(c(unlist(double_df$retention_time)),collapse = ';')
    df$p_value  <- paste(c(unlist(double_df$p_value)),collapse = ';')
    df$statistic  <- paste(c(unlist(double_df$statistic)),collapse = ';')
    df$CompoundID_from_user <- paste(c(unlist(double_df$CompoundID_from_user)),collapse = ';')
    df <- as.data.frame(df)

    df_all <- rbind(df_all, df)
}

#' checks
dim(df_all)
df_all[, 1:3]
colnames(df_all)
length(unique(as.list(df_all$input_row)))

#' subset original mummichog results table with rowid nr exactly 1 time occur in df
df_unique <- subset(userInput_to_EmpiricalCompounds, !(duplicated(userInput_to_EmpiricalCompounds$input_row) | duplicated(userInput_to_EmpiricalCompounds$input_row, fromLast = TRUE)))
dim(df_unique)
colnames(df_unique)

#' merge the non-duplic and the solved into 1 df again
clean_df <- rbind(df_unique, df_all)
clean_df <- clean_df[order(clean_df$input_row),]
dim(clean_df)
rownames(clean_df) <- clean_df$input_row

#' write for testing
handle <- file.path(DATA_PATH, 'cleanup_mummichog_output.txt')
write.table(clean_df, file=handle, sep ="\t", row.names = FALSE, col.names = TRUE)

#fyi, top metab from lm FT01399 == glutamate according to mummichog

```

todo!!!! see in clean df still some rowIDs merged in the untargeted metab df!!!


The computational 'EmpiricalCompound' and other information for a tentative
metabolite is now linked to the feature ID from the untargeted metabolomics
dataset. The candidate metabolite can group multiple ions, and be any of the
isobaric/isomeric species. See the original Mummichog 'results.html' for the
visualisation and for the EmpiricalCompounds prioritized by
pathway/module analysis.  

TODO: how add these to umetab dataset? using summExp object add description
maybe? FINDME!!! (todo after lcms semi quant untarg added, see there how).


# Subset the significant metabolites from the lm, if contain annotation

The input data is loaded in the `lm_output_results_metab` object. After 
subsetting to keep only the significant untargeted metabolites, There are 8 
significantly associated metabolites for control vs. grade 3 steatosis.

the significant metabolites are then retrieved in the cleaned up table with
putative annotations is in the `clean_df` object.

```{r}
#subset lm output with only the significant
#dim(lm_output_results_metab)
#colnames(lm_output_results_metab)
lm_signif <- lm_output_results_metab[lm_output_results_metab[, paste0('significant_', TARGET_LMNAME)], ]  
print('nr of significant metabolites:')
dim(lm_signif)
print('significant metabolites from lm, both annotated and not-annotated:')
lm_signif$AID

#keep only the annotated by subsetting clean_df
lm_signif_with_annotation <- clean_df[clean_df$CompoundID_from_user %in% lm_signif$AID, ]
#head(lm_signif_with_annotation)
print('putative annotated significant metabolites:')
lm_signif_with_annotation$compound_names

#all info together of significant:
lm_signif_with_annotation

```


resuts pasted, which are found from 8 signif (steatosis explore: control vs 
grade 3) that also have a putative ID:

[1] Top 4 = **FT01935** = "**N-Acetylmethionine**"       
row1935	E606	row1935_M+Na[1+] C02712	N-Acetylmethionine
check: 214.1 m/z == 191.25 g/mol + 23 Da
check: 207 rtmed != 161 seconds methionine_13C_15N (approx)
coef = DOWN                                                                                            

[2] Top1 = **FT00689** = "5-Oxoproline$L-1-Pyrroline-3-hydroxy-5-carboxylate$**1-Pyrroline-4-hydroxy-2-carboxylate**", 
the exact molecule is dependent on the adduct:  
According to HTML mumm table: 
row689  E171  row1019_M+Na[1+];row1695_M+HCOONa[1+];row673_M+H[1+];row674_M+H[1+];row688_M(C13)+H[1+];**row689_M(C13)+H[1+]**.  C01879;C04281;C04282  5-Oxoproline$L-1-Pyrroline-3-hydroxy-5-carboxylate$**1-Pyrroline-4-hydroxy-2-carboxylate** 
check: 131.1 m/z == 129.11 g/mol of 1-Pyrroline-4-hydroxy-2-carboxylate + 2 Da M(C13)+H[1+]
check: 169.4 rtmed == similar to proline at 167 sec (by Mar).
coef = UP

[3] Top 5 = **FT00688** = "5-Oxoproline$L-1-Pyrroline-3-hydroxy-5-carboxylate$**1-Pyrroline-4-hydroxy-2-carboxylate**"  
row688	E171	row1019_M+Na[1+];row1695_M+HCOONa[1+];row673_M+H[1+];row674_M+H[1+];**row688_M(C13)+H[1+]**;row689_M(C13)+H[1+]	C01879;C04281;C04282	5-Oxoproline$L-1-Pyrroline-3-hydroxy-5-carboxylate$**1-Pyrroline-4-hydroxy-2-carboxylate**
check: FT00688 is 131.1 m/z == 129.11 g/mol of 1-Pyrroline-4-hydroxy-2-carboxylate + 2 Da M(C13)+H[1+]
check: 175.8 rtmed != similar to proline at 167 sec (by Mar).
coef = DOWN
!! note: feature same m/z Top1 but reverse coef. other rt, so other ID. also not
the isomer -5-. no way of knowing ID...                            

[4] Top 6 = **FT00949** = "D-Glutamate; D-Glutamic acid; D-Glutaminic acid; D-2-Aminoglutaric acid$L-Glutamate$**L-4-Hydroxyglutamate semialdehyde (best match, C12)**"
row949	E14	**row949_M+H[1+]**;row968_M(C13)+H[1+]	C00217;C00025;C05938	**D-Glutamate**; D-Glutamic acid; D-Glutaminic acid; D-2-Aminoglutaric acid$L-Glutamate$L-4-Hydroxyglutamate semialdehyde
check: 148.1 m/z != 146.12 g/mol Glutamate + 1 Da M+H[1+]
check: 148.1 m/z == 147.13 g/mol L-4-Hydroxyglutamate semialdehyde g/mol + 1 Da M+H[1+]
maybe better to double check in lc-msms data the RT of glutamate - By Mar: 171 sec.
check: 169.3 rtmed == 171 seconds glutamic_13C_15N expected approx. but L-4-Hydroxyglutamate semialdehyde very sim rt expected if looking at structure.
FOR OWN EXPERIENCE, this feature is probably the **C12 of L-4-Hydroxyglutamate semialdehyde**
coef = UP

[5] Top 7 = **FT00968** = "D-Glutamate; D-Glutamic acid; D-Glutaminic acid; D-2-Aminoglutaric acid$L-Glutamate$**L-4-Hydroxyglutamate semialdehyde (C13)**"
row968	E14	row949_M+H[1+];**row968_M(C13)+H[1+]**	C00217;C00025;C05938	D-Glutamate; D-Glutamic acid; D-Glutaminic acid; D-2-Aminoglutaric acid$L-Glutamate$**L-4-Hydroxyglutamate semialdehyde**
check: 149.1 m/z == 147.13 g/mol L-4-Hydroxyglutamate semialdehyde + 2 Da M(C13)+H[1+]
check: 169.3 rtmed == 171 seconds glutamic_13C_15N expected approx.
coef = UP





# Acknowlegments

- ref mummichig


# Session information

```{r}
sessionInfo()

```
