---
title: "Steatosis associated metabolites, lipids and proteins"
author: "Marilyn De Graeve, Johannes Rainer, Christoph Grander"
affiliation: "Eurac Research, Bolzano, Italy"
date: "2025-02-27"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
---

**Modified**: `r file.info("NAFLD_steatosis_associations.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r style, message = FALSE, echo = FALSE, warning = FALSE, results = "asis"}
library("BiocStyle")
library("kableExtra")
library("knitr")
library("rmarkdown")
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now, units = "secs")
      # return a character string to show the time
      paste("    Time for this code chunk to run:", round(res,
        2), "seconds")
    }
  }
}))
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7, time_it = TRUE)
set.seed(1)
```

# Introduction

During this workflow, exploratory analyis (Spearman correlations) are performed
on the relatively quantified Scanning SWATH proteomics data (148 proteins) from
the metabolic dysfunction-associated steatotic liver disease (MASLD)
cohort. MASLD, previously known as NAFLD is the most chronic liver disease,
affecting more than 30% of global population. The MAFLD cohort is a subset of
approximatley 353 # participants from the Cooperative Health Research in South
Tyrol (CHRIS) study, which were requested to return for a second resampling and
additional phenotyping. The NAFLD cohort represent an unhealty group with
various degrees of steatotic (non-alcoholic) liver disease classifications. See
[nafld_data_overview.Rmd](nafld_data_overview.Rmd) (respectively the rendered
html report) for an overview of the data set and the definition of the
phenotype.

This documents describes the association analysis between circulating
metabolites, lipids and proteins with liver steatosis.

See the `nafld_data_overview 1.html` report for further
information regarding the SLD sub-classification.

Regarding steatosis, i.e. Abnormal retention of fat (lipids) within a cell or
organ. In this scope: liver steatosis.
The variables containing this parameter is:

Presence of steatotic liver disease (SLD): sonographic evaluation, steatose
score of 2 – 3.

- *naab13*: liver fat. Can be binarized at an absolute value of 288. "liver_fat"
- *naab14*: subjective, visual evaluation of the liver fat. Categorical from 1
  to 3. Values 2 and 3 can be considered steatotic. "visual_fat"

This parameter will be used as the target for a biomarker search in *MASLD*
participants.


# Setup

## Directories

```{r directories, echo = FALSE}
#' General settings
filename <- "NAFLD_steatosis_associations"

#' Path to save images to; remove if exists.
IMAGE_PATH <- paste0("images/", filename, "/")
dir.create(IMAGE_PATH, recursive = TRUE)

#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)

#' Path to the data
#DATA_PATH <- '/home/mdegraeve/Documents/Files/Work_Eurac/Data/NAFLD' #FINDME!
DATA_PATH <- "."


DATASET <- 'CHRIS-NAFLD, none and MASLD'

```

## Packages

```{r packages, message=FALSE}
#' Load libraries
library(chrisUtils)
library(DT)
library(ggfortify)
library(ggplot2)
library(pander)
library(pheatmap)
library(plotly)
library(RColorBrewer)
library(readxl)
library(reshape2)
library(tidyfr)
library(vioplot)
library(writexl)

```


# Data import

## Circulating proteins (tprot) dataset (Scanning SWATH)

The data from proteins was converted into the TDF format and can hence be
loaded and imported using the `tidyfr` R package. Now we read the full data
which includes general information for each measurement.

```{r load_prot_dataset, message = FALSE}
#' load TDF using tidyfr
#list_data_modules(path = DATA_PATH)
md1 <- data_module(name = "chris_scanningswath_proteins", version = "1.0.0.3",
                   path = DATA_PATH)
prot_data <- data(md1)
prot_ann <- labels(md1)
stopifnot(all(colnames(prot_data) == rownames(prot_ann)))
colnames(prot_data) <- prot_ann$description
rownames(prot_ann) <- prot_ann$description

#' define the columns with protein concentrations
proteins <- rownames(prot_ann)[prot_ann$type != "categorical"]
proteins <- rownames(prot_ann)[prot_ann$Genes != ""]

#' select proteins dataframe
prot_data <- prot_data[, proteins]

```

The loaded data contains the samples as rows, and the protein names as
colums. The measurements of the `prot_data` object are the relative
concentrations of protein per sample. No data transformation is performed as the
abundances are already reported in log2 scale, a normal distribution of the
proteomics is assumed.

Before proceeding with the data analysis we remove the POOL samples. In addition
we restrict to samples from the NAFLD study. The number of samples and proteins
are:

```{r subset to NAFLD set, message = FALSE, warning = FALSE}
prot_data <- prot_data[grep("^008", rownames(prot_data)), ]
dim(prot_data)
```


## Targeted metabolomics (tmetab) dataset (Biocrates p180 kit)

The data from metabolites was converted into the TDF format and can hence be
loaded and imported using the `tidyfr` R package. Now we read the full data
which includes general information for each measurement.

```{r load_tmetb_dataset, message = FALSE}
#' load TDF using tidyfr
#list_data_modules(path = DATA_PATH)
p180 <- data_module(name = "metabolomics_p180", version = "1.0.1.3",
                    path = DATA_PATH)
met_data <- data(p180)
met_ann <- labels(p180)
stopifnot(all(colnames(met_data) == rownames(met_ann)))
colnames(met_data) <- met_ann$description
rownames(met_ann) <- met_ann$description

#' define the columns with metabolite concentrations
metabolites <- rownames(met_ann)[met_ann$type != "categorical"]

#' select metabolites dataframe
met_data <- met_data[, metabolites]
#' Remove the individuals with more than 80% missing values
rem <- apply(met_data, MARGIN = 1, function(z) {
    sum(is.na(z)) > length(z) * 0.8
})
met_data <- met_data[!rem, ]


#' Impute missing values using an uniform distribution, after gap filling
#na_unidis <- function(z) {
#    na <- is.na(z)
#    if (any(na)) {
#        min = min(z, na.rm = TRUE)
#        z[na] <- runif(sum(na), min = min/2, max = min)
#    }
#    z
#}

#' Row-wise impute missing values and add the data as a new assay
#tmp <- apply(met_data, MARGIN = 1, na_unidis)
#met_data <- t(tmp)
```

The loaded data contains the samples as rows, and the metabolite names as
colums. The measurements of the `met_data` object are the absolute
concentrations of metabolite per sample, in natural scale.

To ensure a normal distribution of the metabololits, we perform log2
transformation.

```{r m log2-transformation, message = FALSE}
#' log2 scaling of absolute abundances metabolites
met_data <- log2(met_data)
```

Before proceeding with the data analysis we remove the POOL samples. In addition
we restrict to samples from the NAFLD study. The number of samples and
metabolites are listed below.

```{r m subset to NAFLD set, message = FALSE, warning = FALSE}
met_data <- met_data[grep("^008", rownames(met_data)), ]
dim(met_data)
```


## Diet GA2LEN data

Dietary information, i.e. GA2LEN food frequency questionaires, from the MASLD
study were loaded and included in the analysis.

DO NOT RUN next 2 chuncks, we use the newly created dietary indices from Essi
(10/07/2025).

```{r diet-data, eval = FALSE}
#' 1. load the metadata regarding the GA2LEN FFQs
library(nafldData)
#nafldTraitAnnotations()|> head()

galan <- nafldData("FFQ")
rownames(galan) <- galan$AID

#variables, subset:
#naff001 - naff229 are galan FFQ, and the rest = not!

galan <- galan[, 1:230] #incl AID
ffq_items <- colnames(galan)[2:230]

#' change to numerical scores instead of FFQ answers
#' solve using category encoder, in R see:
#' https://www.r-bloggers.com/2020/02/a-guide-to-encoding-categorical-features-using-r/

#' select categor variables
categor_variables <- NULL
for (x in ffq_items){
  if (class(galan[[x]]) == 'factor' | class(galan[[x]]) == 'character'){  #factor and character
    categor_variables <- c(categor_variables, x)
  }
}

#' using ordinal encoding on categorical vars
encode_ordinal <- function(x) {
  if (is.factor(x)) {
    order <- levels(x)   #retain defined factor level order, wo NA but next step ignored so fine
    x <- as.numeric(factor(x, levels = order, exclude = NA)) #as numeric, keep NA in dataset
  } else if (is.character(x)) {
    print(x) #*
  }
  x
}

#* text info: NOT
    #naff246 - naff249b
    #naff260 - naff263a

#' testing
summary(galan$naff001)

#' apply encoding on categor vars
galan[, categor_variables] <- lapply(galan[, categor_variables], encode_ordinal)
```

Corresponding fake CV-values (no 'pool' samples) for diet. Too many AIDs, but
ok, since the filtering on included rownames is done in the lm.

```{r kora-cv, eval = FALSE}
#' 1. load the metadata regarding the GA2LEN FFQs
library(nafldData)
#nafldTraitAnnotations()|> head()

nafldAnn <- nafldTraitAnnotations()
kora_ann <- nafldAnn[ffq_items, ]
rownames(kora_ann) <- ffq_items
#head(kora_ann)

#' TEMP, just accept all features now as having good cv
cv_values <- 0
#' save in the met_ann as column
kora_ann$cv_qc_chris <- cv_values
```

The newly created dietary indices from Essi are saved under
NAFLD_dietary_indices_data_Marilyn. And note in the
lm: "you should also correct your models for total energy intake (just have it
as covariate in the model, nanu006a)". This was added in the section of dietary
lms.

```{r NAFLD-indices}
#' load the NAFLD_dietary_indices_data
handle <- file.path(DATA_PATH, 'GALAN', 'NAFLD_dietary_indices_data_Marilyn.txt')
galan <- read.table(handle, header = TRUE, sep = "\t")  |>
    as.data.frame()

#'galan is with , instead of decimal point so change
comma_to_point <- function(x) {
  x <- scan(text = x, dec = ",", sep = ".")
  x
}
galan <- apply(galan, MARGIN = 2, comma_to_point) |>
    as.data.frame()
rownames(galan) <- paste0('00', as.character(galan$AID))
#dim(galan)
#colnames(galan)
#galan$AID[1:5]
#head(galan)

#' replace the aids columns with 00 start and as a string
galan$AID <- rownames(galan)

#' variables
ffq_items <- colnames(galan)[2:ncol(galan)] #excl AID col

```

Also add the 4 macros and add to the `ffq_items`.

```{r NAFLD_food_macros}
library(nafldData)
#need these one
#nanu003 (protein), nanu004 (total fat), nanu005 (total carbs), nanu019b (saturated fatty acids), nanu023b (monounsaturated fatty acids), nanu025b (polyunsaturated fatty acids), nanu028b (trans fatty acids)

ffq <- nafldData("FFQ")
rownames(ffq) <- ffq$AID

galan$protein <- ffq[rownames(galan), 'nanu003']
galan$total_fat <- ffq[rownames(galan), 'nanu004']
galan$total_carbs <- ffq[rownames(galan), 'nanu005']
galan$saturated_fatty_acids <- ffq[rownames(galan), 'nanu019b']
galan$monounsaturated_fatty_acids <- ffq[rownames(galan), 'nanu023b']
galan$polyunsaturated_fatty_acids <- ffq[rownames(galan), 'nanu025b']
galan$trans_fatty_acids <- ffq[rownames(galan), 'nanu028b']

#' variables
ffq_items <- colnames(galan)[2:ncol(galan)] #excl AID col

```


## Metadata (excel)

All available trait (phenotype) data was defined in the
[nafld_data_overview.Rmd](nafld_data_overview.Rmd), is loaded below and merged
with the log2 transformed proteomics data of the NAFLD cohort.

```{r add phenotype dataa}
#' load xlsx file instead of below
handle <- 'masld_phenotypes.xlsx'
masld_phenotype_df <- read_excel(file.path(DATA_PATH, handle)) |>
    as.data.frame()
rownames(masld_phenotype_df) <- masld_phenotype_df$AID
#head(masld_phenotype_df, 5)
metadata <- colnames(masld_phenotype_df)[3:ncol(masld_phenotype_df)] #info wo AIDs
#dim(masld_phenotype_df)

#' make sure sld_type is a factor with group 'none' as the baseline
masld_phenotype_df$sld_type <- factor(masld_phenotype_df$sld_type,
                                      levels = c('none', 'MASLD',
                                                 'MetALD', 'SLD'))
print('the sld types are:')
levels(masld_phenotype_df$sld_type)
print('nr of samples per sld types:')
table(masld_phenotype_df$sld_type)
```


# Merge tmetab, tprot, kora and metadata

Merging all data tables. Note that restrict the data to participants for which
we have proteomics **and** metabolomics data. This also ensures that any
potential study drop-outs are no longer considered in the data (as their
proteomics or metabolomics data was removed).

There were less AIDs present in the dietary data, so we restrict with the data
to participants for which we have proteomics **and** metabolomics **and**
dietary data. ??? TODO findme!!!

```{r merge-dfs}
#AIDs <- intersect(rownames(prot_data), rownames(met_data))
AIDs1 <- intersect(rownames(prot_data), rownames(met_data))
AIDs2 <- intersect(rownames(masld_phenotype_df), rownames(galan))
AIDs <- intersect(AIDs1, AIDs2)

#' Subset all datasets to the same IDs
prot_data <- prot_data[AIDs, ]
met_data <- met_data[AIDs, ]
masld_phenotype_df <- masld_phenotype_df[AIDs, ]
galan <- galan[AIDs, ]

#' merge with proteomics data
prot_data$AID <- rownames(prot_data)
prot_data <- merge(prot_data, masld_phenotype_df, by = 'AID')

#' merge all sets
met_data <- as.data.frame(met_data)
met_data$AID <- rownames(met_data)
met_prot_data <- merge(prot_data, met_data, by = 'AID')

#' merge with metabolomics data only
met_data <- merge(met_data, masld_phenotype_df, by = 'AID')

#' merge all sets, kora too
met_prot_data <- as.data.frame(met_prot_data)
rownames(met_prot_data) <- met_prot_data$AID
gal_met_prot_data <- merge(met_prot_data, galan, by = 'AID') #kora has only 001 AIDs
rownames(gal_met_prot_data) <- gal_met_prot_data$AID
#gal_met_prot_data[1:5, 1:5]

#' write to file for manuel inspection
handle <- 'data/gal_met_prot_data.txt'
write.table(gal_met_prot_data, file = handle, sep = "\t",
            row.names = FALSE, col.names = TRUE)
```

All data sets are subset to `r length(AIDs)` study participants.


# Clean data and restrict to relevant participants

The data is further processed and cleaned to remove unreliable
measurements. Subsequently the data is restricted to relevant participants. We
thus next:

- remove unreliable liver fat measurements
- exclude participants with MetALD and SLD to ensure a more homogeneous data set
- remove participants with missing values in relevant variables
- define variables and phenotype(s) for the analysis

Add the information from the [nafld_data_overview.Rmd](nafld_data_overview.Rmd)
report regarding the steatosis, using the parameter naab11
(higher values are worse). Values for participants that have a naab12
(liver stiffness IQR) > 30 should be either removed or taken with care. Below we
remove these unreliable measurements.

```{r load-liver-param2}
#' remove samples with  naab12 (liver stiffness IQR) > 30
gal_met_prot_data$liver_stiffness[
                      which(gal_met_prot_data$liver_stiffnes_iqr > 30)] <- NA
```

We next remove the MetALD and SLD participants.

```{r keep only none and MASLD types}
#' because problem 2 proteins only NAs in some sld types,
#' keep only none and MASLD
gal_met_prot_data <- gal_met_prot_data[gal_met_prot_data$sld_type %in%
                                       c('none', 'MASLD'), ] #
gal_met_prot_data$sld_type <- droplevels(gal_met_prot_data$sld_type)
gal_met_prot_data$sld_type <- factor(gal_met_prot_data$sld_type)
```

We next remove participants with missing values for relevant variables:
- liver fat measurement

```{r}
#' Define variables for which we don't allow missing values
no_na_var <- c("liver_fat")
has_na <- lapply(gal_met_prot_data[, no_na_var, drop = FALSE], is.na) |>
    do.call(what = cbind)
#' Remove rows (participants) with a missing value in any of the above
#' variables
gal_met_prot_data <- gal_met_prot_data[rowSums(has_na) == 0, , drop = FALSE]
```

Remove fibrosis outlier values: **CURRENTLY NOT DONE**.
Previous outlier removal removed a participant with a FIB4 score of 7.8. This
participant was originally removed. The second outlier removal excluded
a participant with a liver_stiffness value of 34. Both participants are still
present, as fibrosis is not the phenotype of interest anymore.

We next define the categorical variables from numeric measurements:

Add a categoric liver_stiffness category too, using both the definition of
Christoph (7, 15 cutoffs) and in literature (7, 12.5 kPa cutoffs).

```{r liver-stiffness-category}
#' naab11 according to Christoph
gal_met_prot_data$liver_stiffness_cat1 <- 2
gal_met_prot_data$liver_stiffness_cat1[
                      is.na(gal_met_prot_data$liver_stiffness)] <- NA
gal_met_prot_data$liver_stiffness_cat1[
                      which(gal_met_prot_data$liver_stiffness < 7)] <- 1
gal_met_prot_data$liver_stiffness_cat1[
                      which(gal_met_prot_data$liver_stiffness > 15)] <- 3
gal_met_prot_data$liver_stiffness_cat1 <-
    factor(gal_met_prot_data$liver_stiffness_cat1)

#' naab11 according to literature
gal_met_prot_data$liver_stiffness_cat2 <- 2
gal_met_prot_data$liver_stiffness_cat2[
                      is.na(gal_met_prot_data$liver_stiffness)] <- NA
gal_met_prot_data$liver_stiffness_cat2[
                      which(gal_met_prot_data$liver_stiffness < 7)] <- 1     #mild or absent fibrosis
gal_met_prot_data$liver_stiffness_cat2[
                      which(gal_met_prot_data$liver_stiffness > 12.5)] <- 3  #cirrhosis
gal_met_prot_data$liver_stiffness_cat2 <-
    factor(gal_met_prot_data$liver_stiffness_cat2)
#' appearantly differnet cutoffs, depending what background...
#' https://www.mskcc.org/cancer-care/patient-education/understanding-your-fibroscan-results
```

Add a categoric liver_fat (i.e. steatosis) category, using the following
cutoffs:

- "normal": A score below 238 dB/m means the amount of fatty change in your
  liver is not higher than normal.
- grade steatosis S1: 238 to 260 dB/m
- grade steatosis S2: 260 to 290 dB/m
- grade steatosis S3: 290 to 400 dB/m

Depending on the distribution, look only at the extreme normal - steatosis.

```{r liver-fat-category}
#' naab13 make 4 groups
gal_met_prot_data$liver_fat_cat1 <- NA
gal_met_prot_data$liver_fat_cat1[which(gal_met_prot_data$liver_fat < 238)] <- 1 #normal
gal_met_prot_data$liver_fat_cat1[which(gal_met_prot_data$liver_fat >= 238)] <- 2
gal_met_prot_data$liver_fat_cat1[which(gal_met_prot_data$liver_fat > 260)] <- 3
gal_met_prot_data$liver_fat_cat1[which(gal_met_prot_data$liver_fat > 290)] <- 4
gal_met_prot_data$liver_fat_cat1 <- factor(gal_met_prot_data$liver_fat_cat1)

#' naab13 make 2 groups
gal_met_prot_data$liver_fat_cat2 <- NA
gal_met_prot_data$liver_fat_cat2[which(gal_met_prot_data$liver_fat < 238)] <- 1 #normal
gal_met_prot_data$liver_fat_cat2[which(gal_met_prot_data$liver_fat > 290)] <- 2  #steatosis, high S3

#' appearantly differnet cutoffs, depending what background...
#' https://www.mskcc.org/cancer-care/patient-education/understanding-your-fibroscan-results
```

At last we ensure categorical variables are encoded as a factor. Also, we create
additional variable for age per 10y, because coefficients and effect sizes
correspond to one year difference, and the anual difference is
neglectible. Dividing age with 10, difference in coefficents and effect sizes
would correspond to 10 years difference and it would be easier to define
meaningful cut-off. Finally, we add a visc_subcut_fat variable on the dataset,
comprising the ratio of visceral to subcutaneous fat thickness.


```{r define-age10-variable, message = FALSE}
#' Convert categorical variables to factors
col_cat <- c("case", "self_DM", "med_DM", "fasting", "no_drinking",
             "med_BP", "med_lip", "med_MS", "T2D", "Sex", "MASLD")
for (i in col_cat)
    gal_met_prot_data[, i] <- as.factor(gal_met_prot_data[, i])

#' define age per 10y group
gal_met_prot_data$Age_10 <- gal_met_prot_data$Age / 10
gal_met_prot_data$visc_subcut_fat <- gal_met_prot_data$visc_fat /
    gal_met_prot_data$body_fat
```


## Summary of the final data set

```{r}
summary(gal_met_prot_data[, c("Age", "Sex", "T2D", "MASLD", "FIB4", "BMI",
                              "liver_stiffness", "liver_fat", "liver_fat_cat1",
                              "visc_subcut_fat")])
```

The variables that are not allowed to conatin NAs are: sex, age and liver_fat.
Participants with NA in BMI, and those 2 with extreme liver stiffness/FIB4
scores are not removed.


# Output for the paper

## Violin plot BMI

Relationship between liver fat and BMI. The BMI distribution for the different
liver fat categories is shown below. There is a clear relationship between the
BMI and the liver fat, which is increasing both for female and male
participants.

```{r violin-bmi-vs-sex-target, fig.path = IMAGE_PATH, fig.height=8, fig.width=7, dpi = 300}
vioplot(split(gal_met_prot_data$BMI,
              paste0(gal_met_prot_data$Sex, gal_met_prot_data$liver_fat_cat1)),
        ylab = "BMI",
        col = c("#e1b0af", "#e1b0af", "#e1b0af", "#e1b0af",
                "#aabed0", "#aabed0", "#aabed0", "#aabed0"))
grid()
```

We also directly correlate BMI with the liver fat measurements.

```{r}
col_sex <- rep("#ff0000", nrow(gal_met_prot_data))
col_sex[gal_met_prot_data$Sex == "Male"] <- "#0000ff"
plot(y = gal_met_prot_data$BMI, x = gal_met_prot_data$liver_fat,
     col = paste0(col_sex, 80), pch = 21, bg = paste0(col_sex, 40))
grid()
l <- lm(BMI ~ liver_fat, data = gal_met_prot_data)
abline(l)
```

The regression of BMI on liver fat values is highly significant:

```{r}
summary(l)
```

## Violin plot Age

Not for in the paper, but for the sentence regarding if age increases with grade
3 steatosis

```{r violin-age-vs-sex-target, fig.path = IMAGE_PATH, fig.height=8, fig.width=7, dpi = 300}
vioplot(split(gal_met_prot_data$Age,
              paste0(gal_met_prot_data$Sex, gal_met_prot_data$liver_fat_cat1)),
        ylab = "Age")
grid()
```


## table all proteins

```{r, results = "asis"}
#148 proteins
length(proteins)

#The Divisors of 148 are as follows:
#1, 2, 4, 37, 74, and 148.
length(proteins) / 3
length(proteins) / 3 + length(proteins) / 3
length(proteins) / 3 + length(proteins) / 3 + length(proteins) / 3
length(proteins) / 3 + length(proteins) / 3 + length(proteins) / 3 + length(proteins) / 3

#' make table
df <- NULL
df$col1 <- proteins[1:50]
df$col2 <- proteins[51:100]
df$col3 <- proteins[101:150]
df <- as.data.frame(df)

#' print table
library(pander)
pandoc.table(
  df, style = "rmarkdown",
  caption = "Relatively quantified proteins.")

#' export table
handle <- file.path('paper', 'SI_table1.txt')
write.table(df, file = handle, sep = "\t", row.names = FALSE, col.names = TRUE)

```


## table all metabolites and lipids

```{r, results = "asis"}
#175 metabolites
length(metabolites)

#The Divisors of 175 are as follows:
#1, 5, 7, 25, 35, and 175.
#So not by 4,
#=> /4 and with empty rows in last col.
length(metabolites) / 4
length(metabolites) / 4 + length(metabolites) / 4
length(metabolites) / 4 + length(metabolites) / 4 + length(metabolites) / 4
length(metabolites) / 4 + length(metabolites) / 4 + length(metabolites) / 4 + length(metabolites) / 4

#' make table
df <- NULL
df$col1 <- metabolites[1:44]
df$col2 <- metabolites[45:88]
df$col3 <- metabolites[89:132]
df$col4 <- metabolites[132:175]
df <- as.data.frame(df)

#' print table
library(pander)
pandoc.table(
  df, style = "rmarkdown",
  caption = "Absolutely quantified metabolites and lipids.")

#' export table
handle <- file.path('paper', 'SI_table2.txt')
write.table(df, file = handle, sep = "\t", row.names = FALSE, col.names = TRUE)

```


# Correlation naab13 liver_fat and sld_type

```{r}
plot(gal_met_prot_data$sld_type, gal_met_prot_data$liver_stiffness)
```


# Correlation naab13 vs naab14

scatterplot with naab13 = liver_fat vs naab14 = visual_fat: see previous report.
Below, boxplot with the 2 categorical variables naab13_cat1/2 and naab14.
data was not transformed, it was subsetted as above and described in the report.

```{r box-naab13_cat vs naab14}
boxplot(gal_met_prot_data$visual_fat ~ gal_met_prot_data$liver_fat_cat1)
boxplot(gal_met_prot_data$visual_fat ~ gal_met_prot_data$liver_fat_cat2)

```


# Correlations (all)

No transformations was done beforehand. proteomics are relative abundances.
metadata with numeric data are left as such.
metadata with categoric data is converted to integers eg sex 0/1.
Before correlation analyis, we do standard scaling on all, iter over columns
(variables).

```{r transform-categories}
#rownames(gal_met_prot_data) <- gal_met_prot_data$AID #nok, duplic nownames present...
#gal_met_prot_data[1:10, 149:ncol(gal_met_prot_data)]

#' bools to bin
gal_met_prot_data$MetS <- as.numeric(gal_met_prot_data$MetS)
gal_met_prot_data$MetALD <- as.numeric(gal_met_prot_data$MetALD)
gal_met_prot_data$MASLD <- as.numeric(gal_met_prot_data$MASLD)
gal_met_prot_data$SLD <- as.numeric(gal_met_prot_data$SLD)
summary(gal_met_prot_data$sld_type) #gives NA in corr since all values are same for

#' remove AID and AID_chris column; rename df
variables <- colnames(gal_met_prot_data)
#met_data <- gal_met_prot_data
met_data <- gal_met_prot_data[, -which(colnames(gal_met_prot_data) == 'AID')] #drop 2x aid, and aid_chris
met_data <- met_data[, -which(colnames(met_data) == 'AID_chris')] #drop 2x aid, and aid_chris
dim(met_data)

#' optional: only with the clinical metadata
met_data <- met_data[, metadata]

#' select categor variables
categor_variables <- NULL
for (x in variables){
  if (class(met_data[[x]]) == 'factor' | class(met_data[[x]]) == 'character'){  #factor and character
    categor_variables <- c(categor_variables, x)
  }
}
categor_variables #ok

#' using ordinal encoding on categorical vars
encode_ordinal <- function(x) {
  if (is.factor(x)) {
    order <- levels(x)   #retain defined factor level order, wo NA but next step ignored so fine
    x <- as.numeric(factor(x, levels = order, exclude = NA)) #as numeric, keep NA in dataset
  } else if (is.character(x)) {
    #print(x)
    tmp <- x[order(x)]
    order <- unique(tmp) #don't change order, only use to chronol/alphab rank
    x <- as.numeric(factor(x, levels = order, exclude = NA)) #ignore NA, stays NA in dataset
  }
  x
}

#' apply encoding on categor vars
met_data[, categor_variables] <- lapply(met_data[, categor_variables], encode_ordinal)
#met_data[1:10, 150:181]
dim(met_data)


#' evaluate distribution, transform if needed
test <- unlist(c(met_data))
hist(test, col = rgb(1,0,0,1/4), main = paste0("Histogram variables of ", DATASET),
     breaks = 64)

#' scaling before correlation analysis
scaledData <- scale(met_data, scale = TRUE)

#' transform the numeric values w z-scores
#' #FYI, same as z-scores (center = TRUE, scale = TRUE)
test <- scale(scaledData, scale = TRUE)
hist(test, col = rgb(0,0,1,1/4), main = "Histogram after autoscale - test",
     breaks = 64) #OK, nice distrib!

#head(scaledData)

```

correlations.
FYI, for the SLD, MASLD, MetALD and sld_type column, the correltation
coefficients will become NA. Because all options are same eg sld_type == MASLD.
Only for MetS, both true/false present.

```{r correlation-samples}
#' spearman correlation
#cor(met_data$Age, met_data$FIB4, use = "complete.obs", method = "spearman")
correlation_scaledmatrix <- round(cor(met_data, use = "complete.obs", method = "spearman"), 3)
suppressMessages(library(psych))
Pvalues_correlation_scaledmatrix <- corr.test(met_data, use = "complete", method = "spearman", adjust = "fdr", alpha = 0.05)
Pvalues_correlation_scaledmatrix <- round(Pvalues_correlation_scaledmatrix$p, 3)
detach(package:psych)


#write all info as list
correlation_df <- as.data.frame(correlation_scaledmatrix)
correlation_df$CompID <- rownames(correlation_df) #depending on which name (system or nic name), as chosen above
data_melt <- melt(correlation_df, id = c("CompID"))

Pvalues_correlation_df <- as.data.frame(Pvalues_correlation_scaledmatrix)
Pvalues_correlation_df$CompID <- rownames(Pvalues_correlation_df) #depending on which name (system or nic name), as chosen above
data_melt_Pvalues <- melt(Pvalues_correlation_df, id = c("CompID"))

corr_info <- cbind(data_melt, data_melt_Pvalues$value)
colnames(corr_info) <- c("CompID1", "CompID2", "Corr_coef", "adj_p_value")
head(corr_info)

write.table(corr_info, 'data/corr_info.txt', sep = '\t')

```

print the hierarchical dendogram / order of corrs.

```{r hier-clustering}
#1feat all same value => NAN after corr so add rn nan here as well
correlation_scaledmatrix[is.nan(correlation_scaledmatrix)] <- 0
correlation_scaledmatrix[is.na(correlation_scaledmatrix)] <- 0

#clustering samples order
ord <- hclust(dist(correlation_scaledmatrix))$order #info about order in object ord$order
#default clustering: Cluster method: complete; Distance: euclidean see doc: https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/hclust
#ord
dendrogram <- as.dendrogram(hclust(dist(correlation_scaledmatrix)))
png(file.path("images", "dendogram_clust.png"), width = 17, height = 10, units = "in", res = 300)
par(cex = 0.50)
plot(dendrogram)
dev.off()

plot(dendrogram)

```

keep only for signif p-adj values and for
correlations with absolute rho-values >= 0.2.

```{r top-rho}
save <- corr_info

#' keep signif and corr higher then 0.2
nrow(corr_info)
corr_info <- corr_info[corr_info$adj_p_value < 0.05, ]
nrow(corr_info) #6202
corr_info <- corr_info[abs(corr_info$Corr_coef) >= 0.2, ]
nrow(corr_info) #5263

#' discard association with self
corr_info <- corr_info[corr_info$CompID1 != corr_info$CompID2, ]
nrow(corr_info) #5080

#sort based on order run samples
corr_info$ABSrho <- abs(as.numeric(corr_info$Corr_coef))
corr_info <- corr_info[order(corr_info$ABSrho, decreasing = TRUE), ]

```

venn diagram between naab13 and naab14.

```{r overlapping-rho-naab13-naab14, eval = FALSE}
#liver_fat naab13
tmp <- corr_info[corr_info$CompID1 == 'liver_fat_cat2', ]
liver_fat_associations <- tmp$CompID2
liver_fat_associations <- liver_fat_associations[!is.na(liver_fat_associations)]
print('associated with cat naab13:')
liver_fat_associations

#naab14 visual fat
tmp <- corr_info[corr_info$CompID1 == 'visual_fat', ]
visual_fat_associations <- tmp$CompID2
visual_fat_associations <- visual_fat_associations[!is.na(visual_fat_associations)]
print('associated with cat naab14:')
visual_fat_associations

#' overlapping
print('overlapping between naab13 and naab14:')
intersect(visual_fat_associations, liver_fat_associations)

x <- list(visual_fat_associations, liver_fat_associations)

#Venn diagram
library(gplots)
venn(x)

```

venn diagram between naab13 and FIB4.

```{r overlapping-rho-naab13-fib4, eval = FALSE}
#liver_fat naab13
tmp <- corr_info[corr_info$CompID1 == 'liver_fat_cat2', ]
liver_fat_associations <- tmp$CompID2
liver_fat_associations <- liver_fat_associations[!is.na(liver_fat_associations)]
print('associated with cat naab13:')
liver_fat_associations

#FIB4
tmp <- corr_info[corr_info$CompID1 == 'FIB4', ]
FIB4_associations <- tmp$CompID2
FIB4_associations <- FIB4_associations[!is.na(FIB4_associations)]
print('associated with FIB4:')
FIB4_associations

#' overlapping
print('overlapping between naab13 and FIB4:')
intersect(FIB4_associations, liver_fat_associations)

x <- list(FIB4_associations, liver_fat_associations)

#Venn diagram
library(gplots)
venn(x)

```

# Association analyses

We next perform the association analyses between food items, metabolites and
proteins with the liver fat categories.

## Association liver fat (steatosis) with diet

We perform linear regression analyses to evaluate whether there are
differences in dietary habits between the liver fat categories.

Feedback from Essi: load food indices and summarized food groups (eg pizza),
and include the "estimated daily nutrient intake" as additional covariate.

```{r}
library(nafldData)
#need this one
#nafldAnn['nanu006a', ]   #Estimated daily nutrient intake

ffq <- nafldData("FFQ")
rownames(ffq) <- ffq$AID

gal_met_prot_data$estim_daily_nutr_intake <- ffq[rownames(gal_met_prot_data), 'nanu006a']

```

Next feedback: remove food groups, keep the macronutrients only.
Afterwards, keep only PDI and run again, no adj p-value.
Also, for the heatmap plot, can run macronutrients + PDI.

```{r}
#' selection of diet items of interest
print(ffq_items)
ffq_items <- c("protein", "total_fat",
               "total_carbs", "saturated_fatty_acids",
               "monounsaturated_fatty_acids", "polyunsaturated_fatty_acids",
               "trans_fatty_acids")

#' selection for the run with 1 indices, ran manual
#ffq_items <- "pdi_score"

#' selection with macronutrients and PDI for the heat map in paper
#ffq_items <- c("protein", "total_fat",
#               "total_carbs", "saturated_fatty_acids",
#               "monounsaturated_fatty_acids", "polyunsaturated_fatty_acids",
#               "trans_fatty_acids", "pdi_score")

#' print final selection of diet items
print(ffq_items)

```

linear models setup:

```{r define-lm-covariates-and-target-diet}
print('covariates of the linear regression model:')
COVARIATES <- c("Sex", "Age_10", "estim_daily_nutr_intake")
print(COVARIATES)

print('target of the linear regression model:')
TARGET <- 'liver_fat_cat1'
print("liver_fat_cat1")

DATASET <- 'NAFLD'
print(paste0('final variables of the ', DATASET, ' dataset used:'))

print('nr of variables inspected:')
print(length(ffq_items))

print('nr of individuals inspected:')
print(nrow(gal_met_prot_data))

```

To identify ffq_items related with the 2 (none, MASLD) of the 4 SLD types (none,
MASLD, MetALD, SLD), seperate linear regression models are fitted for all
ffq_items separately (as a response) with sex, age and the steatosis categories
as explanatory variables.

Model coefficients and p-values are provided. Raw p-values are adjusted for
multiple hypothesis testing with the method from Benjamini & Hochberg.

```{r linear model-diet with CIs, message=FALSE}

#' Define a function to fit the linear model:
#' - fit the linear model
#' - extract the coefficients and p-values skipping the intercept
#' - return the result as a one-row matrix, columns being coefficients
#'   and p-values
funForLinReg <- function(analyte, data) {
    y <- data[, analyte]
    data <- data[, c(COVARIATES, TARGET)]
    res <- coef(summary(lm(y ~ ., data = data)))[-1, ]
    pvals <- matrix(c(res[, 1L], res[, 4L]), nrow = 1)
    rownames(pvals) <- analyte
    colnames(pvals) <- c(paste0("coef_", rownames(res)),
                         paste0("p-value_", rownames(res)))
    pvals
}

#' tests
#gal_met_prot_data[1:5, c("Sex", "Age", "FIB4")]
#gal_met_prot_data[1:5, ffq_items[1]]
#data <- gal_met_prot_data
#analyte <- ffq_items[1]
#funForLinReg(analyte, gal_met_prot_data)
#analyte <- ffq_items[14]
#funForLinReg(analyte, gal_met_prot_data)
#for (analyte in ffq_items) {
#  print(analyte)
#  funForLinReg(analyte, gal_met_prot_data)
#}

#' Fit linear models and extract coefficients and p-values on each protein
lm_output <- lapply(ffq_items, FUN = funForLinReg, data = gal_met_prot_data) |>
    do.call(what = rbind)

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)

```

Add confidence intervals reorting in the LM for the not-significant dietary
intake variables. keep R chunk above for the next lms (proteins, metabolites).

```{r lm-with-CIs-diet}

#' Define a function to fit the linear model:
#' - fit the linear model
#' - extract the coefficients and p-values skipping the intercept
#' - extract the 95% confidence interval for coefficients in lm
#' - return the result as a one-row matrix, columns being coefficients
#'   and p-values
funForLinReg_CIs <- function(analyte, data) {
    y <- data[, analyte]
    data <- data[, c(COVARIATES, TARGET)]
    res <- coef(summary(lm(y ~ ., data = data)))[-1, ]
    CIs <- confint(lm(y ~ ., data = data))[-1, ]
    pvals <- matrix(c(res[, 1L], res[, 4L], CIs[, 1L], CIs[, 2L]), nrow = 1)
    rownames(pvals) <- analyte
    colnames(pvals) <- c(paste0("coef_", rownames(res)),
                         paste0("p-value_", rownames(res)),
                         paste0("CI_lower_", rownames(res)),
                         paste0("CI_upper_", rownames(res))
                         )
    pvals
}

#data <- gal_met_prot_data
#analyte <- ffq_items[1]
#funForLinReg_CIs(analyte, gal_met_prot_data)

#' Fit linear models and extract coefficients and p-values on each protein
lm_output <- lapply(ffq_items, FUN = funForLinReg_CIs, data = gal_met_prot_data) |>
    do.call(what = rbind)

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)

```

Next we calculate the effect sizes.

To obtain comparable coefficients, prior to linear modelling all the ffq_items
are standardized to zero mean and unit variance (note that this does not
influence the linear models since they are affine equivariant). In such
autoscaled data, difference of 1 corresponds to a standard deviation of 1. This
parameter will be called "effect size" and will be use to evaluate group
differences.

```{r effectsize-diet, message=FALSE}
#' Perform autoscaling for effect size (es) calculation
gal_met_prot_data_scaled <- gal_met_prot_data
gal_met_prot_data_scaled[, ffq_items] <- gal_met_prot_data[, ffq_items] |>
    as.matrix() |>
    scale(scale = TRUE)

#' fit linear models and extract the coefficients
es <- lapply(ffq_items, FUN = funForLinReg, data = gal_met_prot_data_scaled) |>
    do.call(what = rbind)
es <- es[, grep("coef", colnames(es))]
colnames(es) <- sub("coef", "effectSize", colnames(es))
lm_output <- cbind(lm_output, es)
```

For diet, we can not have cv-based cut-off, therefore, we use "just" the
statistical significance criteria (just the adjusted p-value).

```{r define significance criteria-diet}
#' define the different coefficients as comparisons
comps <- colnames(lm_output)[grep("coef", colnames(lm_output))]

#' define which ffq_items are considered statistically significant
sign_p <- lm_output[, grep("p-adj", colnames(lm_output))] < 0.05
colnames(sign_p) <- sub("p-adj_", "significant_", colnames(sign_p))
#colSums(sign_p)

#' Combine with result from linear model converting to a data.frame to
#' avoid logicals being converted to numeric
lm_output <- data.frame(lm_output, sign_p)
```

The number of ffq_items called significant or with an adjusted p-value smaller
than 0.05 are shown below.

```{r table-sig-diet, results = "asis"}
#' merge to p-adj to cv significance summary
sig_tab <- data.frame(
    significant = colSums(lm_output[, sub("coef_", "significant_", comps)]),
    check.names = FALSE)

pandoc.table(
    sig_tab, style = "rmarkdown",
    caption = paste0("Number of significant ffq_items per tested covariate. ",
                     " *significant*: number of ffq_items with an adjusted ",
                     "p-value smaller than 0.05"))
```

For ease of evaluation, add a column describing if the CI is not in zero.

```{r CI_eval}
#' nice output for paper, CIs
lm_output$CI_range_fat_cat14 <- paste0(
    "[",
    format(round(lm_output$CI_lower_liver_fat_cat14, 4), 4),
    ", ",
    format(round(lm_output$CI_upper_liver_fat_cat14, 4), 4),
    "]"
)

#we want, for each variable, to see if the CI is not zero (for singif difference variables).
lm_output$CI_No_0_liver_fat_cat14 <-((lm_output$CI_lower_liver_fat_cat14 > 0 & lm_output$CI_upper_liver_fat_cat14 > 0) | (lm_output$CI_lower_liver_fat_cat14 < 0 & lm_output$CI_upper_liver_fat_cat14 < 0)) #returns true if CI NOT in 0, wanted to find 'sign' different vars!

#' nice explain:
#' https://www.econometrics-with-r.org/5.2-cifrc.html
#' if the interval [CI_lower, CI_upper] does not contain the value zero, this leads to the rejection of the null hypothesis β1,0=0.
#' A Confidence Interval for βi
#' Imagine you could draw all possible random samples of given size. The interval that contains the true value βi in 95%
#' of all samples is given by the expression
#' CIβi0.95=[^βi−1.96×SE(^βi),^βi+1.96×SE(^βi)].
#' Equivalently, this interval can be seen as the set of null hypotheses for which a 5% two-sided hypothesis test does not reject.

```

Export results for inspection:

```{r export-results-diet}
results_diet <- data.frame(
    #kora_ann[rownames(lm_output), c("domain", "name")],  #with indices essi, info NA
    lm_output
)
#' write to file for manual inspection,
#' inclusing inspection of the CV-value of each protein
handle <- file.path(RDATA_PATH, 'lm_output_results_ffq_items-macronutrients.txt')
write.table(results_diet, file = handle, sep = "\t", row.names = FALSE,
            col.names = TRUE)
```

```{r, eval = TRUE}
rem <- grepl("effect|significant|p.value|CI", colnames(results_diet))
tmp <- results_diet[, !rem]
datatable(tmp) |>
    formatRound(columns = which(vapply(tmp, is.numeric, logical(1))),
                digits = 3)
```

We visualize the associations in a heatmap. The effect size are added for the
associations, for the significant ffq_items (min 1 significance to
covariate). also for the not significant associations, the effect size is shown.

```{r heatmap-diet, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for ffq_items with at least one significant association."}
#' Extract effect sizes
es_results <- results_diet[, grep("coef_", colnames(results_diet))] |>
    as.matrix()

#' Restrict to those significant for at least one of the groups (sex, age, SLD type)
sig_results <- results_diet[, grep("significant_", colnames(results_diet))] |>
    as.matrix()
es_results <- es_results[rowSums(sig_results) > 0, ]
print('nr of ffq_items with >= 1 significant association:')
nrow(es_results)

#' hierarchical clustering
clust1 <- hclust(dist((es_results)), method = "ward.D")
clust2 <- hclust(dist(t(es_results)), method = "ward.D")

#' clean roundup nrs for plot
es_results[es_results == 0] <- NA
es_results <- round(es_results, 2)

#' info regarding FIB4
#print('nr of ffq_items with a not-zero effect size in the FIB4 group:')
#sum(es_results[, 'effectSize_FIB4'] != 0)

#' in long format
es_results_long <- melt(as.matrix(es_results))

#' ggplot
plt <- es_results_long |>
    ggplot(aes(Var2, Var1, fill = value, label = value)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Effect Size") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 2.5) +
    theme_classic() +
    scale_x_discrete(
        expand = c(0, 0), limits = colnames(es_results)[clust2$order]) +
    scale_y_discrete(
        expand = c(0, 0), limits = rownames(es_results)[clust1$order]) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 10),
          axis.text.y = element_text(hjust = 1, size = 8),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.position = "none")

ggsave(paste0(IMAGE_PATH, "protein_SLDtypes_effect_sizes.png"),
       plot = plt, dpi = 600, width = 12,
       height = 12, units = "cm", scale = 2.5)
plt

```

In second plot, only the significant effect size labels were added to the
heat map. also, the dendograms after hierarchical clustering are shown.

```{r heatmap2-diet, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for significant associations the value of the effect size is shown."}
#' keep just the effect sizes of the significant associations
es_results_sig <- es_results
for (i in colnames(es_results_sig)) {
    es_results_sig[!results_diet[rownames(es_results_sig),
                                 sub("coef", "significant", i)], i] <- ""
}
#' plot
colnames(es_results_sig) <- sub("coef_", "", colnames(es_results_sig))
pheatmap(es_results, display_numbers = es_results_sig,
         number_color = "#444444", fontsize_number = 7)

```

In the third heat map, we plot the coefficients of molecules with >= 1
significant association in the comparisons of naab13_cat1:

This is also Figure S2 (edit in ppt to merge macro + PDI into 1 figure).

```{r diet-heatmap3, fig.path = IMAGE_PATH, res = 150, fig.cap = "Associations between covariates. Shown are coeficients"}

#' Extract coefficients
coef_results_diet <- results_diet[, grep("coef_", colnames(results_diet))] |>
    as.matrix()

#' drop sex, age
coef_results_diet <- coef_results_diet[, c("coef_liver_fat_cat12", "coef_liver_fat_cat13", "coef_liver_fat_cat14")] |>
    as.matrix()

#' DO NOT Restrict to those significant for at least one of the groups (sex, age, SLD type)
sig_results_diet <- results_diet[, c("significant_liver_fat_cat12", "significant_liver_fat_cat13", "significant_liver_fat_cat14")] |>
    as.matrix()
#coef_results_diet <- coef_results_diet[rowSums(sig_results_diet) > 0, ] #keep all
print('nr of proteins with >= 1 significant association:')
nrow(coef_results_diet)

#' hierarchical clustering
clust1 <- hclust(dist((coef_results_diet)), method = "ward.D")
clust2 <- hclust(dist(t(coef_results_diet)), method = "ward.D") #0 signif

#' clean roundup nrs for plot
coef_results_diet[coef_results_diet == 0] <- NA
coef_results_diet <- round(coef_results_diet, 2)

#' in long format
coef_results_diet_long <- melt(as.matrix(coef_results_diet))

#' ggplot
plt <- coef_results_diet_long |>
    ggplot(aes(Var2, Var1, fill = value, label = value)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Coefficient") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 2.5) +
    theme_classic() +
    #scale_x_discrete(
        #expand = c(0, 0), limits = colnames(coef_results_diet)[clust2$order]) +
    #scale_y_discrete(
        #expand = c(0, 0), limits = rownames(coef_results_diet)[clust1$order]) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 10),
          axis.text.y = element_text(hjust = 1, size = 8),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.position = "none")

#ggsave(paste0(IMAGE_PATH, "diet_naab13cat1_HMcoef.png"),
#       plot = plt, dpi = 600, width = 12,
#       height = 12, units = "cm", scale = 2.5)
#plt  #0 singificant


#' keep just the Coefficients of the significant associations
#coef_results_diet_sig <- coef_results_diet
#coef_results_diet_sig[as.matrix(sig_results_diet)[rownames(coef_results_diet), ] == 0] <- ""

#' plot
#colnames(coef_results_diet) <- sub("Coefficient_", "", colnames(coef_results_diet))
#pheatmap(coef_results_diet, display_numbers = es_results_diet_sig,
#         number_color = "#444444", fontsize_number = 7,
#         cluster_cols = FALSE)

```

A considerable number of ffq_items are significant both for sex and age,
in many cases even with the same sign.


### Summary

- Strong age and sex differences.
- No food item is significantly different between the steatosis categories.
- There are food items with relatively large coefficient, but not significant.


## Association between circulating proteins and liver fat (steatosis)

Model setup:

```{r define-lm--covariates-and-target}
print('covariates of the linear regression model:')
COVARIATES <- c("Sex", "Age_10")
print(COVARIATES)

print('target of the linear regression model:')
TARGET <- 'liver_fat_cat1'
print(TARGET)

DATASET <- 'NAFLD'
print(paste0('final variables of the ', DATASET, ' dataset used:'))
#' variables are the output y in lm(y ~ .):
#print(proteins)

print('nr of variables inspected:')
print(length(proteins))

print('nr of individuals inspected:')
print(nrow(gal_met_prot_data))

```

To identify proteins associated with liver steatosis, seperate linear regression
models are fitted for all proteins separately using their (log2) abundance as
response variable with sex, age and the (categorical) steatosis as a covariates.

Model coefficients and p-values are provided. Raw p-values are adjusted for
multiple hypothesis testing with the method from Benjamini & Hochberg.

```{r linear-model-prot}

#' Fit linear models and extract coefficients and p-values on each protein
lm_output <- lapply(proteins, FUN = funForLinReg, data = gal_met_prot_data) |>
    do.call(what = rbind)

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)
```

Next we calculate the effect sizes.

To obtain comparable coefficients, prior to linear modelling all the
proteins
are standardized to zero mean and unit variance (note that this does not
influence the linear models since they are affine equivariant). In such
autoscaled data, difference of 1 corresponds to a standard deviation of 1. This
parameter will be called "effect size" and will be use to evaluate group
differences.

```{r lm--effectsize, message=FALSE}
#' Perform autoscaling for effect size (es) calculation
gal_met_prot_data_scaled <- gal_met_prot_data
gal_met_prot_data_scaled[, proteins] <- gal_met_prot_data[, proteins] |>
    as.matrix() |>
    scale(scale = TRUE)

#' fit linear models and extract the coefficients
es <- lapply(proteins, FUN = funForLinReg, data = gal_met_prot_data_scaled) |>
    do.call(what = rbind)
es <- es[, grep("coef", colnames(es))]
colnames(es) <- sub("coef", "effectSize", colnames(es))
lm_output <- cbind(lm_output, es)
```

To call a proteins *significant* we consider, in addition to the *statistical*
significance levels also the magnitude of its coefficient, relative to the
coefficient of variation (CV) for that protein. The CV, calculated on
study-internal QC samples (pools of all samples) are a measure of the technical
noise specific for that protein in the present data set. Thus, with this method
we prioritize proteins with a larger difference between the observed
(biological) difference and the (separately estimated) technical variance.

Below we add thus evaluate for each protein and comparison whether it
fulfills that criteria.

```{r lm--define significance criteria}
#' define the different coefficients as comparisons
comps <- colnames(lm_output)[grep("coef", colnames(lm_output))]

#' define which proteins are considered statistically significant
sign_p <- lm_output[, grep("p-adj", colnames(lm_output))] < 0.05
colnames(sign_p) <- sub("p-adj_", "significant_", colnames(sign_p))
#colSums(sign_p)

#' Convert the (absolute) difference in abundance into a difference in
#' percentage based on a (log2) coefficient.
diff_percentage <- function(x) {
    (2^abs(x) - 1) * 100
}
diff_perc <- apply(lm_output[, comps], MARGIN = 2, diff_percentage)
colnames(diff_perc) <- sub("coef", "diff_perc", colnames(diff_perc))

#' Define for which proteins the difference in concentration is larger
#' than a certain proportion of the CV
cv_cut <- lapply(comps, function(z) {
    z <- sub("coef_", "", z)
    diff_perc[, paste0("diff_perc_", z)] > 50 *  ##after eval, set to 50
        prot_ann[proteins, "cv_qc_chris"]
}) |>
    do.call(what = cbind)
colnames(cv_cut) <- sub("coef_", "cv_cut_", comps)

#' Use statistical significance for non-categorical traits instead of bool
sign_cv <- sign_p & cv_cut
sign_cv[, "significant_Age_10"] <- sign_p[, "significant_Age_10"]

#' Combine with result from linear model converting to a data.frame to
#' avoid logicals being converted to numeric
lm_output <- data.frame(lm_output, sign_cv)
```

The number of proteins called significant or with an adjusted p-value smaller
than 0.05 for each of the extracted coefficients are shown below. For a complete
comparison, the number of proteins complying to either the adjusted p value
or either the CV-value, as single threshold criteria, are calculated and
summarized here as well.

```{r lm--table-sig, results = "asis"}
#' merge to p-adj to cv significance summary
sig_tab <- data.frame(
    `p_adj < 0.05` = colSums(sign_p),
    `cv thresh` = colSums(cv_cut),
    significant = colSums(lm_output[, sub("coef_", "significant_", comps)]),
    check.names = FALSE)

pandoc.table(
    sig_tab, style = "rmarkdown",
    caption = paste0("Number of significant proteins per tested covariate. ",
                     " *p_adj < 0.05*: number of statistically significant",
                     " proteins.",
                     " *cv*: number of proteins with percentage abundance",
                     " difference above the CV-based threshold.",
                     " *significant*: number of proteins that fulfill ",
                     "the CV-based criteria in addition to the p-adj. threshold"))
```

Export results for inspection, and for loading e.g. PDI influence as covariate:

```{r lm--export-results}
results_prot <- data.frame(
    prot_ann[rownames(lm_output),
             c("description", "long_description", "PeptideNN",
               "PeptideNames", "Protein.Ids", "Protein.Names",
               "Genes", "Modified.Sequence", "Stripped.Sequence",
               "Precursor.Id", "cv_qc_chris")],
    lm_output,
    diff_perc
)

#' inclusing inspection of the CV-value of each protein
handle <- file.path(RDATA_PATH, 'lm_output_results_proteins.txt')
write.table(results_prot, file=handle, sep ="\t",
            row.names = FALSE, col.names = TRUE)
```

Export for the paper SI Table S1:

"Table S1. Relatively quantified Scanning SWATH proteomics data. Alternative names of the proteins and associated genes can be consulted at https://www.uniprot.org."

```{r lm--export-Table-S1, warning = FALSE}
#' write Table S1
results <- data.frame(
    prot_ann[rownames(lm_output),
             c("description", "long_description", "PeptideNN",
               "PeptideNames", "Protein.Ids", "Protein.Names",
               "Genes", "Modified.Sequence", "Stripped.Sequence",
               "Precursor.Id")])

#' Add average concentrations for most important groups.
dr <- paste0("data/xlsx/", filename)
dir.create(dr, showWarnings = FALSE, recursive = TRUE)

#' rename so more readable
colnames(results) <- sub("^description$", "Uniprot ID", colnames(results))
colnames(results) <- sub("^long_description$", "Description", colnames(results))
colnames(results) <- sub("^Genes$", "Gene", colnames(results))

write_xlsx(results, path = paste0(dr, "/SI_Table_S1.xlsx"))
```

Export for the paper SI Table S3:

"Table S3. Significant proteins when comparing control (n=141, CAP<238 dB/m) versus different grades of steatosis: grade 1 (n=51, CAP>238 dB/m and CAP>260 dB/m), grade 2 (n=50, CAP>260 dB/m and CAP<290 dB/m), grade 3 steatosis (n=79, naab13>290 dB/m) using linear regression analysis with age and sex as covariates. CVQC, coefficient of variance of pooled quality control samples from the CHRIS study. Alternative nomenclature can be consulted on UniProt (https://www.uniprot.org/)."

```{r lm--export-Table-S3, warning = FALSE}
#' write Table S3
results <- data.frame(
    prot_ann[proteins, c("description", "Genes", "cv_qc_chris")],
    lm_output,
    diff_perc)
results$cv_qc_chris <- 50 * results$cv_qc_chris

#' Add average concentrations for most important groups.
dr <- paste0("data/xlsx/", filename)

#' rename so more readable
colnames(results) <- sub("^description$", "Uniprot ID", colnames(results))
colnames(results) <- sub("^long_description$", "Description", colnames(results))
colnames(results) <- sub("^Genes$", "Gene", colnames(results))

write_xlsx(results, path = paste0(dr, "/SI_Table_S3.xlsx"))
```

The full results are shown below.

```{r, eval = TRUE}
keep <- grepl("description|Genes|coef|p.adj", colnames(results_prot))
tmp <- results_prot[, keep]
datatable(tmp) |>
    formatRound(columns = which(vapply(tmp, is.numeric, logical(1))),
                digits = 3)
```

We visualize the associations in a heatmap. The effect size are added for the
associations, for the significant proteins (min 1 significance to
covariate). also for the not significant associations, the effect size is shown.

```{r prot-lm--heatmap, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for proteins with at least one significant association."}
#' Extract effect sizes
es_results <- results_prot[, grep("effectSize_", colnames(results_prot))] |>
    as.matrix()

#' Restrict to those significant for at least one of the groups (sex, age, SLD type)
sig_results <- results_prot[, grep("significant_", colnames(results_prot))] |>
    as.matrix()
es_results <- es_results[rowSums(sig_results) > 0, ]
print('nr of proteins with >= 1 significant association:')
nrow(es_results)

#' hierarchical clustering
clust1 <- hclust(dist((es_results)), method = "ward.D")
clust2 <- hclust(dist(t(es_results)), method = "ward.D")

#' clean roundup nrs for plot
es_results[es_results == 0] <- NA
es_results <- round(es_results, 2)

#' info regarding FIB4
#print('nr of proteins with a not-zero effect size in the FIB4 group:')
#sum(es_results[, 'effectSize_FIB4'] != 0)

#' in long format
es_results_long <- melt(as.matrix(es_results))

#' ggplot
plt <- es_results_long |>
    ggplot(aes(Var2, Var1, fill = value, label = value)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Effect Size") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 2.5) +
    theme_classic() +
    scale_x_discrete(
        expand = c(0, 0), limits = colnames(es_results)[clust2$order]) +
    scale_y_discrete(
        expand = c(0, 0), limits = rownames(es_results)[clust1$order]) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 10),
          axis.text.y = element_text(hjust = 1, size = 8),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.position = "none")

ggsave(paste0(IMAGE_PATH, "protein_SLDtypes_effect_sizes.png"),
       plot = plt, dpi = 600, width = 12,
       height = 12, units = "cm", scale = 2.5)
plt

```

In second plot, only the significant effect size labels were added to the
heat map. also, the dendograms after hierarchical clustering are shown.

```{r prot-lm--heatmap2, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for significant associations the value of the effect size is shown."}
#' keep just the effect sizes of the significant associations
es_results_sig <- es_results
es_results_sig[as.matrix(sig_results)[rownames(es_results), ] == 0] <- ""

#' plot
colnames(es_results) <- sub("effectSize_", "", colnames(es_results))
pheatmap(es_results, display_numbers = es_results_sig,
         number_color = "#444444", fontsize_number = 7)

```

As expected, the strongest effects are visible in the comparison between
steatosis grade 4 and 1.

In the third heat map, we plot the coefficients of molecules with >= 1
significant association in the comparisons of naab13_cat1:

This is also Figure 2B.

```{r proteins-heatmap3, fig.path = IMAGE_PATH, res = 150, fig.cap = "Associations between covariates. Shown are the coeficients."}

#' Extract coefficients
coef_results_prot <- results_prot[, grep("coef_", colnames(results_prot))] |>
    as.matrix()

#' drop sex, age
coef_results_prot <- coef_results_prot[, c("coef_liver_fat_cat12", "coef_liver_fat_cat13", "coef_liver_fat_cat14")] |>
    as.matrix()

#' Restrict to those significant for at least one of the groups (sex, age, SLD type)
sig_results_prot <- results_prot[, c("significant_liver_fat_cat12", "significant_liver_fat_cat13", "significant_liver_fat_cat14")] |>
    as.matrix()
coef_results_prot <- coef_results_prot[rowSums(sig_results_prot) > 0, ]
print('nr of proteins with >= 1 significant association:')
nrow(coef_results_prot)

#' hierarchical clustering
clust1 <- hclust(dist((coef_results_prot)), method = "ward.D")
clust2 <- hclust(dist(t(coef_results_prot)), method = "ward.D")

#' clean roundup nrs for plot
coef_results_prot[coef_results_prot == 0] <- NA
coef_results_prot <- round(coef_results_prot, 2)

#' in long format
coef_results_prot_long <- melt(as.matrix(coef_results_prot))

#' ggplot
plt <- coef_results_prot_long |>
    ggplot(aes(Var2, Var1, fill = value, label = value)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Coefficient") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 3) +
    theme_classic() +
    #scale_x_discrete(
    #    expand = c(0, 0), limits = colnames(coef_results_prot)[clust2$order]) +
    scale_y_discrete(
        expand = c(0, 0), limits = rownames(coef_results_prot)[clust1$order]) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 5),
          axis.text.y = element_text(hjust = 1, size = 10),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.position = "none")

#ggsave(paste0(IMAGE_PATH, "protein_naab13cat1_HMcoef.png"),
#       plot = plt, dpi = 600, width = 12,
#       height = 12, units = "cm", scale = 2.5)
#plt


#' keep just the Coefficients of the significant associations
coef_results_prot_sig <- coef_results_prot
coef_results_prot_sig[as.matrix(sig_results_prot)[rownames(coef_results_prot), ] == 0] <- ""

#' plot
#colnames(coef_results_prot) <- sub("Coefficient_", "", colnames(coef_results_prot))
#pheatmap(coef_results_prot, display_numbers = es_results_prot_sig,
#         number_color = "#444444", fontsize_number = 7,
#         cluster_cols = FALSE)

```

request with the significance also depicted on the heat map. In the fourth heat
map, we plot the coefficients of molecules with >= 1 significant association in
the comparisons of naab13_cat1:

This is also Figure 2B (2).

```{r proteins-heatmap4, fig.path = IMAGE_PATH, res = 150, fig.cap = "Associations between covariates. Shown are the coeficients, with asteriks for the significants variables."}


#' Extract coefficients
coef_results_prot <- results_prot[, grep("coef_", colnames(results_prot))] |>
    as.matrix()

#' drop sex, age
coef_results_prot <- coef_results_prot[, c("coef_liver_fat_cat12", "coef_liver_fat_cat13", "coef_liver_fat_cat14")] |>
    as.matrix()

#' Restrict to those significant for at least one of the groups (sex, age, SLD type)
sig_results_prot <- results_prot[, c("significant_liver_fat_cat12", "significant_liver_fat_cat13", "significant_liver_fat_cat14")] |>
    as.matrix()
coef_results_prot <- coef_results_prot[rowSums(sig_results_prot) > 0, ]
sig_results_prot  <- sig_results_prot[rowSums(sig_results_prot) > 0, ] #for the asterisks
print('nr of proteins with >= 1 significant association:')
nrow(coef_results_prot)

#' hierarchical clustering
clust1 <- hclust(dist((coef_results_prot)), method = "ward.D")
clust2 <- hclust(dist(t(coef_results_prot)), method = "ward.D")

#' clean roundup nrs for plot
coef_results_prot[coef_results_prot == 0] <- NA
coef_results_prot <- round(coef_results_prot, 2)

#' add asterisks for the significant
coef_with_stars <- coef_results_prot
coef_with_stars[] <- ifelse(sig_results_prot, paste0(round(coef_results_prot, 2), " *"), paste0(round(coef_results_prot, 2), "  "))

#' in long format
coef_results_prot_long <- melt(as.matrix(coef_results_prot))
coef_results_prot_long$label <- as.vector(melt(coef_with_stars)$value)

#' ggplot
plt <- coef_results_prot_long |>
    ggplot(aes(Var2, Var1, fill = value, label = label)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Coefficient") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 3) +
    theme_classic() +
    #scale_x_discrete(
    #    expand = c(0, 0), limits = colnames(coef_results_prot)[clust2$order]) +
    scale_y_discrete(
        expand = c(0, 0), limits = rownames(coef_results_prot)[clust1$order]) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 5),
          axis.text.y = element_text(hjust = 1, size = 10),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.position = "none")

plt
```


We next create volcano plots showing the results for the individual steatosis
categories.

This is also Figure S1A.

```{r prot-FIG-volcano-steatosis, fig.path = IMAGE_PATH, fig.width = 15, fig.height = 5, fig.cap = "Volcano plots for the protein associations with the various steatosis categories (compared to category 1)."}
plot_volcano <- function(x, comparison = character(), ...) {
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    plot(X, -log10(Y), pch = 21,
         col = "#000000ce", bg = "#00000060",
         xlab = "coefficient", ylab = expression(-log[10](p[adjusted])),
         ...)
}
significant_points <- function(x, comparison = character(),
                               diff_perc = 2, col = "#E41A1CCE",
                               bg = "#E41A1C40", ...) {
    s <- x[, paste0("significant_", comparison)]
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    if (any(s))
        points(X[s], -log10(Y[s]), col = col, bg = bg, pch = 21)
}

par(mfrow = c(1, 3))
xl <- range(results_prot[, c("coef_liver_fat_cat12", "coef_liver_fat_cat13",
                             "coef_liver_fat_cat14")])
yl <- c(0, -log10(min(results_prot[, c("p.adj_liver_fat_cat12",
                                       "p.adj_liver_fat_cat13",
                                       "p.adj_liver_fat_cat14")])))
sigs <- results_prot$significant_liver_fat_cat12 |
    results_prot$significant_liver_fat_cat13 |
    results_prot$significant_liver_fat_cat14

plot_volcano(results_prot, "liver_fat_cat12", main = "Category 2",
             xlim = xl, ylim = yl)
points(results_prot$coef_liver_fat_cat12[sigs],
       -log10(results_prot$p.adj_liver_fat_cat12[sigs]),
       pch= 21, col = "#E41A1CCE", bg = "#E41A1C40")
grid()
abline(h = -log10(0.05), lty = 2, col = "grey")
plot_volcano(results_prot, "liver_fat_cat13", main = "Category 3",
             xlim = xl, ylim = yl)
points(results_prot$coef_liver_fat_cat13[sigs],
       -log10(results_prot$p.adj_liver_fat_cat13[sigs]),
       pch= 21, col = "#E41A1CCE", bg = "#E41A1C40")
grid()
abline(h = -log10(0.05), lty = 2, col = "grey")
plot_volcano(results_prot, "liver_fat_cat14", main = "Category 4",
             xlim = xl, ylim = yl)
points(results_prot$coef_liver_fat_cat14[sigs],
       -log10(results_prot$p.adj_liver_fat_cat14[sigs]),
       pch= 21, col = "#E41A1CCE", bg = "#E41A1C40")
grid()
abline(h = -log10(0.05), lty = 2, col = "grey")

```

Difference of abundance increases with the steatosis grade. Also, the
significant proteins for steatosis grade 4 show a trend of difference already in
steatosis grade 2 and 3 participants.

Proteins with significant different abundance for at least one steatosis grade
class are shown below.

```{r lm--table-sig-sex, results = "asis", echo = FALSE}
tmp <- results_prot[sigs, c("Genes",
                       "coef_SexMale", "p.adj_SexMale",
                       "coef_Age_10", "p.adj_Age_10",
                       "coef_liver_fat_cat12", "p.adj_liver_fat_cat12",
                       "coef_liver_fat_cat13", "p.adj_liver_fat_cat13",
                       "coef_liver_fat_cat14", "p.adj_liver_fat_cat14")]
tmp <- tmp[order(tmp$p.adj_liver_fat_cat14), ]

pandoc.table(
    tmp, style = "rmarkdown", split.table = Inf,
    caption = paste0("Protiens significantly associated with at least ",
                     "on steatosis category"))
```

Significant proteins associated with steatosis (top 5 strongest associated
proteins) are visualized below. Points are colored according to sample sex.

```{r lm--prep for plotting}
#' prep for plotting
PLOTLY <- FALSE
ggplotly_or_not <- function(x, interactive = TRUE) {
    if (interactive) ggplotly(x)
    else print(x)
}
myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))

box_plot <- function(data, aes, analyte) {
    ggplot(data, aes) +
        geom_jitter(position = position_jitter(0.24),
                    aes(colour = Sex, alpha = 0.5 , size = (BMI/10))) +
        geom_violin(trim = FALSE, aes(alpha = 0.5)) +
        stat_summary(fun = median, geom = "point", shape = 23, size = 2) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) + labs(title = analyte) +
        theme_bw ()
}

scatter_plot <- function(data, aes, analyte) {
    ggplot(data, aes) +
        geom_jitter(position = position_jitter(0.24),
                    aes(colour = Sex, alpha = 0.5 , size = 0.1)) +
        stat_smooth(method = lm, se = TRUE) + #with standard error
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) + labs(title = analyte) +
        theme_bw ()
}

#' for boxplot func to work automatically
tmp_colnr <- which(colnames(gal_met_prot_data) == TARGET)
gal_met_prot_data$tmp_target <- gal_met_prot_data[, tmp_colnr]   #value of eg BMI
```

```{r prot-lm-top1, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[1]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]  #intensities of metabolite
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r prot-lm-top2, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[2]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r prot-lm-top3, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[3]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r prot-lm-top4, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[4]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r prot-lm-top5, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[5]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

### Summary

- A considerable number of proteins is associated with steatosis.
- The extend of difference, compared to steatosis grade 1 increases with the
  steatosis category. Consequently, the largest number of proteins with
  significant differences were found between steatosis grade 4 and 1.
- We restrict thus interpretation and further analyses on these.


## Association between serum metabolites and liver fat (steatosis)

Perform multiple linear regression analyses for each metabolite to identify
those with significant differences in abundances between steatosis categories.

```{r define-lm-covariates-and-target}
print('covariates of the linear regression model:')
COVARIATES <- c("Sex", "Age_10")
print(COVARIATES)

print('target of the linear regression model:')
TARGET <- 'liver_fat_cat1'
print(TARGET)

DATASET <- 'NAFLD'
print(paste0('final variables of the ', DATASET, ' dataset used:'))

print('nr of variables inspected:')
print(length(metabolites))

print('nr of individuals inspected:')
print(nrow(gal_met_prot_data))
```

Linear regression models are fitted for all metabolites separately (as a
response) with sex, age and the (categorical) steatosis category as a
covariates.

Model coefficients and p-values are provided. Raw p-values are adjusted for
multiple hypothesis testing with the method from Benjamini & Hochberg.

```{r linear model, message=FALSE}

#' Fit linear models and extract coefficients and p-values on each protein
lm_output <- lapply(metabolites, FUN = funForLinReg,
                    data = gal_met_prot_data) |>
    do.call(what = rbind)

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)
```

Next we calculate the effect sizes.

To obtain comparable coefficients, prior to linear modelling all the metabolites
are standardized to zero mean and unit variance (note that this does not
influence the linear models since they are affine equivariant). In such
autoscaled data, difference of 1 corresponds to a standard deviation of 1. This
parameter will be called "effect size" and will be use to evaluate group
differences.

```{r effectsize, message=FALSE}
#' Perform autoscaling for effect size (es) calculation
gal_met_prot_data_scaled <- gal_met_prot_data
gal_met_prot_data_scaled[, metabolites] <- gal_met_prot_data[, metabolites] |>
    as.matrix() |>
    scale(scale = TRUE)

#' fit linear models and extract the coefficients
es <- lapply(metabolites, FUN = funForLinReg,
             data = gal_met_prot_data_scaled) |>
    do.call(what = rbind)
es <- es[, grep("coef", colnames(es))]
colnames(es) <- sub("coef", "effectSize", colnames(es))
lm_output <- cbind(lm_output, es)
```

To call a metabolites *significant* we consider, in addition to the
*statistical* significance levels also the magnitude of its coefficient,
relative to the coefficient of variation (CV) for that protein. The CV,
calculated on study-internal QC samples (pools of all samples) are a measure of
the technical noise specific for that protein in the present data set. Thus,
with this method we prioritize metabolites with a larger difference between the
observed (biological) difference and the (separately estimated) technical
variance.

Below we add thus evaluate for each protein and comparison whether it
fulfills that criteria.

```{r define significance criteria}
#' define the different coefficients as comparisons
comps <- colnames(lm_output)[grep("coef", colnames(lm_output))]

#' define which metabolites are considered statistically significant
sign_p <- lm_output[, grep("p-adj", colnames(lm_output))] < 0.05
colnames(sign_p) <- sub("p-adj_", "significant_", colnames(sign_p))
#colSums(sign_p)

#' Convert the (absolute) difference in abundance into a difference in
#' percentage based on a (log2) coefficient.
diff_percentage <- function(x) {
    (2^abs(x) - 1) * 100
}
diff_perc <- apply(lm_output[, comps], MARGIN = 2, diff_percentage)
colnames(diff_perc) <- sub("coef", "diff_perc", colnames(diff_perc))

#' Define for which metabolites the difference in concentration is larger
#' than a certain proportion of the CV
cv_cut <- lapply(comps, function(z) {
    z <- sub("coef_", "", z)
    diff_perc[, paste0("diff_perc_", z)] > 50 *  ##after eval, set to 50
        met_ann[metabolites, "cv_qc_chris"]
}) |>
    do.call(what = cbind)
colnames(cv_cut) <- sub("coef_", "cv_cut_", comps)

#' Use statistical significance for non-categorical traits instead of bool
sign_cv <- sign_p & cv_cut
sign_cv[, "significant_Age_10"] <- sign_p[, "significant_Age_10"]

#' Combine with result from linear model converting to a data.frame to
#' avoid logicals being converted to numeric
lm_output <- data.frame(lm_output, sign_cv)
```

The number of metabolites called significant or with an adjusted p-value smaller
than 0.05 for each of the extracted coefficients are shown below. For a complete
comparison, the number of metabolites complying to either the adjusted p value
or either the CV-value, as single threshold criteria, are calculated and
summarized here as well.

```{r table-sig, results = "asis"}
#' merge to p-adj to cv significance summary
sig_tab <- data.frame(
    `p_adj < 0.05` = colSums(sign_p),
    `cv thresh` = colSums(cv_cut),
    significant = colSums(lm_output[, sub("coef_", "significant_", comps)]),
    check.names = FALSE)

pandoc.table(
    sig_tab, style = "rmarkdown",
    caption = paste0("Number of significant metabolites per tested covariate. ",
                     " *p_adj < 0.05*: number of statistically significant",
                     " metabolites.",
                     " *cv*: number of metabolites with percentage abundance",
                     " difference above the CV-based threshold.",
                     " *significant*: number of metabolites that fulfill ",
                     "the CV-based criteria in addition to the p-adj. threshold"))
```

Export results for inspection, and for loading e.g. PDI influence as covariate:

```{r export-results}
results_met <- data.frame(
    met_ann[rownames(lm_output), c("analyte_name", "analyte_class",
                                   "biochemical_name", "cv_qc_chris")],
    lm_output,
    diff_perc
)

#' write to file for manual inspection,
#' inclusing inspection of the CV-value of each protein
handle <- file.path(RDATA_PATH, 'lm_output_results_metabolites.txt')
write.table(results_met, file=handle, sep ="\t",
            row.names = FALSE, col.names = TRUE)
```

Export for the paper SI Table S2:

"Table S2. Absolutely quantified metabolites and lipids. Alternative names of the analytes can be consulted with the provider (Biocrates AbsoluteIDQ® p180 kit)."

```{r lm--export-Table-S2, warning = FALSE}
#' write Table S2
results <- data.frame(
    met_ann[rownames(lm_output), c("analyte_name", "analyte_class",
                                    "biochemical_name")])

#' Add average concentrations for most important groups.
dr <- paste0("data/xlsx/", filename)
dir.create(dr, showWarnings = FALSE, recursive = TRUE)

#' rename so more readable
colnames(results) <- sub("^analyte_name$", "Analyte name", colnames(results))
colnames(results) <- sub("^analyte_class$", "Analyte class", colnames(results))
colnames(results) <- sub("^biochemical_name$", "Biochemical name", colnames(results))

write_xlsx(results, path = paste0(dr, "/SI_Table_S2.xlsx"))
```


Export for the paper SI Table S4:

"Table S4. Significant metabolites and lipids when comparing control (n=141, CAP<238 dB/m) versus different grades of steatosis: grade 1 (n=51, CAP>238 dB/m and CAP>260 dB/m), grade 2 (n=50, CAP>260 dB/m and CAP<290 dB/m), grade 3 steatosis (n=79, naab13>290 dB/m) using linear regression analysis with age and sex as covariates. CVQC, coefficient of variance of pooled quality control samples from the CHRIS study. Alternative nomenclature can be consulted with the provider (Biocrates p180 kit)."

```{r lm--export-Table-S4, warning = FALSE}
#' write Table S4
results <- data.frame(
    met_ann[rownames(lm_output), c("analyte_name", "analyte_class",
                                   "biochemical_name", "cv_qc_chris")],
    lm_output,
    diff_perc
)
results$cv_qc_chris <- 50 * results$cv_qc_chris

#' Add average concentrations for most important groups.
dr <- paste0("data/xlsx/", filename)

write_xlsx(results, path = paste0(dr, "/SI_Table_S4.xlsx"))
```

The full results are shown below.

```{r, eval = TRUE}
keep <- grepl("analyte_name|coef|p.adj", colnames(results_met))
tmp <- results_met[, keep]
datatable(tmp) |>
    formatRound(columns = which(vapply(tmp, is.numeric, logical(1))),
                digits = 3)
```

We visualize the associations in a heatmap. The effect size are added for the
associations, for the significant metabolites (min 1 significance to
covariate). also for the not significant associations, the effect size is shown.

```{r met-heatmap, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for metabolites with at least one significant association.", eval = FALSE}
#' Extract effect sizes
es_results <- results_met[, grep("effectSize_", colnames(results_met))] |>
    as.matrix()

#' Restrict to those significant for at least one of the groups (sex, age, SLD type)
sig_results <- results_met[, grep("significant_", colnames(results_met))] |>
    as.matrix()
es_results <- es_results[rowSums(sig_results) > 0, ]
print('nr of metabolites with >= 1 significant association:')
nrow(es_results)

#' hierarchical clustering
clust1 <- hclust(dist((es_results)), method = "ward.D")
clust2 <- hclust(dist(t(es_results)), method = "ward.D")

#' clean roundup nrs for plot
es_results[es_results == 0] <- NA
es_results <- round(es_results, 2)

#' info regarding FIB4
#print('nr of metabolites with a not-zero effect size in the FIB4 group:')
#sum(es_results[, 'effectSize_FIB4'] != 0)

#' in long format
es_results_long <- melt(as.matrix(es_results))

#' ggplot
plt <- es_results_long |>
    ggplot(aes(Var2, Var1, fill = value, label = value)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Effect Size") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 2.5) +
    theme_classic() +
    scale_x_discrete(
        expand = c(0, 0), limits = colnames(es_results)[clust2$order]) +
    scale_y_discrete(
        expand = c(0, 0), limits = rownames(es_results)[clust1$order]) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 10),
          axis.text.y = element_text(hjust = 1, size = 8),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.position = "none")

ggsave(paste0(IMAGE_PATH, "protein_SLDtypes_effect_sizes.png"),
       plot = plt, dpi = 600, width = 12,
       height = 12, units = "cm", scale = 2.5)
plt

```

In second plot, only the significant effect size labels were added to the
heat map. also, the dendograms after hierarchical clustering are shown.

```{r met-heatmap2, fig.width = 14, fig.height = 14, fig.cap = "Associations between covariates. Shown are the effect size for the associations, for significant associations the value of the effect size is shown.", eval = FALSE}
#' keep just the effect sizes of the significant associations
es_results_sig <- es_results
es_results_sig[as.matrix(sig_results)[rownames(es_results), ] == 0] <- ""

#' plot
colnames(es_results) <- sub("effectSize_", "", colnames(es_results))
pheatmap(es_results, display_numbers = es_results_sig,
         number_color = "#444444", fontsize_number = 7)
```

In the third heat map, we plot the coefficients of molecules with >= 1
significant association in the comparisons of naab13_cat1:

This is also Figure 2C.

```{r metabolites-heatmap3, fig.path = IMAGE_PATH, res = 150, fig.cap = "Associations between covariates. Shown are the coeficients."}

#' Extract coefficients
coef_results_met <- results_met[, grep("coef_", colnames(results_met))] |>
    as.matrix()

#' drop sex, age
coef_results_met <- coef_results_met[, c("coef_liver_fat_cat12", "coef_liver_fat_cat13", "coef_liver_fat_cat14")] |>
    as.matrix()

#' Restrict to those significant for at least one of the groups (sex, age, SLD type)
sig_results_met <- results_met[, c("significant_liver_fat_cat12", "significant_liver_fat_cat13", "significant_liver_fat_cat14")] |>
    as.matrix()
coef_results_met <- coef_results_met[rowSums(sig_results_met) > 0, ]
print('nr of proteins with >= 1 significant association:')
nrow(coef_results_met)

#' hierarchical clustering
clust1 <- hclust(dist((coef_results_met)), method = "ward.D")
clust2 <- hclust(dist(t(coef_results_met)), method = "ward.D")

#' clean roundup nrs for plot
coef_results_met[coef_results_met == 0] <- NA
coef_results_met <- round(coef_results_met, 2)

#' in long format
coef_results_met_long <- melt(as.matrix(coef_results_met))

#' ggplot
plt <- coef_results_met_long |>
    ggplot(aes(Var2, Var1, fill = value, label = value)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Coefficient") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 1.5) +
    theme_classic() +
    #scale_x_discrete(
    #    expand = c(0, 0), limits = colnames(coef_results_met)[clust2$order]) +
    scale_y_discrete(
        expand = c(0, 0), limits = rownames(coef_results_met)[clust1$order]) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 5),
          axis.text.y = element_text(hjust = 1, size = 5),
          plot.margin = margin(t = 5, l = 145, r = 145, b = 5), #top left right bottom
          legend.position = "none")

#ggsave(paste0(IMAGE_PATH, "metab_naab13cat1_HMcoef.png"),
#       plot = plt, dpi = 600, width = 12,
#       height = 12, units = "cm", scale = 2.5)
#plt


#' keep just the Coefficients of the significant associations
coef_results_met_sig <- coef_results_met
coef_results_met_sig[as.matrix(sig_results_met)[rownames(coef_results_met), ] == 0] <- ""

#' plot
#colnames(coef_results_met) <- sub("Coefficient_", "", colnames(coef_results_met))
#pheatmap(coef_results_met, display_numbers = es_results_met_sig,
#         number_color = "#444444", fontsize_number = 7,
#         cluster_cols = FALSE)

```

request with the significance also depicted on the heat map. In the fourth heat
map, we plot the coefficients of molecules with >= 1 significant association in
the comparisons of naab13_cat1:

This is also Figure 2C (2).

```{r metabolites-heatmap4, fig.path = IMAGE_PATH, res = 150, fig.cap = "Associations between covariates. Shown are the coeficients, with asteriks for signifcant variables."}


#' Extract coefficients
coef_results_met <- results_met[, grep("coef_", colnames(results_met))] |>
    as.matrix()

#' drop sex, age
coef_results_met <- coef_results_met[, c("coef_liver_fat_cat12", "coef_liver_fat_cat13", "coef_liver_fat_cat14")] |>
    as.matrix()

#' Restrict to those significant for at least one of the groups (sex, age, SLD type)
sig_results_met <- results_met[, c("significant_liver_fat_cat12", "significant_liver_fat_cat13", "significant_liver_fat_cat14")] |>
    as.matrix()
coef_results_met <- coef_results_met[rowSums(sig_results_met) > 0, ]
sig_results_met  <- sig_results_met[rowSums(sig_results_met) > 0, ] #for the asterisks
print('nr of proteins with >= 1 significant association:')
nrow(coef_results_met)

#' hierarchical clustering
clust1 <- hclust(dist((coef_results_met)), method = "ward.D")
clust2 <- hclust(dist(t(coef_results_met)), method = "ward.D")

#' clean roundup nrs for plot
coef_results_met[coef_results_met == 0] <- NA
coef_results_met <- round(coef_results_met, 2)

#' add asterisks for the significant
coef_with_stars <- coef_results_met
coef_with_stars[] <- ifelse(sig_results_met, paste0(round(coef_results_met, 2), " *"), paste0(round(coef_results_met, 2), "  "))

#' in long format
coef_results_met_long <- melt(as.matrix(coef_results_met))
coef_results_met_long$label <- as.vector(melt(coef_with_stars)$value)

#' ggplot
plt <- coef_results_met_long |>
    ggplot(aes(Var2, Var1, fill = value, label = label)) +
    geom_tile() +
    labs(x = NULL, y = NULL, fill = "Coefficient") +
    scale_fill_gradient2(mid = "#FBFEF9", low = "#0C6291", high = "#A63446") +
    geom_text(size = 3) +
    theme_classic() +
    #scale_x_discrete(
    #    expand = c(0, 0), limits = colnames(coef_results_met)[clust2$order]) +
    scale_y_discrete(
        expand = c(0, 0), limits = rownames(coef_results_met)[clust1$order]) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 5),
          axis.text.y = element_text(hjust = 1, size = 10),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.position = "none")

plt
```

We investigate the associations for MASLD, i.e. the difference in average
concentrations between participants of SLD types *MASLD* and *none*. We show the
results for that coefficient using a volcano plot, coloring significant
associations in red.

This is also Figure S1B.

```{r met-FIG-volcano-steatosis, fig.path = IMAGE_PATH, fig.width = 15, fig.height = 5, fig.cap = "Volcano plots for associations of metabolites with steatosis categories."}
par(mfrow = c(1, 3))
xl <- range(results_met[, c("coef_liver_fat_cat12", "coef_liver_fat_cat13",
                            "coef_liver_fat_cat14")])
yl <- c(0, -log10(min(results_met[, c("p.adj_liver_fat_cat12",
                                      "p.adj_liver_fat_cat13",
                                      "p.adj_liver_fat_cat14")])))
sigs <- results_met$significant_liver_fat_cat12 |
    results_met$significant_liver_fat_cat13 |
    results_met$significant_liver_fat_cat14

plot_volcano(results_met, "liver_fat_cat12", main = "Category 2",
             xlim = xl, ylim = yl)
points(results_met$coef_liver_fat_cat12[sigs],
       -log10(results_met$p.adj_liver_fat_cat12[sigs]),
       pch= 21, col = "#E41A1CCE", bg = "#E41A1C40")
grid()
abline(h = -log10(0.05), lty = 2, col = "grey")
plot_volcano(results_met, "liver_fat_cat13", main = "Category 3",
             xlim = xl, ylim = yl)
points(results_met$coef_liver_fat_cat13[sigs],
       -log10(results_met$p.adj_liver_fat_cat13[sigs]),
       pch= 21, col = "#E41A1CCE", bg = "#E41A1C40")
grid()
abline(h = -log10(0.05), lty = 2, col = "grey")
plot_volcano(results_met, "liver_fat_cat14", main = "Category 4",
             xlim = xl, ylim = yl)
points(results_met$coef_liver_fat_cat14[sigs],
       -log10(results_met$p.adj_liver_fat_cat14[sigs]),
       pch= 21, col = "#E41A1CCE", bg = "#E41A1C40")
grid()
abline(h = -log10(0.05), lty = 2, col = "grey")
```

Similar to the proteomics analysis, significance of difference in abundance
increases. For metabolites we see however a large number of significant
metabolites for the comparison between steatosis grade 3 and 1.

Metabolites with significant different abundance for at least one steatosis
grade class are shown below.

```{r, results = "asis", echo = FALSE}
tmp <- results_met[sigs, c("analyte_name",
                           "coef_SexMale", "p.adj_SexMale",
                           "coef_Age_10", "p.adj_Age_10",
                           "coef_liver_fat_cat12", "p.adj_liver_fat_cat12",
                           "coef_liver_fat_cat13", "p.adj_liver_fat_cat13",
                           "coef_liver_fat_cat14", "p.adj_liver_fat_cat14")]
tmp <- tmp[order(tmp$p.adj_liver_fat_cat14), ]

pandoc.table(
    tmp, style = "rmarkdown", split.table = Inf,
    caption = paste0("Metabolites significantly associated with at least ",
                     "on steatosis category"))
```

Significant metabolites associated with steatosis (top 5 strongest associated
metabolites) are visualized below. Points are colored according to sample sex.

```{r met-lm-top1, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[1]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]  #intensities of metabolite
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r met-lm-top2, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[2]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r met-lm-top3, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[3]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r met-lm-top4, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[4]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

```{r met-lm-top5, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
top_analyte <- rownames(tmp)[5]
gal_met_prot_data$top_analyte <- gal_met_prot_data[, top_analyte]
s_1 <- scatter_plot(gal_met_prot_data, aes(x = tmp_target, y = top_analyte),
                analyte = top_analyte)
ggplotly_or_not(s_1, PLOTLY)
```

### Summary

- Similar to the proteomics analysis we see the most significant differences
  between steatosis grade 4 and 1.
- However, we see already differences between grade 3 and 1.


# Check some trends of proteins/metabolites of interest

## Proteins

Alpha-1-antitrypsin (A1AT, P01009). from Christoph, not signif/changing
anywhere.

```{r explore-prot1}
#colnames(gal_met_prot_data)
#' A1AT
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$liver_fat, y = gal_met_prot_data$P01009, main = 'liver_fat')
plot(x = gal_met_prot_data$visual_fat, y = gal_met_prot_data$P01009, main = 'visual_fat')
plot(x = gal_met_prot_data$liver_fat_cat1, y = gal_met_prot_data$P01009, main = 'liver_fat_cat1')
plot(x = gal_met_prot_data$liver_fat_cat2, y = gal_met_prot_data$P01009, main = 'liver_fat_cat2')
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$Sex, y = gal_met_prot_data$P01009, main = 'sex')
plot(x = gal_met_prot_data$Age_10, y = gal_met_prot_data$P01009, main = 'Age_10')
plot(x = gal_met_prot_data$BMI, y = gal_met_prot_data$P01009, main = 'BMI')
plot(x = gal_met_prot_data$T2D, y = gal_met_prot_data$P01009, main = 'T2D')

```

P19827 = ITIH1 = Inter-alpha-trypsin inhibitor heavy chain H1. signif in the
fib4.

```{r explore-prot2}
#' ITIH1
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$liver_fat, y = gal_met_prot_data$P19827, main = 'liver_fat')
plot(x = gal_met_prot_data$visual_fat, y = gal_met_prot_data$P19827, main = 'visual_fat')
plot(x = gal_met_prot_data$liver_fat_cat1, y = gal_met_prot_data$P19827, main = 'liver_fat_cat1')
plot(x = gal_met_prot_data$liver_fat_cat2, y = gal_met_prot_data$P19827, main = 'liver_fat_cat2')
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$Sex, y = gal_met_prot_data$P19827, main = 'sex')
plot(x = gal_met_prot_data$Age_10, y = gal_met_prot_data$P19827, main = 'Age_10')
plot(x = gal_met_prot_data$BMI, y = gal_met_prot_data$P19827, main = 'BMI')
plot(x = gal_met_prot_data$T2D, y = gal_met_prot_data$P19827, main = 'T2D')

```

P02774;P02774-3 = GC = Vitamin D-binding protein. signif in the naab11_cat2.

```{r explore-prot3}
#' GC
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$liver_fat, y = gal_met_prot_data$`P02774;P02774-3`, main = 'liver_fat')
plot(x = gal_met_prot_data$visual_fat, y = gal_met_prot_data$`P02774;P02774-3`, main = 'visual_fat')
plot(x = gal_met_prot_data$liver_fat_cat1, y = gal_met_prot_data$`P02774;P02774-3`, main = 'liver_fat_cat1')
plot(x = gal_met_prot_data$liver_fat_cat2, y = gal_met_prot_data$`P02774;P02774-3`, main = 'liver_fat_cat2')
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$Sex, y = gal_met_prot_data$`P02774;P02774-3`, main = 'sex')
plot(x = gal_met_prot_data$Age_10, y = gal_met_prot_data$`P02774;P02774-3`, main = 'Age_10')
plot(x = gal_met_prot_data$BMI, y = gal_met_prot_data$`P02774;P02774-3`, main = 'BMI')
plot(x = gal_met_prot_data$T2D, y = gal_met_prot_data$`P02774;P02774-3`, main = 'T2D')

```

P04004 = Vitronectin. neg corr for fib4, pos corr for naab11.
Vitronectin is a glycoprotein of the hemopexin family which is synthesized and
excreted by the liver, and abundantly found in serum, the extracellular matrix
and bone. In humans it is encoded by the VTN gene. Vitronectin binds to
integrin alpha-V beta-3 and thus promotes cell adhesion and spreading.

```{r explore-prot4}
#' Vitronectin
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$liver_fat, y = gal_met_prot_data$P04004, main = 'liver_fat')
plot(x = gal_met_prot_data$visual_fat, y = gal_met_prot_data$P04004, main = 'visual_fat')
plot(x = gal_met_prot_data$liver_fat_cat1, y = gal_met_prot_data$P04004, main = 'liver_fat_cat1')
plot(x = gal_met_prot_data$liver_fat_cat2, y = gal_met_prot_data$P04004, main = 'liver_fat_cat2')
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$Sex, y = gal_met_prot_data$P04004, main = 'sex')
plot(x = gal_met_prot_data$Age_10, y = gal_met_prot_data$P04004, main = 'Age_10')
plot(x = gal_met_prot_data$BMI, y = gal_met_prot_data$P04004, main = 'BMI')
plot(x = gal_met_prot_data$T2D, y = gal_met_prot_data$P04004, main = 'T2D')

```

P05090 = Apolipoprotein D. pos corr with fib4, weak neg corr with naab11.

```{r explore-prot5}
#' Apolipoprotein D
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$liver_fat, y = gal_met_prot_data$P05090, main = 'liver_fat')
plot(x = gal_met_prot_data$visual_fat, y = gal_met_prot_data$P05090, main = 'visual_fat')
plot(x = gal_met_prot_data$liver_fat_cat1, y = gal_met_prot_data$P05090, main = 'liver_fat_cat1')
plot(x = gal_met_prot_data$liver_fat_cat2, y = gal_met_prot_data$P05090, main = 'liver_fat_cat2')
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$Sex, y = gal_met_prot_data$P05090, main = 'sex')
plot(x = gal_met_prot_data$Age_10, y = gal_met_prot_data$P05090, main = 'Age_10')
plot(x = gal_met_prot_data$BMI, y = gal_met_prot_data$P05090, main = 'BMI')
plot(x = gal_met_prot_data$T2D, y = gal_met_prot_data$P05090, main = 'T2D')

```

"P02750" = Leucine-rich alpha-2-glycoprotein

```{r}
#' P02750
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$liver_fat, y = gal_met_prot_data$P02750, main = 'liver_fat')
plot(x = gal_met_prot_data$visual_fat, y = gal_met_prot_data$P02750, main = 'visual_fat')
plot(x = gal_met_prot_data$liver_fat_cat1, y = gal_met_prot_data$P02750, main = 'liver_fat_cat1')
plot(x = gal_met_prot_data$liver_fat_cat2, y = gal_met_prot_data$P02750, main = 'liver_fat_cat2')
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$Sex, y = gal_met_prot_data$P02750, main = 'sex')
plot(x = gal_met_prot_data$Age_10, y = gal_met_prot_data$P02750, main = 'Age_10')
plot(x = gal_met_prot_data$BMI, y = gal_met_prot_data$P02750, main = 'BMI')
plot(x = gal_met_prot_data$T2D, y = gal_met_prot_data$P02750, main = 'T2D')

```

O95445 = APOM = Apolipoprotein M

```{r}
#' O95445
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$liver_fat, y = gal_met_prot_data$O95445, main = 'liver_fat')
plot(x = gal_met_prot_data$visual_fat, y = gal_met_prot_data$O95445, main = 'visual_fat')
plot(x = gal_met_prot_data$liver_fat_cat1, y = gal_met_prot_data$O95445, main = 'liver_fat_cat1')
plot(x = gal_met_prot_data$liver_fat_cat2, y = gal_met_prot_data$O95445, main = 'liver_fat_cat2')
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$Sex, y = gal_met_prot_data$O95445, main = 'sex')
plot(x = gal_met_prot_data$Age_10, y = gal_met_prot_data$O95445, main = 'Age_10')
plot(x = gal_met_prot_data$BMI, y = gal_met_prot_data$O95445, main = 'BMI')
plot(x = gal_met_prot_data$T2D, y = gal_met_prot_data$O95445, main = 'T2D')

```

Color/size also control-MASLD, steatosis grade (liver_fat_cat1). TODO

```{r scatters-Apo M}
plot(x = gal_met_prot_data$liver_fat, y = gal_met_prot_data$O95445)

#' prep for plotting
PLOTLY <- FALSE
ggplotly_or_not <- function(x, interactive = TRUE) {
    if (interactive) ggplotly(x)
    else print(x)
}
myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))


#' scatterplot with sex as color, BMI as size
scatter_plot <- function(data, aes, analyte) {
    ggplot(data, aes) +
        geom_point(aes(colour = Sex, alpha = 0.5, size = BMI)) +
        #geom_jitter(position = position_jitter(0.24),
        #            aes(colour = Sex, alpha = 0.5 , size = 0.1)) +
        stat_smooth(method = lm, se = TRUE) + #with standard error
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) + labs(title = analyte) +
        theme_bw ()
}

s_1 <- scatter_plot(gal_met_prot_data, aes(x = liver_fat, y = O95445),
                analyte = 'O95445')
ggplotly_or_not(s_1, PLOTLY)

#' scatterplot with T2D as color, Age as size
scatter_plot <- function(data, aes, analyte) {
    ggplot(data, aes) +
        geom_point(aes(colour = T2D, alpha = 0.5, size = Age_10)) +
        #geom_jitter(position = position_jitter(0.24),
        #            aes(colour = Sex, alpha = 0.5 , size = 0.1)) +
        stat_smooth(method = lm, se = TRUE) + #with standard error
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) + labs(title = analyte) +
        theme_bw ()
}

s_1 <- scatter_plot(gal_met_prot_data, aes(x = liver_fat, y = O95445),
                analyte = 'O95445')
ggplotly_or_not(s_1, PLOTLY)

#' scatterplot with sld_type as color, liver_fat_cat1 as size
scatter_plot <- function(data, aes, analyte) {
    ggplot(data, aes) +
        geom_point(aes(colour = sld_type, alpha = 0.5, size = as.numeric(visual_fat))) +
        #geom_jitter(position = position_jitter(0.24),
        #            aes(colour = Sex, alpha = 0.5 , size = 0.1)) +
        stat_smooth(method = lm, se = TRUE) + #with standard error
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) + labs(title = analyte) +
        theme_bw ()
}

s_1 <- scatter_plot(gal_met_prot_data, aes(x = liver_fat, y = O95445),
                analyte = 'O95445')
ggplotly_or_not(s_1, PLOTLY)

```


A0A0B4J1U7, IGHV6-1 = Immunoglobulin heavy variable 6-1

```{r}
#' A0A0B4J1U7
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$liver_fat, y = gal_met_prot_data$A0A0B4J1U7, main = 'liver_fat')
plot(x = gal_met_prot_data$visual_fat, y = gal_met_prot_data$A0A0B4J1U7, main = 'visual_fat')
plot(x = gal_met_prot_data$liver_fat_cat1, y = gal_met_prot_data$A0A0B4J1U7, main = 'liver_fat_cat1')
plot(x = gal_met_prot_data$liver_fat_cat2, y = gal_met_prot_data$A0A0B4J1U7, main = 'liver_fat_cat2')
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$Sex, y = gal_met_prot_data$A0A0B4J1U7, main = 'sex')
plot(x = gal_met_prot_data$Age_10, y = gal_met_prot_data$A0A0B4J1U7, main = 'Age_10')
plot(x = gal_met_prot_data$BMI, y = gal_met_prot_data$A0A0B4J1U7, main = 'BMI')
plot(x = gal_met_prot_data$T2D, y = gal_met_prot_data$A0A0B4J1U7, main = 'T2D')

```


## Metabolites

Taurine

```{r explore-prot6}
#' Taurine
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$liver_fat, y = gal_met_prot_data$Taurine, main = 'liver_fat')
plot(x = gal_met_prot_data$visual_fat, y = gal_met_prot_data$Taurine, main = 'visual_fat')
plot(x = gal_met_prot_data$liver_fat_cat1, y = gal_met_prot_data$Taurine, main = 'liver_fat_cat1')
plot(x = gal_met_prot_data$liver_fat_cat2, y = gal_met_prot_data$Taurine, main = 'liver_fat_cat2')
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$Sex, y = gal_met_prot_data$Taurine, main = 'sex')
plot(x = gal_met_prot_data$Age_10, y = gal_met_prot_data$Taurine, main = 'Age_10')
plot(x = gal_met_prot_data$BMI, y = gal_met_prot_data$Taurine, main = 'BMI')
plot(x = gal_met_prot_data$T2D, y = gal_met_prot_data$Taurine, main = 'T2D')

```

Tyrosine

```{r explore-prot7}
#' Tyrosine
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$liver_fat, y = gal_met_prot_data$Tyrosine, main = 'liver_fat')
plot(x = gal_met_prot_data$visual_fat, y = gal_met_prot_data$Tyrosine, main = 'visual_fat')
plot(x = gal_met_prot_data$liver_fat_cat1, y = gal_met_prot_data$Tyrosine, main = 'liver_fat_cat1')
plot(x = gal_met_prot_data$liver_fat_cat2, y = gal_met_prot_data$Tyrosine, main = 'liver_fat_cat2')
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$Sex, y = gal_met_prot_data$Tyrosine, main = 'sex')
plot(x = gal_met_prot_data$Age_10, y = gal_met_prot_data$Tyrosine, main = 'Age_10')
plot(x = gal_met_prot_data$BMI, y = gal_met_prot_data$Tyrosine, main = 'BMI')
plot(x = gal_met_prot_data$T2D, y = gal_met_prot_data$Tyrosine, main = 'T2D')

```

lysoPC a C24:0

```{r explore-prot8}
#' lysoPC a C24:0
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$liver_fat, y = gal_met_prot_data$`lysoPC a C24:0`, main = 'liver_fat')
plot(x = gal_met_prot_data$visual_fat, y = gal_met_prot_data$`lysoPC a C24:0`, main = 'visual_fat')
plot(x = gal_met_prot_data$liver_fat_cat1, y = gal_met_prot_data$`lysoPC a C24:0`, main = 'liver_fat_cat1')
plot(x = gal_met_prot_data$liver_fat_cat2, y = gal_met_prot_data$`lysoPC a C24:0`, main = 'liver_fat_cat2')
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$Sex, y = gal_met_prot_data$`lysoPC a C24:0`, main = 'sex')
plot(x = gal_met_prot_data$Age_10, y = gal_met_prot_data$`lysoPC a C24:0`, main = 'Age_10')
plot(x = gal_met_prot_data$BMI, y = gal_met_prot_data$`lysoPC a C24:0`, main = 'BMI')
plot(x = gal_met_prot_data$T2D, y = gal_met_prot_data$`lysoPC a C24:0`, main = 'T2D')

```

PC ae C42:3

```{r}
#' lysoPC a C24:0
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$liver_fat, y = gal_met_prot_data$`PC ae C42:3`, main = 'liver_fat')
plot(x = gal_met_prot_data$visual_fat, y = gal_met_prot_data$`PC ae C42:3`, main = 'visual_fat')
plot(x = gal_met_prot_data$liver_fat_cat1, y = gal_met_prot_data$`PC ae C42:3`, main = 'liver_fat_cat1')
plot(x = gal_met_prot_data$liver_fat_cat2, y = gal_met_prot_data$`PC ae C42:3`, main = 'liver_fat_cat2')
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$Sex, y = gal_met_prot_data$`PC ae C42:3`, main = 'sex')
plot(x = gal_met_prot_data$Age_10, y = gal_met_prot_data$`PC ae C42:3`, main = 'Age_10')
plot(x = gal_met_prot_data$BMI, y = gal_met_prot_data$`PC ae C42:3`, main = 'BMI')
plot(x = gal_met_prot_data$T2D, y = gal_met_prot_data$`PC ae C42:3`, main = 'T2D')

```

H1

```{r}
#' H1
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$liver_fat, y = gal_met_prot_data$H1, main = 'liver_fat')
plot(x = gal_met_prot_data$visual_fat, y = gal_met_prot_data$H1, main = 'visual_fat')
plot(x = gal_met_prot_data$liver_fat_cat1, y = gal_met_prot_data$H1, main = 'liver_fat_cat1')
plot(x = gal_met_prot_data$liver_fat_cat2, y = gal_met_prot_data$H1, main = 'liver_fat_cat2')
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$Sex, y = gal_met_prot_data$H1, main = 'sex')
plot(x = gal_met_prot_data$Age_10, y = gal_met_prot_data$H1, main = 'Age_10')
plot(x = gal_met_prot_data$BMI, y = gal_met_prot_data$H1, main = 'BMI')
plot(x = gal_met_prot_data$T2D, y = gal_met_prot_data$H1, main = 'T2D')

```

glucose

```{r}
#' glucose
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$liver_fat, y = gal_met_prot_data$glucose, main = 'liver_fat')
plot(x = gal_met_prot_data$visual_fat, y = gal_met_prot_data$glucose, main = 'visual_fat')
plot(x = gal_met_prot_data$liver_fat_cat1, y = gal_met_prot_data$glucose, main = 'liver_fat_cat1')
plot(x = gal_met_prot_data$liver_fat_cat2, y = gal_met_prot_data$glucose, main = 'liver_fat_cat2')
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$Sex, y = gal_met_prot_data$glucose, main = 'sex')
plot(x = gal_met_prot_data$Age_10, y = gal_met_prot_data$glucose, main = 'Age_10')
plot(x = gal_met_prot_data$BMI, y = gal_met_prot_data$glucose, main = 'BMI')
plot(x = gal_met_prot_data$T2D, y = gal_met_prot_data$glucose, main = 'T2D')

```


## Diet

Butter

```{r explore-diet1}
#colnames(gal_met_prot_data)
#' A1AT
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$liver_fat, y = gal_met_prot_data$Butter, main = 'liver_fat')
plot(x = gal_met_prot_data$visual_fat, y = gal_met_prot_data$Butter, main = 'visual_fat')
plot(x = gal_met_prot_data$liver_fat_cat1, y = gal_met_prot_data$Butter, main = 'liver_fat_cat1')
plot(x = gal_met_prot_data$liver_fat_cat2, y = gal_met_prot_data$Butter, main = 'liver_fat_cat2')
par(mfrow = c(2, 2))
plot(x = gal_met_prot_data$Sex, y = gal_met_prot_data$Butter, main = 'sex')
plot(x = gal_met_prot_data$Age_10, y = gal_met_prot_data$Butter, main = 'Age_10')
plot(x = gal_met_prot_data$BMI, y = gal_met_prot_data$Butter, main = 'BMI')
plot(x = gal_met_prot_data$T2D, y = gal_met_prot_data$Butter, main = 'T2D')

```


# Other visualization of `r TARGET` phenotype

## Multi splitted violin plots with all significant

```{r}
#' filter to keep only the singif proteins and metabolites
#load(file.path(RDATA_PATH, 'lm_output_results_proteins.txt'))
#lm_output_prot <- lm_output
lm_output_prot <- results_prot
lm_output_prot <- lm_output_prot[lm_output_prot$significant_liver_fat_cat14 == 1, ]
sign_prot <- rownames(lm_output_prot)
sign_gene <- lm_output_prot$Genes
sign_prot_gene <- c(rbind(sign_prot, sign_gene)) #zipped

#load(file.path(RDATA_PATH, 'lm_output_results_metabolites.txt'))
#lm_output_met <- lm_output
lm_output_met <- results_met
lm_output_met <- lm_output_met[lm_output_met$significant_liver_fat_cat14 == 1, ]
sign_met <- rownames(lm_output_met)

#testing
#significant_data <- gal_met_prot_data[, c('liver_fat_cat2', sign_prot, sign_met)]
#plot(as.factor(significant_data$liver_fat_cat2), significant_data[,2], main = 'liver_fat_cat2')

#vioplot(significant_data[,2] ~ as.factor(significant_data$liver_fat_cat2),
#        xlab = "Control versus Grade 3 Steatosis", ylab = "Abundance", main = "",
#        method = "jitter", cex.names = 0.6)

```

Fancy splitted (`r TARGET`) violin plot for each significant feature.

```{r functions-splitted-violin-plot}
#' inspiration see: https://stackoverflow.com/questions/47409086/how-to-plot-two-violin-plot-series-on-the-same-graph-using-seaborn
#' code from: https://stackoverflow.com/questions/35717353/split-violin-plot-with-ggplot2

GeomSplitViolin <- ggproto("GeomSplitViolin", GeomViolin,
                           draw_group = function(self, data, ..., draw_quantiles = NULL) {
  data <- transform(data, xminv = x - violinwidth * (x - xmin), xmaxv = x + violinwidth * (xmax - x))
  grp <- data[1, "group"]
  newdata <- plyr::arrange(transform(data, x = if (grp %% 2 == 1) xminv else xmaxv), if (grp %% 2 == 1) y else -y)
  newdata <- rbind(newdata[1, ], newdata, newdata[nrow(newdata), ], newdata[1, ])
  newdata[c(1, nrow(newdata) - 1, nrow(newdata)), "x"] <- round(newdata[1, "x"])

  if (length(draw_quantiles) > 0 & !scales::zero_range(range(data$y))) {
    stopifnot(all(draw_quantiles >= 0), all(draw_quantiles <=
      1))
    quantiles <- ggplot2:::create_quantile_segment_frame(data, draw_quantiles)
    aesthetics <- data[rep(1, nrow(quantiles)), setdiff(names(data), c("x", "y")), drop = FALSE]
    aesthetics$alpha <- rep(1, nrow(quantiles))
    both <- cbind(quantiles, aesthetics)
    quantile_grob <- GeomPath$draw_panel(both, ...)
    ggplot2:::ggname("geom_split_violin", grid::grobTree(GeomPolygon$draw_panel(newdata, ...), quantile_grob))
  }
  else {
    ggplot2:::ggname("geom_split_violin", GeomPolygon$draw_panel(newdata, ...))
  }
})

geom_split_violin <- function(mapping = NULL, data = NULL, stat = "ydensity", position = "identity", ...,
                              draw_quantiles = NULL, trim = TRUE, scale = "area", na.rm = FALSE,
                              show.legend = NA, inherit.aes = TRUE) {
  layer(data = data, mapping = mapping, stat = stat, geom = GeomSplitViolin,
        position = position, show.legend = show.legend, inherit.aes = inherit.aes,
        params = list(trim = trim, scale = scale, draw_quantiles = draw_quantiles, na.rm = na.rm, ...))
}

#ggplot(df_long, aes(variable, value, fill = liver_fat_cat2)) + geom_split_violin()

```

Plot the splitted violin plots.

TODO: somehow this does no longer work!

```{r plot-splitviolin, eval = FALSE}
#' Splitted violin plots
#' correct format df, for proteins
df_long <- melt(gal_met_prot_data[, c('liver_fat_cat1', sign_prot)])

#rename protein to gene names
library(stringr)
for (i in 1:length(sign_prot)){
    df_long$variable <- str_replace(as.character(df_long$variable),
                                    sign_prot[i], sign_gene[i])
}
unique(df_long$variable)

plt <- ggplot(df_long, aes(variable, value, fill = liver_fat_cat1)) +
    geom_split_violin() +
    labs(subtitle = "Control versus Grade 3 Steatosis") +
    scale_fill_manual(labels = c("Control", "Grade 3 Steatosis"),
                      values = c("#0C6291", "#A63446")) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 10),
          axis.text.y = element_text(hjust = 1, size = 8),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.title=element_blank())

ggsave(paste0(IMAGE_PATH, "multiviolin_prot.png"),
       plot = plt, dpi = 600, width = 11,
       height = 12, units = "cm", scale = 2.5)
plt


#' redo for metabolites
df_long <- melt(gal_met_prot_data[, c('liver_fat_cat2', sign_met)])


plt <- ggplot(df_long, aes(variable, value, fill = liver_fat_cat2)) +
    geom_split_violin() +
    scale_fill_manual(labels = c("Control", "Grade 3 Steatosis"),
                      values = c("#0C6291", "#A63446")) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 10),
          axis.text.y = element_text(hjust = 1, size = 8),
          plot.margin = margin(t = 5, l = 45, r = 5, b = 5),
          legend.position = "none")

ggsave(paste0(IMAGE_PATH, "multiviolin_metab.png"),
       plot = plt, dpi = 600, width = 18,
       height = 12, units = "cm", scale = 2.5)
plt

```



## Pathway analysis: proteins and metab, other

plan:

- [x] find related pathways and perform pathway enrichment with Ramp-DB.
      https://rampdb.nih.gov/pathway-enrichment. see rampdb-test.ods
- [x] manually curate and define the pathway clusters, also split again proteins
      and metabolites (since no overlap).
- [x] floorplan of coagulation and AA metabolism, make/paste in ppt.
- [x] add barplots of measured metab/prot with background (prot/metab) and
     control, grade S3 steatosis 2 bars, with '*' if significant (lm's).
     For now, boxplots and only of the significant ones, can change in the
     future.
- [x] optional: add umetab results from mummichog for validation of the
     targeted p180 metabolites results. Or just as a sentence (incl in M&M)
     that this was confirmed and SI table of found pathways.


Box plots for on the plot pathway:

Proteins:

1) Response to elevated platelet cytosolic Ca2+

```{r}
#' Response to elevated platelet cytosolic Ca2+
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P01011, main = 'SERPINA3')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Q14624, main = 'ITIH4')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`Q14624;Q14624-2`, main = 'ITIH4')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`Q06033;Q06033-2`, main = 'ITIH3')

```


2) Network map of SARS-CoV-2 signaling pathway

```{r}
#' Network map of SARS-CoV-2 signaling pathway
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Q14624, main = 'ITIH4')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`Q14624;Q14624-2`, main = 'ITIH4')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`Q06033;Q06033-2`, main = 'ITIH3')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P02750, main = 'LRG1')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P05156, main = 'CFI')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P08603, main = 'CFH')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P05090, main = 'APOD')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$O95445, main = 'APOM')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`P06396;P06396-2`, main = 'GSN')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P04217, main = 'A1BG')

```

3) Post-translational protein phosphorylation

```{r}
#' Post-translational protein phosphorylation
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P01008, main = 'SERPINC1')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P05546, main = 'SERPIND1')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P01024, main = 'C3')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P00450, main = 'CP')

```

4) Complement and coagulation cascades

```{r}
#' Complement and coagulation cascades
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P01008, main = 'SERPINC1', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P05546, main = 'SERPIND1', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P01024, main = 'C3', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P02748, main = 'C9', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P04004, main = 'VTN', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P05156, main = 'CFI', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$P08603, main = 'CFH', col = c("#0C6291", "#A63446"))

```

Metabolites and lipids:

5) Lipids biosynthesis and degradation

As all except one has same direction, and only 1 dot on the pathways, we only
add one PC and one lysoPC for on the map. We take the top-ranking ones. idem for
the top SM.

```{r}
#' Lipids synthesis/degradation
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`PC ae C34:3`, main = 'PC ae C34:3')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`lysoPC a C24:0`, main = 'lysoPC a C24:0')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`lysoPC a C20:4`, main = 'lysoPC a C20:4')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`SM C16:0`, main = 'SM C16:0')

```

6) AAs biosynthesis and degradation

```{r}
#' AAs synthesis/degradation
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Glu, main = 'Glu', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Tyr, main = 'Tyr', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Val, main = 'Val', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Ile, main = 'Ile', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Ala, main = 'Ala', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Leu, main = 'Leu', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Gly, main = 'Gly', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Phe, main = 'Phe', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Trp, main = 'Trp', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Gln, main = 'Gln', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`alpha-AAA`, main = 'alpha-AAA', col = c("#0C6291", "#A63446"))
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$SDMA, main = 'SDMA', col = c("#0C6291", "#A63446"))

```

7) carnitines beta oxidation

```{r}
#' carnitines beta-oxidat
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$C0, main = 'C0')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$C3, main = 'C3')

```

8) BCAA and carnitines

```{r}
#' BCAA and carnitines
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$C5, main = 'C5')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$C4, main = 'C4')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$C3, main = 'C3')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`C3:1`, main = 'C3:1')
#plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`C4:1`, main = 'C4:1')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Leu, main = 'Leu')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Val, main = 'Val')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$Ile, main = 'Ile')

```

9) SM metabolism

```{r}
#' SM metabolism
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`SM (OH) C22:2`, main = 'SM (OH) C22:2')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`SM C24:1`, main = 'SM C24:1')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`SM C16:0`, main = 'SM C16:0')
plot(x = as.factor(gal_met_prot_data$liver_fat_cat2), y = gal_met_prot_data$`SM C20:2`, main = 'SM C20:2')

```



# Sensitivity analysis to investigate influence of T2D on the results

See document: NAFLD_T2D_sensititvity_analysis.Rmd, the original lms will be
loaded from above. The lms with T2D as added covariate are calculated in this
paragraph and also saved as Rdata for further sensititvity_analysis.

We first evaluate the number of T2D cases in each steatosis category:

```{r}
t2d <- split(gal_met_prot_data$T2D, gal_met_prot_data$liver_fat_cat1)
lapply(t2d, summary)
```

The ratio between T2D and non T2D seems to change with the steatosis category -
this might have an impact on the analysis, in particular for the higher
categories.

## Proteomics data

We apply the same linear regression analyses as above, but with the *T2D*
variable as additional covariate.

linear models (classification) setup:

```{r define-lm--covariates-and-target-t2d}
print('covariates of the linear regression model:')
COVARIATES <- c("Sex", "Age_10", "T2D")
print(COVARIATES)

print('target of the linear regression model:')
TARGET <- 'liver_fat_cat1'
print(TARGET)

DATASET <- 'NAFLD'
print(paste0('final variables of the ', DATASET, ' dataset used:'))

print('nr of variables inspected:')
print(length(proteins))

print('nr of individuals inspected:')
print(nrow(gal_met_prot_data))
```

```{r linear-model-t2d}

#' Fit linear models and extract coefficients and p-values on each protein
lm_output <- lapply(proteins, FUN = funForLinReg, data = gal_met_prot_data) |>
    do.call(what = rbind)

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)
lm_output <- as.data.frame(lm_output)
```

The impact of including *T2D* in the linear models on the steatosis-related
protein abundance differences are shown below.

Export for the paper SI Figre S0A:

```{r sensitivity-proteomics, fig.path = IMAGE_PATH, fig.width = 12, fig.height = 4}
par(mfrow = c(1, 3))
lim <- range(lm_output$coef_liver_fat_cat12, results_prot$coef_liver_fat_cat12)
plot(results_prot$coef_liver_fat_cat12, lm_output$coef_liver_fat_cat12,
     xlim = lim, ylim = lim, xlab = expression(coef),
     ylab = expression(coef[+T2D]), main = "Steatosis category 2 vs 1",
     pch = 21, col = "#00000080", bg = "#00000040")
grid()
l <- lm(lm_output$coef_liver_fat_cat12 ~ results_prot$coef_liver_fat_cat12)
abline(l)

lim <- range(lm_output$coef_liver_fat_cat13, results_prot$coef_liver_fat_cat13)
plot(results_prot$coef_liver_fat_cat13, lm_output$coef_liver_fat_cat13,
     xlim = lim, ylim = lim, xlab = expression(coef),
     ylab = expression(coef[+T2D]), main = "Steatosis category 3 vs 1",
     pch = 21, col = "#00000080", bg = "#00000040")
grid()
l <- lm(lm_output$coef_liver_fat_cat13 ~ results_prot$coef_liver_fat_cat13)
abline(l)

lim <- range(lm_output$coef_liver_fat_cat14, results_prot$coef_liver_fat_cat14)
plot(results_prot$coef_liver_fat_cat14, lm_output$coef_liver_fat_cat14,
     xlim = lim, ylim = lim, xlab = expression(coef),
     ylab = expression(coef[+T2D]), main = "Steatosis category 4 vs 1",
     pch = 21, col = "#00000080", bg = "#00000040")
grid()
l <- lm(lm_output$coef_liver_fat_cat14 ~ results_prot$coef_liver_fat_cat14)
abline(l)

```

## Metabolomics data

We apply the same linear regression analyses as above, but with the *T2D*
variable as additional covariate.

linear models (classification) setup:

```{r}
print('covariates of the linear regression model:')
COVARIATES <- c("Sex", "Age_10", "T2D")
print(COVARIATES)

print('target of the linear regression model:')
TARGET <- 'liver_fat_cat1'
print(TARGET)

DATASET <- 'NAFLD'
print(paste0('final variables of the ', DATASET, ' dataset used:'))

print('nr of variables inspected:')
print(length(metabolites))

print('nr of individuals inspected:')
print(nrow(gal_met_prot_data))
```

```{r}

#' Fit linear models and extract coefficients and p-values on each metabolite
lm_output <- lapply(metabolites, FUN = funForLinReg,
                    data = gal_met_prot_data) |>
    do.call(what = rbind)

#' Adjust p-values for multiple hypothesis testing
adjp <- apply(lm_output[, grep("p-value", colnames(lm_output))],
              MARGIN = 2, p.adjust, method = "BH")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
lm_output <- cbind(lm_output, adjp)
lm_output <- as.data.frame(lm_output)
```

The impact of including *T2D* in the linear models on the steatosis-related
metabolite abundance differences are shown below.

Export for the paper SI Figre S0B:

```{r sensitivity-metabolomics, fig.path = IMAGE_PATH, fig.width = 12, fig.height = 4}
par(mfrow = c(1, 3))
lim <- range(lm_output$coef_liver_fat_cat12, results_met$coef_liver_fat_cat12)
plot(results_met$coef_liver_fat_cat12, lm_output$coef_liver_fat_cat12,
     xlim = lim, ylim = lim, xlab = expression(coef),
     ylab = expression(coef[+T2D]), main = "Steatosis category 2 vs 1",
     pch = 21, col = "#00000080", bg = "#00000040")
grid()
l <- lm(lm_output$coef_liver_fat_cat12 ~ results_met$coef_liver_fat_cat12)
abline(l)

lim <- range(lm_output$coef_liver_fat_cat13, results_met$coef_liver_fat_cat13)
plot(results_met$coef_liver_fat_cat13, lm_output$coef_liver_fat_cat13,
     xlim = lim, ylim = lim, xlab = expression(coef),
     ylab = expression(coef[+T2D]), main = "Steatosis category 3 vs 1",
     pch = 21, col = "#00000080", bg = "#00000040")
grid()
l <- lm(lm_output$coef_liver_fat_cat13 ~ results_met$coef_liver_fat_cat13)
abline(l)

lim <- range(lm_output$coef_liver_fat_cat14, results_met$coef_liver_fat_cat14)
plot(results_met$coef_liver_fat_cat14, lm_output$coef_liver_fat_cat14,
     xlim = lim, ylim = lim, xlab = expression(coef),
     ylab = expression(coef[+T2D]), main = "Steatosis category 4 vs 1",
     pch = 21, col = "#00000080", bg = "#00000040")
grid()
l <- lm(lm_output$coef_liver_fat_cat14 ~ results_met$coef_liver_fat_cat14)
abline(l)

```

## Summary

For steatosis category 2 there is no difference, but coefficients deviate from
the identity line for steatosis grade 3 and 4.

TODO: ask Cristian for opinion. Whether it's better to account for T2D or not
here.



# Acknowledgments

Consulted sources used as basis for code:

- ~/nafld_proteo_metabolomics/NAFLD_metab_targeted_lm.Rmd
- ~?metabo_chris_untargeted/CHRIS_lm_pos.Rmd
- kora: CHRIS_KORA_tmetab_lm_binary.Rmd


# Session information

R packages used for the analysis:

```{r}
sessionInfo()
```
