# Compile the NAFLD untargeted metabolomics data set

```{r, echo = FALSE}
library(BiocStyle)
```

This document describes how the untargeted data set for the NAFLD project is
compiled and how it can be used. We will store the full MS data into a SQLite
database which can be accessed and used with the `r Biocpkg("MsBackendSql")`
package.

We first load the information on the raw data files generated from each study
participant.

```{r}
library(readxl)
library(Spectra)
library(MsBackendSql)

pd <- as.data.frame(read_xlsx("data/nafld_hilic_files.xlsx"))
pd$time_stamp <- as.character(pd$time_stamp)
```

## Store MS data into a SQL database

Next we store the MS data into a RSQLite database.

```{r}
fls <- file.path("data", pd$mzML_file)

library(RSQLite)
con <- dbConnect(SQLite(), dbname = "data/NAFLD_untargeted.sqlite")

system.time(
    createMsBackendSqlDatabase(con, fls, MsBackendMzR(), chunksize = 10)
)
dbDisconnect(con)
```

Add sample annotations to the database.

```{r}
sps <- Spectra("data/NAFLD_untargeted.sqlite", source = MsBackendOfflineSql(),
               drv = SQLite())
sps$file <- basename(dataOrigin(sps))
writeSampleData(sps, pd, colname = "file_name", spectraVariable = "file")
```

Check that it worked.

```{r}
con <- dbconn(sps@backend)
dbListTables(con)
dbGetQuery(con, "select * from sample_data limit 4")
dbGetQuery(con, "select * from sample_to_msms_spectrum limit 4")
```

## Create a `MsExperiment` representing the data


```{r}
library(MsExperiment)
library(RSQLite)
library(readxl)
pd <- as.data.frame(read_xlsx("data/nafld_hilic_files.xlsx"))
pd$time_stamp <- as.character(pd$time_stamp)

#' NEW way using the readMsExperiment with the backend.
nafld_1 <- readMsExperiment(
    spectraBackend = MsBackendOfflineSql(),
    sampleData = pd, spectraFiles = pd$file_name,
    drv = SQLite(), dbname = "data/NAFLD_untargeted.sqlite")

#' Read all in one; requires we have samples in the database
nafld_2 <- readMsExperiment(
    spectraBackend = MsBackendOfflineSql(), drv = SQLite(),
    dbname = "data/NAFLD_untargeted.sqlite")


#' OLD way
#' Load the spectra data from the database
sps <- Spectra("data/NAFLD_untargeted.sqlite", drv = SQLite(),
               source = MsBackendOfflineSql())

#' Create MsExperiment and fill with data
nafld <- MsExperiment()
sampleData(nafld) <- DataFrame(pd)

sps$file <- basename(dataOrigin(sps))
spectra(nafld) <- sps

#' Link spectra to samples
nafld <- linkSampleData(nafld, with = "sampleData.file_name = spectra.file")

```

Compare that data is the same.

```{r}
library(testthat)
#' sampleData
expect_equal(sampleData(nafld_1), sampleData(nafld))
expect_equal(sampleData(nafld_2)[, colnames(sampleData(nafld))],
             sampleData(nafld))
#' mapping
expect_equal(nafld_1@sampleDataLinks[["spectra"]],
             nafld@sampleDataLinks[["spectra"]])
expect_equal(nafld_2@sampleDataLinks[["spectra"]],
             nafld@sampleDataLinks[["spectra"]])
#' Spectra
a <- spectraData(spectra(nafld))
b <- spectraData(spectra(nafld_1))
d <- spectraData(spectra(nafld_2))
cols <- intersect(colnames(a), colnames(b))

expect_equal(a[, cols], b[, cols])
expect_equal(a[, cols], d[, cols])

a <- peaksData(spectra(nafld[1:10]))
b <- peaksData(spectra(nafld_1[1:10]))
d <- peaksData(spectra(nafld_2[1:10]))
expect_equal(a, b)
expect_equal(a, d)
```

Subset to data from a specific polarity.

```{r}
nafld_pos <- nafld[sampleData(nafld)$polarity == "POS"]
nafld_neg <- nafld[sampleData(nafld)$polarity == "NEG"]
```

# Add sample link in SQL database

```{r}
#' Install Johannes branch method - change this once it's merged and published
#' remotes::install_github("RforMassSpectrometry/MsExperiment", ref = "jomain")

sps <- Spectra("data/NAFLD_untargeted.sqlite", drv = SQLite(),
               source = MsBackendOfflineSql())
sps$file_name <- basename(dataOrigin(sps))
pd <- as.data.frame(read_xlsx("data/nafld_hilic_files.xlsx"))

#' Create MsExperiment and fill with data
nafld <- MsExperiment()
sampleData(nafld) <- DataFrame(pd)
spectra(nafld) <- sps

nafld <- linkSampleData(nafld, with = "sampleData.file_name = spectra.file_name")

#' Write sample data to the database
dbWriteSampleData(nafld)
```


To load after that

```{r}
#' Load the spectra data from the database
sps <- Spectra("data/NAFLD_untargeted.sqlite", drv = SQLite(),
               source = MsBackendOfflineSql())

nafld <- MsExperiment(spectra = sps)
nafld
```

# Session information

```{r}
sessionInfo()
```
