# Compile the NAFLD untargeted metabolomics data set

```{r, echo = FALSE}
library(BiocStyle)
```

This document describes how the untargeted data set for the NAFLD project is
compiled and how it can be used. We will store the full MS data into a SQLite
database which can be accessed and used with the `r Biocpkg("MsBackendSql")`
package.

We first load the information on the raw data files generated from each study
participant.

```{r}
library(readxl)
library(Spectra)
library(MsBackendSql)

pd <- as.data.frame(read_xlsx("data/nafld_untargeted_metabolomics.xlsx"))
```

## Store MS data into a SQL database

Next we store the MS data into a RSQLite database.

```{r}
fls <- file.path("data", pd$mzML_file)

library(RSQLite)
con <- dbConnect(SQLite(), dbname = "data/NAFLD_untargeted.sqlite")

system.time(
    createMsBackendSqlDatabase(con, fls, MsBackendMzR(), chunksize = 10)
)
dbDisconnect(con)
```


## Create a `MsExperiment` representing the data


```{r}
library(MsExperiment)
library(RSQLite)

#' Load the spectra data from the database
sps <- Spectra("data/NAFLD_untargeted.sqlite", drv = SQLite(),
               source = MsBackendOfflineSql())

#' Subset to positive polarity
pd <- pd[pd$polarity == "POS", ]

#' Subset the Spectra to the respective files
sps <- sps[basename(sps$dataOrigin) %in% pd$file]
sps$file <- basename(sps$dataOrigin)

#' Create MsExperiment and fill with data
nafld <- MsExperiment()
sampleData(nafld) <- DataFrame(pd)
spectra(nafld) <- sps

#' Link spectra to samples
nafld <- linkSampleData(nafld, with = "sampleData.file = spectra.file")
```



# Session information

```{r}
sessionInfo()
```
