---
title: "Comparison of adjustedRTIme strategies"
author: "Philippine Louail"
date: "2023-12-04"
output: html_document
---

```{r packages used, message=FALSE, warning=FALSE}
library(MsExperiment)
library(xcms)
library(Spectra)
library(RColorBrewer)
library(pander)
library(readxl)
library(MetaboCoreUtils)
library(pheatmap)
library(CompMetaboTools) # might want to get rid of this one though.
library(MetaboAnnotation)
library(MatrixGenerics)
``` 

This Document summarize the different strategy I tried for adjsuting the retention time of the NAFLD dataset as the classic parameters was 
not satisfying. 

To note "A mix standard list" was created specifically for nafld Rtime adjustment. It is comprised of the IS that we spiked the samples with and some common compounds that I filtered visually. this helps the anchor peaks ot be more distributed along the retention time axis and thus for better alignment. 

Here is a summary of the strategy 

- Strategy A:  Normal parameters, one run, correspondence analysis followed by adjustment based on pool samples only 
- Strategy B: First run with a defined peak matrix for all sample ( - blanks), followed by second run with correspondence and adjustment based on pool samples only 
- Strategy C: First run with a defined peak matrix for pool samples, followed by second run with correspondence and adjustment based on pool samples only 
- Strategy D: Only 1 run based on peak matrix 
(these are ran in separate rmd files)

First create a table of chrompeaks name for each samples for each IS. 

- table of IS for alignment. Here jsut showing how I build it.
```{r eval=FALSE, include=FALSE}
#' the inital table was a complete mix of IS spiked and standard manually selected from known compounds. only peaks that were clearly defined were selected. 
alignment_standard <- read.delim("Mix_IS_standard.txt") 
alignment_standard <- alignment_standard[!is.na(alignment_standard$POS), ]
rownames(alignment_standard) <- alignment_standard$abbreviation

#'generate calcualte formula
alignment_standard$mz <- mapply(alignment_standard$formula, 
                             alignment_standard$POS,
                             FUN = mass2mz)

#' replace RT by obs 
alignment_standard$RT <- c(167, 183, 179, 208, 170, 167, 185, 184, 169,
                               166, 164, 35, 37, 176, 189, 177, 30, 170, 145,
                               176, 167, 174, 163,  35, 35, 163, 35)
#' not sure i am choosing the right peak for pipecolic acid but it is very
#'  clean so still nice to observe and base alignment on this

#' Fit for each standards
alignment_standard$mzmin <- alignment_standard$mz - 0.05
alignment_standard$mzmax <- alignment_standard$mz + 0.05
alignment_standard$rtmin <- alignment_standard$RT - 15
alignment_standard$rtmax <- alignment_standard$RT + 15

#' Save this table (also alignment for Strategy B/C/D is based on this)
write.table(alignment_standard, "Mix_alignment.txt", sep = "\t")
```

load raw NAFLD after peak detection/refinement and load the standard
table used for this alignment. 

```{r}
load("nafld_after_refine")
standard <- read.delim("Mix_alignment.txt")

# get chromPeak data
fdef <- as.data.frame(chromPeaks(nafld))
fdef$feature_id <- rownames(fdef)

#' prep result of loop
final_table <- matrix(
    ncol = 530, 
    nrow = nrow(standard), 
    dimnames = list(c(row.names(standard)), c(1:530))
)

#' Keep only matches with highest intensity 
rem_dups <- function(x) {
    if (nrow(matches(x)) > 1)
        filterMatches(x, index = integer())
    else x
}

#' Get chromPeaks ID 
for (i in unique(fdef$sample)){
    tmp <- fdef[fdef[,"sample"]== i, ]
    if(sampleData(nafld)$sample_type[i] != "Blank") {
        match_intern_standard <- matchValues(
        query = standard,
        target = tmp,
        mzColname = c("mz", "mz"),
        rtColname = c("RT", "rt" ),
        param = MzRtParam(ppm = 50, toleranceRt = 20))
    match_intern_standard <- MetaboAnnotation::endoapply(match_intern_standard,
                                                     FUN = rem_dups)
    final_table[,i] <- match_intern_standard$target_feature_id
    } 
}
#' if need to use only the pool
index <- sampleData(nafld)$sample_type != "Blank"
ID_table <- final_table[, index]
```


Make a matrix of the Rt of these chrompeaks in each alignment strategy (+raw)

```{r}
#Function to create rt dataframe
rtdf <- function(nafld, ID_table) {
    indices <- as.vector(ID_table)
    fdef <- fdef <- as.data.frame(chromPeaks(nafld))
    fdef$feature_id <- rownames(fdef)
    x <- fdef[indices, "rt"]
    dim(x) <- dim(ID_table)
    rownames(x) <- rownames(ID_table)
    colnames(x) <- colnames(ID_table)
    x
}

### Raw 
RTR <- rtdf(nafld, ID_table)

### A
load("strategyA")
RTA <- rtdf(nafld, ID_table)

### B
load("strategyB")
RTB <- rtdf(nafld, ID_table)

### C
load("strategyC")
RTC <- rtdf(nafld, ID_table)

###D
load("strategyD")
RTD <- rtdf(nafld, ID_table)
```

```{r}
sumNA <- data.frame(
    Raw = apply(RTR, 1, function(row) {sum(is.na(row))}),
    A = apply(RTA, 1, function(row) {sum(is.na(row))}), 
    B = apply(RTB, 1, function(row) {sum(is.na(row))}),
    C = apply(RTC, 1, function(row) {sum(is.na(row))}), 
    D = apply(RTD, 1, function(row) {sum(is.na(row))})
)

sumNA
```

```{r}
Sdsdf <- data.frame(
    Raw = rowSds(RTR, na.rm = TRUE),
    A = rowSds(RTA, na.rm = TRUE),
    B = rowSds(RTB, na.rm = TRUE),
    C = rowSds(RTC, na.rm = TRUE),
    D = rowSds(RTD, na.rm = TRUE)
)

pandoc.table(Sdsdf, style = "rmarkdown")
boxplot(Sdsdf, outline = FALSE, xlab = "RT standard deviation")
```

