---
title: "CHRIS_fib4_explore"
author: "Marilyn De Graeve, Essi Hantikainen, Johannes Rainer"
affiliation: "Eurac Research, Bolzano, Italy"
date: "2024-04-05"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
editor_options:
  markdown:
    wrap: 72
bibliography: references.bib
---

**Modified**: `r file.info("CHRIS_fib4_explore.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r style, message = FALSE, echo = FALSE, warning = FALSE, results = "asis"}
library("BiocStyle")
library("kableExtra")
library("knitr")
library("rmarkdown")
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now, units = "secs")
      # return a character string to show the time
      paste("    Time for this code chunk to run:", round(res,
        2), "seconds")
    }
  }
}))
opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE,
               cache = FALSE, fig.width = 7, fig.height = 7, time_it = TRUE)

```

# Introduction

Load CHRIS data (001...) and calculate the fib4 from proteomics data.


# Setup

## Directories

```{r directories, echo = FALSE}
#' General settings
filename <- "CHRIS_fib4_explore"

#' Path to save images to; remove if exists.
IMAGE_PATH <- paste0("images/", filename, "/")
#if (file.exists(IMAGE_PATH))
#    unlink(IMAGE_PATH, recursive = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE)

#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)

#' PathS to the data
DATA_INPUT_PATH <- '/home/mdegraeve/Documents/Files/Work_Eurac/Data/NAFLD'  #FINDME!
DATA_OUTPUT_PATH <- "data"
```

## Packages

Data packages (*chrisData* and *chrisUtils*) and the ScanningSWATH TDFF data
module *chris_scanningswath_proteins* are not publicly available. Any such data
access needs to got through a *CHRIS Access Committee* request.

```{r packages, message=FALSE}
#' Load libraries
library(atc) # https://github.com/jorainer/atc
library(chron)
library(chrisData)
library(chrisUtils)
library(DT)
library(FamAgg)
library(ggfortify)
library(ggplot2)
library(ggpubr)
library(magick)
library(mixOmics)
library(pander)
library(pheatmap)
library(plotly)
library(RColorBrewer)
library(readxl)
library(reshape2)
options(rgl.useNULL = TRUE)
library(rgl)
library(tidyfr)
library(UpSetR)
library(vioplot)
library(writexl)

```

# Load clinical data CHRIS 001 set

From the clinical data, take the following:

```{r fib4-calc}
#' Biochemical traits (labdata)
#' lab param in Table 1.1: Blood parameters
#' find the right name: chrisDataFile()
lab <- chrisData("labdata")
rownames(lab) <- lab$AID
#colnames(lab)
#rownames(lab)[1:10]
#lab$AID[1:10]
#nrow(lab)
#head(lab)


#' retaining if interesting variable, don't keep AID col
#' blood
blood_params <- c('x0lp02aq',
                    'x0lp03', 'x0lp04q', 'x0lp05', 'x0lp06b', 'x0lp07', 'x0lp08', 'x0lp09', 'x0lp10',
                    'x0lp11q', 'x0lp12q', 'x0lp13', 'x0lp14bq', 'x0lp15q', 'x0lp16', 'x0lp17', 'x0lp18', 'x0lp19',
                    'x0lp20', 'x0lp21', 'x0lp22', 'x0lp23a', 'x0lp24', 'x0lp25aq', 'x0lp26a', 'x0lp27', 'x0lp28q', 'x0lp29',
                    'x0lp30', 'x0lp31', 'x0lp32q', 'x0lp33q', 'x0lp35', 'x0lp38', 'x0lp39',
                    'x0lp40', 'x0lp41', 'x0lp42', 'x0lp43', 'x0lp44', 'x0lp45', 'x0lp46', 'x0lp47', 'x0lp48',
                    'x0lp50f', 'x0lp50g', 'x0lp50h', 'x0lp50i', 'x0lp50j', 'x0lp51a', 'x0lp52q',
                    'x0lp82', 'x0lp83', 'x0lp84')


"0010224047" %in% rownames(lab) #T
"0080208536" %in% rownames(lab) #F

#lab$x0lp47

```

For correct age of 001:

```{r}
chris_general <- chrisData("general")
rownames(chris_general) <- chris_general$AID

```


# Metadata (excel)

All available trait (phenotype) data was defined in the
[nafld_data_overview.Rmd](nafld_data_overview.Rmd), is loaded below and merged
with the log2 transformed proteomics data of the NAFLD cohort.

```{r add phenotype dataa}
#' load xlsx file instead of below
handle <- 'masld_phenotypes.xlsx'
masld_phenotype_df <- read_excel(file.path(DATA_INPUT_PATH, handle)) |>
    as.data.frame()


#' make sure sld_type is a factor with group 'none' as the baseline
masld_phenotype_df$sld_type <- factor(masld_phenotype_df$sld_type,
                                      levels = c('none', 'MASLD',
                                                 'MetALD', 'SLD'))
print('the sld types are:')
levels(masld_phenotype_df$sld_type)
print('nr of samples per sld types:')
table(masld_phenotype_df$sld_type)

```

# FIB4 calcualtion and comparison with 008 set

```{r}
#' Fib4_chris

#handle <- 'data/kor_met_prot_data.txt'
#df <- read.table(file = handle, sep = "\t", header = TRUE)

df <- masld_phenotype_df
#df$AID_chris
#df$AID

colnames(df)

# keep only overlapping with chris in masld df
lab <- lab[df$AID_chris, ]
chris_general <- chris_general[df$AID_chris, ]

#' add remaining for fib4 (incl 001 age):
df$Age_chris001 <- chris_general$x0_age
df$Platelet_Count_chris001 <- lab$x0lp47     #Platelet Count (10^9/L) == x0lp47   Platelets (PLT) – x1000/µL
df$AST_chris001 <- lab$x0lp10                #AST level (U/L) == x0lp10   Aspartate aminotransferase (AST) – U/L
df$ALT_chris001 <- lab$x0lp11q               #ALT level (U/L) == x0lp11   Alanine aminotransferase (ALT) – U/L;
                                    #x0lp11q    Alanine aminotransferase (ALT) – U/L
                                    #x0lp11q contains numeric values
                                    #only; limit of detection values in
                                    #x0lp11 were replaced with the next
                                    #lower valid value

df$FIB4_chris001 <- ((df$Age_chris001 * df$AST_chris001)/(df$Platelet_Count_chris001 * sqrt(df$ALT_chris001)))
print('FIB$ of 001 chris baseline:')
summary(df$FIB4_chris001)
print('FIB$ of 008 nafld:')
summary(df$FIB4)

plot(df$FIB4_chris001, df$FIB4)

```

Correlate very well, fib4 from chris001 was calcualted, while the fib4 from
nafld008 comes from the excel (and cannot be recalculated, see next code chunk).
While the age differs, actually less effect than initially expected.

# FIB4 calculate again to check 008

```{r, eval = FALSE}
library(nafldData)
#nafldTraitAnnotations()|> head()

lab <- nafldData("labdata")
rownames(lab) <- lab$AID
colnames(lab)

"0010224047" %in% rownames(lab) #F
"0080208536" %in% rownames(lab) #T
nalp47

# keep only overlapping with chris in masld df
df <- masld_phenotype_df
lab <- lab[df$AID_chris, ]

#' add remaining for fib4:
df$Platelet_Count_chris008 <- lab$nalp47     #PLT (x1000/µl)
#df$AST_chris008 <- NO       #AST level (U/L) == x0lp10   Aspartate aminotransferase (AST) – U/L
#df$ALT_chris008 <- NO       #ALT level (U/L) == x0lp11   Alanine aminotransferase (ALT) – U/L;

#but age anyway older....
#df$FIB4_chris008 <- ((df$Age_chris008 * df$AST_chris008)/(df$Platelet_Count_chris008 * sqrt(df$ALT_chris008)))


#' add remaining for fib4:
#df$Platelet_Count <- lab$x0lp47

```

# Correlate age 

```{r}

print('Age of 001 chris baseline:')
summary(df$Age_chris001)
print('Age of 008 nafld:')
summary(df$Age)

plot(df$Age_chris001, df$Age)

```